---
title: "Assessment diagnostics and sensitivity analyses"
format: html
execute:
  eval: true
  warning: false
  message: false
vignette: >
  %\VignetteIndexEntry{Assessment diagnostics and sensitivity analyses}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

This guide demonstrates biological checks, one-step-ahead (OSA) residuals,
posterior summaries, parameter profiles, model grids, and equilibrium reference
points. Each analysis retains its settings and source identity in an `opal_obj`.
Changing its scientific configuration clears dependent analyses. Saving and
reading the object preserves the results.

The diagnostic design follows the [SBT assessment examples](https://www.quantifish.co.nz/sbt/articles/sbt.html).
The combined SDNR figure places each dataset on a separate row and compares its
OSA residual dispersion with a reference value of one.

## The dataset-level OSA figure

Start with the bundled Opakapaka optimum. Reading it does not optimise or sample.
This legacy optimum is retained for comparison with the numerical reference; the
[fit-review article](opakapaka-fit-review.html) explains its remaining gradient.
An observation diagnostic can be inspected even when a fit check fails.

```{r}
#| label: opaka-osa
library(opal)
opaka <- opal_read(system.file("extdata", "opaka_quickstart_fit.rds", package = "opal"))
opaka <- opal_osa(opaka, seed = 123)
```

```{r}
#| label: fig-opaka-sdnr
#| fig-cap: "Opakapaka OSA residual SDNR by dataset. The dashed line is one, and the horizontal intervals are approximate 95% intervals. Parameters, including estimated recruitment deviations, are held at their fitted values in this fixed-effects model."
#| fig-width: 8
#| fig-height: 4.5
plot_osa_sdnr(opaka)
```

```{r}
#| label: opaka-osa-table
knitr::kable(opal_derived(opaka, "osa")$summary[,
  c("dataset", "family", "n", "sdnr", "lower", "upper", "excluded", "failures")], digits = 3)
```

SDNR is the standard deviation of normalised OSA residuals. Values above one
indicate more residual variation than expected under the fitted observation
model; values below one indicate less. This example has relatively small CPUE
SDNRs and a larger research-fleet composition SDNR. These are prompts to examine
variance assumptions, weighting, and model flexibility. They are not automatic
instructions to reweight the data. The chi-squared intervals assume independent
normal residuals with known model parameters, so fitted-parameter effects make
them approximate.

Each CPUE index has its own row, even when several indices use the same fishery.
Length and weight compositions are grouped by fishery. Inspect time patterns and
normal quantiles as well as SDNR:

```{r}
#| label: fig-osa-qq
#| fig-cap: "Normal quantile plots for the same OSA residuals."
#| fig-width: 8
#| fig-height: 5
plot_osa_residuals(opaka, type = "qq")
```

### Likelihood and conditioning conventions

| Observation likelihood | Conditional distribution used by OSA |
|---|---|
| Lognormal CPUE | Normal distribution of log CPUE, with the fitted total log-scale SD |
| Multinomial composition | Sequential binomial distributions |
| Dirichlet composition | Sequential beta distributions |
| Dirichlet–multinomial composition | Sequential beta-binomial distributions |

All three composition likelihoods work for both length and weight data. The
calculation uses the prepared observations, fitted probabilities, effective
sample sizes, and concentration parameters. It introduces no extra pseudocounts.
Multinomial rounding follows the likelihood. Discrete residuals randomise within
the observed count's probability mass; an explicit seed makes them reproducible
without changing the caller's random-number state.

For fixed-effect models, residuals condition on the fitted parameters. When the
model declares random effects through `opal_obj(random = ...)`, Opal uses RTMB's
sequential, Laplace-based integration of those latent states. Other observation
streams remain observed while each stream is processed. This distinction matters
for recruitment deviations: estimated fixed effects and integrated random
effects are different diagnostic targets.

Within a stream, input order defines observation order, followed by ascending
composition bin. The final bin is constrained and has no residual. Removed
fisheries, zero sample sizes, exhausted counts, and any numerical failures remain
in `opal_derived(x, "osa")$residuals` with an explicit `reason`. Failed residuals
are reported in the summary; they must not be mistaken for a successful check.
Catch is conditioned on, rather than modelled with an observation likelihood, so
its reconstruction differences are not OSA residuals. Recruitment priors are also
outside this observation diagnostic.

## A small, reproducible assessment

The remaining examples use simulated data for two fisheries, 12 years, and two
seasons. `opal_example_inputs()` supplies CPUE and both composition streams; its
`family` argument selects any of the three composition distributions. Three
parameters are estimated, and biology and recruitment deviations are fixed.
This is a software tutorial, not an assessment of a real stock.

```{r}
#| label: synthetic-inputs
inputs <- opal_example_inputs("multinomial", seed = 712)
assessment <- opal_obj(inputs$data, inputs$parameters, inputs$map)
assessment <- opal_fit(assessment)
summary(assessment)
```

`cpue_data$ts` is a one-based, flattened year/season index:
`(year_index - 1) * n_season + season_index`. The first year's seasons are 1 and
2; the second year's seasons are 3 and 4. Composition years are model-year
indices because their predictions use catch accumulated over that year.
`data$years` gives display labels. Biological vectors are at age unless the model
explicitly supplies length-based biology; weight and spawning-output units must
be consistent with the supplied inputs. `catch_units_f` distinguishes weight
from numbers. Use `plot_catch(weight_units = ...)` and
`plot_biomass_spawning(units = ..., scale = ...)` for explicit display units.

### Biological acceptance

```{r}
#| label: biological-checks
opal_diagnose(assessment)
```

A passing fit requires convergence, the gradient and Hessian checks, parameter
bounds, and biological feasibility. The latter includes initial-equilibrium and
harvest penalties, abundance, mortality, maturity, spawning potential, harvest
fractions, catch reconstruction, and depletion. Positive continuation keeps an
invalid trial state numerically evaluable; a positive penalty does not establish
a valid equilibrium.

Posterior checking evaluates **every retained joint draw**, including its bounds
and draw-specific biology, in addition to R-hat, ESS, divergences, and tree depth.
Failures retain their chain, iteration, and reason. Marginal draws that omit
random effects cannot pass the full biological posterior check. Old validation
records become stale when the validation contract changes.

### Other composition distributions

These examples fit both composition streams for each likelihood, then collect the
same SDNR statistic. The seed controls independent simulated observations and
residual randomisation; results need not be identical across distributions.

```{r}
#| label: distribution-examples
family_diagnostics <- lapply(c("multinomial", "Dirichlet", "Dirichlet-multinomial"), function(family) {
  z <- opal_example_inputs(family)
  fit <- opal_fit(opal_obj(z$data, z$parameters, z$map))
  fit <- opal_osa(fit)
  out <- opal_derived(fit, "osa")$summary
  out$composition_family <- family
  out
})
knitr::kable(do.call(rbind, family_diagnostics)[,
  c("composition_family", "dataset", "n", "sdnr", "failures")], digits = 3)
```

```{r}
#| label: fig-composition-fit
#| fig-cap: "Observed length-composition proportions and fitted probabilities for the first fishery in the simulated example."
#| fig-width: 8
#| fig-height: 6
plot_composition(assessment, type = "lf", fishery = 1)
```

## Posterior analysis

The bundled simulated posterior was generated with four chains, 1,000 warm-up
iterations, and 1,000 retained iterations per chain. Its generation script is
`data-raw/generate-assessment-example.R`. That script requires both fit and MCMC
checks to pass before saving. Website builds read the result instead of sampling.
To reproduce it yourself:

```{r}
#| eval: false
assessment <- opal_mcmc(assessment, seed = 927, chains = 4, cores = 1,
                       num_warmup = 1000, num_samples = 1000, adapt_delta = 0.95)
assessment <- opal_check(assessment, "mcmc")
opal_save(assessment, "simulated-assessment.rds")
```

```{r}
#| label: read-posterior
sampled <- opal_read(system.file("extdata", "assessment_example.rds", package = "opal"))
summary(sampled)
knitr::kable(sampled$validation$mcmc$metrics$parameters, digits = 3)
sampled$validation$mcmc$metrics$biology[c("passes", "checked", "scope")]
```

Summaries use each draw's own model report and preserve the selected chain and
iteration identifiers. Non-finite values stop the calculation rather than being
silently removed. Running summaries does not turn an unchecked posterior into an
accepted one.

```{r}
#| label: posterior-summaries
sampled <- opal_posterior(sampled,
  quantities = c("B0", "spawning_biomass_y", "static_depletion_y"))
posterior <- opal_derived(sampled, "posterior")
knitr::kable(posterior$parameters, digits = 3)
```

```{r}
#| label: fig-posterior-depletion
#| fig-cap: "Median and 95% posterior intervals for simulated depletion. These intervals are conditional on the model's fixed biology and recruitment deviations."
#| fig-width: 8
#| fig-height: 4.5
depletion <- subset(posterior$reports, quantity == "static_depletion_y")
depletion$year <- c(sampled$data$years, sampled$data$last_yr + 1)
ggplot2::ggplot(depletion, ggplot2::aes(x = year, y = q0.500)) +
  ggplot2::geom_ribbon(ggplot2::aes(ymin = q0.025, ymax = q0.975), alpha = 0.2) +
  ggplot2::geom_line() + ggplot2::labs(x = "Year", y = "Spawning output / unfished output") +
  ggplot2::theme_bw()
```

```{r}
#| label: fig-prior-posterior
#| fig-cap: "Specified log-spawning-output prior and the simulated example's posterior. Densities are on the stored parameter scale."
plot_prior_posterior(sampled)
```

The plotted prior is its specified, untruncated density; bounds may further
restrict the target. Shared-map parameters with multiple prior contributions are
omitted with a warning rather than displayed as a misleading single prior.
SparseNUTS diagnostics can use `opal_as_tmbfit(sampled)` for additional chain plots.

## Profiles and model grids

`opal_profile()` fixes a parameter value and re-optimises the remaining active
parameters. It profiles the **penalised objective**, including priors, unless
those terms are absent. Do not interpret it automatically as a likelihood-ratio
confidence interval. Inspect the saved convergence, gradient, biological checks,
and component contributions for every point.

```{r}
#| label: profile-example
centre <- assessment$fit$parameters$log_B0
assessment <- opal_profile(assessment, "log_B0", centre + seq(-0.5, 0.5, length.out = 9))
knitr::kable(opal_derived(assessment, "profile")$table, digits = 3)
plot_opal_profile(assessment)
```

A grid makes alternative assumptions explicit. Checkpoints are reusable only
when the scientific target, starting state, fitting controls, and validation
contract match. Failed scenarios remain in the result.

```{r}
#| label: model-grid
higher_mortality <- inputs$data
higher_mortality$M <- rep(0.35, inputs$data$n_age)
grid <- opal_grid(opal_obj(inputs$data, inputs$parameters, inputs$map),
  scenarios = list(base = list(), higher_M = list(data = higher_mortality)))
knitr::kable(grid$summary, digits = 3)
```

For a larger analysis, supply `directory = "sensitivity-checkpoints"`. Sampling
and balanced model selection are explicit subsequent steps:

```{r}
#| eval: false
grid <- opal_grid_mcmc(grid, seed = 927, chains = 4, cores = 4,
                       num_warmup = 1000, num_samples = 1000)
selection <- opal_grid_draws(grid, n_per_model = 400, seed = 123)
```

Balanced selection requires a current passing posterior check for **every** grid
member. It samples without replacement, balances chains, and attaches equal
model weights. Equal model weights are a scientific assumption, not a conclusion
from comparing the fitted objectives.

## Equilibrium reference points

`opal_msy()` calculates deterministic equilibrium yield using a stated fleet
allocation, fixed reference-year selectivity, and Beverton–Holt recruitment.
Here the fleets receive equal harvest-intensity weights. This calculation is
independent of the projection engine.

```{r}
#| label: reference-points
assessment <- opal_msy(assessment, fleet_weights = c(0.5, 0.5))
knitr::kable(opal_derived(assessment, "msy")$draws, digits = 3)
```

`u_msy` is a scalar **seasonal harvest intensity**, not an instantaneous annual
fishing mortality. Fleet/age harvest fractions are
`u * fleet_weight[f] * selectivity[f,a]`. Yield uses model weight units, while
`b_msy` uses model spawning-output units. The engine matches the historical
seasonal survival and plus-group convention and solves equilibrium recruitment
without recruitment deviations or positive continuation. Upper-bound maxima
are flagged as unresolved.

```{r}
#| label: posterior-reference-points
# Deliberately small, balanced subset for a quick tutorial calculation.
selected <- unlist(lapply(0:3, function(chain) chain * 1000 + seq(1, 901, by = 100)))
sampled <- opal_msy(sampled, fleet_weights = c(0.5, 0.5),
                    uncertainty = "mcmc", draws = selected)
msy <- opal_derived(sampled, "msy")
knitr::kable(head(msy$draws), digits = 3)
msy$probability_below_bmsy
```

Every selected draw supplies its own biology. This 40-draw result demonstrates
the calculation; use an appropriately sized posterior analysis for inference.
These reference points are conditional on the fleet mix, selectivity, biology,
and deterministic recruitment assumptions. They are not stochastic MSY or
management advice.

Projection development, estimated selectivity time blocks, and close-kin
components are outside the scope of this tutorial and feature addition.
