---
title: "The BET model"
format:
  html:
    embed-resources: false
    html-math-method: mathml
    code-fold: true
    code-summary: "Show code"
    lightbox:
      match: auto
      effect: zoom
execute:
  eval: true
  warning: false
vignette: >
  %\VignetteIndexEntry{The BET model}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{quarto::html}
editor_options: 
  chunk_output_type: console
---

```{r}
#| include: false
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

# Introduction

`opal` is an open, modular R package for fisheries stock assessment built on 
[RTMB](https://github.com/kaskr/RTMB) for automatic differentiation. It includes 
bundled data from the western and central Pacific Ocean (WCPO) bigeye tuna (BET) 
stock assessment. This vignette demonstrates model setup, fitting, and 
diagnostics using those data.

# Load inputs

Load the `opal` package and the `RTMB` dependency. The `ggplot2` package is
used for plotting.

```{r}
#| label: load-pkg
#| echo: true
#| message: false
library(opal)
library(tidyverse)
library(RTMB)

theme_set(theme_bw())
```

The bundled data object `wcpo_bet_data` contains all biological parameters,
catch and CPUE observations, length structure, and prior specifications needed
for the BET assessment model:

```{r}
#| label: load-data
#| echo: true
#| message: false
data(wcpo_bet_data)
data <- wcpo_bet_data
if (is.null(data$n_index)) data$n_index <- 1L
if (!"index" %in% names(data$cpue_data)) data$cpue_data$index <- rep(1L, nrow(data$cpue_data))
names(data)
```

Key dimensions of the data:

```{r}
#| label: data-dims
#| echo: true
cat("Number of ages:", data$n_age, "\n")
cat("Number of years:", data$n_year, "\n")
cat("Number of fisheries:", data$n_fishery, "\n")
cat("Number of length bins:", data$n_len, "\n")
```

## Biological inputs

Natural mortality at age, maturity at length, and length-weight parameters are
all contained in the data object. The two primary biological inputs are plotted
below.

```{r}
#| label: fig-biology
#| echo: true
#| message: false
#| fig-height: 4
#| fig-width: 7
#| fig-cap: Natural mortality-at-age and maturity-at-length used in the BET model.
ages <- data$age_a
real_age <- ages / 4

# Natural mortality
ggplot(data = data.frame(age = real_age, M = data$M), aes(x = age, y = M)) +
  geom_line() + geom_point() +
  scale_y_continuous(limits = c(0, NA), expand = expansion(mult = c(0, 0.05))) +
  labs(x = "Age (years)", y = "M (quarterly rate)", 
       title = "Natural Mortality-at-Age")

# Maturity at length
len_mid <- seq(data$len_bin_start + data$len_bin_width / 2,
               by = data$len_bin_width, length.out = data$n_len)
ggplot(data = data.frame(length = len_mid, maturity = data$maturity), 
       aes(x = length, y = maturity)) +
  geom_line() + geom_point() +
  scale_y_continuous(limits = c(0, NA), expand = expansion(mult = c(0, 0.05))) +
  labs(x = "Length (cm)", y = "Maturity", title = "Maturity-at-Length")

fleet_names <- c("F01_LL.NORTH", "F02_LL.US", "F03_LL.OFFSH",
                 "F04_LL.EQUAT", "F05_LL.WEST", "F06_LL.SOUTH", "F07_LL.AUS",
                 "F08_PS.ASSOC", "F09_PS.UNASS", "F10_DOM.MISC",
                 "F11_DOM.HL", "F12_JP.PS.N", "F13_JP.PL", "F14_EQ.PL",
                 "S01_INDEX")
fleet_palette <- setNames(
  grDevices::hcl.colors(data$n_fishery, palette = "Dark 3"),
  fleet_names
)
```

## CPUE data

```{r}
#| label: prep-cpue
#| echo: true
#| message: false
data$cpue_data$index <- as.integer(factor(data$cpue_data$month, levels = c(2, 5, 8, 11)))
data$n_index <- 4L
```

## Length composition data

Length composition observations are loaded from the bundled `wcpo_bet_lf`
long-format data object and transformed into model-ready arrays.

```{r}
#| label: prep-lf
#| echo: true
#| message: false
data(wcpo_bet_lf)

# Pivot to wide format: one row per fishery x timestep, bins as columns
# Note: LF data exist for fisheries 8-14. Fishery 8 is retained here, but is
# strongly downweighted relative to the other active LF fisheries.
lf_wide <- wcpo_bet_lf %>%
  pivot_wider(
    id_cols = c(fishery, year, month, ts),
    names_from = bin,
    values_from = value,
    values_fill = 0
  ) %>%
  arrange(fishery, ts)

unique(lf_wide$fishery)

# define variance adjustment scalars for each fishery if using lf_switch = 1 (multinomial)
var_adjust_scalars <- rep(1, data$n_fishery)
var_adjust_scalars[] <- 80
var_adjust_scalars[8] <- 5000 # keep LF8 present but much less influential
# var_adjust_scalars <- 1/rep(20000,data$n_fishery)
# var_adjust_scalars[c(1, 4, 5, 6, 15)] <- 1 / 40000
# var_adjust_scalars[] <- 1 / 4e9

data <- prep_lf_data(data = data, lf_wide = lf_wide, 
                     lf_keep_fisheries = c(8, 9, 10, 11, 12, 13, 14),
                     lf_var_adjust = var_adjust_scalars)
lf_year_lookup <- lf_wide %>%
  distinct(ts, year, month)
lf_year_match <- match(data$lf_year, lf_year_lookup$ts)
data$lf_calendar_year <- lf_year_lookup$year[lf_year_match] +
  (lf_year_lookup$month[lf_year_match] - 1) / 12
data$lf_calendar_year_fi <- split(data$lf_calendar_year, data$lf_fishery)

data$lf_fishery
range(data$lf_n) # aim for tens to low thousands per obs
sum(data$lf_n) # aim for ~1e4–1e5 total
```

## Weight composition data

Weight composition observations are loaded from the bundled `wcpo_bet_wf`
long-format data object and transformed into model-ready arrays.

```{r}
#| label: prep-wf
#| echo: true
#| message: false
data(wcpo_bet_wf)

# Weight bin scalars (1 kg bins, 1–200 kg)
data$wt_bin_start <- 1
data$wt_bin_width <- 1
data$n_wt <- 200L

# Pivot to wide format: one row per fishery x timestep, bins as columns
wf_wide <- wcpo_bet_wf |>
  pivot_wider(
    id_cols = c(fishery, year, month, ts),
    names_from = bin,
    values_from = value,
    values_fill = 0
  ) |>
  arrange(fishery, ts)

unique(wf_wide$fishery)

wf_var_adjust_scalars <- rep(2000, data$n_fishery)

data <- prep_wf_data(data = data, wf_wide = wf_wide,
                     wf_keep_fisheries = c(1, 2, 3, 4, 6, 7, 15),
                     wf_switch = 1L,
                     wf_var_adjust = wf_var_adjust_scalars)
wf_year_lookup <- wf_wide %>%
  distinct(ts, year, month)
wf_year_match <- match(data$wf_year, wf_year_lookup$ts)
data$wf_calendar_year <- wf_year_lookup$year[wf_year_match] +
  (wf_year_lookup$month[wf_year_match] - 1) / 12
data$wf_calendar_year_fi <- split(data$wf_calendar_year, data$wf_fishery)
```

# Model setup

## Parameters

Define the initial parameter values. Growth parameters (`log_L1`, `log_L2`,
`log_k`, `log_CV1`, `log_CV2`) and selectivity (`par_sel`) are initialised at
reasonable starting values. Load these from the bundled `wcpo_bet_parameters` 
data object:

```{r}
#| label: get-pars
#| echo: true
#| message: false
data(wcpo_bet_parameters)
par_sel_raw <- as.matrix(wcpo_bet_parameters$par_sel)
par_sel_start <- par_sel_raw

# BET double-normal width starts are stored on the legacy opal width scale.
# The current double-normal uses exp(width) * sd(length)^2, so shift width
# columns by log(sd(length)) to preserve the same natural-scale denominator.
double_normal_f <- data$sel_type_f == 2L
par_sel_start[double_normal_f, 3:4] <- par_sel_start[double_normal_f, 3:4] - log(sd(data$len_mid))

parameters <- list(
  # LF likelihoods are invalid from the bundled log_B0 = 12 start because
  # implied harvest rates can make predicted LF probabilities non-finite.
  log_B0 = 15,
  log_h = as.numeric(wcpo_bet_parameters$log_h),
  log_sigma_r = as.numeric(wcpo_bet_parameters$log_sigma_r),
  log_cpue_q = rep(0, data$n_index),
  cpue_creep = rep(as.numeric(wcpo_bet_parameters$cpue_creep), data$n_index),
  log_cpue_tau = rep(log(0.1), data$n_index),
  log_cpue_omega = rep(as.numeric(wcpo_bet_parameters$log_cpue_omega), data$n_index),
  log_lf_tau = as.numeric(log(rep(0.1, data$n_fishery))),
  log_wf_tau = rep(0, data$n_fishery),
  log_L1 = as.numeric(wcpo_bet_parameters$log_L1),
  log_L2 = as.numeric(wcpo_bet_parameters$log_L2),
  log_k = as.numeric(wcpo_bet_parameters$log_k),
  log_CV1 = as.numeric(wcpo_bet_parameters$log_CV1),
  log_CV2 = as.numeric(wcpo_bet_parameters$log_CV2),
  par_sel = par_sel_start,
  rdev_y = as.numeric(wcpo_bet_parameters$rdev_y)
)
```

The bundled `wcpo_bet_parameters$par_sel` values use the legacy opal width
scale for double-normal selectivity. They are converted to the current
`sel_double_normal()` width scale before fitting. @tbl-selectivity-scale shows
the raw bundled values and the converted fitting starts.

```{r}
#| label: tbl-selectivity-scale
#| echo: true
#| message: false
#| tbl-cap: Raw bundled selectivity parameters and opal-scale fitting starts.
parameter_names <- c("peak", "top", "asc_width", "desc_width", "init", "final")
bind_rows(
  as_tibble(par_sel_raw, .name_repair = ~ parameter_names) %>%
    mutate(fleet = row_number(), scale = "bundled"),
  as_tibble(parameters$par_sel, .name_repair = ~ parameter_names) %>%
    mutate(fleet = row_number(), scale = "opal_start")
) %>%
  relocate(fleet, scale) %>%
  mutate(across(where(is.numeric), ~ round(.x, 3))) %>%
  knitr::kable()
```

## Priors

The BET diagnostic fit is run without parameter priors. This matches the older
diagnostic setup and avoids centering the `log_B0` prior on the deliberately
broad starting value.

```{r}
#| label: get-priors
#| echo: true
#| message: false
data$priors <- list()
0
```

## Parameter map

Use RTMB's `map` option to turn parameters on/off. Parameters mapped to
`factor(NA)` are fixed at their initial values. Recruitment deviations are
estimated for all model time steps. Initial age deviations are not included in
the BET diagnostic fit, so the model starts from equilibrium initial age
structure.

```{r}
#| label: get-map
#| echo: true
#| message: false
parameters$par_sel
wcpo_bet_parameters$par_sel

map_sel <- matrix(NA_integer_, nrow(parameters$par_sel), ncol(parameters$par_sel))

# Retained diagnostic set after sequential LF/WF exploration, using a
# selectivity map closer to SS3's 02-fix-sel setup:
# - LF8 is included with strong downweighting (lf_var_adjust = 5000).
# - WF5 remains inactive; an SS3-style WF5 screen still had large gradients.
# - SS3-fixed width parameters are fixed here too. Several longline final
#   selectivity logits are estimated because SS3 let those right-tail terms move.
next_sel_level <- 1L
free_sel <- function(f, cols) {
  levels <- seq.int(next_sel_level, length.out = length(cols))
  map_sel[f, cols] <<- levels
  next_sel_level <<- next_sel_level + length(cols)
}

free_sel(8, c(1, 3, 4))   # peak, asc-width, desc-width
free_sel(9, c(1, 3))      # peak, asc-width; desc-width fixed as in SS3
free_sel(10, c(1, 3, 4))  # peak, asc-width, desc-width
free_sel(11, 1:2)         # logistic inflection and width
free_sel(12, c(1, 3))     # peak, asc-width; desc-width fixed as in SS3
free_sel(13, c(1, 4))     # peak, desc-width; asc-width fixed as in SS3
free_sel(14, c(1, 3, 4))  # peak, asc-width, desc-width; SS3 desc-width hit HI

free_sel(1, c(1, 4, 6))   # peak, desc-width, final selectivity; asc-width fixed
free_sel(2, c(1, 4, 6))   # peak, desc-width, final selectivity; asc-width fixed
free_sel(3, 1)            # peak only; widths fixed and final logit did not move in SS3
free_sel(4, 1)            # peak only; widths fixed and final logit did not move in SS3
free_sel(6, c(1, 6))      # peak and final selectivity; widths fixed
free_sel(7, c(1, 6))      # peak and final selectivity; widths fixed
free_sel(15, 1:2)         # index-fishery logistic inflection and width

map_sel

map_rdev <- seq_along(parameters$rdev_y)

map_lf_tau <- rep(NA, length(parameters$log_lf_tau))
map_lf_tau[c(8, 9, 10, 11, 12, 13, 14)] <- NA

map <- list(
  # log_B0 = factor(NA),
  log_h = factor(NA),
  log_sigma_r = factor(NA),
  log_cpue_q = factor(seq_len(data$n_index)),
  cpue_creep = factor(rep(NA, data$n_index)),
  log_cpue_tau = factor(rep(NA, data$n_index)),
  log_cpue_omega = factor(rep(NA, data$n_index)),
  log_lf_tau = factor(map_lf_tau),
  log_wf_tau = factor(rep(NA, data$n_fishery)),
  log_L1  = factor(NA),
  log_L2  = factor(NA),
  log_k   = factor(NA),
  log_CV1 = factor(NA),
  log_CV2 = factor(NA),
  par_sel = factor(map_sel),
  rdev_y = factor(map_rdev)
)
```

## Build the AD object

Using the `data`, the `parameters`, the parameter `map`, and the model
(`opal_model`), the AD object is created using RTMB's `MakeADFun` function.
Optionally, we could also specify random effects here 
(e.g. `random = "rdev_y"`), but we'll start with a simpler fixed-effects model 
to check everything is working first.

```{r}
#| label: make-adfun
#| echo: true
#| message: false
# data$lf_switch <- 0 # skip length comps (removal only)
data$lf_switch <- 1 # multinomial likelihood on flat counts (default)
# data$lf_switch <- 2 # fails while fitting, need to sort out log_lf_tau for this, use simulate to tune and find
# data$lf_switch <- 3 # fails - wants integers - think this should be an issue to RTMBdist guys

# Note: wf_switch was already set by prep_wf_data(); it is re-stated here for
# clarity alongside the equivalent lf_switch assignment.
data$wf_switch <- 1 # multinomial likelihood on flat counts (default)
# data$wf_switch <- 0 # skip weight comps (removal only)
# data$wf_switch <- 2 # Dirichlet
# data$wf_switch <- 3 # Dirichlet-multinomial

obj <- MakeADFun(func = cmb(opal_model, data), parameters = parameters, map = map)
obj$env$tracemgc <- FALSE
obj$par
names(obj$par)
unique(names(obj$par))
obj$fn()
obj$gr()
# TMB::benchmark(obj, n = 500, expr = expression(obj$gr()))
```

Inspect initial model outputs. The initial biomass trajectory is shown in
@fig-init-biomass, and the predicted catch checks are shown in @fig-init-catch.

```{r}
#| label: fig-init-biomass
#| echo: true
#| message: false
#| fig-height: 4
#| fig-width: 7
#| fig-cap: Initial spawning biomass trajectory before optimization.
plot(obj$report()$spawning_biomass_y, type = "l",
     xlab = "Time step", ylab = "Spawning biomass (mt)",
     main = "Initial spawning biomass trajectory")
```

```{r}
#| label: fig-init-catch
#| echo: true
#| message: false
#| fig-height: 4
#| fig-width: 7
#| fig-cap: Initial predicted catch by fleet compared with observed catch.
plot_catch(data = data, obj = obj)
```

## Parameter bounds

```{r}
#| label: get-par-bounds
#| echo: true
#| message: false
Lwr <- rep(-Inf, length(obj$par))
Upr <- rep(Inf, length(obj$par))
Lwr[grep("log_B0", names(obj$par))] <- 12
Upr[grep("log_B0", names(obj$par))] <- 22
Lwr[grep("log_cpue_q", names(obj$par))] <- log(0.1)
Upr[grep("log_cpue_q", names(obj$par))] <- log(10)
Lwr[grep("log_lf_tau", names(obj$par))] <- rep(-9, length(grep("log_lf_tau", names(obj$par))))
Upr[grep("log_lf_tau", names(obj$par))] <- rep(9, length(grep("log_lf_tau", names(obj$par))))
Lwr[grep("rdev_y", names(obj$par))] <- rep(-5, length(grep("rdev_y", names(obj$par))))
Upr[grep("rdev_y", names(obj$par))] <- rep(5, length(grep("rdev_y", names(obj$par))))

mu_len <- mean(data$len_mid)
sd_len <- sd(data$len_mid)

par_labels <- names(obj$par)
par_sel_pos <- which(names(obj$par) == "par_sel")
par_sel_level_pos <- setNames(par_sel_pos, levels(map$par_sel))

for (f in seq_len(data$n_fishery)) {
  idx <- rep(NA_integer_, ncol(map_sel))
  for (p in seq_len(ncol(map_sel))) {
    level <- as.character(map_sel[f, p])
    if (!is.na(level)) {
      idx[p] <- par_sel_level_pos[[level]]
      par_labels[idx[p]] <- paste0("par_sel[", f, ",", p, "]")
    }
  }

  if (data$sel_type_f[f] == 1L) {
    # Logistic: inflection [5,200], width [4,500] cm
    if (!is.na(idx[1])) {
      Lwr[idx[1]] <- (5   - mu_len) / sd_len
      Upr[idx[1]] <- (200 - mu_len) / sd_len
    }

    if (!is.na(idx[2])) {
      Lwr[idx[2]] <- log(4    / sd_len)
      Upr[idx[2]] <- log(500  / sd_len)
    }
  } else {
    # Double-normal SS3 bounds
    if (!is.na(idx[1])) {
      Lwr[idx[1]] <- (10.1 - mu_len) / sd_len
      Upr[idx[1]] <- (200  - mu_len) / sd_len
    }

    if (!is.na(idx[2])) {
      Lwr[idx[2]] <- -7
      Upr[idx[2]] <-  7
    }

    # for sel_double_normal using exp(c)*sd^2 and exp(d)*sd^2:
    # keep widths positive on a practical natural scale while avoiding
    # artificial convergence on the transformed denominator bound.
    shift <- 2 * log(sd_len)
    width_log_lower <- log(4)
    width_log_upper <- 8
    if (!is.na(idx[3])) {
      Lwr[idx[3]] <- width_log_lower - shift
      Upr[idx[3]] <-  width_log_upper - shift
    }
    if (!is.na(idx[4])) {
      Lwr[idx[4]] <- width_log_lower - shift
      Upr[idx[4]] <-  width_log_upper - shift
    }

    if (!is.na(idx[5])) {
      Lwr[idx[5]] <- -9 # practical bound (instead of -999)
      Upr[idx[5]] <-  9
    }
    if (!is.na(idx[6])) {
      Lwr[idx[6]] <- -9
      Upr[idx[6]] <-  9
    }
  }
}

bounds <- data.frame(par = par_labels, lower = Lwr, upper = Upr)
```

# Optimisation

Optimise using the `nlminb` function:

```{r}
#| label: run-nlminb
#| echo: true
#| results: hide
#| eval: true
control <- list(eval.max = 10000, iter.max = 10000)
# control <- list(eval.max = 100, iter.max = 100)
obj$env$tracemgc <- FALSE
opt <- list(par = obj$par)
fit_grad <- Inf
for (i in seq_len(4)) {
  # For this expanded BET fit, forming obj$he dominates runtime and makes the
  # vignette render impractical; the diagnostic scripts use gradient-only nlminb.
  opt <- nlminb(start = opt$par, objective = obj$fn, gradient = obj$gr,
                lower = Lwr, upper = Upr, control = control)
  fit_grad <- max(abs(obj$gr(opt$par)))
  if (fit_grad < 1e-2) break
}
fit_nll <- obj$fn(opt$par)
obj$env$last.par.best <- opt$par
fit_grad <- max(abs(obj$gr(opt$par)))
fit_kkt_grad <- {
  grad <- obj$gr(opt$par)
  grad_constrained <- abs(grad)
  at_lower <- is.finite(Lwr) & opt$par <= Lwr + 1e-7
  at_upper <- is.finite(Upr) & opt$par >= Upr - 1e-7
  grad_constrained[at_lower] <- pmax(0, -grad[at_lower])
  grad_constrained[at_upper] <- pmax(0, grad[at_upper])
  max(grad_constrained)
}
fit_rep <- obj$report(obj$env$last.par.best)
# TMBhelper::check_estimability(obj = obj)
if (fit_grad < 1e-2) {
  check_estimability(obj = obj)
} else {
  message("Skipping check_estimability(); max |gradient| = ", signif(fit_grad, 4), ". Improve optimization first.")
}
```

```{r}
#| label: tbl-fit-summary
#| echo: true
#| message: false
#| tbl-cap: Optimization and likelihood-component summary for the fitted BET model.
tibble(
  metric = c(
    "nlminb convergence",
    "objective",
    "max gradient",
    "max KKT gradient",
    "B0",
    "lp_prior",
    "lp_penalty",
    "lp_rec",
    "lp_cpue",
    "lp_lf",
    "lp_wf"
  ),
  value = c(
    opt$convergence,
    fit_nll,
    fit_grad,
    fit_kkt_grad,
    fit_rep$B0,
    fit_rep$lp_prior,
    fit_rep$lp_penalty,
    fit_rep$lp_rec,
    sum(fit_rep$lp_cpue),
    sum(fit_rep$lp_lf),
    sum(fit_rep$lp_wf)
  )
) %>%
  mutate(value = signif(value, 6)) %>%
  knitr::kable()
```

Compare initial and estimated parameter values:

```{r}
#| label: opt-results
#| echo: true
#| message: false
#| eval: true
par_table <- get_par_table(obj, parameters, map, lower = Lwr, upper = Upr, grad_tol = 1e-2)
par_table
```

The selectivity diagnostics below isolate the active selectivity parameters with
large raw gradients or estimates close to bounds. A large raw gradient on a
bound usually indicates that the constrained optimum wants to move farther in
that direction; the KKT-style gradient is the relevant convergence check under
active bounds.

```{r}
#| label: tbl-selectivity-diagnostics
#| echo: true
#| message: false
#| tbl-cap: Active selectivity parameters with large gradients or bound checks.
sel_component <- c(
  "peak/inflection",
  "top/width",
  "ascending width",
  "descending width",
  "initial selectivity",
  "final selectivity"
)

sel_units <- function(f, p) {
  if (data$sel_type_f[f] == 1L) {
    return(c("cm", "cm", NA, NA, NA, NA)[p])
  }
  c("cm", "logit", "cm^2 denominator", "cm^2 denominator",
    "proportion", "proportion")[p]
}

sel_to_natural <- function(x, f, p) {
  out <- rep(NA_real_, length(x))
  finite <- is.finite(x)
  if (!any(finite)) return(out)

  if (data$sel_type_f[f] == 1L) {
    if (p == 1L) out[finite] <- mu_len + x[finite] * sd_len
    if (p == 2L) out[finite] <- exp(x[finite]) * sd_len
  } else {
    if (p == 1L) out[finite] <- mu_len + x[finite] * sd_len
    if (p == 2L) out[finite] <- x[finite]
    if (p %in% 3:4) out[finite] <- exp(x[finite] + 2 * log(sd_len))
    if (p %in% 5:6) out[finite] <- plogis(x[finite])
  }
  out
}

sel_issue_label <- function(component, bound_status, raw_bad) {
  if (bound_status == "lower" && grepl("width", component)) {
    return("minimum width bound")
  }
  if (bound_status == "upper" && grepl("width", component)) {
    return("maximum width bound")
  }
  if (bound_status == "lower") {
    return("lower bound")
  }
  if (bound_status == "upper") {
    return("upper bound")
  }
  if (raw_bad) {
    return("interior raw gradient")
  }
  "ok"
}

sel_diag <- map_dfr(seq_len(nrow(map_sel)), function(f) {
  map_dfr(seq_len(ncol(map_sel)), function(p) {
    level <- as.character(map_sel[f, p])
    if (is.na(level)) return(tibble())

    idx <- par_sel_level_pos[[level]]
    estimate <- obj$env$last.par.best[idx]
    gradient <- obj$gr(obj$env$last.par.best)[idx]
    lower <- Lwr[idx]
    upper <- Upr[idx]
    bound_status <- case_when(
      is.finite(lower) && estimate <= lower + 1e-7 ~ "lower",
      is.finite(upper) && estimate >= upper - 1e-7 ~ "upper",
      TRUE ~ "free"
    )
    kkt_gradient <- case_when(
      bound_status == "lower" ~ max(0, -gradient),
      bound_status == "upper" ~ max(0, gradient),
      TRUE ~ abs(gradient)
    )
    raw_bad <- abs(gradient) > 1e-2

    tibble(
      fishery = f,
      fleet = fleet_names[f],
      type = if_else(data$sel_type_f[f] == 1L, "logistic", "double-normal"),
      component = sel_component[p],
      estimate = estimate,
      lower = lower,
      upper = upper,
      gradient = gradient,
      kkt_gradient = kkt_gradient,
      bound_status = bound_status,
      natural_estimate = sel_to_natural(estimate, f, p),
      natural_lower = sel_to_natural(lower, f, p),
      natural_upper = sel_to_natural(upper, f, p),
      natural_units = sel_units(f, p),
      diagnostic = sel_issue_label(sel_component[p], bound_status, raw_bad)
    )
  })
})

sel_diag_issues <- sel_diag %>%
  filter(abs(gradient) > 1e-2 | bound_status != "free") %>%
  arrange(fishery, component) %>%
  mutate(across(
    c(estimate, lower, upper, gradient, kkt_gradient,
      natural_estimate, natural_lower, natural_upper),
    ~ signif(.x, 4)
  ))

if (nrow(sel_diag_issues) == 0L) {
  cat("No active selectivity parameters have large gradients or bound checks.\n")
} else {
  sel_diag_issues %>%
    select(fishery, fleet, type, component, estimate, lower, upper,
           gradient, kkt_gradient, bound_status, natural_estimate,
           natural_lower, natural_upper, natural_units, diagnostic) %>%
    knitr::kable()
}
```

Inspect estimated recruitment deviations:

```{r}
#| label: fig-rdev
#| echo: true
#| message: false
#| fig-height: 6
#| fig-width: 9
#| fig-cap: Estimated recruitment deviations split by assessment season.
rdev_est <- obj$env$parList(obj$env$last.par.best)$rdev_y
rdev_month_label <- function(month) {
  out <- paste("month", month)
  valid <- month %in% seq_along(month.abb)
  out[valid] <- month.abb[month[valid]]
  out
}

rdev_time_lookup <- NULL
if (all(c("ts", "year", "month") %in% names(data$cpue_data))) {
  rdev_time_lookup <- data$cpue_data %>%
    distinct(time_step = ts, calendar_year = year, month) %>%
    arrange(time_step)
}

if (!is.null(rdev_time_lookup) &&
    nrow(rdev_time_lookup) == length(rdev_est) &&
    all(rdev_time_lookup$time_step == seq_along(rdev_est))) {
  rdev_season_lookup <- rdev_time_lookup %>%
    distinct(month) %>%
    arrange(month) %>%
    mutate(
      season = row_number(),
      season_label = paste0("Season ", season, " (", rdev_month_label(month), ")")
    )

  rdev_df <- rdev_time_lookup %>%
    left_join(rdev_season_lookup, by = "month")
} else {
  # BET stores quarters as sequential model time steps in this data object
  # (n_season = 1), so recover seasons from the quarterly ordering.
  rdev_n_season <- 4L
  rdev_months <- if ("month" %in% names(data$cpue_data)) {
    sort(unique(data$cpue_data$month))
  } else {
    seq_len(rdev_n_season)
  }
  if (length(rdev_months) != rdev_n_season) rdev_months <- seq_len(rdev_n_season)
  rdev_first_year <- if ("year" %in% names(data$cpue_data)) {
    min(data$cpue_data$year, na.rm = TRUE)
  } else {
    data$first_yr
  }

  rdev_season_lookup <- tibble(
    season = seq_len(rdev_n_season),
    month = rdev_months,
    season_label = paste0("Season ", season, " (", rdev_month_label(month), ")")
  )
  rdev_df <- tibble(
    time_step = seq_along(rdev_est),
    calendar_year = rdev_first_year + (time_step - 1L) %/% rdev_n_season,
    season = ((time_step - 1L) %% rdev_n_season) + 1L
  ) %>%
    left_join(rdev_season_lookup, by = "season")
}

rdev_df <- rdev_df %>%
  mutate(
    season_label = factor(season_label, levels = rdev_season_lookup$season_label),
    rdev = as.numeric(rdev_est),
    estimated = !is.na(map_rdev)
  )

ggplot(rdev_df, aes(x = calendar_year, y = rdev)) +
  geom_hline(yintercept = 0, linewidth = 0.3, colour = "grey40") +
  geom_line() +
  geom_point(aes(colour = estimated), size = 0.8) +
  facet_wrap(~ season_label, ncol = 2) +
  scale_colour_manual(values = c(`FALSE` = "grey40", `TRUE` = "firebrick")) +
  scale_x_continuous(breaks = scales::breaks_pretty(n = 6)) +
  labs(x = "Calendar year", y = "Recruitment deviation", colour = "Estimated")
```

## Selectivity

Visualise the selectivity-at-length curves by fleet in @fig-selectivity-length.

```{r}
#| label: fig-selectivity-length
#| echo: true
#| message: false
#| fig-height: 8
#| fig-width: 10
#| fig-cap: Selectivity-at-length by fleet. Coloured lines use the current parameter values; dashed black lines show the bundled reference.
par_sel_est <- obj$env$parList(obj$env$last.par.best)$par_sel
par_sel_ref <- parameters$par_sel

map_sel_mat <- matrix(map$par_sel, nrow = nrow(par_sel_est), ncol = ncol(par_sel_est))
fishery_estimated <- apply(map_sel_mat, 1, function(row) any(!is.na(row)))

sel_est_df <- do.call(rbind, lapply(seq_len(data$n_fishery), function(f) {
  sel_l <- if (data$sel_type_f[f] == 1L) {
    sel_logistic(len_mid, par_sel_est[f, ])
  } else {
    sel_double_normal(len_mid, par_sel_est[f, ])
  }
  data.frame(
    fishery = f,
    fleet_name = fleet_names[f],
    length = len_mid,
    selectivity = as.numeric(sel_l),
    estimated = fishery_estimated[f]
  )
}))

sel_orig_df <- do.call(rbind, lapply(seq_len(data$n_fishery), function(f) {
  sel_l <- if (data$sel_type_f[f] == 1L) {
    sel_logistic(len_mid, par_sel_ref[f, ])
  } else {
    sel_double_normal(len_mid, par_sel_ref[f, ])
  }
  data.frame(
    fishery = f,
    fleet_name = fleet_names[f],
    length = len_mid,
    selectivity = as.numeric(sel_l)
  )
}))

ggplot(data = sel_est_df, aes(x = length, y = selectivity, colour = estimated)) +
  geom_line() +
  geom_line(
    data = sel_orig_df,
    aes(x = length, y = selectivity),
    inherit.aes = FALSE,
    colour = "black",
    linetype = "dashed",
    linewidth = 0.7
  ) +
  facet_wrap(fleet_name ~ ., ncol = 4) +
  labs(x = "Length (cm)", y = "Selectivity") +
  coord_cartesian(ylim = c(0, 1))
```

The corresponding selectivity-at-age curves are shown in @fig-selectivity-age.
For comparison with SS3's reported age-selectivity output, the plotted curves
are scaled to a maximum of 1 within each fleet. The model still uses the
unscaled PLA-integrated values in `rep$sel_fya`.

```{r}
#| label: fig-selectivity-age
#| echo: true
#| message: false
#| fig-height: 8
#| fig-width: 10
#| fig-cap: Reported selectivity-at-age by fleet after integrating length selectivity over the age-length key and scaling each fleet to a maximum of 1 for SS3-style plotting.
rep <- obj$report(obj$env$last.par.best)
sel_fya <- rep$sel_fya

sel_df <- expand.grid(fishery = 1:data$n_fishery, age = 1:data$n_age)
sel_df$selectivity <- sapply(1:nrow(sel_df), function(i) {
  sel_fya[sel_df$fishery[i], 1, sel_df$age[i]]
})
sel_df$real_age <- sel_df$age / 4
sel_df$fleet_name <- fleet_names[sel_df$fishery]
sel_df <- sel_df %>%
  group_by(fishery) %>%
  mutate(selectivity_reported = selectivity / max(selectivity, na.rm = TRUE)) %>%
  ungroup()

ggplot(sel_df, aes(x = real_age, y = selectivity_reported)) +
  geom_line() +
  facet_wrap(~fleet_name, ncol = 4) +
  labs(x = "Age (years)", y = "Selectivity") +
  ylim(0, 1)
```

@tbl-sel-par summarises the selectivity parameter values and the map status.

```{r}
#| label: tbl-sel-par
#| echo: true
#| message: false
#| tbl-cap: Selectivity parameter values, bundled reference values, and map status.
# Three sources to compare side by side:
#   1. Initial values used to build obj   (parameters$par_sel)
#   2. Bundled reference values            (wcpo_bet_parameters$par_sel)
#   3. Estimated / mapped status           (map$par_sel)
#
# par_sel is [n_fishery x 6] with columns (a, b, c, d, e, f) for double-normal
# or (par1, par2) for logistic (cols 3-6 ignored). `factor(map_sel)` flattens
# matrices column-major, so wrap the map vector back into the same shape.

n_f <- nrow(parameters$par_sel)
n_p <- ncol(parameters$par_sel)
col_lbls <- c("a (peak)", "b (plateau)", "c (asc width)", "d (desc width)", "e (init sel)", "f (final sel)")

# Estimated/fixed status from the map
map_sel_mat <- if (is.null(map$par_sel)) {
  matrix(seq_len(n_f * n_p), n_f, n_p) # everything estimated if no map
} else {
  matrix(map$par_sel, n_f, n_p)
}
estimated_mat <- !is.na(map_sel_mat)

# Reference values (may not have row/col names — assume same shape)
ref_mat <- as.matrix(wcpo_bet_parameters$par_sel)
stopifnot("wcpo_bet_parameters$par_sel must match parameters$par_sel shape" = all(dim(ref_mat) == c(n_f, n_p)))

# Selectivity type (logistic = 1, double-normal = 2)
sel_type_lbl <- ifelse(data$sel_type_f == 1L, "logistic", "double-normal")

# For logistic fisheries, columns 3-6 are unused -> NA-out for clarity
init_mat <- parameters$par_sel
for (f in seq_len(n_f)) {
  if (data$sel_type_f[f] == 1L) {
    init_mat[f, 3:6] <- NA
    ref_mat[f, 3:6] <- NA
    estimated_mat[f, 3:6] <- FALSE
  }
}

# Assemble long table
sel_par_tbl <- tibble(
  fishery    = rep(seq_len(n_f), times = n_p),
  fleet_name = rep(fleet_names[seq_len(n_f)], times = n_p),
  sel_type   = rep(sel_type_lbl, times = n_p),
  param      = rep(col_lbls, each = n_f),
  init       = as.numeric(init_mat),
  reference  = as.numeric(ref_mat),
  estimated  = as.vector(estimated_mat)
) %>%
  mutate(diff = init - reference) %>%
  arrange(fishery, match(param, col_lbls))

sel_par_tbl %>%
  mutate(
    init      = ifelse(is.na(init),      "—", formatC(init,      digits = 3, format = "f")),
    reference = ifelse(is.na(reference), "—", formatC(reference, digits = 3, format = "f")),
    diff      = ifelse(is.na(diff),      "—", formatC(diff,      digits = 3, format = "f")),
    estimated = ifelse(estimated, "✓", "")
  ) %>%
  select(Fleet = fleet_name,
         Type = sel_type,
         Param = param,
         Initial = init,
         `Reference (bundled)` = reference,
         `Δ (init − ref)` = diff,
         `Est?` = estimated) %>%
  knitr::kable(align   = c("l", "l", "l", "r", "r", "r", "c"))
```

## Diagnostics

Check that all parameters are estimable using the `check_estimability` function:

```{r}
#| label: check-estimability
#| echo: true
#| message: false
#| error: true
#| eval: true
obj$env$tracemgc <- FALSE
fit_grad <- max(abs(obj$gr(obj$env$last.par.best)))
if (fit_grad < 1e-2) {
  check_estimability(obj = obj)
} else {
  message("Skipping check_estimability(); max |gradient| = ", signif(fit_grad, 4),
          ". Improve optimization first.")
}
```

Look for any parameters with high correlations (absolute correlation > 0.95) using the `get_cor_pairs` function:

```{r}
#| label: check-cor
#| echo: true
#| message: false
#| eval: true
get_cor_pairs(obj = obj, threshold = 0.95)
```

Parameter standard errors can be computed with `sdreport(obj)` when needed. It
is not run here because the diagnostic figures below use fitted trajectories
directly from `obj$report(obj$env$last.par.best)`.

## Data fits

Inspect fitted CPUE and catch. CPUE fits are shown in @fig-cpue-fit; catch fits
and residuals are plotted below.

```{r}
#| label: fig-fleet-palette
#| echo: false
#| message: false
#| fig-height: 2.4
#| fig-width: 7
#| fig-cap: Fleet colour palette used for fishery-level diagnostic plots.
palette_df <- tibble(
  fishery = seq_len(data$n_fishery),
  fleet_name = fleet_names,
  fill = unname(fleet_palette),
  col = (fishery - 1) %% 5,
  row = -((fishery - 1) %/% 5)
)
palette_rgb <- grDevices::col2rgb(palette_df$fill) / 255
palette_df$text_colour <- ifelse(
  0.299 * palette_rgb[1, ] + 0.587 * palette_rgb[2, ] + 0.114 * palette_rgb[3, ] > 0.55,
  "black",
  "white"
)

ggplot(palette_df, aes(x = col, y = row)) +
  geom_tile(aes(fill = fill), colour = "white", linewidth = 0.5) +
  geom_text(
    aes(label = paste0(fishery, "\n", fleet_name), colour = text_colour),
    size = 2.5,
    lineheight = 0.85
  ) +
  scale_fill_identity() +
  scale_colour_identity() +
  scale_x_continuous(expand = c(0, 0), limits = c(-0.5, 4.5)) +
  scale_y_continuous(expand = c(0, 0), limits = c(-2.5, 0.5)) +
  theme_void()
```

```{r}
#| label: fig-cpue-fit
#| echo: true
#| message: false
#| fig-height: 4
#| fig-width: 7
#| fig-cap: Observed and predicted CPUE by fishery and quarter, with simulated replicate series.
if (data$cpue_switch == 0) {
  cat("CPUE likelihood is switched off (cpue_switch = 0), so no predicted CPUE to show.\n")
} else {
  rep <- obj$report(obj$env$last.par.best)
  
  df <- data$cpue_data %>%
    mutate(obs_id = row_number(), pred = rep$cpue_pred, sigma = rep$cpue_sigma)
  
  # Simulations
  n_sim <- 0
  df_sim <- if (n_sim > 0) {
    map_dfr(seq_len(n_sim), function(s) {
      sim_log <- obj$simulate(obj$env$last.par.best)$cpue_log_obs
      tibble(
        obs_id = seq_along(sim_log),
        year = data$cpue_data$year,
        month = data$cpue_data$month,
        fishery = data$cpue_data$fishery,
        sim_id = s,
        sim = exp(sim_log)
      )
    })
  } else {
    tibble(
      obs_id = integer(),
      year = numeric(),
      month = integer(),
      fishery = integer(),
      sim_id = integer(),
      sim = numeric()
    )
  }
  
  month_levels <- sort(unique(data$cpue_data$month))
  fish_levels <- sort(unique(data$cpue_data$fishery))
  format_facets <- function(d) {
    d %>%
      mutate(
        month = factor(month, levels = month_levels, labels = paste0("Month: ", month_levels)),
        fishery = factor(fishery, levels = fish_levels, labels = paste0("Fishery: ", fish_levels))
      )
  }
  df <- format_facets(df)
  df_sim <- format_facets(df_sim)
  
  ggplot(data = df, aes(x = year)) +
    {if (n_sim > 0) geom_line(data = df_sim, aes(y = sim, group = sim_id),
                              colour = "lightblue", alpha = 0.5, linewidth = 0.3)} +
    geom_linerange(aes(ymin = exp(log(value) - sigma),
                       ymax = exp(log(value) + sigma),
                       colour = "Observed"), alpha = 0.6) +
    geom_point(aes(y = value, colour = "Observed")) +
    geom_line(aes(y = pred, colour = "Predicted"), linewidth = 0.8) +
    facet_grid(fishery ~ month, scales = "free_y") +
    scale_y_continuous(limits = c(0, NA), expand = expansion(mult = c(0, 0.05))) +
    scale_x_continuous(breaks = scales::pretty_breaks()) +
    scale_colour_manual(values = c(Observed = "black", Predicted = "red3")) +
    labs(x = "Year", y = "CPUE", colour = NULL,
         subtitle = if (n_sim > 0) paste0(n_sim, " fitted-model simulations (grey)") else NULL) +
    theme(legend.position = "bottom")
}
```

```{r}
#| label: fig-catch-fit
#| echo: true
#| message: false
#| fig-height: 4
#| fig-width: 7
#| fig-cap: Predicted catch and catch residual diagnostics for the initial model state.
sum(fit_rep$catch_pred_ysf - data$catch_obs_ysf)
plot_catch(data = data, obj = obj)
plot_catch(data = data, obj = obj, plot_resid = TRUE)
```

Inspect predicted vs observed length compositions when the length-composition
likelihood is active.

```{r}
#| label: fig-lf-diagnostics
#| echo: true
#| message: false
#| fig-height: 10
#| fig-width: 9
#| fig-cap: Mean observed, predicted, and simulated length composition proportions for active length-composition fisheries.
# Convert model timestep -> "YYYY Qn" label
ts_to_label <- function(ts, first_yr = 1952, n_season = 4) {
  yr <- first_yr + (ts - 1L) %/% n_season
  qn <- ((ts - 1L) %% n_season) + 1L
  paste0(yr, " Q", qn)
}

if (data$lf_switch == 0) {
  cat("Length composition likelihood is switched off (lf_switch = 0), so no predicted length compositions to show.\n")
} else {
  lf_rep <- obj$report(obj$env$last.par.best)
  lf_pred <- lf_rep$lf_pred
  names(lf_pred) <- data$lf_fishery_f

  flat_to_comp_list <- function(v, n_f, fishery_f, minbin, maxbin) {
    out <- vector("list", length(n_f))
    offset <- 0L
    for (k in seq_along(n_f)) {
      f <- fishery_f[k]
      nbins <- maxbin[f] - minbin[f] + 1L
      n_values <- n_f[k] * nbins
      m <- matrix(v[seq.int(offset + 1L, length.out = n_values)],
                  nrow = n_f[k], ncol = nbins, byrow = TRUE)
      rs <- rowSums(m)
      rs[rs == 0] <- 1
      out[[k]] <- m / rs
      offset <- offset + n_values
    }
    out
  }

  comp_to_lf_long <- function(comp_list, fishery_f) {
    map_dfr(seq_along(comp_list), function(i) {
      f <- fishery_f[i]
      bins <- data$len_mid[data$lf_minbin[f]:data$lf_maxbin[f]]
      m <- comp_list[[i]]
      colnames(m) <- bins
      as.data.frame.table(m, responseName = "proportion",
                          stringsAsFactors = FALSE) %>%
        transmute(
          fishery = f,
          length = as.numeric(Var2),
          proportion = as.numeric(proportion)
        )
    })
  }

  for (k in seq_along(lf_pred)) {
    f <- data$lf_fishery_f[k]
    colnames(lf_pred[[k]]) <- data$len_mid[data$lf_minbin[f]:data$lf_maxbin[f]]
  }

  lf_obs_vec <- switch(as.character(data$lf_switch),
                       `1` = data$lf_obs_flat,
                       `2` = data$lf_obs_prop,
                       `3` = data$lf_obs_ints)
  lf_obs_list <- flat_to_comp_list(lf_obs_vec, data$lf_n_f, data$lf_fishery_f,
                                   data$lf_minbin, data$lf_maxbin)

  df_pred_mean <- comp_to_lf_long(lf_pred, data$lf_fishery_f) %>%
    group_by(fishery, length) %>%
    summarise(proportion = mean(proportion, na.rm = TRUE), .groups = "drop") %>%
    mutate(series = "Predicted")

  df_obs_mean <- comp_to_lf_long(lf_obs_list, data$lf_fishery_f) %>%
    group_by(fishery, length) %>%
    summarise(proportion = mean(proportion, na.rm = TRUE), .groups = "drop") %>%
    mutate(series = "Observed")

  n_sim <- 5
  df_sim_mean <- map_dfr(seq_len(n_sim), function(s) {
    sim <- obj$simulate(obj$env$last.par.best)
    sim_vec <- switch(as.character(data$lf_switch),
                      `1` = sim$lf_obs_flat,
                      `2` = sim$lf_obs_prop,
                      `3` = sim$lf_obs_ints)
    sim_list <- flat_to_comp_list(sim_vec, data$lf_n_f, data$lf_fishery_f,
                                  data$lf_minbin, data$lf_maxbin)
    comp_to_lf_long(sim_list, data$lf_fishery_f) %>%
      group_by(fishery, length) %>%
      summarise(proportion = mean(proportion, na.rm = TRUE), .groups = "drop") %>%
      mutate(sim_id = s)
  })

  fishery_levels <- paste("Fishery", sort(unique(data$lf_fishery_f)))
  df_lf_mean <- bind_rows(df_obs_mean, df_pred_mean) %>%
    mutate(fishery = factor(paste("Fishery", fishery), levels = fishery_levels))
  df_sim_mean <- df_sim_mean %>%
    mutate(fishery = factor(paste("Fishery", fishery), levels = fishery_levels))

  ggplot() +
    geom_col(data = filter(df_lf_mean, series == "Observed"),
             aes(x = length, y = proportion), fill = "grey75", width = 1.8) +
    geom_line(data = df_sim_mean,
              aes(x = length, y = proportion, group = sim_id),
              colour = "steelblue", alpha = 0.35, linewidth = 0.35) +
    geom_line(data = filter(df_lf_mean, series == "Predicted"),
              aes(x = length, y = proportion), colour = "red3", linewidth = 0.8) +
    facet_wrap(~ fishery, ncol = 2, scales = "free_y") +
    scale_y_continuous(limits = c(0, NA), expand = expansion(mult = c(0, 0.05))) +
    labs(x = "Length (cm)", y = "Mean proportion") +
    theme_minimal()
}
```

Inspect predicted vs observed weight compositions when the weight-composition
likelihood is enabled.

```{r}
#| label: fig-wf-diagnostics
#| echo: true
#| message: false
#| fig-height: 10
#| fig-width: 9
#| fig-cap: Mean observed, predicted, and simulated weight composition proportions for active weight-composition fisheries.
if (data$wf_switch == 0) {
  cat("Weight composition likelihood is switched off (wf_switch = 0), so no predicted weight compositions to show.\n")
} else {
  wf_rep <- obj$report(obj$env$last.par.best)
  wf_pred <- wf_rep$wf_pred
  names(wf_pred) <- data$wf_fishery_f

  flat_to_comp_list <- function(v, n_f, fishery_f, minbin, maxbin) {
    out <- vector("list", length(n_f))
    offset <- 0L
    for (k in seq_along(n_f)) {
      f <- fishery_f[k]
      nbins <- maxbin[f] - minbin[f] + 1L
      n_values <- n_f[k] * nbins
      m <- matrix(v[seq.int(offset + 1L, length.out = n_values)],
                  nrow = n_f[k], ncol = nbins, byrow = TRUE)
      rs <- rowSums(m)
      rs[rs == 0] <- 1
      out[[k]] <- m / rs
      offset <- offset + n_values
    }
    out
  }

  comp_to_wf_long <- function(comp_list, fishery_f) {
    map_dfr(seq_along(comp_list), function(i) {
      f <- fishery_f[i]
      bins <- data$wt_mid[data$wf_minbin[f]:data$wf_maxbin[f]]
      m <- comp_list[[i]]
      colnames(m) <- bins
      as.data.frame.table(m, responseName = "proportion",
                          stringsAsFactors = FALSE) %>%
        transmute(
          fishery = f,
          weight = as.numeric(Var2),
          proportion = as.numeric(proportion)
        )
    })
  }

  for (k in seq_along(wf_pred)) {
    f <- data$wf_fishery_f[k]
    colnames(wf_pred[[k]]) <- data$wt_mid[data$wf_minbin[f]:data$wf_maxbin[f]]
  }

  wf_obs_vec <- switch(as.character(data$wf_switch),
                       `1` = data$wf_obs_flat,
                       `2` = data$wf_obs_prop,
                       `3` = data$wf_obs_ints)
  wf_obs_list <- flat_to_comp_list(wf_obs_vec, data$wf_n_f, data$wf_fishery_f,
                                   data$wf_minbin, data$wf_maxbin)

  df_wf_pred_mean <- comp_to_wf_long(wf_pred, data$wf_fishery_f) %>%
    group_by(fishery, weight) %>%
    summarise(proportion = mean(proportion, na.rm = TRUE), .groups = "drop") %>%
    mutate(series = "Predicted")

  df_wf_obs_mean <- comp_to_wf_long(wf_obs_list, data$wf_fishery_f) %>%
    group_by(fishery, weight) %>%
    summarise(proportion = mean(proportion, na.rm = TRUE), .groups = "drop") %>%
    mutate(series = "Observed")

  n_sim <- 5
  df_wf_sim_mean <- map_dfr(seq_len(n_sim), function(s) {
    sim <- obj$simulate(obj$env$last.par.best)
    sim_vec <- switch(as.character(data$wf_switch),
                      `1` = sim$wf_obs_flat,
                      `2` = sim$wf_obs_prop,
                      `3` = sim$wf_obs_ints)
    sim_list <- flat_to_comp_list(sim_vec, data$wf_n_f, data$wf_fishery_f,
                                  data$wf_minbin, data$wf_maxbin)
    comp_to_wf_long(sim_list, data$wf_fishery_f) %>%
      group_by(fishery, weight) %>%
      summarise(proportion = mean(proportion, na.rm = TRUE), .groups = "drop") %>%
      mutate(sim_id = s)
  })

  fishery_levels <- paste("Fishery", sort(unique(data$wf_fishery_f)))
  df_wf_mean <- bind_rows(df_wf_obs_mean, df_wf_pred_mean) %>%
    mutate(fishery = factor(paste("Fishery", fishery), levels = fishery_levels))
  df_wf_sim_mean <- df_wf_sim_mean %>%
    mutate(fishery = factor(paste("Fishery", fishery), levels = fishery_levels))

  ggplot() +
    geom_col(data = filter(df_wf_mean, series == "Observed"),
             aes(x = weight, y = proportion), fill = "grey75", width = 0.9) +
    geom_line(data = df_wf_sim_mean,
              aes(x = weight, y = proportion, group = sim_id),
              colour = "steelblue", alpha = 0.35, linewidth = 0.35) +
    geom_line(data = filter(df_wf_mean, series == "Predicted"),
              aes(x = weight, y = proportion), colour = "red3", linewidth = 0.8) +
    facet_wrap(~ fishery, ncol = 2, scales = "free_y") +
    scale_y_continuous(limits = c(0, NA), expand = expansion(mult = c(0, 0.05))) +
    labs(x = "Weight (kg)", y = "Mean proportion") +
    theme_minimal()
}
```

Inspect estimated spawning biomass trajectory:

```{r}
#| label: fig-opt-spbio
#| echo: true
#| message: false
#| fig-height: 4
#| fig-width: 7
#| fig-cap: Estimated spawning biomass trajectory.
#| eval: true
sb_df <- tibble(
  time_step = seq_along(fit_rep$spawning_biomass_y),
  spawning_biomass = as.numeric(fit_rep$spawning_biomass_y)
)

ggplot(sb_df, aes(x = time_step, y = spawning_biomass)) +
  geom_line(colour = "blue", linewidth = 1) +
  scale_y_continuous(limits = c(0, NA), expand = expansion(mult = c(0, 0.05))) +
  labs(
    x = "Model time step",
    y = "Spawning biomass (mt)",
    title = "Estimated spawning biomass trajectory"
  )
```

Inspect estimated relative spawning biomass ($SB/B_0$). The dynamic depletion
ratio ($SB/SB_{F=0}$) is also reported in @tbl-depletion-metrics because it
uses a different denominator and should not be read as the usual $SB/B_0$
trajectory.

```{r}
#| label: fig-opt-spdep
#| echo: true
#| message: false
#| fig-height: 4
#| fig-width: 7
#| fig-cap: Estimated relative spawning biomass trajectories.
#| eval: true
sb_dep_df <- tibble(
  time_step = seq_along(fit_rep$spawning_biomass_y),
  `SB/B0` = as.numeric(fit_rep$spawning_biomass_y) / fit_rep$B0,
  `SB/SB_F0` = as.numeric(fit_rep$dynamic_depletion_y)
) %>%
  pivot_longer(-time_step, names_to = "quantity", values_to = "value")

ggplot(sb_dep_df, aes(x = time_step, y = value, colour = quantity)) +
  geom_line(linewidth = 0.9) +
  scale_y_continuous(limits = c(0, NA), expand = expansion(mult = c(0, 0.05))) +
  labs(
    x = "Model time step",
    y = "Relative spawning biomass",
    colour = NULL,
    title = "Estimated relative spawning biomass trajectory",
    subtitle = "SB/B0 and dynamic depletion use different denominators"
  )
```

```{r}
#| label: tbl-depletion-metrics
#| echo: true
#| message: false
#| tbl-cap: Final-year relative biomass diagnostics using the estimated fit.
tibble(
  metric = c("SB/B0", "SB/SB_F0"),
  final_value = c(
    tail(fit_rep$spawning_biomass_y, 1) / fit_rep$B0,
    tail(fit_rep$dynamic_depletion_y, 1)
  )
) %>%
  mutate(final_value = round(final_value, 3)) %>%
  knitr::kable()
```

# Simulation

The CPUE series is set up using RTMB's `OBS` mechanism inside the model, which
allows simulation via `obj$simulate()`. The plots below compare observed,
predicted, and simulated log-CPUE.

```{r}
#| label: fig-cpue-sim
#| echo: true
#| message: false
#| fig-height: 5
#| fig-width: 9
#| fig-cap: Observed, predicted, and simulated log-CPUE faceted by index.
if (data$cpue_switch == 0) {
  cat("CPUE likelihood is switched off (cpue_switch = 0), so no simulated CPUE to show.\n")
} else {
  cpue_rep <- obj$report(obj$env$last.par.best)
  cpue_indices <- sort(unique(data$cpue_data$index))
  cpue_months <- sort(unique(data$cpue_data$month))
  cpue_index_labels <- setNames(
    paste0("Month: ", cpue_months),
    as.character(seq_along(cpue_months))
  )

  cpue_df <- data$cpue_data %>%
    mutate(
      obs_id = row_number(),
      observed = log(value),
      predicted = log(cpue_rep$cpue_pred),
      index = factor(index, levels = cpue_indices,
                     labels = cpue_index_labels[as.character(cpue_indices)])
    )

  n_sim <- 10
  cpue_sim_df <- map_dfr(seq_len(n_sim), function(s) {
    tibble(
      obs_id = cpue_df$obs_id,
      year = cpue_df$year,
      index = cpue_df$index,
      sim_id = s,
      simulated = obj$simulate(obj$env$last.par.best)$cpue_log_obs
    )
  })

  ggplot(cpue_df, aes(x = year)) +
    geom_line(data = cpue_sim_df, aes(y = simulated, group = sim_id),
              colour = "grey70", alpha = 0.55, linewidth = 0.35) +
    geom_point(aes(y = observed), colour = "black", size = 1) +
    geom_line(aes(y = predicted), colour = "red3", linewidth = 0.8) +
    facet_wrap(~ index, ncol = 2, scales = "free_y") +
    labs(x = "Year", y = "log(CPUE)",
         subtitle = paste0(n_sim, " fitted-model simulations in grey")) +
    theme_minimal()
}
```

# One step ahead (OSA) residuals

OSA residuals are a replacement for Pearson residuals:

```{r}
#| label: fig-osa-residuals
#| echo: true
#| message: false
#| fig-height: 5
#| fig-width: 8
#| fig-cap: One-step-ahead CPUE residual QQ and residual-index diagnostics.
#| eval: true
if (data$cpue_switch == 0) {
  cat("CPUE likelihood is switched off (cpue_switch = 0), so no OSA residuals to show.\n")
} else {
  osa_cpue <- oneStepPredict(obj = obj, observation.name = "cpue_log_obs",
                             method = "oneStepGeneric", trace = FALSE)
  osa_df <- data$cpue_data %>%
    mutate(residual = osa_cpue$res)

  if ("month" %in% names(osa_df)) {
    cpue_months <- sort(unique(osa_df$month))
    osa_df <- osa_df %>%
      mutate(index = factor(month, levels = cpue_months,
                            labels = month.name[cpue_months]))
  } else {
    cpue_indices <- sort(unique(osa_df$index))
    osa_df <- osa_df %>%
      mutate(index = factor(index, levels = cpue_indices,
                            labels = paste("Index", cpue_indices)))
  }

  osa_qq <- osa_df %>%
    group_by(index) %>%
    arrange(residual, .by_group = TRUE) %>%
    mutate(
      q_theoretical = qnorm(ppoints(n())),
      panel = "Normal Q-Q"
    ) %>%
    ungroup() %>%
    transmute(index, panel, x = q_theoretical, y = residual)

  osa_time <- osa_df %>%
    transmute(index, panel = "Residual by year", x = year, y = residual)

  ggplot(bind_rows(osa_qq, osa_time), aes(x = x, y = y)) +
    geom_hline(yintercept = 0, colour = "grey55", linewidth = 0.35) +
    geom_hline(yintercept = c(-2, 2), linetype = "dashed",
               colour = "grey55", linewidth = 0.35) +
    geom_abline(data = data.frame(panel = "Normal Q-Q"),
                aes(intercept = 0, slope = 1),
                inherit.aes = FALSE, colour = "grey55", linewidth = 0.35) +
    geom_segment(data = osa_time,
                 aes(x = x, xend = x, y = 0, yend = y),
                 inherit.aes = FALSE, colour = "grey65", linewidth = 0.25) +
    geom_point(size = 0.9, alpha = 0.75, colour = "black") +
    facet_grid(index ~ panel, scales = "free_x") +
    labs(x = NULL, y = "OSA residual") +
    theme_minimal()
}
```

# Composition OSA residuals

The BET diagnostic data used here contain length- and weight-composition
observations rather than age-composition observations. The residual plot below
follows the same structure as the SBT composition residual plots, using
`compResidual` row-by-row for the active multinomial composition likelihoods.

```{r}
#| label: fig-comp-residuals
#| echo: true
#| message: false
#| fig-height: 10
#| fig-width: 11
#| fig-cap: Composition OSA residual bubble plots by fishery and Q-Q diagnostics for active length and weight composition likelihoods.
#| eval: true
if (!requireNamespace("compResidual", quietly = TRUE)) {
  cat("Package 'compResidual' is required for composition residual plots.\n")
} else if (data$lf_switch != 1 || data$wf_switch != 1) {
  cat("This diagnostic chunk currently supports multinomial LF/WF likelihoods (switch = 1).\n")
} else {
  normalise_composition <- function(x) {
    total <- sum(x, na.rm = TRUE)
    if (!is.finite(total) || total <= 0) {
      return(rep(NA_real_, length(x)))
    }
    x / total
  }

  quiet_comp_residual_warnings <- function(expr) {
    withCallingHandlers(
      expr,
      warning = function(w) {
        if (grepl("NA/NaN function evaluation", conditionMessage(w), fixed = TRUE)) {
          invokeRestart("muffleWarning")
        }
      }
    )
  }

  ts_to_decimal_year <- function(ts, first_yr = data$first_yr,
                                 n_season = data$n_season) {
    first_yr + (as.integer(ts) - 1L) %/% n_season +
      ((as.integer(ts) - 1L) %% n_season) / n_season
  }

  flat_to_count_list <- function(v, n_f, fishery_f, minbin, maxbin) {
    out <- vector("list", length(n_f))
    offset <- 0L
    for (k in seq_along(n_f)) {
      f <- fishery_f[k]
      nbins <- maxbin[f] - minbin[f] + 1L
      n_values <- n_f[k] * nbins
      out[[k]] <- matrix(
        v[seq.int(offset + 1L, length.out = n_values)],
        nrow = n_f[k],
        ncol = nbins,
        byrow = TRUE
      )
      offset <- offset + n_values
    }
    out
  }

  composition_residual_rows <- function(type_label, y_label, obs_flat,
                                        pred_list, n_by_row, n_f, fishery_f,
                                        year_fi, minbin, maxbin, bins) {
    obs_list <- flat_to_count_list(obs_flat, n_f, fishery_f, minbin, maxbin)
    row_offset <- 0L

    map_dfr(seq_along(pred_list), function(i) {
      f <- fishery_f[i]
      n_rows <- n_f[i]
      row_ids <- seq.int(row_offset + 1L, length.out = n_rows)
      row_offset <<- row_offset + n_rows
      active_bins <- bins[minbin[f]:maxbin[f]]
      pred <- pred_list[[i]]
      obs <- obs_list[[i]]
      years <- year_fi[[as.character(f)]]

      map_dfr(seq_len(n_rows), function(j) {
        obs_prop <- normalise_composition(obs[j, ])
        pred_prop <- normalise_composition(pred[j, ])
        n_eff <- n_by_row[row_ids[j]]

        if (anyNA(obs_prop) || anyNA(pred_prop) ||
            !is.finite(n_eff) || n_eff <= 0) {
          return(tibble())
        }

        pred_prop <- pred_prop + 1e-6
        pred_prop <- pred_prop / sum(pred_prop)
        residual <- as.numeric(quiet_comp_residual_warnings(
          compResidual::resMulti(
            obs = matrix(obs_prop * n_eff, ncol = 1L),
            pred = matrix(pred_prop, ncol = 1L)
          )
        ))
        keep <- seq_along(residual)

        plot_year <- years[j]
        if (!is.finite(plot_year) || plot_year < 1000) {
          plot_year <- ts_to_decimal_year(years[j])
        }

        tibble(
          composition = type_label,
          y_label = y_label,
          fishery_num = f,
          fishery = fleet_names[f],
          row = row_ids[j],
          year = plot_year,
          bin = active_bins[keep],
          observed = obs_prop[keep],
          predicted = pred_prop[keep],
          n_eff = n_eff,
          residual = residual,
          Sign = if_else(residual >= 0, "Positive", "Negative")
        )
      })
    })
  }

  lf_residuals <- composition_residual_rows(
    type_label = "Length",
    y_label = "Length (cm)",
    obs_flat = data$lf_obs_flat,
    pred_list = fit_rep$lf_pred,
    n_by_row = data$lf_n,
    n_f = data$lf_n_f,
    fishery_f = data$lf_fishery_f,
    year_fi = if (!is.null(data$lf_calendar_year_fi)) data$lf_calendar_year_fi else data$lf_year_fi,
    minbin = data$lf_minbin,
    maxbin = data$lf_maxbin,
    bins = data$len_mid
  )

  wf_residuals <- composition_residual_rows(
    type_label = "Weight",
    y_label = "Weight (kg)",
    obs_flat = data$wf_obs_flat,
    pred_list = fit_rep$wf_pred,
    n_by_row = data$wf_n,
    n_f = data$wf_n_f,
    fishery_f = data$wf_fishery_f,
    year_fi = if (!is.null(data$wf_calendar_year_fi)) data$wf_calendar_year_fi else data$wf_year_fi,
    minbin = data$wf_minbin,
    maxbin = data$wf_maxbin,
    bins = data$wt_mid
  )

  comp_residuals <- bind_rows(lf_residuals, wf_residuals) %>%
    filter(is.finite(residual)) %>%
    mutate(
      fishery = factor(fishery, levels = fleet_names[sort(unique(fishery_num))]),
      composition = factor(composition, levels = c("Length", "Weight"))
    )

  residual_summary <- comp_residuals %>%
    summarise(
      n = n(),
      mean = mean(residual),
      SDNR = sd(residual),
      MAR = median(abs(residual)),
      max_abs = max(abs(residual)),
      .by = c(composition, fishery)
    ) %>%
    mutate(across(where(is.numeric), ~ signif(.x, 4)))

  plot_comp_bubbles <- function(plot_data, title, y_label) {
    ggplot(
      plot_data,
      aes(x = year, y = bin, size = abs(residual), fill = Sign)
    ) +
      geom_point(shape = 21, alpha = 0.65, colour = "grey30", stroke = 0.08) +
      facet_wrap(~ fishery, ncol = 2, scales = "free_y") +
      scale_fill_manual(values = c(Positive = "white", Negative = "black")) +
      scale_size_area(
        max_size = 4,
        limits = c(0, 4),
        oob = scales::squish,
        name = "|OSA residual|"
      ) +
      scale_x_continuous(breaks = scales::breaks_pretty(n = 6)) +
      labs(
        title = title,
        x = "Year",
        y = y_label,
        fill = "Residual sign",
        caption = "White = observed greater than predicted; black = observed less than predicted"
      ) +
      theme_bw() +
      theme(
        legend.position = "bottom",
        panel.grid.minor = element_blank(),
        strip.text = element_text(size = 9)
      )
  }

  lf_plot_data <- filter(comp_residuals, composition == "Length")
  wf_plot_data <- filter(comp_residuals, composition == "Weight")

  if (nrow(lf_plot_data) > 0L) {
    print(plot_comp_bubbles(lf_plot_data, "Length Composition OSA Residuals", "Length (cm)"))
  }

  if (nrow(wf_plot_data) > 0L) {
    print(plot_comp_bubbles(wf_plot_data, "Weight Composition OSA Residuals", "Weight (kg)"))
  }

  ggplot(comp_residuals, aes(sample = residual)) +
    stat_qq(size = 0.45, alpha = 0.45) +
    stat_qq_line(colour = "red3", linetype = "dashed") +
    facet_wrap(~ composition, ncol = 2, scales = "free") +
    labs(
      title = "Composition OSA Q-Q Normal Check",
      x = "Theoretical quantiles",
      y = "Sample quantiles"
    ) +
    theme_bw()
}
```

```{r}
#| label: tbl-comp-residual-summary
#| echo: false
#| tbl-cap: Composition OSA residual diagnostics by composition type and fishery.
if (exists("residual_summary") && nrow(residual_summary) > 0L) {
  knitr::kable(residual_summary)
} else {
  cat("No finite composition residuals were available to summarise.\n")
}
```
