---
title: "The BET model"
format: html
execute:
  eval: true
vignette: >
  %\VignetteIndexEntry{The BET model}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{quarto::html}
editor_options: 
  chunk_output_type: console
---

```{r}
#| include: false
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

# Introduction

`opal` is an open, modular R package for fisheries stock assessment built on 
[RTMB](https://github.com/kaskr/RTMB) for automatic differentiation. It includes 
bundled data from the western and central Pacific Ocean (WCPO) bigeye tuna (BET) 
stock assessment. This vignette demonstrates model setup, fitting, and 
diagnostics using those data.

# Load inputs

Load the `opal` package and the `RTMB` dependency. The `ggplot2` package is
used for plotting.

```{r}
#| label: load-pkg
#| echo: true
#| message: false
library(opal)
library(tidyverse)
library(RTMB)

theme_set(theme_bw())
```

The bundled data object `wcpo_bet_data` contains all biological parameters,
catch and CPUE observations, length structure, and prior specifications needed
for the BET assessment model:

```{r}
#| label: load-data
#| echo: true
#| message: false
data(wcpo_bet_data)
data <- wcpo_bet_data
if (is.null(data$n_index)) data$n_index <- 1L
if (is.null(data$cpue_data$index)) data$cpue_data$index <- rep(1L, nrow(data$cpue_data))
names(data)
```

Key dimensions of the data:

```{r}
#| label: data-dims
#| echo: true
cat("Number of ages:", data$n_age, "\n")
cat("Number of years:", data$n_year, "\n")
cat("Number of fisheries:", data$n_fishery, "\n")
cat("Number of length bins:", data$n_len, "\n")
```

## Biological inputs

Natural mortality at age, maturity at length, and length-weight parameters are
all contained in the data object. The two primary biological inputs are plotted
below.

```{r}
#| label: fig-biology
#| echo: true
#| message: false
#| fig-height: 4
#| fig-width: 7
#| fig-cap: Natural mortality-at-age and maturity-at-length used in the BET model.
ages <- data$age_a
real_age <- ages / 4

# Natural mortality
ggplot(data = data.frame(age = real_age, M = data$M), aes(x = age, y = M)) +
  geom_line() + geom_point() +
  labs(x = "Age (years)", y = "M (quarterly rate)", 
       title = "Natural Mortality-at-Age")

# Maturity at length
len_mid <- seq(data$len_bin_start + data$len_bin_width / 2,
               by = data$len_bin_width, length.out = data$n_len)
ggplot(data = data.frame(length = len_mid, maturity = data$maturity), 
       aes(x = length, y = maturity)) +
  geom_line() + geom_point() +
  labs(x = "Length (cm)", y = "Maturity", title = "Maturity-at-Length")

fleet_names <- c("F01_LL.NORTH", "F02_LL.US", "F03_LL.OFFSH",
                 "F04_LL.EQUAT", "F05_LL.WEST", "F06_LL.SOUTH", "F07_LL.AUS",
                 "F08_PS.ASSOC", "F09_PS.UNASS", "F10_DOM.MISC",
                 "F11_DOM.HL", "F12_JP.PS.N", "F13_JP.PL", "F14_EQ.PL",
                 "S01_INDEX")
```

## CPUE data

```{r}
#| label: prep-cpue
#| echo: true
#| message: false
data$cpue_data$index <- as.integer(factor(data$cpue_data$month, levels = c(2, 5, 8, 11)))
data$n_index <- 4L
```

## Length composition data

Length composition observations are loaded from the bundled `wcpo_bet_lf`
long-format data object and transformed into model-ready arrays.

```{r}
#| label: prep-lf
#| echo: true
#| message: false
data(wcpo_bet_lf)

# Pivot to wide format: one row per fishery x timestep, bins as columns
# Note: LF data exist for fisheries 8-14, but we initially use only fisheries 8 & 9
lf_wide <- wcpo_bet_lf %>%
  pivot_wider(
    id_cols = c(fishery, year, month, ts),
    names_from = bin,
    values_from = value,
    values_fill = 0
  ) %>%
  arrange(fishery, ts)

unique(lf_wide$fishery)

# define variance adjustment scalars for each fishery if using lf_switch = 1 (multinomial)
var_adjust_scalars <- rep(1, data$n_fishery)
var_adjust_scalars[] <- 80
# var_adjust_scalars <- 1/rep(20000,data$n_fishery)
# var_adjust_scalars[c(1, 4, 5, 6, 15)] <- 1 / 40000
# var_adjust_scalars[] <- 1 / 4e9

data <- prep_lf_data(data = data, lf_wide = lf_wide, 
                     # lf_keep_fisheries = c(8),
                     # lf_keep_fisheries = c(8, 9, 10),
                     lf_var_adjust = var_adjust_scalars)

data$lf_fishery
range(data$lf_n) # aim for tens to low thousands per obs
sum(data$lf_n) # aim for ~1e4–1e5 total
```

## Weight composition data

Weight composition observations are loaded from the bundled `wcpo_bet_wf`
long-format data object and transformed into model-ready arrays.

```{r}
#| label: prep-wf
#| echo: true
#| message: false
data(wcpo_bet_wf)

# Weight bin scalars (1 kg bins, 1–200 kg)
data$wt_bin_start <- 1
data$wt_bin_width <- 1
data$n_wt <- 200L

# Pivot to wide format: one row per fishery x timestep, bins as columns
wf_wide <- wcpo_bet_wf |>
  pivot_wider(
    id_cols = c(fishery, year, month, ts),
    names_from = bin,
    values_from = value,
    values_fill = 0
  ) |>
  arrange(fishery, ts)

unique(wf_wide$fishery)

data <- prep_wf_data(data = data, wf_wide = wf_wide,
                     wf_keep_fisheries = c(1),
                     wf_switch = 1L)
```

# Model setup

## Parameters

Define the initial parameter values. Growth parameters (`log_L1`, `log_L2`,
`log_k`, `log_CV1`, `log_CV2`) and selectivity (`par_sel`) are initialised at
reasonable starting values. Load these from the bundled `wcpo_bet_parameters` 
data object:

```{r}
#| label: get-pars
#| echo: true
#| message: false
data(wcpo_bet_parameters)
par_sel_raw <- as.matrix(wcpo_bet_parameters$par_sel)
par_sel_start <- par_sel_raw

# BET double-normal width starts are stored on the legacy opal width scale.
# The current double-normal uses exp(width) * sd(length)^2, so shift width
# columns by log(sd(length)) to preserve the same natural-scale denominator.
double_normal_f <- data$sel_type_f == 2L
par_sel_start[double_normal_f, 3:4] <- par_sel_start[double_normal_f, 3:4] - log(sd(data$len_mid))

parameters <- list(
  log_B0 = 20,
  log_h = as.numeric(wcpo_bet_parameters$log_h),
  log_sigma_r = as.numeric(wcpo_bet_parameters$log_sigma_r),
  log_cpue_q = rep(0, data$n_index),
  log_cpue_tau = rep(log(0.1), data$n_index),
  log_cpue_omega = rep(as.numeric(wcpo_bet_parameters$log_cpue_omega), data$n_index),
  log_lf_tau = as.numeric(log(rep(0.1, data$n_fishery))),
  log_wf_tau = rep(0, data$n_fishery),
  log_L1 = as.numeric(wcpo_bet_parameters$log_L1),
  log_L2 = as.numeric(wcpo_bet_parameters$log_L2),
  log_k = as.numeric(wcpo_bet_parameters$log_k),
  log_CV1 = as.numeric(wcpo_bet_parameters$log_CV1),
  log_CV2 = as.numeric(wcpo_bet_parameters$log_CV2),
  par_sel = par_sel_start,
  rdev_y = as.numeric(wcpo_bet_parameters$rdev_y)
)
```

The bundled `wcpo_bet_parameters$par_sel` values use the legacy opal width
scale for double-normal selectivity. They are converted to the current
`sel_double_normal()` width scale before fitting. @tbl-selectivity-scale shows
the raw bundled values and the converted fitting starts.

```{r}
#| label: tbl-selectivity-scale
#| echo: true
#| message: false
#| tbl-cap: Raw bundled selectivity parameters and opal-scale fitting starts.
parameter_names <- c("peak", "top", "asc_width", "desc_width", "init", "final")
bind_rows(
  as_tibble(par_sel_raw, .name_repair = ~ parameter_names) %>%
    mutate(fleet = row_number(), scale = "bundled"),
  as_tibble(parameters$par_sel, .name_repair = ~ parameter_names) %>%
    mutate(fleet = row_number(), scale = "opal_start")
) %>%
  relocate(fleet, scale) %>%
  mutate(across(where(is.numeric), ~ round(.x, 3))) %>%
  knitr::kable()
```

## Priors

Priors are specified using `get_priors()`. The data object already contains
prior center values for growth parameters:

```{r}
#| label: get-priors
#| echo: true
#| message: false
data$priors <- get_priors(parameters = parameters, data = data)
evaluate_priors(parameters = parameters, priors = data$priors)
```

## Parameter map

Use RTMB's `map` option to turn parameters on/off. Parameters mapped to
`factor(NA)` are fixed at their initial values. Recruitment deviations are
fixed here to keep this diagnostic fit focused on scale and selectivity; freeing
all quarterly deviations as fixed effects can create high-gradient early
recruitment spikes and a misleading initial biomass trajectory.

```{r}
#| label: get-map
#| echo: true
#| message: false
parameters$par_sel
wcpo_bet_parameters$par_sel

map_sel <- matrix(NA, nrow(parameters$par_sel), ncol(parameters$par_sel))

# 8-14 have LF data. Estimate peak and width parameters for most LF fleets.
# F13_JP.PL is fixed at the bundled SS3-derived start because estimating it
# pushes the peak to the lower length bound and produces an edge spike.
map_sel[8, c(1, 3, 4)] <- 1:3 # peak, asc-width, desc-width
map_sel[9, c(1, 3, 4)] <- 4:6 # peak, asc-width, desc-width
map_sel[10, c(1, 3, 4)] <- 7:9 # peak, asc-width, desc-width
map_sel[11, 1:2] <- 10:11 # logistic inflection and width
map_sel[12, c(1, 3, 4)] <- 12:14 # peak, asc-width, desc-width
map_sel[14, c(1, 3, 4)] <- 15:17 # peak, asc-width, desc-width

# 1-7 & 15 is WFs
# map_sel[1, c(1, 3, 4, 6)] <- 29:32

map_sel

map_rdev <- rep(NA, length(parameters$rdev_y))

map_lf_tau <- rep(NA, length(parameters$log_lf_tau))
map_lf_tau[c(8, 9)] <- NA

map <- list(
  # log_B0 = factor(NA),
  log_h = factor(NA),
  log_sigma_r = factor(NA),
  # log_cpue_q = factor(NA),
  log_cpue_q = factor(c(NA, 1, 2, 3)),
  log_cpue_tau = factor(rep(NA, data$n_index)),
  log_cpue_omega = factor(rep(NA, data$n_index)),
  log_lf_tau = factor(map_lf_tau),
  log_wf_tau = factor(rep(NA, data$n_fishery)),
  log_L1  = factor(NA),
  log_L2  = factor(NA),
  log_k   = factor(NA),
  log_CV1 = factor(NA),
  log_CV2 = factor(NA),
  par_sel = factor(map_sel),
  rdev_y = factor(map_rdev)
)
```

## Build the AD object

Using the `data`, the `parameters`, the parameter `map`, and the model
(`opal_model`), the AD object is created using RTMB's `MakeADFun` function.
Optionally, we could also specify random effects here 
(e.g. `random = "rdev_y"`), but we'll start with a simpler fixed-effects model 
to check everything is working first.

```{r}
#| label: make-adfun
#| echo: true
#| message: false
# data$lf_switch <- 0 # skip length comps (removal only)
data$lf_switch <- 1 # multinomial likelihood on flat counts (default)
# data$lf_switch <- 2 # fails while fitting, need to sort out log_lf_tau for this, use simulate to tune and find
# data$lf_switch <- 3 # fails - wants integers - think this should be an issue to RTMBdist guys

# Note: wf_switch was already set by prep_wf_data(); it is re-stated here for
# clarity alongside the equivalent lf_switch assignment.
data$wf_switch <- 0 # skip weight comps (removal only)
# data$wf_switch <- 1 # multinomial likelihood on flat counts (default)
# data$wf_switch <- 2 # Dirichlet
# data$wf_switch <- 3 # Dirichlet-multinomial

obj <- MakeADFun(func = cmb(opal_model, data), parameters = parameters, map = map)
obj$env$tracemgc <- FALSE
obj$par
unique(names(obj$par))
obj$fn()
obj$gr()
# TMB::benchmark(obj, n = 500, expr = expression(obj$gr()))
```

Inspect initial model outputs. The initial biomass trajectory is shown in
@fig-init-biomass, and the predicted catch checks are shown in @fig-init-catch.

```{r}
#| label: fig-init-biomass
#| echo: true
#| message: false
#| fig-height: 4
#| fig-width: 7
#| fig-cap: Initial spawning biomass trajectory before optimization.
plot(obj$report()$spawning_biomass_y, type = "l",
     xlab = "Time step", ylab = "Spawning biomass (mt)",
     main = "Initial spawning biomass trajectory")
```

```{r}
#| label: fig-init-catch
#| echo: true
#| message: false
#| fig-height: 4
#| fig-width: 7
#| fig-cap: Initial predicted catch by fleet compared with observed catch.
plot_catch(data = data, obj = obj)
```

## Parameter bounds

```{r}
#| label: get-par-bounds
#| echo: true
#| message: false
Lwr <- rep(-Inf, length(obj$par))
Upr <- rep(Inf, length(obj$par))
Lwr[grep("log_B0", names(obj$par))] <- log(1)
Upr[grep("log_B0", names(obj$par))] <- log(exp(22))
Lwr[grep("log_cpue_q", names(obj$par))] <- log(0.1)
Upr[grep("log_cpue_q", names(obj$par))] <- log(10)
Lwr[grep("log_lf_tau", names(obj$par))] <- rep(-9, length(grep("log_lf_tau", names(obj$par))))
Upr[grep("log_lf_tau", names(obj$par))] <- rep(9, length(grep("log_lf_tau", names(obj$par))))
Lwr[grep("rdev_y", names(obj$par))] <- rep(-5, length(grep("rdev_y", names(obj$par))))
Upr[grep("rdev_y", names(obj$par))] <- rep(5, length(grep("rdev_y", names(obj$par))))

mu_len <- mean(data$len_mid)
sd_len <- sd(data$len_mid)

par_labels <- names(obj$par)
par_sel_pos <- which(names(obj$par) == "par_sel")
par_sel_level_pos <- setNames(par_sel_pos, levels(map$par_sel))

for (f in seq_len(data$n_fishery)) {
  idx <- rep(NA_integer_, ncol(map_sel))
  for (p in seq_len(ncol(map_sel))) {
    level <- as.character(map_sel[f, p])
    if (!is.na(level)) {
      idx[p] <- par_sel_level_pos[[level]]
      par_labels[idx[p]] <- paste0("par_sel[", f, ",", p, "]")
    }
  }

  if (data$sel_type_f[f] == 1L) {
    # Logistic: inflection [5,200], width [>0,500]
    if (!is.na(idx[1])) {
      Lwr[idx[1]] <- (5   - mu_len) / sd_len
      Upr[idx[1]] <- (200 - mu_len) / sd_len
    }

    if (!is.na(idx[2])) {
      Lwr[idx[2]] <- log(1e-6 / sd_len) # avoid log(0)
      Upr[idx[2]] <- log(500  / sd_len)
    }
  } else {
    # Double-normal SS3 bounds
    if (!is.na(idx[1])) {
      Lwr[idx[1]] <- (10.1 - mu_len) / sd_len
      Upr[idx[1]] <- (200  - mu_len) / sd_len
    }

    if (!is.na(idx[2])) {
      Lwr[idx[2]] <- -7
      Upr[idx[2]] <-  7
    }

    # for sel_double_normal using exp(c)*sd^2 and exp(d)*sd^2:
    # the upper width bound is slightly wider than the SS3-style default to
    # avoid artificial convergence on the transformed denominator bound.
    shift <- 2 * log(sd_len)
    width_log_upper <- 8
    if (!is.na(idx[3])) {
      Lwr[idx[3]] <- -7 - shift
      Upr[idx[3]] <-  width_log_upper - shift
    }
    if (!is.na(idx[4])) {
      Lwr[idx[4]] <- -7 - shift
      Upr[idx[4]] <-  width_log_upper - shift
    }

    if (!is.na(idx[5])) {
      Lwr[idx[5]] <- -9 # practical bound (instead of -999)
      Upr[idx[5]] <-  9
    }
    if (!is.na(idx[6])) {
      Lwr[idx[6]] <- -9
      Upr[idx[6]] <-  9
    }
  }
}

bounds <- data.frame(par = par_labels, lower = Lwr, upper = Upr)
```

# Optimisation

Optimise using the `nlminb` function:

```{r}
#| label: run-nlminb
#| echo: true
#| results: hide
#| eval: false
control <- list(eval.max = 10000, iter.max = 10000)
# control <- list(eval.max = 100, iter.max = 100)
obj$env$tracemgc <- FALSE
opt <- list(par = obj$par)
fit_grad <- Inf
for (i in seq_len(4)) {
  opt <- nlminb(start = opt$par, objective = obj$fn, gradient = obj$gr,
                hessian = obj$he, lower = Lwr, upper = Upr, control = control)
  fit_grad <- max(abs(obj$gr(opt$par)))
  if (fit_grad < 1e-2) break
}
fit_nll <- obj$fn(opt$par)
obj$env$last.par.best <- opt$par
fit_grad <- max(abs(obj$gr(opt$par)))
fit_nll
fit_grad
obj$report()$lp_prior
obj$report()$lp_penalty
obj$report()$lp_rec
sum(obj$report()$lp_cpue)
sum(obj$report()$lp_lf)
sum(obj$report()$lp_wf)
# TMBhelper::check_estimability(obj = obj)
if (fit_grad < 1e-2) {
  check_estimability(obj = obj)
} else {
  message("Skipping check_estimability(); max |gradient| = ", signif(fit_grad, 4), ". Improve optimization first.")
}
```

Compare initial and estimated parameter values:

```{r}
#| label: opt-results
#| echo: true
#| message: false
#| eval: false
get_par_table(obj, parameters, map, lower = Lwr, upper = Upr, grad_tol = 1e-2, digits = 2L)
```

## Selectivity

Visualise the selectivity-at-length curves by fleet in @fig-selectivity-length.

```{r}
#| label: fig-selectivity-length
#| echo: true
#| message: false
#| fig-height: 8
#| fig-width: 10
#| fig-cap: Selectivity-at-length by fleet. Coloured lines use the current parameter values; dashed black lines show the bundled reference.
par_sel_est <- obj$env$parList(obj$env$last.par.best)$par_sel
par_sel_ref <- parameters$par_sel

map_sel_mat <- matrix(map$par_sel, nrow = nrow(par_sel_est), ncol = ncol(par_sel_est))
fishery_estimated <- apply(map_sel_mat, 1, function(row) any(!is.na(row)))

sel_est_df <- do.call(rbind, lapply(seq_len(data$n_fishery), function(f) {
  sel_l <- if (data$sel_type_f[f] == 1L) {
    sel_logistic(len_mid, par_sel_est[f, ])
  } else {
    sel_double_normal(len_mid, par_sel_est[f, ])
  }
  data.frame(
    fishery = f,
    fleet_name = fleet_names[f],
    length = len_mid,
    selectivity = as.numeric(sel_l),
    estimated = fishery_estimated[f]
  )
}))

sel_orig_df <- do.call(rbind, lapply(seq_len(data$n_fishery), function(f) {
  sel_l <- if (data$sel_type_f[f] == 1L) {
    sel_logistic(len_mid, par_sel_ref[f, ])
  } else {
    sel_double_normal(len_mid, par_sel_ref[f, ])
  }
  data.frame(
    fishery = f,
    fleet_name = fleet_names[f],
    length = len_mid,
    selectivity = as.numeric(sel_l)
  )
}))

ggplot(data = sel_est_df, aes(x = length, y = selectivity, colour = estimated)) +
  geom_line() +
  geom_line(
    data = sel_orig_df,
    aes(x = length, y = selectivity),
    inherit.aes = FALSE,
    colour = "black",
    linetype = "dashed",
    linewidth = 0.7
  ) +
  facet_wrap(fleet_name ~ ., ncol = 4) +
  labs(x = "Length (cm)", y = "Selectivity") +
  coord_cartesian(ylim = c(0, 1))
```

The corresponding selectivity-at-age curves are shown in @fig-selectivity-age.
For comparison with SS3's reported age-selectivity output, the plotted curves
are scaled to a maximum of 1 within each fleet. The model still uses the
unscaled PLA-integrated values in `rep$sel_fya`.

```{r}
#| label: fig-selectivity-age
#| echo: true
#| message: false
#| fig-height: 8
#| fig-width: 10
#| fig-cap: Reported selectivity-at-age by fleet after integrating length selectivity over the age-length key and scaling each fleet to a maximum of 1 for SS3-style plotting.
rep <- obj$report()
sel_fya <- rep$sel_fya

sel_df <- expand.grid(fishery = 1:data$n_fishery, age = 1:data$n_age)
sel_df$selectivity <- sapply(1:nrow(sel_df), function(i) {
  sel_fya[sel_df$fishery[i], 1, sel_df$age[i]]
})
sel_df$real_age <- sel_df$age / 4
sel_df$fleet_name <- fleet_names[sel_df$fishery]
sel_df <- sel_df %>%
  group_by(fishery) %>%
  mutate(selectivity_reported = selectivity / max(selectivity, na.rm = TRUE)) %>%
  ungroup()

ggplot(sel_df, aes(x = real_age, y = selectivity_reported)) +
  geom_line() +
  facet_wrap(~fleet_name, ncol = 4) +
  labs(x = "Age (years)", y = "Selectivity") +
  ylim(0, 1)
```

@tbl-sel-par summarises the selectivity parameter values and the map status.

```{r}
#| label: tbl-sel-par
#| echo: true
#| message: false
#| tbl-cap: Selectivity parameter values, bundled reference values, and map status.
# Three sources to compare side by side:
#   1. Initial values used to build obj   (parameters$par_sel)
#   2. Bundled reference values            (wcpo_bet_parameters$par_sel)
#   3. Estimated / mapped status           (map$par_sel)
#
# par_sel is [n_fishery x 6] with columns (a, b, c, d, e, f) for double-normal
# or (par1, par2) for logistic (cols 3-6 ignored). `factor(map_sel)` flattens
# matrices column-major, so wrap the map vector back into the same shape.

n_f <- nrow(parameters$par_sel)
n_p <- ncol(parameters$par_sel)
col_lbls <- c("a (peak)", "b (plateau)", "c (asc width)", "d (desc width)", "e (init sel)", "f (final sel)")

# Estimated/fixed status from the map
map_sel_mat <- if (is.null(map$par_sel)) {
  matrix(seq_len(n_f * n_p), n_f, n_p) # everything estimated if no map
} else {
  matrix(map$par_sel, n_f, n_p)
}
estimated_mat <- !is.na(map_sel_mat)

# Reference values (may not have row/col names — assume same shape)
ref_mat <- as.matrix(wcpo_bet_parameters$par_sel)
stopifnot("wcpo_bet_parameters$par_sel must match parameters$par_sel shape" = all(dim(ref_mat) == c(n_f, n_p)))

# Selectivity type (logistic = 1, double-normal = 2)
sel_type_lbl <- ifelse(data$sel_type_f == 1L, "logistic", "double-normal")

# For logistic fisheries, columns 3-6 are unused -> NA-out for clarity
init_mat <- parameters$par_sel
for (f in seq_len(n_f)) {
  if (data$sel_type_f[f] == 1L) {
    init_mat[f, 3:6] <- NA
    ref_mat[f, 3:6] <- NA
    estimated_mat[f, 3:6] <- FALSE
  }
}

# Assemble long table
sel_par_tbl <- tibble(
  fishery    = rep(seq_len(n_f), times = n_p),
  fleet_name = rep(fleet_names[seq_len(n_f)], times = n_p),
  sel_type   = rep(sel_type_lbl, times = n_p),
  param      = rep(col_lbls, each = n_f),
  init       = as.numeric(init_mat),
  reference  = as.numeric(ref_mat),
  estimated  = as.vector(estimated_mat)
) %>%
  mutate(diff = init - reference) %>%
  arrange(fishery, match(param, col_lbls))

sel_par_tbl %>%
  mutate(
    init      = ifelse(is.na(init),      "—", formatC(init,      digits = 3, format = "f")),
    reference = ifelse(is.na(reference), "—", formatC(reference, digits = 3, format = "f")),
    diff      = ifelse(is.na(diff),      "—", formatC(diff,      digits = 3, format = "f")),
    estimated = ifelse(estimated, "✓", "")
  ) %>%
  select(Fleet = fleet_name,
         Type = sel_type,
         Param = param,
         Initial = init,
         `Reference (bundled)` = reference,
         `Δ (init − ref)` = diff,
         `Est?` = estimated) %>%
  knitr::kable(align   = c("l", "l", "l", "r", "r", "r", "c"))
```

## Diagnostics

Check that all parameters are estimable using the `check_estimability` function:

```{r}
#| label: check-estimability
#| echo: true
#| message: false
#| error: true
#| eval: false
obj$env$tracemgc <- FALSE
fit_grad <- max(abs(obj$gr(obj$env$last.par.best)))
if (fit_grad < 1e-2) {
  check_estimability(obj = obj)
} else {
  message("Skipping check_estimability(); max |gradient| = ", signif(fit_grad, 4),
          ". Improve optimization first.")
}
```

Look for any parameters with high correlations (absolute correlation > 0.95) using the `get_cor_pairs` function:

```{r}
#| label: check-cor
#| echo: true
#| message: false
#| eval: false
get_cor_pairs(obj = obj, threshold = 0.95)
```

Calculate standard deviations of all model parameters:

```{r}
#| label: run-sdreport
#| echo: true
#| message: false
#| eval: false
Report <- sdreport(obj)
```

## Data fits

Inspect fitted CPUE and catch. CPUE fits are shown in @fig-cpue-fit; catch fits
and residuals are plotted below.

```{r}
#| label: fig-cpue-fit
#| echo: true
#| message: false
#| fig-height: 4
#| fig-width: 7
#| fig-cap: Observed and predicted CPUE by fishery and quarter, with simulated replicate series.
if (data$cpue_switch == 0) {
  cat("CPUE likelihood is switched off (cpue_switch = 0), so no predicted CPUE to show.\n")
} else {
  rep <- obj$report(obj$env$last.par.best)
  
  df <- data$cpue_data %>%
    mutate(obs_id = row_number(), pred = rep$cpue_pred, sigma = rep$cpue_sigma)
  
  # Simulations
  n_sim <- 25
  df_sim <- map_dfr(seq_len(n_sim), function(s) {
    sim_log <- obj$simulate()$cpue_log_obs
    tibble(
      obs_id = seq_along(sim_log),
      year = data$cpue_data$year,
      month = data$cpue_data$month,
      fishery = data$cpue_data$fishery,
      sim_id = s,
      sim = exp(sim_log)
    )
  })
  
  month_levels <- sort(unique(data$cpue_data$month))
  fish_levels <- sort(unique(data$cpue_data$fishery))
  format_facets <- function(d) {
    d %>%
      mutate(
        month = factor(month, levels = month_levels, labels = paste0("Month: ", month_levels)),
        fishery = factor(fishery, levels = fish_levels, labels = paste0("Fishery: ", fish_levels))
      )
  }
  df <- format_facets(df)
  df_sim <- format_facets(df_sim)
  
  ggplot(data = df, aes(x = year)) +
    geom_line(data = df_sim, aes(y = sim, group = sim_id),
              colour = "lightblue", alpha = 0.5, linewidth = 0.3) +
    geom_linerange(aes(ymin = exp(log(value) - sigma),
                       ymax = exp(log(value) + sigma),
                       colour = "Observed"), alpha = 0.6) +
    geom_point(aes(y = value, colour = "Observed")) +
    geom_line(aes(y = pred, colour = "Predicted"), linewidth = 0.8) +
    facet_grid(fishery ~ month, scales = "free_y") +
    scale_y_continuous(limits = c(0, NA), expand = expansion(mult = c(0, 0.05))) +
    scale_x_continuous(breaks = scales::pretty_breaks()) +
    scale_colour_manual(values = c(Observed = "black", Predicted = "red3")) +
    labs(x = "Year", y = "CPUE", colour = NULL, subtitle = paste0(n_sim, " posterior-predictive simulations (grey)")) +
    theme(legend.position = "bottom")
}
```

```{r}
#| label: fig-catch-fit
#| echo: true
#| message: false
#| fig-height: 4
#| fig-width: 7
#| fig-cap: Predicted catch and catch residual diagnostics for the initial model state.
sum(obj$report()$catch_pred_ysf - data$catch_obs_ysf)
plot_catch(data = data, obj = obj)
plot_catch(data = data, obj = obj, plot_resid = TRUE)
```

Inspect predicted vs observed length compositions in @fig-lf-diagnostics.

```{r}
#| label: fig-lf-diagnostics
#| echo: true
#| message: false
#| fig-height: 6
#| fig-width: 7
#| fig-cap: Observed, predicted, and simulated length composition proportions for the selected fishery.
# Convert model timestep -> "YYYY Qn" label
ts_to_label <- function(ts, first_yr = 1952, n_season = 4) {
  yr <- first_yr + (ts - 1L) %/% n_season
  qn <- ((ts - 1L) %% n_season) + 1L
  paste0(yr, " Q", qn)
}

if (data$lf_switch == 0) {
  cat("Length composition likelihood is switched off (lf_switch = 0), so no predicted length compositions to show.\n")
} else {
  lf_rep <- obj$report()
  lf_pred <- lf_rep$lf_pred
  nbins <- data$lf_maxbin[1] - data$lf_minbin[1] + 1
  # Name the list elements and columns
  names(lf_pred) <- data$lf_fishery_f
  for (k in seq_along(lf_pred)) colnames(lf_pred[[k]]) <- data$len_mid
  
  # Helper: convert a flat OBS vector back to list-of-matrices (proportions)
  flat_to_list <- function(v, lf_n_f, nbins) {
    out <- vector("list", length(lf_n_f))
    offset <- 0L
    for (k in seq_along(lf_n_f)) {
      m <- matrix(v[(offset + 1):(offset + lf_n_f[k] * nbins)], nrow = lf_n_f[k], ncol = nbins, byrow = TRUE)
      # Normalise each row to proportions
      rs <- rowSums(m)
      rs[rs == 0] <- 1
      out[[k]] <- m / rs
      offset <- offset + lf_n_f[k] * nbins
    }
    out
  }
  
  # Split year index by fishery (must match length of lf_year)
  lf_year_list <- split(data$lf_year, data$lf_fishery)
  
  # ---- Predicted: long data frame ----
  df_pred <- map_dfr(seq_along(lf_pred), function(i) {
    m <- lf_pred[[i]]
    yrs <- lf_year_list[[i]]
    as.data.frame.table(m, responseName = "pred", stringsAsFactors = FALSE) %>%
      rename(id = Var1, length = Var2) %>%
      mutate(fishery = names(lf_pred)[i],
             id = as.integer(factor(id, levels = unique(id))),
             year = yrs[id],
             length = as.numeric(length))
  })
  
  # ---- Observed: long data frame ----
  df_obs <- data.frame(data$lf_obs_in)
  names(df_obs) <- data$len_mid
  df_obs <- df_obs %>%
    mutate(year = data$lf_year, fishery = as.character(data$lf_fishery)) %>%
    group_by(fishery) %>%
    mutate(id = row_number()) %>%
    ungroup() %>%
    pivot_longer(cols = -c(year, fishery, id), names_to = "length", values_to = "obs") %>%
    mutate(length = as.numeric(length))
  
  # ---- Simulations: generate n_sim draws and convert to long format ----
  n_sim <- 5
  df_sim_all <- map_dfr(seq_len(n_sim), function(s) {
    # Determine which OBS vector was active based on lf_switch
    sim <- obj$simulate()
    if (data$lf_switch == 1) {
      sim_vec <- sim$lf_obs_flat
    } else if (data$lf_switch == 2) {
      sim_vec <- sim$lf_obs_prop
    } else {
      sim_vec <- sim$lf_obs_ints
    }
    sim_list <- flat_to_list(sim_vec, data$lf_n_f, nbins)
    names(sim_list) <- data$lf_fishery_f
    for (k in seq_along(sim_list)) colnames(sim_list[[k]]) <- data$len_mid
    map_dfr(seq_along(sim_list), function(i) {
      m <- sim_list[[i]]
      yrs <- lf_year_list[[i]]
      as.data.frame.table(m, responseName = "sim", stringsAsFactors = FALSE) %>%
        rename(id = Var1, length = Var2) %>%
        mutate(fishery = names(sim_list)[i],
               id = as.integer(factor(id, levels = unique(id))),
               year = yrs[id],
               length = as.numeric(length),
               sim_id = s)
    })
  })
  df_sim_all <- df_sim_all %>% mutate(year = ts_to_label(year))
  
  df <- left_join(df_obs, df_pred, by = c("id", "fishery", "length")) %>%
    select(-year.y) %>%
    # rename(year = year.x) %>% 
    mutate(year = ts_to_label(year.x))
  
  # yrs_plot <- 146:188
  # yrs_plot <- ts_to_label(146:188)
  yrs_plot <- sample(ts_to_label(unique(data$lf_year)), 20)
  yrs_plot <- ts_to_label(unique(data$lf_year))
  fishery_plot <- "14" # 8:14
  ggplot(df %>% filter(fishery == fishery_plot, year %in% yrs_plot), aes(x = length)) +
    geom_col(aes(y = obs), fill = "grey70", width = 2) +
    geom_line(data = df_sim_all %>% filter(fishery == fishery_plot, year %in% yrs_plot),
              aes(y = sim, group = sim_id), colour = "steelblue", alpha = 0.3, linewidth = 0.4) +
    geom_line(aes(y = pred), colour = "red3", linewidth = 0.7) +
    facet_wrap(~ year, scales = "free_y", ncol = 6) +
    labs(x = "Length (cm)", y = "Proportion",
         title = paste0("Length composition: Fishery ", fishery_plot),
         subtitle = "Grey bars = observed, red = predicted, blue = simulated") +
    theme(strip.text = element_text(size = 7), axis.text = element_text(size = 6))
}
```

Inspect predicted vs observed weight compositions when the weight-composition
likelihood is enabled.

```{r}
#| label: fig-wf-diagnostics
#| echo: true
#| message: false
#| fig-height: 5
#| fig-width: 7
#| fig-cap: Observed and predicted weight composition proportions for the first active weight-composition fishery.
if (data$wf_switch == 0) {
  cat("Weight composition likelihood is switched off (wf_switch = 0), so no predicted weight compositions to show.\n")
} else {
  wf_rep <- obj$report()
  wf_pred <- wf_rep$wf_pred
  wt_mid_trim <- data$wt_mid[data$wf_minbin[data$wf_fishery_f[1]]:data$wf_maxbin[data$wf_fishery_f[1]]]
  names(wf_pred) <- data$wf_fishery_f
  for (k in seq_along(wf_pred)) colnames(wf_pred[[k]]) <- wt_mid_trim
  wf_year_list <- split(data$wf_year, data$wf_fishery)
  # ---- Predicted: long data frame ----
  df_wf_pred <- map_dfr(seq_along(wf_pred), function(i) {
    m <- wf_pred[[i]]
    yrs <- wf_year_list[[i]]
    as.data.frame.table(m, responseName = "pred", stringsAsFactors = FALSE) %>%
      rename(id = Var1, weight = Var2) %>%
      mutate(fishery = names(wf_pred)[i],
             id = as.integer(factor(id, levels = unique(id))),
             year = yrs[id],
             weight = as.numeric(weight))
  })
  
  # ---- Observed: long data frame (first WF fishery only for this plot) ----
  df_wf_obs <- data.frame(data$wf_obs_in[
    seq_len(data$wf_n_f[1]), , drop = FALSE])
  names(df_wf_obs) <- data$wt_mid
  df_wf_obs <- df_wf_obs %>%
    mutate(year    = wf_year_list[[1]],
           fishery = as.character(data$wf_fishery_f[1])) %>%
    mutate(id = row_number()) %>%
    pivot_longer(cols = -c(year, fishery, id),
                 names_to = "weight", values_to = "obs") %>%
    mutate(weight = as.numeric(weight)) %>%
    filter(weight >= wt_mid_trim[1], weight <= wt_mid_trim[length(wt_mid_trim)])
  
  df_wf <- left_join(df_wf_obs, df_wf_pred, by = c("id", "fishery", "year", "weight"))
  
  # ---- Plot first fishery with WF data ----
  f_plot <- as.character(data$wf_fishery_f[1])
  max_yrs_wf_plot <- 24 # show at most 24 time steps for readability
  yrs_wf_plot <- sort(unique(df_wf$year))[seq_len(min(max_yrs_wf_plot, length(unique(df_wf$year))))]
  ggplot(df_wf %>% filter(fishery == f_plot, year %in% yrs_wf_plot), aes(x = weight)) +
    geom_col(aes(y = obs), fill = "grey70") +
    geom_line(aes(y = pred), colour = "red3", linewidth = 0.7) +
    facet_wrap(~ year, scales = "free_y", ncol = 6) +
    labs(x = "Weight (kg)", y = "Proportion",
         title = paste0("Weight composition: Fishery ", f_plot),
         subtitle = "Grey bars = observed, red = predicted") +
    theme(strip.text = element_text(size = 7), axis.text  = element_text(size = 6))
}
```

Inspect estimated spawning biomass trajectory:

```{r}
#| label: fig-opt-spbio
#| echo: true
#| message: false
#| fig-height: 4
#| fig-width: 7
#| fig-cap: Estimated spawning biomass with lognormal 50% and 95% uncertainty intervals.
#| eval: false
# Use sdreport output for uncertainty (Report from `sdreport(obj)`).
# Report$value and Report$sd are named vectors with repeated spawning_biomass_y entries.
sb_name <- "spawning_biomass_y"
sb_idx <- which(names(Report$value) == sb_name)
sb_mean <- Report$value[sb_idx]
sb_sd <- Report$sd[sb_idx]

# Build data frame with uncertainty intervals computed under a lognormal assumption
sb_df <- data.frame(
  year = seq_along(sb_mean),
  mean = sb_mean,
  sd = sb_sd
) %>%
  mutate(
    # lognormal parameters via delta method
    log_mu = log(mean^2 / sqrt(mean^2 + sd^2)),
    log_sigma = sqrt(log(1 + (sd^2 / mean^2))),
    median = exp(log_mu),
    lower50 = exp(log_mu - 0.6745 * log_sigma),
    upper50 = exp(log_mu + 0.6745 * log_sigma),
    lower95 = exp(log_mu - 1.96 * log_sigma),
    upper95 = exp(log_mu + 1.96 * log_sigma)
  )

ggplot(sb_df, aes(x = year)) +
  geom_ribbon(aes(ymin = lower95, ymax = upper95), fill = "steelblue", alpha = 0.15) +
  geom_ribbon(aes(ymin = lower50, ymax = upper50), fill = "steelblue", alpha = 0.30) +
  geom_line(aes(y = median), colour = "blue", linewidth = 1) +
  labs(
    x = "Time step",
    y = "Spawning biomass (mt)",
    title = "Estimated spawning biomass trajectory",
    subtitle = "Median with lognormal 50% and 95% uncertainty intervals"
  )
```

Inspect estimated dynamic depletion ($SB/SB_{F=0}$) trajectory:

```{r}
#| label: fig-opt-spdep
#| echo: true
#| message: false
#| fig-height: 4
#| fig-width: 7
#| fig-cap: Estimated dynamic depletion with lognormal 50% and 95% uncertainty intervals.
#| eval: false
# Use sdreport output for uncertainty (Report from `sdreport(obj)`).
# Report$value and Report$sd are named vectors with repeated dynamic_depletion_y entries.
sb_name <- "dynamic_depletion_y"
sb_idx <- which(names(Report$value) == sb_name)
sb_mean <- Report$value[sb_idx]
sb_sd <- Report$sd[sb_idx]

# Build data frame with uncertainty intervals computed under a lognormal assumption
sb_df <- data.frame(
  year = seq_along(sb_mean),
  mean = sb_mean,
  sd = sb_sd
) %>%
  mutate(
    # lognormal parameters via delta method
    log_mu = log(mean^2 / sqrt(mean^2 + sd^2)),
    log_sigma = sqrt(log(1 + (sd^2 / mean^2))),
    median = exp(log_mu),
    lower50 = exp(log_mu - 0.6745 * log_sigma),
    upper50 = exp(log_mu + 0.6745 * log_sigma),
    lower95 = exp(log_mu - 1.96 * log_sigma),
    upper95 = exp(log_mu + 1.96 * log_sigma)
  )

# Plot with ribbon bands (lognormal-centered)
ggplot(data = sb_df, aes(x = year)) +
  geom_ribbon(aes(ymin = lower95, ymax = upper95), fill = "steelblue", alpha = 0.15) +
  geom_ribbon(aes(ymin = lower50, ymax = upper50), fill = "steelblue", alpha = 0.30) +
  geom_line(aes(y = median), colour = "blue", size = 1) +
  labs(
    x = "Time step",
    y = "Dynamic depletion (SB/SBF=0)",
    title = "Estimated dynamic depletion trajectory",
    subtitle = "Median with lognormal 50% and 95% uncertainty intervals"
  )
```

# Simulation

The CPUE series is set up using RTMB's `OBS` mechanism inside the model, which
allows simulation via `obj$simulate()`. The plots below compare observed,
predicted, and simulated log-CPUE.

```{r}
#| label: fig-cpue-sim
#| echo: true
#| message: false
#| fig-height: 4
#| fig-width: 7
#| fig-cap: Observed, predicted, and simulated log-CPUE by index.
if (data$cpue_switch == 0) {
  cat("CPUE likelihood is switched off (cpue_switch = 0), so no simulated CPUE to show.\n")
} else {
  for (idx in seq_len(data$n_index)) {
    rows <- which(data$cpue_data$index == idx)
    plot(log(data$cpue_data$value[rows]), col = 2, pch = 16,
         xlab = "Observation", ylab = "log(CPUE)",
         main = paste0("Index ", idx, ": Observed vs predicted CPUE"))
    lines(log(obj$report()$cpue_pred[rows]), lwd = 2)
    for (s in 1:5) lines(obj$simulate()$cpue_log_obs[rows], col = "gray70")
  }
}
```

# One step ahead (OSA) residuals

OSA residuals are a replacement for Pearson residuals:

```{r}
#| label: fig-osa-residuals
#| echo: true
#| message: false
#| fig-height: 4
#| fig-width: 7
#| fig-cap: One-step-ahead CPUE residual QQ and residual-index diagnostics.
#| eval: false
if (data$cpue_switch == 0) {
  cat("CPUE likelihood is switched off (cpue_switch = 0), so no OSA residuals to show.\n")
} else {
  osa_cpue <- oneStepPredict(obj = obj, observation.name = "cpue_log_obs",
                             method = "oneStepGeneric", trace = FALSE)
  for (idx in seq_len(data$n_index)) {
    rows <- which(data$cpue_data$index == idx)
    qqnorm(osa_cpue$res[rows], main = paste0("Index ", idx, ": OSA QQ"))
    abline(0, 1)
    plot(osa_cpue$res[rows], xlab = "Observation", ylab = "OSA residual",
         main = paste0("Index ", idx, ": OSA residuals"))
    abline(h = c(-2, 0, 2), lty = c(2, 1, 2))
  }
}
```
