---
title: "The BET model"
format: html
execute:
  eval: false
vignette: >
  %\VignetteIndexEntry{The BET model}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{quarto::html}
editor_options: 
  chunk_output_type: console
---

```{r, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

# Introduction

`opal` is an open, modular R package for fisheries stock assessment built on 
[RTMB](https://github.com/kaskr/RTMB) for automatic differentiation. It includes 
bundled data from the western and central Pacific Ocean (WCPO) bigeye tuna (BET) 
stock assessment. This vignette demonstrates model setup, fitting, and 
diagnostics using those data.

# Load inputs

Load the `opal` package and the `RTMB` dependency. The `ggplot2` package is
used for plotting.

```{r load-pkg, echo=TRUE, message=FALSE}
library(opal)
library(tidyverse)
library(RTMB)

theme_set(theme_bw())
```

The bundled data object `wcpo_bet_data` contains all biological parameters,
catch and CPUE observations, length structure, and prior specifications needed
for the BET assessment model:

```{r load-data, echo=TRUE, message=FALSE}
data(wcpo_bet_data)
data <- wcpo_bet_data
if (is.null(data$n_index)) data$n_index <- 1L
if (is.null(data$cpue_data$index)) data$cpue_data$index <- rep(1L, nrow(data$cpue_data))
names(data)
```

Key dimensions of the data:

```{r data-dims, echo=TRUE}
cat("Number of ages:", data$n_age, "\n")
cat("Number of years:", data$n_year, "\n")
cat("Number of fisheries:", data$n_fishery, "\n")
cat("Number of length bins:", data$n_len, "\n")
```

## Biological inputs

Natural mortality at age, maturity at length, and length-weight parameters are
all contained in the data object.

```{r plot-biology, echo=TRUE, message=FALSE, fig.height=4, fig.width=7}
ages <- data$age_a
real_age <- ages / 4

# Natural mortality
ggplot(data = data.frame(age = real_age, M = data$M), aes(x = age, y = M)) +
  geom_line() + geom_point() +
  labs(x = "Age (years)", y = "M (quarterly rate)", 
       title = "Natural Mortality-at-Age")

# Maturity at length
len_mid <- seq(data$len_bin_start + data$len_bin_width / 2,
               by = data$len_bin_width, length.out = data$n_len)
ggplot(data = data.frame(length = len_mid, maturity = data$maturity), 
       aes(x = length, y = maturity)) +
  geom_line() + geom_point() +
  labs(x = "Length (cm)", y = "Maturity", title = "Maturity-at-Length")

fleet_names <- c("F01_LL.NORTH", "F02_LL.US", "F03_LL.OFFSH",
                 "F04_LL.EQUAT", "F05_LL.WEST", "F06_LL.SOUTH", "F07_LL.AUS",
                 "F08_PS.ASSOC", "F09_PS.UNASS", "F10_DOM.MISC",
                 "F11_DOM.HL", "F12_JP.PS.N", "F13_JP.PL", "F14_EQ.PL",
                 "S01_INDEX")
```

## CPUE data

```{r prep-cpue, echo=TRUE, message=FALSE}
data$cpue_data$index <- as.integer(factor(data$cpue_data$month, levels = c(2, 5, 8, 11)))
data$n_index <- 4L
```

## Length composition data

Length composition observations are loaded from the bundled `wcpo_bet_lf`
long-format data object and transformed into model-ready arrays.

```{r prep-lf, echo=TRUE, message=FALSE}
data(wcpo_bet_lf)

# Pivot to wide format: one row per fishery x timestep, bins as columns
# Note: LF data exist for fisheries 8-14, but we initially use only fisheries 8 & 9
lf_wide <- wcpo_bet_lf %>%
  pivot_wider(
    id_cols = c(fishery, year, month, ts),
    names_from = bin,
    values_from = value,
    values_fill = 0
  ) %>%
  arrange(fishery, ts)

unique(lf_wide$fishery)

# define variance adjustment scalars for each fishery if using lf_switch = 1 (multinomial)
var_adjust_scalars <- rep(1, data$n_fishery)
var_adjust_scalars[] <- 80
# var_adjust_scalars <- 1/rep(20000,data$n_fishery)
# var_adjust_scalars[c(1, 4, 5, 6, 15)] <- 1 / 40000
# var_adjust_scalars[] <- 1 / 4e9

data <- prep_lf_data(data = data, lf_wide = lf_wide, 
                     # lf_keep_fisheries = c(8),
                     # lf_keep_fisheries = c(8, 9, 10),
                     lf_var_adjust = var_adjust_scalars)

data$lf_fishery
range(data$lf_n) # aim for tens to low thousands per obs
sum(data$lf_n) # aim for ~1e4–1e5 total
```

## Weight composition data

Weight composition observations are loaded from the bundled `wcpo_bet_wf`
long-format data object and transformed into model-ready arrays.

```{r prep-wf, echo=TRUE, message=FALSE}
data(wcpo_bet_wf)

# Weight bin scalars (1 kg bins, 1–200 kg)
data$wt_bin_start <- 1
data$wt_bin_width <- 1
data$n_wt <- 200L

# Pivot to wide format: one row per fishery x timestep, bins as columns
wf_wide <- wcpo_bet_wf |>
  pivot_wider(
    id_cols = c(fishery, year, month, ts),
    names_from = bin,
    values_from = value,
    values_fill = 0
  ) |>
  arrange(fishery, ts)

unique(wf_wide$fishery)

data <- prep_wf_data(data = data, wf_wide = wf_wide,
                     wf_keep_fisheries = c(1),
                     wf_switch = 1L)
```

# Model setup

## Parameters

Define the initial parameter values. Growth parameters (`log_L1`, `log_L2`,
`log_k`, `log_CV1`, `log_CV2`) and selectivity (`par_sel`) are initialised at
reasonable starting values. Load these from the bundled `wcpo_bet_parameters` 
data object:

```{r get-pars, echo=TRUE, message=FALSE}
data(wcpo_bet_parameters)
wcpo_bet_parameters$par_sel[, 3:4] <- wcpo_bet_parameters$par_sel[, 3:4] - log(sd(data$len_mid))

parameters <- list(
  log_B0 = 20,
  log_h = as.numeric(wcpo_bet_parameters$log_h),
  log_sigma_r = as.numeric(wcpo_bet_parameters$log_sigma_r),
  log_cpue_q = rep(0, data$n_index),
  log_cpue_tau = rep(log(0.1), data$n_index),
  log_cpue_omega = rep(as.numeric(wcpo_bet_parameters$log_cpue_omega), data$n_index),
  log_lf_tau = as.numeric(log(rep(0.1, data$n_fishery))),
  log_wf_tau = rep(0, data$n_fishery),
  log_L1 = as.numeric(wcpo_bet_parameters$log_L1),
  log_L2 = as.numeric(wcpo_bet_parameters$log_L2),
  log_k = as.numeric(wcpo_bet_parameters$log_k),
  log_CV1 = as.numeric(wcpo_bet_parameters$log_CV1),
  log_CV2 = as.numeric(wcpo_bet_parameters$log_CV2),
  par_sel = as.matrix(wcpo_bet_parameters$par_sel),
  rdev_y = as.numeric(wcpo_bet_parameters$rdev_y)
)
```

## Priors

Priors are specified using `get_priors()`. The data object already contains
prior center values for growth parameters:

```{r get-priors, echo=TRUE, message=FALSE}
data$priors <- get_priors(parameters = parameters, data = data)
evaluate_priors(parameters = parameters, priors = data$priors)
```

## Parameter map

Use RTMB's `map` option to turn parameters on/off. Parameters mapped to
`factor(NA)` are fixed at their initial values:

```{r get-map, echo=TRUE, message=FALSE}
parameters$par_sel
wcpo_bet_parameters$par_sel

map_sel <- matrix(NA, nrow(parameters$par_sel), ncol(parameters$par_sel))

# 8-14 is LFs
map_sel[8, c(1, 3, 4)] <- 1:3 # peak, plateau, asc-width, desc-width
map_sel[9, c(1, 3, 4)] <- 4:6 # peak, plateau, asc-width, desc-width
map_sel[10, c(1, 3, 4)] <- 7:9 # peak, plateau, asc-width, desc-width
map_sel[11, 1:2] <- 10:11 # peak, plateau, asc-width, desc-width
map_sel[12, c(1, 3, 4)] <- 12:14 # peak, plateau, asc-width, desc-width
map_sel[13, c(1, 3, 4)] <- 15:17 # peak, plateau, asc-width, desc-width
map_sel[14, c(1, 3, 4)] <- 18:20 # peak, plateau, asc-width, desc-width

# 1-7 & 15 is WFs
# map_sel[1, c(1, 3, 4, 6)] <- 29:32

map_sel

map_rdev <- rep(NA, length(parameters$rdev_y))

map_lf_tau <- rep(NA, length(parameters$log_lf_tau))
map_lf_tau[c(8, 9)] <- NA

map <- list(
  # log_B0 = factor(NA),
  log_h = factor(NA),
  log_sigma_r = factor(NA),
  # log_cpue_q = factor(NA),
  log_cpue_q = factor(c(NA, 1, 2, 3)),
  log_cpue_tau = factor(rep(NA, data$n_index)),
  log_cpue_omega = factor(rep(NA, data$n_index)),
  log_lf_tau = factor(map_lf_tau),
  log_wf_tau = factor(rep(NA, data$n_fishery)),
  log_L1  = factor(NA),
  log_L2  = factor(NA),
  log_k   = factor(NA),
  log_CV1 = factor(NA),
  log_CV2 = factor(NA),
  par_sel = factor(map_sel)
  # rdev_y = as.factor(map_rdev),
)
```

## Build the AD object

Using the `data`, the `parameters`, the parameter `map`, and the model
(`opal_model`), the AD object is created using RTMB's `MakeADFun` function.
Optionally, we could also specify random effects here 
(e.g. `random = "rdev_y"`), but we'll start with a simpler fixed-effects model 
to check everything is working first.

```{r make-adfun, echo=TRUE, message=FALSE}
# data$lf_switch <- 0 # skip length comps (removal only)
data$lf_switch <- 1 # multinomial likelihood on flat counts (default)
# data$lf_switch <- 2 # fails while fitting, need to sort out log_lf_tau for this, use simulate to tune and find
# data$lf_switch <- 3 # fails - wants integers - think this should be an issue to RTMBdist guys

# Note: wf_switch was already set by prep_wf_data(); it is re-stated here for
# clarity alongside the equivalent lf_switch assignment.
data$wf_switch <- 0 # skip weight comps (removal only)
# data$wf_switch <- 1 # multinomial likelihood on flat counts (default)
# data$wf_switch <- 2 # Dirichlet
# data$wf_switch <- 3 # Dirichlet-multinomial

obj <- MakeADFun(func = cmb(opal_model, data), parameters = parameters, map = map)
obj$par
unique(names(obj$par))
obj$fn()
obj$gr()
# TMB::benchmark(obj, n = 500, expr = expression(obj$gr()))
```

Inspect initial model outputs:

```{r init-checks-1, echo=TRUE, message=FALSE, fig.height=4, fig.width=7}
plot(obj$report()$spawning_biomass_y, type = "l",
     xlab = "Time step", ylab = "Spawning biomass (mt)",
     main = "Initial spawning biomass trajectory")
```

```{r init-checks-2, echo=TRUE, message=FALSE, fig.height=4, fig.width=7}
plot_catch(data = data, obj = obj)
```

## Parameter bounds

```{r get-par-bounds, echo=TRUE, message=FALSE}
Lwr <- rep(-Inf, length(obj$par))
Upr <- rep(Inf, length(obj$par))
Lwr[grep("log_B0", names(obj$par))] <- log(1)
Upr[grep("log_B0", names(obj$par))] <- log(exp(22))
Lwr[grep("log_cpue_q", names(obj$par))] <- log(0.1)
Upr[grep("log_cpue_q", names(obj$par))] <- log(10)
Lwr[grep("log_lf_tau", names(obj$par))] <- rep(-9, length(grep("log_lf_tau", names(obj$par))))
Upr[grep("log_lf_tau", names(obj$par))] <- rep(9, length(grep("log_lf_tau", names(obj$par))))
Lwr[grep("par_sel", names(obj$par))] <- rep(-15, length(grep("par_sel", names(obj$par))))
Upr[grep("par_sel", names(obj$par))] <- rep(15, length(grep("par_sel", names(obj$par))))
Lwr[grep("rdev_y", names(obj$par))] <- rep(-5, length(grep("rdev_y", names(obj$par))))
Upr[grep("rdev_y", names(obj$par))] <- rep(5, length(grep("rdev_y", names(obj$par))))


# remove generic par_sel bounds
Lwr[grep("par_sel", names(obj$par))] <- -Inf
Upr[grep("par_sel", names(obj$par))] <-  Inf

mu_len <- mean(data$len_mid)
sd_len <- sd(data$len_mid)

for (f in seq_len(data$n_fishery)) {
  idx <- grep(paste0("^par_sel\\[", f, ","), names(obj$par))

  if (data$sel_type_f[f] == 1L) {
    # Logistic: inflection [5,200], width [>0,500]
    Lwr[idx[1]] <- (5   - mu_len) / sd_len
    Upr[idx[1]] <- (200 - mu_len) / sd_len

    Lwr[idx[2]] <- log(1e-6 / sd_len)   # avoid log(0)
    Upr[idx[2]] <- log(500  / sd_len)

    # unused logistic slots
    Lwr[idx[3:6]] <- -Inf
    Upr[idx[3:6]] <-  Inf
  } else {
    # Double-normal SS3 bounds
    Lwr[idx[1]] <- (10.1 - mu_len) / sd_len
    Upr[idx[1]] <- (200  - mu_len) / sd_len

    Lwr[idx[2]] <- -7
    Upr[idx[2]] <-  7

    # for sel_double_normal using exp(c)*sd^2 and exp(d)*sd^2:
    shift <- 2 * log(sd_len)
    Lwr[idx[3]] <- -7 - shift
    Upr[idx[3]] <-  7 - shift
    Lwr[idx[4]] <- -7 - shift
    Upr[idx[4]] <-  7 - shift

    Lwr[idx[5]] <- -9    # practical bound (instead of -999)
    Upr[idx[5]] <-  9
    Lwr[idx[6]] <- -9
    Upr[idx[6]] <-  9
  }
}

bounds <- data.frame(par = names(obj$par), lower = Lwr, upper = Upr)
```

# Optimisation

Optimise using the `nlminb` function:

```{r run-nlminb, echo=TRUE, results="hide"}
control <- list(eval.max = 10000, iter.max = 10000)
# control <- list(eval.max = 100, iter.max = 100)
opt <- nlminb(start = obj$par, objective = obj$fn, gradient = obj$gr,
              hessian = obj$he, lower = Lwr, upper = Upr, control = control)
opt <- nlminb(start = opt$par, objective = obj$fn, gradient = obj$gr,
              hessian = obj$he, lower = Lwr, upper = Upr, control = control)
obj$fn()
max(abs(obj$gr()))
obj$report()$lp_prior
obj$report()$lp_penalty
obj$report()$lp_rec
sum(obj$report()$lp_cpue)
sum(obj$report()$lp_lf)
sum(obj$report()$lp_wf)
check_estimability(obj = obj)
```

Compare initial and estimated parameter values:

```{r opt-results, echo=TRUE, message=FALSE}
get_par_table(obj, parameters, map, lower = Lwr, upper = Upr, grad_tol = 1e-2, digits = 2L)
```

## Selectivity

Visualise the selectivity curves by fleet (at length):

```{r plot-selectivity-length, echo=TRUE, message=FALSE, fig.height=8, fig.width=10, fig.cap="Selectivity-at-Length by Fleet"}
par_sel_est  <- obj$env$parList(obj$env$last.par.best)$par_sel
par_sel_orig <- wcpo_bet_parameters$par_sel

map_sel_mat <- matrix(map$par_sel, nrow = nrow(par_sel_est), ncol = ncol(par_sel_est))
fishery_estimated <- apply(map_sel_mat, 1, function(row) any(!is.na(row)))

sel_est_df <- do.call(rbind, lapply(seq_len(data$n_fishery), function(f) {
  sel_l <- if (data$sel_type_f[f] == 1L) {
    sel_logistic(len_mid, par_sel_est[f, ])
  } else {
    sel_double_normal(len_mid, par_sel_est[f, ])
  }
  data.frame(
    fishery = f,
    fleet_name = fleet_names[f],
    length = len_mid,
    selectivity = as.numeric(sel_l),
    estimated = fishery_estimated[f]
  )
}))

sel_orig_df <- do.call(rbind, lapply(seq_len(data$n_fishery), function(f) {
  sel_l <- if (data$sel_type_f[f] == 1L) {
    sel_logistic(len_mid, par_sel_orig[f, ])
  } else {
    sel_double_normal(len_mid, par_sel_orig[f, ])
  }
  data.frame(
    fishery = f,
    fleet_name = fleet_names[f],
    length = len_mid,
    selectivity = as.numeric(sel_l)
  )
}))

ggplot(data = sel_est_df, aes(x = length, y = selectivity, colour = estimated)) +
  geom_line() +
  geom_line(
    data = sel_orig_df,
    aes(x = length, y = selectivity),
    inherit.aes = FALSE,
    colour = "black",
    linetype = "dashed",
    linewidth = 0.7
  ) +
  facet_wrap(fleet_name ~ ., ncol = 4) +
  labs(x = "Length (cm)", y = "Selectivity") +
  coord_cartesian(ylim = c(0, 1))
```

Visualise the selectivity curves by fleet (at age):

```{r plot-selectivity, echo=TRUE, message=FALSE, fig.height=8, fig.width=10, fig.cap="Selectivity-at-Age by Fleet"}
rep <- obj$report()
sel_fya <- rep$sel_fya

sel_df <- expand.grid(fishery = 1:data$n_fishery, age = 1:data$n_age)
sel_df$selectivity <- sapply(1:nrow(sel_df), function(i) {
  sel_fya[sel_df$fishery[i], 1, sel_df$age[i]]
})
sel_df$real_age <- sel_df$age / 4
sel_df$fleet_name <- fleet_names[sel_df$fishery]

ggplot(sel_df, aes(x = real_age, y = selectivity)) +
  geom_line() +
  facet_wrap(~fleet_name, ncol = 4) +
  labs(x = "Age (years)", y = "Selectivity") +
  ylim(0, 1)
```

Build a long-format table of selectivity parameters.

```{r sel-par-table, echo=TRUE, message=FALSE}
# Three sources to compare side by side:
#   1. Initial values used to build obj   (parameters$par_sel)
#   2. Bundled reference values            (wcpo_bet_parameters$par_sel)
#   3. Estimated / mapped status           (map$par_sel)
#
# par_sel is [n_fishery x 6] with columns (a, b, c, d, e, f) for double-normal
# or (par1, par2) for logistic (cols 3-6 ignored). `factor(map_sel)` flattens
# matrices column-major, so wrap the map vector back into the same shape.

n_f <- nrow(parameters$par_sel)
n_p <- ncol(parameters$par_sel)
col_lbls <- c("a (peak)", "b (plateau)", "c (asc width)", "d (desc width)", "e (init sel)", "f (final sel)")

# Estimated/fixed status from the map
map_sel_mat <- if (is.null(map$par_sel)) {
  matrix(seq_len(n_f * n_p), n_f, n_p) # everything estimated if no map
} else {
  matrix(map$par_sel, n_f, n_p)
}
estimated_mat <- !is.na(map_sel_mat)

# Reference values (may not have row/col names — assume same shape)
ref_mat <- as.matrix(wcpo_bet_parameters$par_sel)
stopifnot("wcpo_bet_parameters$par_sel must match parameters$par_sel shape" = all(dim(ref_mat) == c(n_f, n_p)))

# Selectivity type (logistic = 1, double-normal = 2)
sel_type_lbl <- ifelse(data$sel_type_f == 1L, "logistic", "double-normal")

# For logistic fisheries, columns 3-6 are unused -> NA-out for clarity
init_mat <- parameters$par_sel
for (f in seq_len(n_f)) {
  if (data$sel_type_f[f] == 1L) {
    init_mat[f, 3:6] <- NA
    ref_mat[f, 3:6] <- NA
    estimated_mat[f, 3:6] <- FALSE
  }
}

# Assemble long table
sel_par_tbl <- tibble(
  fishery    = rep(seq_len(n_f), times = n_p),
  fleet_name = rep(fleet_names[seq_len(n_f)], times = n_p),
  sel_type   = rep(sel_type_lbl, times = n_p),
  param      = rep(col_lbls, each = n_f),
  init       = as.numeric(init_mat),
  reference  = as.numeric(ref_mat),
  estimated  = as.vector(estimated_mat)
) %>%
  mutate(diff = init - reference) %>%
  arrange(fishery, match(param, col_lbls))

sel_par_tbl %>%
  mutate(
    init      = ifelse(is.na(init),      "—", formatC(init,      digits = 3, format = "f")),
    reference = ifelse(is.na(reference), "—", formatC(reference, digits = 3, format = "f")),
    diff      = ifelse(is.na(diff),      "—", formatC(diff,      digits = 3, format = "f")),
    estimated = ifelse(estimated, "✓", "")
  ) %>%
  select(Fleet = fleet_name,
         Type = sel_type,
         Param = param,
         Initial = init,
         `Reference (bundled)` = reference,
         `Δ (init − ref)` = diff,
         `Est?` = estimated) %>%
  knitr::kable(
    caption = "Selectivity parameter values: initial vs. bundled reference, with map status.",
    align   = c("l", "l", "l", "r", "r", "r", "c")
  )
```

## Diagnostics

Check that all parameters are estimable using the `check_estimability` function:

```{r check-estimability, echo=TRUE, message=FALSE, error = TRUE}
check_estimability(obj = obj)
```

Look for any parameters with high correlations (absolute correlation > 0.95) using the `get_cor_pairs` function:

```{r check-cor, echo=TRUE, message=FALSE}
get_cor_pairs(obj = obj, threshold = 0.95)
```

Calculate standard deviations of all model parameters:

```{r run-sdreport, echo=TRUE, message=FALSE}
Report <- sdreport(obj)
```

## Data fits

Inspect fitted CPUE and catch:

```{r opt-cpue, echo=TRUE, message=FALSE, fig.height=4, fig.width=7}
if (data$cpue_switch == 0) {
  cat("CPUE likelihood is switched off (cpue_switch = 0), so no predicted CPUE to show.\n")
} else {
  rep <- obj$report(obj$env$last.par.best)
  
  df <- data$cpue_data %>%
    mutate(obs_id = row_number(), pred = rep$cpue_pred, sigma = rep$cpue_sigma)
  
  # Simulations
  n_sim <- 25
  df_sim <- map_dfr(seq_len(n_sim), function(s) {
    sim_log <- obj$simulate()$cpue_log_obs
    tibble(
      obs_id = seq_along(sim_log),
      year = data$cpue_data$year,
      month = data$cpue_data$month,
      fishery = data$cpue_data$fishery,
      sim_id = s,
      sim = exp(sim_log)
    )
  })
  
  month_levels <- sort(unique(data$cpue_data$month))
  fish_levels <- sort(unique(data$cpue_data$fishery))
  format_facets <- function(d) {
    d %>%
      mutate(
        month = factor(month, levels = month_levels, labels = paste0("Month: ", month_levels)),
        fishery = factor(fishery, levels = fish_levels, labels = paste0("Fishery: ", fish_levels))
      )
  }
  df <- format_facets(df)
  df_sim <- format_facets(df_sim)
  
  ggplot(data = df, aes(x = year)) +
    geom_line(data = df_sim, aes(y = sim, group = sim_id),
              colour = "lightblue", alpha = 0.5, linewidth = 0.3) +
    geom_linerange(aes(ymin = exp(log(value) - sigma),
                       ymax = exp(log(value) + sigma),
                       colour = "Observed"), alpha = 0.6) +
    geom_point(aes(y = value, colour = "Observed")) +
    geom_line(aes(y = pred, colour = "Predicted"), linewidth = 0.8) +
    facet_grid(fishery ~ month, scales = "free_y") +
    scale_y_continuous(limits = c(0, NA), expand = expansion(mult = c(0, 0.05))) +
    scale_x_continuous(breaks = scales::pretty_breaks()) +
    scale_colour_manual(values = c(Observed = "black", Predicted = "red3")) +
    labs(x = "Year", y = "CPUE", colour = NULL, subtitle = paste0(n_sim, " posterior-predictive simulations (grey)")) +
    theme(legend.position = "bottom")
}
```

```{r opt-catch, echo=TRUE, message=FALSE, fig.height=4, fig.width=7}
sum(obj$report()$catch_pred_ysf - data$catch_obs_ysf)
plot_catch(data = data, obj = obj)
plot_catch(data = data, obj = obj, plot_resid = TRUE)
```

Inspect predicted vs observed length compositions:

```{r lf-diagnostics, echo=TRUE, message=FALSE, fig.height=6, fig.width=7}
# Convert model timestep -> "YYYY Qn" label
ts_to_label <- function(ts, first_yr = 1952, n_season = 4) {
  yr <- first_yr + (ts - 1L) %/% n_season
  qn <- ((ts - 1L) %% n_season) + 1L
  paste0(yr, " Q", qn)
}

if (data$lf_switch == 0) {
  cat("Length composition likelihood is switched off (lf_switch = 0), so no predicted length compositions to show.\n")
} else {
  lf_rep <- obj$report()
  lf_pred <- lf_rep$lf_pred
  nbins <- data$lf_maxbin[1] - data$lf_minbin[1] + 1
  # Name the list elements and columns
  names(lf_pred) <- data$lf_fishery_f
  for (k in seq_along(lf_pred)) colnames(lf_pred[[k]]) <- data$len_mid
  
  # Helper: convert a flat OBS vector back to list-of-matrices (proportions)
  flat_to_list <- function(v, lf_n_f, nbins) {
    out <- vector("list", length(lf_n_f))
    offset <- 0L
    for (k in seq_along(lf_n_f)) {
      m <- matrix(v[(offset + 1):(offset + lf_n_f[k] * nbins)], nrow = lf_n_f[k], ncol = nbins, byrow = TRUE)
      # Normalise each row to proportions
      rs <- rowSums(m)
      rs[rs == 0] <- 1
      out[[k]] <- m / rs
      offset <- offset + lf_n_f[k] * nbins
    }
    out
  }
  
  # Split year index by fishery (must match length of lf_year)
  lf_year_list <- split(data$lf_year, data$lf_fishery)
  
  # ---- Predicted: long data frame ----
  df_pred <- map_dfr(seq_along(lf_pred), function(i) {
    m <- lf_pred[[i]]
    yrs <- lf_year_list[[i]]
    as.data.frame.table(m, responseName = "pred", stringsAsFactors = FALSE) %>%
      rename(id = Var1, length = Var2) %>%
      mutate(fishery = names(lf_pred)[i],
             id = as.integer(factor(id, levels = unique(id))),
             year = yrs[id],
             length = as.numeric(length))
  })
  
  # ---- Observed: long data frame ----
  df_obs <- data.frame(data$lf_obs_in)
  names(df_obs) <- data$len_mid
  df_obs <- df_obs %>%
    mutate(year = data$lf_year, fishery = as.character(data$lf_fishery)) %>%
    group_by(fishery) %>%
    mutate(id = row_number()) %>%
    ungroup() %>%
    pivot_longer(cols = -c(year, fishery, id), names_to = "length", values_to = "obs") %>%
    mutate(length = as.numeric(length))
  
  # ---- Simulations: generate n_sim draws and convert to long format ----
  n_sim <- 5
  df_sim_all <- map_dfr(seq_len(n_sim), function(s) {
    # Determine which OBS vector was active based on lf_switch
    sim <- obj$simulate()
    if (data$lf_switch == 1) {
      sim_vec <- sim$lf_obs_flat
    } else if (data$lf_switch == 2) {
      sim_vec <- sim$lf_obs_prop
    } else {
      sim_vec <- sim$lf_obs_ints
    }
    sim_list <- flat_to_list(sim_vec, data$lf_n_f, nbins)
    names(sim_list) <- data$lf_fishery_f
    for (k in seq_along(sim_list)) colnames(sim_list[[k]]) <- data$len_mid
    map_dfr(seq_along(sim_list), function(i) {
      m <- sim_list[[i]]
      yrs <- lf_year_list[[i]]
      as.data.frame.table(m, responseName = "sim", stringsAsFactors = FALSE) %>%
        rename(id = Var1, length = Var2) %>%
        mutate(fishery = names(sim_list)[i],
               id = as.integer(factor(id, levels = unique(id))),
               year = yrs[id],
               length = as.numeric(length),
               sim_id = s)
    })
  })
  df_sim_all <- df_sim_all %>% mutate(year = ts_to_label(year))
  
  df <- left_join(df_obs, df_pred, by = c("id", "fishery", "length")) %>%
    select(-year.y) %>%
    # rename(year = year.x) %>% 
    mutate(year = ts_to_label(year.x))
  
  # yrs_plot <- 146:188
  # yrs_plot <- ts_to_label(146:188)
  yrs_plot <- sample(ts_to_label(unique(data$lf_year)), 20)
  yrs_plot <- ts_to_label(unique(data$lf_year))
  fishery_plot <- "14" # 8:14
  ggplot(df %>% filter(fishery == fishery_plot, year %in% yrs_plot), aes(x = length)) +
    geom_col(aes(y = obs), fill = "grey70", width = 2) +
    geom_line(data = df_sim_all %>% filter(fishery == fishery_plot, year %in% yrs_plot),
              aes(y = sim, group = sim_id), colour = "steelblue", alpha = 0.3, linewidth = 0.4) +
    geom_line(aes(y = pred), colour = "red3", linewidth = 0.7) +
    facet_wrap(~ year, scales = "free_y", ncol = 6) +
    labs(x = "Length (cm)", y = "Proportion",
         title = paste0("Length composition: Fishery ", fishery_plot),
         subtitle = "Grey bars = observed, red = predicted, blue = simulated") +
    theme(strip.text = element_text(size = 7), axis.text = element_text(size = 6))
}
```

Inspect predicted vs observed weight compositions:

```{r wf-diagnostics, echo=TRUE, message=FALSE, fig.height=5, fig.width=7}
if (data$wf_switch == 0) {
  cat("Weight composition likelihood is switched off (wf_switch = 0), so no predicted weight compositions to show.\n")
} else {
  wf_rep <- obj$report()
  wf_pred <- wf_rep$wf_pred
  wt_mid_trim <- data$wt_mid[data$wf_minbin[data$wf_fishery_f[1]]:data$wf_maxbin[data$wf_fishery_f[1]]]
  names(wf_pred) <- data$wf_fishery_f
  for (k in seq_along(wf_pred)) colnames(wf_pred[[k]]) <- wt_mid_trim
  wf_year_list <- split(data$wf_year, data$wf_fishery)
  # ---- Predicted: long data frame ----
  df_wf_pred <- map_dfr(seq_along(wf_pred), function(i) {
    m <- wf_pred[[i]]
    yrs <- wf_year_list[[i]]
    as.data.frame.table(m, responseName = "pred", stringsAsFactors = FALSE) %>%
      rename(id = Var1, weight = Var2) %>%
      mutate(fishery = names(wf_pred)[i],
             id = as.integer(factor(id, levels = unique(id))),
             year = yrs[id],
             weight = as.numeric(weight))
  })
  
  # ---- Observed: long data frame (first WF fishery only for this plot) ----
  df_wf_obs <- data.frame(data$wf_obs_in[
    seq_len(data$wf_n_f[1]), , drop = FALSE])
  names(df_wf_obs) <- data$wt_mid
  df_wf_obs <- df_wf_obs %>%
    mutate(year    = wf_year_list[[1]],
           fishery = as.character(data$wf_fishery_f[1])) %>%
    mutate(id = row_number()) %>%
    pivot_longer(cols = -c(year, fishery, id),
                 names_to = "weight", values_to = "obs") %>%
    mutate(weight = as.numeric(weight)) %>%
    filter(weight >= wt_mid_trim[1], weight <= wt_mid_trim[length(wt_mid_trim)])
  
  df_wf <- left_join(df_wf_obs, df_wf_pred, by = c("id", "fishery", "year", "weight"))
  
  # ---- Plot first fishery with WF data ----
  f_plot <- as.character(data$wf_fishery_f[1])
  max_yrs_wf_plot <- 24 # show at most 24 time steps for readability
  yrs_wf_plot <- sort(unique(df_wf$year))[seq_len(min(max_yrs_wf_plot, length(unique(df_wf$year))))]
  ggplot(df_wf %>% filter(fishery == f_plot, year %in% yrs_wf_plot), aes(x = weight)) +
    geom_col(aes(y = obs), fill = "grey70") +
    geom_line(aes(y = pred), colour = "red3", linewidth = 0.7) +
    facet_wrap(~ year, scales = "free_y", ncol = 6) +
    labs(x = "Weight (kg)", y = "Proportion",
         title = paste0("Weight composition: Fishery ", f_plot),
         subtitle = "Grey bars = observed, red = predicted") +
    theme(strip.text = element_text(size = 7), axis.text  = element_text(size = 6))
}
```

Inspect estimated spawning biomass trajectory:

```{r opt-spbio, echo=TRUE, message=FALSE, fig.height=4, fig.width=7}
# Use sdreport output for uncertainty (Report from `sdreport(obj)`).
# Report$value and Report$sd are named vectors with repeated spawning_biomass_y entries.
sb_name <- "spawning_biomass_y"
sb_idx <- which(names(Report$value) == sb_name)
sb_mean <- Report$value[sb_idx]
sb_sd <- Report$sd[sb_idx]

# Build data frame with uncertainty intervals computed under a lognormal assumption
sb_df <- data.frame(
  year = seq_along(sb_mean),
  mean = sb_mean,
  sd = sb_sd
) %>%
  mutate(
    # lognormal parameters via delta method
    log_mu = log(mean^2 / sqrt(mean^2 + sd^2)),
    log_sigma = sqrt(log(1 + (sd^2 / mean^2))),
    median = exp(log_mu),
    lower50 = exp(log_mu - 0.6745 * log_sigma),
    upper50 = exp(log_mu + 0.6745 * log_sigma),
    lower95 = exp(log_mu - 1.96 * log_sigma),
    upper95 = exp(log_mu + 1.96 * log_sigma)
  )

ggplot(sb_df, aes(x = year)) +
  geom_ribbon(aes(ymin = lower95, ymax = upper95), fill = "steelblue", alpha = 0.15) +
  geom_ribbon(aes(ymin = lower50, ymax = upper50), fill = "steelblue", alpha = 0.30) +
  geom_line(aes(y = median), colour = "blue", linewidth = 1) +
  labs(
    x = "Time step",
    y = "Spawning biomass (mt)",
    title = "Estimated spawning biomass trajectory",
    subtitle = "Median with lognormal 50% and 95% uncertainty intervals"
  )
```

Inspect estimated dynamic depletion ($SB/SB_{F=0}$) trajectory:

```{r opt-spdep, echo=TRUE, message=FALSE, fig.height=4, fig.width=7}
# Use sdreport output for uncertainty (Report from `sdreport(obj)`).
# Report$value and Report$sd are named vectors with repeated dynamic_depletion_y entries.
sb_name <- "dynamic_depletion_y"
sb_idx <- which(names(Report$value) == sb_name)
sb_mean <- Report$value[sb_idx]
sb_sd <- Report$sd[sb_idx]

# Build data frame with uncertainty intervals computed under a lognormal assumption
sb_df <- data.frame(
  year = seq_along(sb_mean),
  mean = sb_mean,
  sd = sb_sd
) %>%
  mutate(
    # lognormal parameters via delta method
    log_mu = log(mean^2 / sqrt(mean^2 + sd^2)),
    log_sigma = sqrt(log(1 + (sd^2 / mean^2))),
    median = exp(log_mu),
    lower50 = exp(log_mu - 0.6745 * log_sigma),
    upper50 = exp(log_mu + 0.6745 * log_sigma),
    lower95 = exp(log_mu - 1.96 * log_sigma),
    upper95 = exp(log_mu + 1.96 * log_sigma)
  )

# Plot with ribbon bands (lognormal-centered)
ggplot(data = sb_df, aes(x = year)) +
  geom_ribbon(aes(ymin = lower95, ymax = upper95), fill = "steelblue", alpha = 0.15) +
  geom_ribbon(aes(ymin = lower50, ymax = upper50), fill = "steelblue", alpha = 0.30) +
  geom_line(aes(y = median), colour = "blue", size = 1) +
  labs(
    x = "Time step",
    y = "Dynamic depletion (SB/SBF=0)",
    title = "Estimated dynamic depletion trajectory",
    subtitle = "Median with lognormal 50% and 95% uncertainty intervals"
  )
```

# Simulation

The CPUE series is set up using RTMB's `OBS` mechanism inside the model, which
allows simulation via `obj$simulate()`. For example:

```{r sim-1, echo=TRUE, message=FALSE, fig.height=4, fig.width=7}
if (data$cpue_switch == 0) {
  cat("CPUE likelihood is switched off (cpue_switch = 0), so no simulated CPUE to show.\n")
} else {
  for (idx in seq_len(data$n_index)) {
    rows <- which(data$cpue_data$index == idx)
    plot(log(data$cpue_data$value[rows]), col = 2, pch = 16,
         xlab = "Observation", ylab = "log(CPUE)",
         main = paste0("Index ", idx, ": Observed vs predicted CPUE"))
    lines(log(obj$report()$cpue_pred[rows]), lwd = 2)
    for (s in 1:5) lines(obj$simulate()$cpue_log_obs[rows], col = "gray70")
  }
}
```

# One step ahead (OSA) residuals

OSA residuals are a replacement for Pearson residuals:

```{r osa1, echo=TRUE, message=FALSE, fig.height=4, fig.width=7}
if (data$cpue_switch == 0) {
  cat("CPUE likelihood is switched off (cpue_switch = 0), so no OSA residuals to show.\n")
} else {
  osa_cpue <- oneStepPredict(obj = obj, observation.name = "cpue_log_obs",
                             method = "oneStepGeneric", trace = FALSE)
  for (idx in seq_len(data$n_index)) {
    rows <- which(data$cpue_data$index == idx)
    qqnorm(osa_cpue$res[rows], main = paste0("Index ", idx, ": OSA QQ"))
    abline(0, 1)
    plot(osa_cpue$res[rows], xlab = "Observation", ylab = "OSA residual",
         main = paste0("Index ", idx, ": OSA residuals"))
    abline(h = c(-2, 0, 2), lty = c(2, 1, 2))
  }
}
```
