---
title: "Length-based population dynamics in opal"
format: html
execute:
  eval: true
  warning: false
  message: false
vignette: >
  %\VignetteIndexEntry{Length-based population dynamics in opal}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{quarto::html}
editor_options:
  chunk_output_type: console
---

```{r}
#| include: false
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 8, fig.align = "center")
```

# Overview

opal's population dynamics have historically been **age-structured**: numbers are
tracked at age, and length enters only at the boundaries of the model. A static
probability-of-length-at-age matrix (the age--length key, `get_pla()`) is used to
collapse length-based biology and selectivity onto the age axis before the
population loop runs, and to expand numbers-at-age back into predicted length
compositions in the likelihood. A consequence of this design is that the
distribution of lengths *at a given age is fixed for all time* --- it cannot
respond to size-selective fishing.

This vignette showcases an **opt-in joint age--length dynamics engine** that
relaxes that assumption. The engine carries a full numbers-at-age-and-length
state and advances it each year with a **growth transition matrix**
(`get_growth_matrix()`) --- a length-bin-to-length-bin operator built from the
same von Bertalanffy growth parameters opal already estimates. The approach
mirrors the length-and-age Stan model in the `lbm` project. It is enabled with a
single data switch,

```r
data$length_dynamics <- 1L
```

and is otherwise a drop-in: the age-only path is unchanged when the switch is
absent, and the age marginal of the length state (`number_ysa`) is produced as a
by-product so all downstream tooling (projections, CPUE, `sdreport`) keeps
working.

We first illustrate the growth-transition machinery and the size-selective
length-at-age truncation it can represent, then fit the 'opakapaka
(*Pristipomoides filamentosus*) assessment with both engines to show that the
length engine reproduces the established age-based results, and finish with a
forward projection.

```{r}
#| label: packages
library(opal)
library(RTMB)
library(tidyverse)
library(patchwork)

# Cohesive, cowplot-style theme with faded gridlines; no plot titles.
theme_opal <- function() {
  theme_minimal(base_size = 11) +
    theme(panel.grid.minor = element_blank(),
          panel.grid.major = element_line(colour = "grey90", linewidth = 0.3),
          plot.title = element_blank(), legend.position = "bottom")
}
theme_set(theme_opal())
pal <- c(Age = "grey45", Length = "#2E8B57",
         Unfished = "#4C9BD1", Fished = "#E8862E")
```

# The 'opakapaka model, two ways

The data, parameters, priors, and map are the *opal with length compositions*
configuration from `vignette("opakapaka")`, reproduced compactly. We build two
objects from identical inputs: the standard age engine and the length engine
(`length_dynamics = 1`).

```{r}
#| label: data
data(opaka_data); data <- opaka_data
if (is.null(data$n_index)) data$n_index <- 2L
data$priors  <- NULL
data$A1      <- 0
data$A2      <- data$A1 + (data$n_age - 1)
data$min_age <- 0
ages <- data$age_a <- seq(data$min_age, by = 1, length.out = data$n_age)
data$sex_ratio <- rep(1, data$n_age)

data(opaka_lf)
lf_wide <- opaka_lf %>%
  tidyr::pivot_wider(id_cols = c(fishery, year, month, ts),
                     names_from = bin, values_from = value, values_fill = 0) %>%
  arrange(fishery, ts)
data <- prep_lf_data(data, lf_wide, lf_keep_fisheries = c(1, 3),
                     lf_var_adjust = 1 / rep(1, data$n_fishery),
                     lf_addtocomp = 1e-10)

calendar_years    <- (1949 - 1) + data$years
init_cohort_years <- 1949 - (data$age_a - 1)
data$init_bias_adj_a <- get_bias_adj_vector(init_cohort_years, 1,
  c(1931, 1974, 2021.88, 2022.43), 0.869)
data$bias_adj_y <- get_bias_adj_vector(calendar_years, 1,
  c(1931, 1974, 2021.88, 2022.43), 0.869)
```

```{r}
#| label: parameters
data(opaka_parameters)
init_rdev_a <- rev(c(0.0206807,0.00321579,0.00372620,0.0043174,0.00500263,
  0.00579674,0.00671727,0.00778422,0.00902081,0.0104539,0.0121143,0.0140375,
  0.0162641,0.0188404,0.0218182,0.0252561,0.0292182,0.0337726,0.0389898,
  0.0449362,0.0516662,0.0592049,0.0675236,0.0764992,0.0858443,0.0950047,
  0.102706,0.107687,0.107712,0.0998349,0.0808137,0.0482339,0.00234443,
  -0.0517563,-0.102395,-0.131458,-0.121101,-0.0714826,-0.00742166,-0.0109308,
  -0.0385029,0.0366483,0.0165105,-0.166906))
init_rdev_a[1] <- opaka_parameters$rdev_y[1]
rdev_y_shifted <- c(opaka_parameters$rdev_y[-1], 0.0)

parameters <- list(
  log_B0 = 9, log_h = as.numeric(opaka_parameters$log_h),
  log_sigma_r = as.numeric(opaka_parameters$log_sigma_r),
  log_cpue_q = as.numeric(opaka_parameters$log_cpue_q),
  log_cpue_tau = rep(log(0.1), data$n_index),
  log_cpue_omega = as.numeric(rep(opaka_parameters$log_cpue_omega, data$n_index)),
  log_lf_tau = as.numeric(log(rep(0.1, data$n_fishery))),
  log_L1 = as.numeric(opaka_parameters$log_L1),
  log_L2 = as.numeric(opaka_parameters$log_L2),
  log_k = as.numeric(opaka_parameters$log_k),
  log_CV1 = as.numeric(opaka_parameters$log_CV1),
  log_CV2 = as.numeric(opaka_parameters$log_CV2),
  par_sel = as.matrix(opaka_parameters$par_sel),
  log_init_F_f = c(log(0.0119122), rep(log(1e-8), data$n_fishery - 1)),
  rdev_y = rdev_y_shifted, init_rdev_a = init_rdev_a)
parameters$par_sel <- convert_ss3_selex_to_rtmb(parameters$par_sel,
                                                data$sel_type_f, data$len_mid)
data$priors <- get_priors(parameters = parameters, data = data)

map_sel <- matrix(NA, nrow(parameters$par_sel), ncol(parameters$par_sel))
map_sel[1, 1:2]        <- c(1, 2)
map_sel[3, c(1, 4, 6)] <- c(3, 4, 5)
map_sel <- as.vector(map_sel)
map <- list(
  log_h = factor(NA), log_sigma_r = factor(NA),
  log_cpue_tau = factor(rep(NA, data$n_index)),
  log_cpue_omega = factor(rep(NA, data$n_index)),
  log_lf_tau = factor(rep(NA, length(parameters$log_lf_tau))),
  log_L1 = factor(NA), log_L2 = factor(NA), log_k = factor(NA),
  log_CV1 = factor(NA), log_CV2 = factor(NA),
  par_sel = factor(map_sel),
  log_init_F_f = rep(factor(NA), data$n_fishery),
  init_rdev_a = as.factor(seq_len(data$n_age)),
  rdev_y = as.factor(seq_along(parameters$rdev_y)))

data$lf_switch <- 1L
```

```{r}
#| label: fit
#| output: false
fit_opal <- function(dat) {
  obj <- MakeADFun(cmb(opal_model, dat), parameters, map = map, silent = TRUE)
  nm <- names(obj$par); Lwr <- rep(-Inf, length(nm)); Upr <- rep(Inf, length(nm))
  Lwr[grep("log_B0", nm)] <- 0;          Upr[grep("log_B0", nm)] <- 22
  Lwr[grep("log_cpue_q", nm)] <- log(1e-3); Upr[grep("log_cpue_q", nm)] <- log(10)
  Lwr[grep("rdev_y|init_rdev_a", nm)] <- -5; Upr[grep("rdev_y|init_rdev_a", nm)] <- 5
  Lwr[grep("par_sel", nm)] <- -7;        Upr[grep("par_sel", nm)] <- 7
  ctrl <- list(eval.max = 1e4, iter.max = 1e4)
  opt <- nlminb(obj$par, obj$fn, obj$gr, lower = Lwr, upper = Upr, control = ctrl)
  opt <- nlminb(opt$par, obj$fn, obj$gr, lower = Lwr, upper = Upr, control = ctrl)
  opt <- nlminb(opt$par, obj$fn, obj$gr, lower = Lwr, upper = Upr, control = ctrl)
  list(obj = obj, opt = opt)
}

data_age <- data
data_len <- data; data_len$length_dynamics <- 1L

fit_age <- fit_opal(data_age)
fit_len <- fit_opal(data_len)
rep_age <- fit_age$obj$report()
rep_len <- fit_len$obj$report()

len_mid <- data$len_mid
n_age   <- data$n_age
n_len   <- data$n_len
n_year  <- data$n_year
```

# The growth transition matrix

The engine's core object is the growth transition matrix `G`, of dimension
`[n_age, n_len, n_len]`. Slice `G[a, , ]` gives, for each source length bin, the
probability distribution of a fish's length one year later, discretised from the
von Bertalanffy one-step increment $E[L_{a+1}\,|\,L_a=l]=L_\infty(1-\rho)+\rho l$
with a normal spread. It is the transition analogue of the static age--length
key: rather than assuming a fixed length-at-age, it *advances* an evolving
length distribution. The transition standard deviation is set (conditional-SD
mode) so that, propagated forward under no fishing, the marginal length-at-age
reproduces the static age--length key exactly.

```{r}
#| label: fig-transition
#| fig-height: 3.6
#| fig-cap: "The growth transition matrix from the fitted 'opakapaka length model. Left: one age-step slice G[a] as a probability field --- each column (a source length) maps to a distribution of destination lengths one year later, sitting above the 1:1 line because fish grow. Right: a single recruitment cohort propagated through G (no mortality); the length distribution spreads and its mean tracks the von Bertalanffy curve as the cohort ages."
G <- rep_len$G
a_show <- 6L
Gdf <- expand.grid(dest = seq_len(n_len), src = seq_len(n_len))
Gdf$p <- as.vector(G[a_show, , ])
Gdf$src_len  <- len_mid[Gdf$src]
Gdf$dest_len <- len_mid[Gdf$dest]

p_mat <- ggplot(Gdf, aes(src_len, dest_len, fill = p)) +
  geom_tile() +
  geom_abline(slope = 1, intercept = 0, colour = "grey60", linetype = "dashed",
              linewidth = 0.3) +
  scale_fill_gradient(low = "white", high = "#2E8B57", name = "P") +
  labs(x = "Length this year (cm)", y = "Length next year (cm)")

# Propagate one cohort of recruits through G with no mortality
rec <- rep_len$recruit_dist_l
ages_show <- c(1, 3, 6, 10, 16, 30)
v <- rec
coh <- list(); mu_track <- numeric(n_age); mu_track[1] <- sum(v * len_mid)
if (1 %in% ages_show) coh[["1"]] <- data.frame(age = 1, len = len_mid, p = v)
for (a in 2:n_age) {
  v <- as.vector(G[a - 1, , ] %*% v)
  mu_track[a] <- sum(v * len_mid)
  if (a %in% ages_show) coh[[as.character(a)]] <- data.frame(age = a, len = len_mid, p = v)
}
coh_df <- bind_rows(coh)
coh_df$age <- factor(coh_df$age, levels = ages_show,
                     labels = paste0("age ", ages_show - 1))

p_coh <- ggplot(coh_df, aes(len, p, colour = age)) +
  geom_line(linewidth = 0.7) +
  scale_colour_viridis_d(option = "D", end = 0.9, name = NULL) +
  labs(x = "Length (cm)", y = "Proportion")

p_mat + p_coh
```

# Why length dynamics matter: size-selective truncation

The payoff of a length-structured state is that size-selective fishing can
reshape the length distribution *within* an age class --- larger, faster-growing
individuals of a cohort are removed preferentially, lowering the mean length at
age. This is Lee's phenomenon, and the static age--length key cannot represent
it: it fixes the length-at-age distribution regardless of fishing.

To make the mechanism visible we trace a single cohort through the fitted growth
transition under an *illustrative* strongly size-selective fishery (logistic in
length, half-selected near $0.5\,L_\infty$). The effect concentrates at the
intermediate, partially-selected ages --- where a cohort still spans the
selectivity ridge, so its larger members are removed --- and tapers at older
ages once the survivors are uniformly large. On the lightly-fished 'opakapaka
stock itself the effect is small (which is why the two engines' assessments in
the next section nearly coincide); it grows with fishing pressure and gear
size-selectivity.

```{r}
#| label: fig-truncation
#| fig-height: 3.8
#| fig-cap: "Size-selective truncation of length-at-age for a single cohort under an illustrative size-selective fishery (half-selected near 0.5 L-infinity), using the fitted 'opakapaka growth transition. Left: mean length-at-age at three fishing levels; the fished curves fall below the unfished curve at the partially-selected ages. Right: the reduction relative to the unfished cohort, peaking at intermediate ages. The static age-length key of the age-only model is identical to the unfished (F = 0) curve at every fishing level."
L1f <- exp(parameters$log_L1); L2f <- exp(parameters$log_L2); kf <- exp(parameters$log_k)
Linf <- L1f + (L2f - L1f) / (1 - exp(-kf * (data$A2 - data$A1)))
sel_ill <- 1 / (1 + exp(-0.3 * (len_mid - 0.5 * Linf)))   # illustrative gear
M_a <- data$M

cohort_mla <- function(Fmax) {
  v <- rec; m <- numeric(n_age); m[1] <- sum(v * len_mid)
  for (a in 2:n_age) {
    surv <- v * exp(-M_a[a - 1]) * exp(-Fmax * sel_ill)
    v <- as.vector(G[a - 1, , ] %*% surv)
    m[a] <- sum(v * len_mid) / sum(v)
  }
  m
}
Fs   <- c(0, 0.6, 1.2)
m0   <- cohort_mla(0)
mla_df <- bind_rows(lapply(Fs, function(f)
  data.frame(age = ages, mla = cohort_mla(f), F = factor(f))))
diff_df <- bind_rows(lapply(Fs[-1], function(f)
  data.frame(age = ages, red = m0 - cohort_mla(f), F = factor(f))))

pA <- ggplot(mla_df, aes(age, mla, colour = F)) +
  geom_line(linewidth = 0.8) +
  scale_colour_viridis_d(option = "D", end = 0.85, name = expression(F~(yr^-1))) +
  labs(x = "Age (years)", y = "Mean length at age (cm)")
pB <- ggplot(diff_df, aes(age, red, colour = F)) +
  geom_line(linewidth = 0.8) +
  scale_colour_viridis_d(option = "D", begin = 0.35, end = 0.85,
                         name = expression(F~(yr^-1))) +
  labs(x = "Age (years)", y = "Reduction vs unfished (cm)")
pA + pB + plot_layout(guides = "collect") & theme(legend.position = "bottom")
```

# Estimation: the length engine reproduces the age-based assessment

Fitting the same data with both engines gives a consistency check. The
length engine recovers essentially the same stock scale and status as the
established age model, while fitting the length compositions slightly better
(they are now predicted from a dynamically consistent length state rather than
expanded through a static key).

```{r}
#| label: tbl-compare
compare_tbl <- tibble(
  Model      = c("Age (static PLA)", "Length (growth matrix)"),
  B0         = c(rep_age$B0, rep_len$B0),
  R0         = c(rep_age$R0, rep_len$R0),
  `Terminal SB` = c(tail(rep_age$spawning_biomass_y, 1),
                    tail(rep_len$spawning_biomass_y, 1)),
  Depletion  = c(tail(rep_age$spawning_biomass_y, 1) / rep_age$spawning_biomass_y[1],
                 tail(rep_len$spawning_biomass_y, 1) / rep_len$spawning_biomass_y[1]),
  NLL        = c(fit_age$opt$objective, fit_len$opt$objective))
knitr::kable(compare_tbl, digits = c(0, 0, 1, 0, 3, 2),
  caption = "'Opakapaka estimates from the age and length engines fitted to identical data.")
```

```{r}
#| label: fig-ssb
#| fig-height: 3.8
#| fig-cap: "Estimated 'opakapaka spawning-biomass trajectories from the age engine (static age-length key) and the length engine (growth transition matrix). The two nearly coincide over the full modelled period, confirming the length engine reproduces the age-based assessment on this stock."
sb_df <- bind_rows(
  data.frame(year = 1948 + seq_along(rep_age$spawning_biomass_y),
             sbio = rep_age$spawning_biomass_y, Model = "Age"),
  data.frame(year = 1948 + seq_along(rep_len$spawning_biomass_y),
             sbio = rep_len$spawning_biomass_y, Model = "Length"))
ggplot(sb_df, aes(year, sbio, colour = Model)) +
  geom_line(linewidth = 0.9) +
  scale_colour_manual(values = pal[c("Age", "Length")]) +
  coord_cartesian(ylim = c(0, NA)) +
  labs(x = "Year", y = "Spawning biomass (mt)")
```

# The length-structured state

Unlike the age engine, the length engine reports the full numbers-at-length
surface through time (`NL_yl`). The relative length composition of the stock is
shown below over the modelled period --- the standing image of the state the
engine now tracks and from which length compositions are predicted directly.

```{r}
#| label: fig-nl-surface
#| fig-height: 3.8
#| fig-cap: "Relative numbers-at-length of the 'opakapaka stock through time, from the fitted length engine (each year's length distribution normalised to sum to one). The recruitment pulse enters at small lengths and propagates up through the size classes via the growth transition, while the population thins at larger sizes under fishing."
NL <- rep_len$NL_yl[seq_len(n_year), ]         # start-of-year numbers-at-length
NLp <- NL / rowSums(NL)
nl_df <- expand.grid(year = 1948 + seq_len(n_year), len = len_mid)
nl_df$p <- as.vector(NLp)
ggplot(nl_df, aes(year, len, fill = p)) +
  geom_tile() +
  scale_fill_gradient(low = "white", high = "#7D3C98", name = "Prop.") +
  labs(x = "Year", y = "Length (cm)")
```

# Structural check: reduction to the static key

Because the conditional-SD transition is calibrated to reproduce the age--length
key, the length engine collapses onto the age engine when growth variability is
removed. Evaluating both engines at a common parameter set --- once at the
estimated growth CVs, once with the CVs shrunk toward zero --- makes this
explicit: at negligible CV the two spawning-biomass trajectories are
indistinguishable.

```{r}
#| label: fig-equivalence
#| fig-height: 3.8
#| fig-cap: "Reduction to the age model. Spawning biomass evaluated at a common parameter set for both engines, at the estimated growth CVs (left) and with the length-at-age CVs shrunk toward zero (right). As growth variability vanishes the length engine's trajectory converges onto the age engine's, confirming the two are consistent by construction."
eval_sby <- function(dat, pars) {
  m <- lapply(pars, function(v) factor(rep(NA, length(v))))
  o <- MakeADFun(cmb(opal_model, dat), pars, map = m, silent = TRUE)
  o$report()$spawning_biomass_y
}
p_est <- parameters; p_est$log_B0 <- log(rep_len$B0)
p_deg <- p_est; p_deg$log_CV1 <- log(1e-3); p_deg$log_CV2 <- log(1e-3)

eq_df <- bind_rows(
  data.frame(year = 1948 + seq_len(n_year + 1), sbio = eval_sby(data_age, p_est),
             Model = "Age", panel = "Estimated CV"),
  data.frame(year = 1948 + seq_len(n_year + 1), sbio = eval_sby(data_len, p_est),
             Model = "Length", panel = "Estimated CV"),
  data.frame(year = 1948 + seq_len(n_year + 1), sbio = eval_sby(data_age, p_deg),
             Model = "Age", panel = "CV -> 0"),
  data.frame(year = 1948 + seq_len(n_year + 1), sbio = eval_sby(data_len, p_deg),
             Model = "Length", panel = "CV -> 0"))
eq_df$panel <- factor(eq_df$panel, levels = c("Estimated CV", "CV -> 0"))
ggplot(eq_df, aes(year, sbio, colour = Model)) +
  geom_line(linewidth = 0.8) +
  facet_wrap(~ panel) +
  scale_colour_manual(values = pal[c("Age", "Length")]) +
  coord_cartesian(ylim = c(0, NA)) +
  labs(x = "Year", y = "Spawning biomass (mt)")
```

# Forward projection with the length engine

Projections work through the same `project_dynamics()` interface as the age
engine; it dispatches on the `length_dynamics` switch and carries the terminal
age--length state forward. Here the stock is projected ten years under constant
terminal catch, with parameter uncertainty drawn from the Hessian
(multivariate-normal) and the historical uncertainty shown by the delta method
(`sdreport`, via the `ADREPORT`ed spawning biomass inside the length engine).

```{r}
#| label: projection
#| output: false
obj_len <- fit_len$obj
obj_len$fn(fit_len$opt$par)              # set last.par.best to the optimum
data_len$first_yr <- 1949L; data_len$last_yr <- 2023L
data_len$removal_switch_f <- rep(0L, data_len$n_fishery)
data_len$age_a <- ages

n_proj <- 10L; n_iter <- 100L
rdev_proj <- project_rec_devs(data_len, obj_len, first_yr = data_len$first_yr,
                              n_proj = n_proj, n_iter = n_iter)
sel_proj  <- project_selectivity_length(data_len, obj_len, first_yr = data_len$last_yr - 9,
                                        n_proj = n_proj, n_iter = n_iter)
catch_ysf <- array(0, dim = c(n_proj, data_len$n_season, data_len$n_fishery))
for (y in seq_len(n_proj)) catch_ysf[y, , ] <- data_len$catch_obs_ysf[data_len$n_year, , ]

proj <- project_dynamics(data_len, obj_len, mcmc = NULL, n_proj = n_proj,
                         n_iter = n_iter, rdev_y = rdev_proj$rdev_y,
                         sel_fya = sel_proj, catch_ysf = catch_ysf, return_hist = TRUE)

# Historical delta-method ribbon
sdr <- RTMB::sdreport(obj_len)
ss  <- summary(sdr, "report")
rows <- which(rownames(ss) == "spawning_biomass_y")
sb_mle <- ss[rows, "Estimate"]; sb_se <- ss[rows, "Std. Error"]
```

```{r}
#| label: fig-projection
#| fig-height: 4
#| fig-cap: "'Opakapaka spawning biomass from the length engine: last 50 years of the estimated trajectory with delta-method historical uncertainty (grey), bridged into a 10-year constant-catch projection with multivariate-normal parameter uncertainty (green; 50% and 90% envelopes over 100 draws). The vertical line marks the end of the historical period."
spl <- rep_len$spawning_potential_l
sby_proj <- sapply(proj$dyn, function(z) sapply(1:(n_proj + 1),
                    function(y) sum(z$NL_yl[y, ] * spl)))   # [n_proj+1, n_iter]

term <- n_year + 1
n_show <- 50L
hidx <- (term - n_show + 1):term
hist_df <- data.frame(year = 1948 + hidx, med = sb_mle[hidx],
  lo90 = sb_mle[hidx] - qnorm(0.95) * sb_se[hidx],
  hi90 = sb_mle[hidx] + qnorm(0.95) * sb_se[hidx],
  lo50 = sb_mle[hidx] - qnorm(0.75) * sb_se[hidx],
  hi50 = sb_mle[hidx] + qnorm(0.75) * sb_se[hidx])

proj_df <- data.frame(year = 2024L + 0:n_proj,
  med  = apply(sby_proj, 1, median),
  lo90 = apply(sby_proj, 1, quantile, 0.05),
  hi90 = apply(sby_proj, 1, quantile, 0.95),
  lo50 = apply(sby_proj, 1, quantile, 0.25),
  hi50 = apply(sby_proj, 1, quantile, 0.75))

ggplot() +
  geom_ribbon(data = hist_df, aes(year, ymin = lo90, ymax = hi90), fill = "grey70", alpha = 0.4) +
  geom_ribbon(data = hist_df, aes(year, ymin = lo50, ymax = hi50), fill = "grey55", alpha = 0.5) +
  geom_line(data = hist_df, aes(year, med), colour = "grey25", linewidth = 0.9) +
  geom_vline(xintercept = 2024, linetype = "dashed", colour = "grey50") +
  geom_ribbon(data = proj_df, aes(year, ymin = lo90, ymax = hi90), fill = "#2E8B57", alpha = 0.18) +
  geom_ribbon(data = proj_df, aes(year, ymin = lo50, ymax = hi50), fill = "#2E8B57", alpha = 0.35) +
  geom_line(data = proj_df, aes(year, med), colour = "#2E8B57", linewidth = 1) +
  coord_cartesian(ylim = c(0, NA)) +
  labs(x = "Year", y = "Spawning biomass (mt)")
```

# A high-contrast simulation: depletion and recovery

'Opakapaka is only lightly fished, so its size structure barely moves and the two
engines agree closely --- a reassuring consistency result, but not a vivid
demonstration. To exercise the length engine we simulate the *same stock* (the
fitted $B_0$, growth, natural mortality, and maturity) forward under a much
heavier, size-selective fishing history: two decades of escalating catch that
draws the stock down, followed by a sharp reduction and rebuild. Recruitment is
deterministic here so the size-structure signal is not masked by process noise.

For this operating model we also make growth **slower** (a lower von Bertalanffy
$k$) and **tighter** (smaller length-at-age CVs) than 'opakapaka's. Both changes
let a cohort keep its size identity for longer before growth remixes the length
classes, so the size structure --- and the within-age truncation from
size-selective removal --- resolve much more sharply. This is the regime in
which a length-structured model earns its keep.

```{r}
#| label: sim-run
#| output: false
B0s   <- rep_len$B0
sp_l  <- rep_len$spawning_potential_l
wt_l  <- rep_len$weight_l
h_s   <- exp(parameters$log_h)
sr_s  <- exp(parameters$log_sigma_r)

# Operating-model growth: slower (k x 0.5) and tighter (CV x 0.4) than 'opakapaka.
k_mult <- 0.5; cv_mult <- 0.4
L1v <- exp(parameters$log_L1); L2v <- exp(parameters$log_L2)
log_k_sim <- parameters$log_k + log(k_mult)
mu_sim <- get_growth(n_age, data$A1, data$A2, L1v, L2v, log_k_sim)
sd_sim <- get_sd_at_age(mu_sim, L1v, L2v,
                        parameters$log_CV1 + log(cv_mult), parameters$log_CV2 + log(cv_mult))
plaL   <- get_pla(data$len_lower, data$len_upper, mu_sim, sd_sim)
G_sim  <- get_growth_matrix(data$len_lower, data$len_upper, len_mid, sd_sim,
                            L1v, L2v, log_k_sim, data$A1, data$A2)
rec_sim <- get_recruit_length_dist(data$len_lower, data$len_upper, mu_sim[1], sd_sim[1])

# Size-selective fishery for the operating model (half-selected near 0.5 L-inf).
Linf_sim <- L1v + (L2v - L1v) / (1 - exp(-exp(log_k_sim) * (data$A2 - data$A1)))
sel_sim  <- 1 / (1 + exp(-0.3 * (len_mid - 0.5 * Linf_sim)))

n_sim <- 60L
# Catch history (fraction of B0): low -> escalating -> cut -> rebuild.
catch_hist <- c(rep(0.010 * B0s, 10),
                seq(0.02 * B0s, 0.10 * B0s, length.out = 20),
                rep(0.010 * B0s, 30))
sim_data <- list(n_age = n_age, n_len = n_len, n_year = n_sim, n_season = 1L,
                 n_fishery = 1L, first_yr = 1, first_yr_catch = 1,
                 catch_units_f = 1L, catch_obs_ysf = array(catch_hist, dim = c(n_sim, 1, 1)))
sim_par  <- list(rdev_y = rep(0, n_sim))

wtf <- array(0, c(1, n_sim, n_len)); slf <- array(0, c(1, n_sim, n_len))
for (y in seq_len(n_sim)) { wtf[1, y, ] <- wt_l; slf[1, y, ] <- sel_sim }

iS <- get_initial_numbers_length(B0s, h_s, M_a, sp_l, plaL, G_sim, rec_sim,
                                 init_F_f = 0, sel_fl = matrix(sel_sim, 1))
dS <- do_dynamics_length(sim_data, sim_par, B0 = B0s, R0 = iS$R0, alpha = iS$alpha,
        beta = iS$beta, h = h_s, sigma_r = sr_s, M_a = M_a,
        spawning_potential_l = sp_l, weight_fyl = wtf, recruit_dist_l = rec_sim, G = G_sim,
        init_number_al = iS$Ninit_al, init_number0_al = iS$Ninit0_al, sel_fyl = slf)

depl   <- dS$spawning_biomass_y / dS$spawning_biomass_y[1]
trough <- which.min(depl)
```

The stock is drawn down to `r sprintf("%.0f%%", 100*min(depl))` of $B_0$ at the
fishing peak (year `r trough`) and rebuilds toward `r sprintf("%.0f%%", 100*tail(depl,1))`
of $B_0$ once catch is cut.

```{r}
#| label: fig-sim
#| fig-height: 3.8
#| fig-cap: "Simulated depletion and recovery. Left: relative spawning biomass over the 60-year history (dotted = 20% B0). Right: the relative numbers-at-length surface from the length engine --- the large size classes are progressively stripped out as fishing escalates, then refill from below as the stock rebuilds. This size-structure erosion and recovery is tracked natively by the length state; the age-only engine represents the stock only through its age marginal."
depl_df <- data.frame(year = seq_len(n_sim + 1), depl = depl)
p_depl <- ggplot(depl_df, aes(year, depl)) +
  geom_line(colour = "#2E8B57", linewidth = 0.9) +
  geom_hline(yintercept = 0.2, linetype = "dotted", colour = "red") +
  coord_cartesian(ylim = c(0, 1)) +
  labs(x = "Year", y = expression("Depletion (SB / B"[0]*")"))

NLs  <- dS$NL_yl[seq_len(n_sim), ]
NLsp <- NLs / rowSums(NLs)
surf_df <- expand.grid(year = seq_len(n_sim), len = len_mid)
surf_df$p <- as.vector(NLsp)
p_surf <- ggplot(surf_df, aes(year, len, fill = p)) +
  geom_tile() +
  scale_fill_gradient(low = "white", high = "#7D3C98", name = "Prop.") +
  labs(x = "Year", y = "Length (cm)")

p_depl + p_surf
```

```{r}
#| label: fig-sim-comp
#| fig-height: 3.6
#| fig-cap: "Population length composition at three points of the simulation: the unfished start, the depletion trough, and after rebuilding. Under heavy size-selective fishing the composition collapses toward the smaller, less-selected sizes; during recovery the large size classes are restored. Each curve is the relative numbers-at-length in that year."
snap <- c(Start = 1L, Trough = trough, Recovered = n_sim + 1L)
comp_df <- bind_rows(lapply(names(snap), function(nm) {
  v <- dS$NL_yl[snap[[nm]], ]; data.frame(phase = nm, len = len_mid, p = v / sum(v))
}))
comp_df$phase <- factor(comp_df$phase, levels = names(snap))
ggplot(comp_df, aes(len, p, colour = phase)) +
  geom_line(linewidth = 0.9) +
  scale_colour_manual(values = c(Start = "#4C9BD1", Trough = "#E8862E",
                                 Recovered = "#2E8B57"), name = NULL) +
  labs(x = "Length (cm)", y = "Proportion")
```

## Where the two models disagree

The size structure above is what the length model tracks; an age-only model must
represent the same stock through numbers-at-age and a *static* age--length key.
The two therefore disagree on the length-at-age itself. The tile plot below shows
that disagreement over age and time --- the difference between the length model's
evolving mean length-at-age and the age model's fixed key. It is near zero before
fishing, then opens up as the fishery escalates: the older, fully-selected ages
are truncated toward smaller lengths (blue), and because fishing continues at a
low level through the rebuild, that truncation persists rather than closing.

```{r}
#| label: fig-laa-diff
#| fig-height: 3.6
#| fig-cap: "Difference in mean length-at-age between the length model and the age model's static age-length key, by age class (vertical) and year (horizontal), for the simulated stock. Blue = the length model carries smaller fish at that age than the static key assumes; the downward truncation is strong and persistent at the older, fully-selected ages (up to ~4 cm), with a weaker opposite-sign band at the intermediate ages as the within-age distribution is reshaped. The age-only model, by construction, sees none of this time-varying length-at-age structure."
static_laa <- sapply(seq_len(n_age), function(a) sum(plaL[, a] * len_mid))
diff_mat <- matrix(NA_real_, n_age, n_sim)
for (y in seq_len(n_sim)) for (a in seq_len(n_age)) {
  v <- dS$number_ysal[y, 1, a, ]
  if (sum(v) > 1e-8) diff_mat[a, y] <- sum(v * len_mid) / sum(v) - static_laa[a]
}
laa_df <- expand.grid(age = 0:(n_age - 1), year = seq_len(n_sim))
laa_df$d <- as.vector(diff_mat)
ggplot(laa_df, aes(year, age, fill = d)) +
  geom_tile() +
  scale_fill_gradient2(low = "#4C9BD1", mid = "white", high = "#E8862E",
                       midpoint = 0, name = "Length\nmodel -\nkey (cm)") +
  labs(x = "Year", y = "Age (years)")
```

# Fitting both models to simulated data: age-model bias

The real test is estimation. We treat the length engine as the **operating
model** (the truth), draw fishery observations from it, and fit *both* a
correctly-specified length model and a misspecified age-only model to exactly the
same data --- then compare their estimates against the known truth. Both models
are given the true growth, natural mortality, steepness, and recruitment
variability; both estimate unfished biomass $B_0$, the catchability of a
relative-abundance index, the two logistic selectivity parameters, and the
recruitment deviations.

The simulated data are the operating model's catch-at-length compositions (scaled
to an effective sample size) and a relative-abundance index proportional to
vulnerable biomass.

```{r}
#| label: sim-est-fit
#| output: false
effN <- 500L

# Length compositions from the OM catch-at-length, scaled to an effective N.
lf_rows <- list()
for (y in seq_len(n_sim)) {
  cl <- dS$catch_pred_fyl[1, y, ]
  if (sum(cl) > 0)
    lf_rows[[length(lf_rows) + 1]] <-
      c(fishery = 1, year = 1948 + y, month = 1, ts = y, round(cl / sum(cl) * effN))
}
lf_wide_sim <- as.data.frame(do.call(rbind, lf_rows))
colnames(lf_wide_sim) <- c("fishery", "year", "month", "ts",
  as.character(seq(data$len_bin_start, by = data$len_bin_width, length.out = n_len)))

# Relative-abundance index proportional to vulnerable biomass.
vuln <- sapply(seq_len(n_sim), function(y) sum(dS$NL_yl[y, ] * sel_sim * wt_l))
cpue_sim <- data.frame(ts = seq_len(n_sim), fishery = 1L, index = 1L,
                       units = 1L, value = vuln, se = 0.2)

mk_simdata <- function() list(
  n_age = n_age, n_len = n_len, n_year = n_sim, n_season = 1L, n_fishery = 1L, n_index = 1L,
  first_yr = 1, last_yr = n_sim, years = seq_len(n_sim), first_yr_catch = 1, catch_units_f = 1L,
  catch_obs_ysf = array(catch_hist, dim = c(n_sim, 1, 1)),
  len_bin_start = data$len_bin_start, len_bin_width = data$len_bin_width,
  A1 = data$A1, A2 = data$A2, lw_a = data$lw_a, lw_b = data$lw_b,
  maturity = data$maturity, fecundity = data$fecundity, M = data$M, sel_type_f = 1L,
  sex_ratio = rep(1, n_age), cpue_switch = 1L, cpue_data = cpue_sim, priors = NULL)
simdat <- prep_lf_data(mk_simdata(), lf_wide_sim, lf_keep_fisheries = 1,
                       lf_var_adjust = 1, lf_addtocomp = 1e-8)
simdat$lf_switch <- 1L

# Parameters: growth / M / steepness / sigma_R fixed at the truth; estimate the
# rest. Both engines share this configuration.
sel0 <- convert_ss3_selex_to_rtmb(matrix(c(0.5 * Linf_sim, 10, 0, 0, 0, 0), 1), 1L, len_mid)
simpar <- list(log_B0 = log(B0s), log_h = log(h_s), log_sigma_r = log(sr_s),
  log_cpue_q = log(1), log_cpue_tau = log(0.2), log_cpue_omega = log(1), cpue_creep = 0,
  log_lf_tau = log(1), log_L1 = parameters$log_L1, log_L2 = parameters$log_L2,
  log_k = log_k_sim, log_CV1 = parameters$log_CV1 + log(cv_mult),
  log_CV2 = parameters$log_CV2 + log(cv_mult),
  par_sel = sel0, log_init_F_f = log(1e-8), rdev_y = rep(0, n_sim))
simmap <- list(log_h = factor(NA), log_sigma_r = factor(NA), log_cpue_tau = factor(NA),
  log_cpue_omega = factor(NA), cpue_creep = factor(NA), log_lf_tau = factor(NA),
  log_L1 = factor(NA), log_L2 = factor(NA), log_k = factor(NA),
  log_CV1 = factor(NA), log_CV2 = factor(NA),
  par_sel = factor(c(1, 2, NA, NA, NA, NA)), log_init_F_f = factor(NA),
  rdev_y = factor(seq_len(n_sim)))

fit_engine <- function(len_mode) {
  dd <- simdat; if (len_mode) dd$length_dynamics <- 1L
  obj <- MakeADFun(cmb(opal_model, dd), simpar, map = simmap, silent = TRUE)
  nm <- names(obj$par); Lw <- rep(-Inf, length(nm)); Up <- rep(Inf, length(nm))
  Lw[grep("log_B0", nm)] <- log(2500); Up[grep("log_B0", nm)] <- log(5e4)
  Lw[grep("log_cpue_q", nm)] <- log(1e-3); Up[grep("log_cpue_q", nm)] <- log(1e3)
  Lw[grep("rdev_y", nm)] <- -5; Up[grep("rdev_y", nm)] <- 5
  Lw[grep("par_sel", nm)] <- -7; Up[grep("par_sel", nm)] <- 7
  op <- nlminb(obj$par, obj$fn, obj$gr, lower = Lw, upper = Up, control = list(eval.max = 1e4, iter.max = 1e4))
  op <- nlminb(op$par, obj$fn, obj$gr, lower = Lw, upper = Up, control = list(eval.max = 1e4, iter.max = 1e4))
  list(rep = obj$report(), nll = op$objective)
}
fit_L <- fit_engine(TRUE)
fit_A <- fit_engine(FALSE)

true_dep <- dS$spawning_biomass_y / dS$spawning_biomass_y[1]
depL <- fit_L$rep$spawning_biomass_y / fit_L$rep$spawning_biomass_y[1]
depA <- fit_A$rep$spawning_biomass_y / fit_A$rep$spawning_biomass_y[1]
```

```{r}
#| label: tbl-simest
fmt <- function(b0, td, mind) c(sprintf("%.0f", b0), sprintf("%.2f", td), sprintf("%.2f", mind))
tibble(Quantity = c("B0 (mt)", "Terminal depletion (SB/B0)", "Lowest depletion"),
       Truth            = fmt(B0s, tail(true_dep, 1), min(true_dep)),
       `Length model`   = fmt(fit_L$rep$B0, tail(depL, 1), min(depL)),
       `Age-only model` = fmt(fit_A$rep$B0, tail(depA, 1), min(depA))) |>
  knitr::kable(caption = paste0(
    "Estimates from fitting each model to data simulated from the length operating ",
    "model. The length model recovers the truth; the age-only model is biased."))
```

The length model recovers the true $B_0$ almost exactly
(`r sprintf("%.0f", fit_L$rep$B0)` mt vs the true `r sprintf("%.0f", B0s)` mt),
whereas the age-only model estimates $B_0 =$ `r sprintf("%.0f", fit_A$rep$B0)` mt
--- a bias of `r sprintf("%+.0f%%", 100*(fit_A$rep$B0/B0s - 1))` --- and reads
the stock as far healthier at the end of the series than it truly is (terminal
depletion `r sprintf("%.2f", tail(depA,1))` vs the true
`r sprintf("%.2f", tail(true_dep,1))`). Forced to explain length compositions
generated by size-selective truncation with a fixed age--length key, it distorts
the biomass scale and status.

```{r}
#| label: fig-simest
#| fig-height: 4
#| fig-cap: "Estimated depletion trajectories from fitting the length model and the age-only model to identical data simulated from the length operating model. The length model tracks the truth; the age-only model is biased, missing the depth of depletion and over-stating the recovery. Dotted line = 20% B0."
se_df <- bind_rows(
  data.frame(year = seq_len(n_sim + 1), dep = true_dep, Series = "Truth (length OM)"),
  data.frame(year = seq_len(n_sim + 1), dep = depL,     Series = "Length model fit"),
  data.frame(year = seq_len(n_sim + 1), dep = depA,     Series = "Age-only model fit"))
se_df$Series <- factor(se_df$Series,
  levels = c("Truth (length OM)", "Length model fit", "Age-only model fit"))
ggplot(se_df, aes(year, dep, colour = Series)) +
  geom_line(linewidth = 0.9) +
  geom_hline(yintercept = 0.2, linetype = "dotted", colour = "red") +
  scale_colour_manual(values = c("Truth (length OM)" = "grey30",
    "Length model fit" = "#2E8B57", "Age-only model fit" = "#E8862E")) +
  coord_cartesian(ylim = c(0, NA)) +
  labs(x = "Year", y = expression("Depletion (SB / B"[0]*")"), colour = NULL)
```

# Summary

The joint age--length engine adds a length dimension to opal's dynamics through a
growth transition matrix built from the existing estimable growth parameters. It
is opt-in (`data$length_dynamics = 1`) and leaves the age-only path untouched.
On the 'opakapaka stock it reproduces the established age-based assessment while
predicting length compositions from a dynamically consistent length state, and
it can represent size-selective length-at-age truncation that the static
age--length key cannot. In a simulation--estimation test where a slower-growing
stock is fished hard and left to recover, the length model recovers the true
unfished biomass and status, whereas an age-only model fitted to the same data is
biased --- underestimating $B_0$ and over-stating the recovery --- because it
must force size-selectively truncated length compositions through a fixed
age--length key. Projections, MVN and MCMC uncertainty, and `sdreport`
historical uncertainty all flow through the length engine unchanged.
```
