---
title: "LF Likelihood Benchmark: Loop vs Batched"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{LF Likelihood Benchmark: Loop vs Batched}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

# Introduction

This vignette benchmarks two implementations of the length-composition
likelihood on the RTMB AD tape:

- **Loop** — old implementation: one `n_age`-length matrix-vector product
  _per observation_ (`catch_a %*% t(pla)`, recorded as `n_age × n_len`
  scalar multiply-adds per obs on the CppAD tape).
- **Batched** — current implementation: one `[n_obs_f × n_age]` matrix
  multiply _per fishery_ (`catch_ya %*% t(pla)`), letting BLAS handle the
  inner loop and compressing the tape to a single DGEMM operation per fishery.

Two data subsets are crossed with the two implementations:

| Subset | Fisheries | Obs-rows |
|--------|-----------|----------|
| Small  | 8 & 9     | ~210     |
| Full   | all 7 LF  | ~568     |

The benchmark model is intentionally minimal: it estimates double-normal
selectivities (`sel_type_f = 2`) for the active fisheries, computes
catch-at-age from fixed baseline numbers-at-age and estimated selectivity,
then evaluates the LF likelihood. The full population dynamics are bypassed so
all AD tape work comes from the selectivity-to-likelihood pipeline. `obj$fn()`
and `obj$gr()` are timed with `bench::mark`.

# Setup

```{r load-pkg, message=FALSE}
library(dplyr)
library(tidyr)
library(RTMB)
library(bench)
library(opal)
```

## Load raw data

```{r load-data, message=FALSE}
data(wcpo_bet_data)
data(wcpo_bet_lf)
data(opal_baseline_data)
data(opal_baseline_parameters)
data(opal_baseline_map)

base_data <- wcpo_bet_data
if (is.null(base_data$n_index))           base_data$n_index <- 1L
if (is.null(base_data$cpue_data$index))   base_data$cpue_data$index <- rep(1L, nrow(base_data$cpue_data))
```

## Build wide-format LF table

```{r lf-wide}
lf_wide <- wcpo_bet_lf |>
  tidyr::pivot_wider(
    id_cols     = c(fishery, year, month, ts),
    names_from  = bin,
    values_from = value,
    values_fill = 0
  ) |>
  dplyr::arrange(fishery, ts)

lf_fish_all   <- sort(unique(wcpo_bet_lf$fishery))
lf_fish_small <- c(8L, 9L)

cat("All LF fisheries:   ", lf_fish_all, "\n")
cat("Small subset (F8+9):", lf_fish_small, "\n")
```

## Prepare LF data subsets

```{r prep-lf}
d_small <- prep_lf_data(base_data, lf_wide,
                        lf_keep_fisheries = lf_fish_small,
                        lf_switch         = 1L)

d_full  <- prep_lf_data(base_data, lf_wide,
                        lf_keep_fisheries = lf_fish_all,
                        lf_switch         = 1L)

cat("Small: n_lf =", d_small$n_lf, "obs-rows,",
    length(d_small$lf_fishery_f), "fisheries\n")
cat("Full:  n_lf =", d_full$n_lf,  "obs-rows,",
    length(d_full$lf_fishery_f),  "fisheries\n")
```

## Extract MLE quantities from baseline

Build the baseline model once (no optimisation) to get the MLE PLA matrix and
numbers-at-age. These are passed as fixed data to the mini benchmark model so
that the AD tape only records selectivity and likelihood operations.

```{r baseline-build, message=FALSE}
# Guard missing fields (opal_baseline_data may pre-date n_index)
bld <- opal_baseline_data
if (is.null(bld$n_index))         bld$n_index <- 1L
if (is.null(bld$cpue_data$index)) bld$cpue_data$index <- rep(1L, nrow(bld$cpue_data))

base_pars <- opal_baseline_parameters
base_pars$log_cpue_q     <- rep(as.numeric(base_pars$log_cpue_q),     bld$n_index)
base_pars$cpue_creep     <- rep(as.numeric(base_pars$cpue_creep),     bld$n_index)
base_pars$log_cpue_tau   <- rep(as.numeric(base_pars$log_cpue_tau),   bld$n_index)
base_pars$log_cpue_omega <- rep(as.numeric(base_pars$log_cpue_omega), bld$n_index)

base_map <- opal_baseline_map
base_map$cpue_creep     <- factor(rep(NA, bld$n_index))
base_map$log_cpue_tau   <- factor(rep(NA, bld$n_index))
base_map$log_cpue_omega <- factor(rep(NA, bld$n_index))

obj_base <- MakeADFun(func       = cmb(opal_model, bld),
                      parameters = base_pars,
                      map        = base_map)
invisible(obj_base$fn(obj_base$par))
rep_base     <- obj_base$report()
par_mle      <- obj_base$env$parList(obj_base$env$last.par.best)
```

```{r extract-mle}
# PLA at MLE growth parameters (plain double, off the tape)
L1_mle  <- exp(par_mle$log_L1)
L2_mle  <- exp(par_mle$log_L2)
mu_a_mle <- get_growth(base_data$n_age, base_data$A1, base_data$A2,
                       L1_mle, L2_mle, par_mle$log_k)
sd_a_mle <- get_sd_at_age(mu_a_mle, L1_mle, L2_mle,
                          par_mle$log_CV1, par_mle$log_CV2)
pla_mle  <- get_pla(opal_baseline_data$len_lower,
                    opal_baseline_data$len_upper,
                    mu_a_mle, sd_a_mle)

# numbers-at-age from baseline dynamics (plain double, off the tape)
number_ysa_mle <- rep_base$number_ysa

cat("pla_mle dim:      ", dim(pla_mle), "\n")
cat("number_ysa dim:   ", dim(number_ysa_mle), "\n")
```

# Loop implementation

`get_length_like_loop` is structurally identical to the current
`get_length_like` except the batched fishery-level matmul is replaced by a
per-observation matrix-vector product. This is what the code looked like before
the batching optimisation.

```{r def-loop-like}
get_length_like_loop <- function(lf_obs_flat, lf_obs_ints, lf_obs_prop,
                                 catch_pred_fya, pla,
                                 lf_n_f, lf_fishery_f, lf_year_fi, lf_n_fi,
                                 lf_minbin, lf_maxbin, removal_switch_f,
                                 lf_switch, n_len, n_lf, log_lf_tau) {
  "[<-" <- ADoverload("[<-")
  "c"   <- ADoverload("c")

  n_f    <- length(lf_n_f)
  lf_pred <- vector("list", n_f)
  lp     <- numeric(n_lf)

  if (lf_switch == 1) lf_obs_flat <- OBS(lf_obs_flat)
  if (lf_switch == 2) lf_obs_prop <- OBS(lf_obs_prop)
  if (lf_switch == 3) lf_obs_ints <- OBS(lf_obs_ints)

  idx        <- 0L
  obs_offset <- 0L

  for (j in seq_len(n_f)) {
    f     <- lf_fishery_f[j]
    bmin  <- lf_minbin[f]
    bmax  <- lf_maxbin[f]
    nbins <- bmax - bmin + 1L
    lf_pred[[j]] <- matrix(0, lf_n_f[j], nbins)

    for (i in seq_len(lf_n_f[j])) {
      idx <- idx + 1L
      y   <- lf_year_fi[[j]][i]

      # ---- LOOP: one [n_age]-length matvec per observation ----
      catch_a <- catch_pred_fya[f, y, ]       # [n_age] row vector
      pred    <- c(catch_a %*% t(pla))         # [n_len] via row × matrix
      # ----------------------------------------------------------

      if (bmin > 1)    pred[bmin] <- sum(pred[1:bmin])
      if (bmax < n_len) pred[bmax] <- sum(pred[bmax:n_len])
      pred <- pred[bmin:bmax]
      pred <- pred + 1e-8
      pred <- pred / sum(pred)
      lf_pred[[j]][i, ] <- pred

      n_i <- lf_n_fi[[j]][i]
      if (removal_switch_f[f] == 0 & n_i > 0) {
        if (lf_switch == 1) {
          obs_i   <- lf_obs_flat[(obs_offset + 1):(obs_offset + nbins)]
          lp[idx] <- -RTMB::dmultinom(x = obs_i, prob = pred, log = TRUE)
        }
        if (lf_switch == 2) {
          obs_i   <- lf_obs_prop[(obs_offset + 1):(obs_offset + nbins)]
          alpha_i <- pred * n_i * exp(log_lf_tau[f])
          lp[idx] <- -RTMBdist::ddirichlet(x = obs_i, alpha = alpha_i, log = TRUE)
        }
        if (lf_switch == 3) {
          obs_i   <- lf_obs_ints[(obs_offset + 1):(obs_offset + nbins)]
          alpha_i <- pred * exp(log_lf_tau[f])
          lp[idx] <- -RTMBdist::ddirmult(x = obs_i, size = n_i, alpha = alpha_i, log = TRUE)
        }
      }
      obs_offset <- obs_offset + nbins
    }
  }

  REPORT(lf_pred)
  return(lp)
}
```

# Mini benchmark model

A factory that returns a thin RTMB model using the supplied LF likelihood
function. Parameters on the tape are double-normal selectivities (`par_sel`)
for the active fisheries. `number_ysa` and `pla` are fixed (off-tape) data,
so every tape operation comes from the selectivity computation and the LF
likelihood.

`catch_pred_fya[f, y, ] = number_ysa_fixed[y, 1, ] × sel_fya[f, y, ]` is a
proportional catch-at-age that preserves the age-composition shape without
requiring full population dynamics.

```{r make-bench-fn}
make_lf_bench_fn <- function(lf_like_fn) {
  force(lf_like_fn)
  function(parameters, data) {
    "[<-" <- ADoverload("[<-")
    "c"   <- ADoverload("c")
    getAll(data, parameters, warn = FALSE)

    # Selectivity from par_sel (on tape) × PLA (fixed, off tape)
    sel_fya <- get_selectivity(data, par_sel, pla_fixed, len_mid)

    # Proportional catch-at-age: N_y × sel_f,y  (no dynamics)
    catch_pred_fya <- array(0, dim = c(n_fishery, n_year, n_age))
    for (f in seq_len(n_fishery)) {
      for (y in seq_len(n_year)) {
        catch_pred_fya[f, y, ] <- number_ysa_fixed[y, 1, ] * sel_fya[f, y, ]
      }
    }

    lf_year_fi <- split(lf_year, lf_fishery)
    lf_n_fi    <- split(lf_n,    lf_fishery)

    lp <- lf_like_fn(
      lf_obs_flat      = lf_obs_flat,
      lf_obs_ints      = lf_obs_ints,
      lf_obs_prop      = lf_obs_prop,
      catch_pred_fya   = catch_pred_fya,
      pla              = pla_fixed,
      lf_n_f           = lf_n_f,
      lf_fishery_f     = lf_fishery_f,
      lf_year_fi       = lf_year_fi,
      lf_n_fi          = lf_n_fi,
      lf_minbin        = lf_minbin,
      lf_maxbin        = lf_maxbin,
      removal_switch_f = removal_switch_f,
      lf_switch        = lf_switch,
      n_len            = n_len,
      n_lf             = n_lf,
      log_lf_tau       = log_lf_tau
    )
    return(sum(lp))
  }
}
```

## Build data lists

```{r build-data}
# Add fixed (off-tape) arrays and double-normal flag to each data subset
add_bench_fields <- function(d) {
  d$pla_fixed        <- pla_mle
  d$number_ysa_fixed <- number_ysa_mle
  d$sel_type_f       <- rep(2L, d$n_fishery)   # double-normal for all fisheries
  d
}

d_small_b <- add_bench_fields(d_small)
d_full_b  <- add_bench_fields(d_full)
```

## Build parameters and maps

```{r build-pars-maps}
# All par_sel start at zero (double-normal at default centre/shape)
params_bench <- list(
  par_sel    = matrix(0, nrow = base_data$n_fishery, ncol = 6),
  log_lf_tau = rep(0, base_data$n_fishery)
)

# Fix par_sel rows not belonging to the active fisheries; fix all log_lf_tau
make_bench_map <- function(n_fishery, active_fisheries) {
  idx_mat <- matrix(seq_len(n_fishery * 6), n_fishery, 6)
  idx_mat[setdiff(seq_len(n_fishery), active_fisheries), ] <- NA
  list(
    par_sel    = factor(idx_mat),
    log_lf_tau = factor(rep(NA, n_fishery))
  )
}

map_small <- make_bench_map(base_data$n_fishery, lf_fish_small)
map_full  <- make_bench_map(base_data$n_fishery, lf_fish_all)

cat("Small: estimating", sum(!is.na(map_small$par_sel)),
    "par_sel params (", length(lf_fish_small), "fisheries × 6)\n")
cat("Full:  estimating", sum(!is.na(map_full$par_sel)),
    "par_sel params (", length(lf_fish_all), "fisheries × 6)\n")
```

## Build four MakeADFun objects

```{r build-adfun, message=FALSE}
fn_loop    <- make_lf_bench_fn(get_length_like_loop)
fn_batched <- make_lf_bench_fn(get_length_like)

t_build <- system.time({
  obj_loop_small    <- MakeADFun(func = cmb(fn_loop,    d_small_b),
                                 parameters = params_bench, map = map_small)
  obj_batched_small <- MakeADFun(func = cmb(fn_batched, d_small_b),
                                 parameters = params_bench, map = map_small)
  obj_loop_full     <- MakeADFun(func = cmb(fn_loop,    d_full_b),
                                 parameters = params_bench, map = map_full)
  obj_batched_full  <- MakeADFun(func = cmb(fn_batched, d_full_b),
                                 parameters = params_bench, map = map_full)
})
cat("MakeADFun × 4 build time:", round(t_build["elapsed"], 1), "sec\n")
```

## Parameter counts and NLL check

```{r nll-check}
cat("Small (F8+9): ", length(obj_loop_small$par),  "active params\n")
cat("Full (7 fish.):", length(obj_loop_full$par), "active params\n")

cat("\nNLL at default params (loop == batched?):\n")
cat("  loop_small   :", obj_loop_small$fn(obj_loop_small$par),    "\n")
cat("  batched_small:", obj_batched_small$fn(obj_batched_small$par), "\n")
cat("  loop_full    :", obj_loop_full$fn(obj_loop_full$par),      "\n")
cat("  batched_full :", obj_batched_full$fn(obj_batched_full$par),  "\n")
```

## Warm up AD tapes

The first call after `MakeADFun` constructs the tape; all subsequent calls
replay it. Warm up before timing.

```{r warmup}
for (obj in list(obj_loop_small, obj_batched_small,
                 obj_loop_full,  obj_batched_full)) {
  invisible(obj$fn(obj$par))
  invisible(obj$gr(obj$par))
}
```

# Benchmarks

## `obj$fn()` — forward pass

```{r bench-fn}
bm_fn <- bench::mark(
  loop_small    = obj_loop_small$fn(obj_loop_small$par),
  batched_small = obj_batched_small$fn(obj_batched_small$par),
  loop_full     = obj_loop_full$fn(obj_loop_full$par),
  batched_full  = obj_batched_full$fn(obj_batched_full$par),
  iterations    = 50,
  check         = FALSE
)
bm_fn[, c("expression", "min", "median", "itr/sec", "total_time")]
```

## `obj$gr()` — reverse-mode gradient

Reverse-mode AD cost is determined by tape size (total recorded scalar
operations). Batching reduces the number of operations the tape must replay.

```{r bench-gr}
bm_gr <- bench::mark(
  loop_small    = obj_loop_small$gr(obj_loop_small$par),
  batched_small = obj_batched_small$gr(obj_batched_small$par),
  loop_full     = obj_loop_full$gr(obj_loop_full$par),
  batched_full  = obj_batched_full$gr(obj_batched_full$par),
  iterations    = 50,
  check         = FALSE
)
bm_gr[, c("expression", "min", "median", "itr/sec", "total_time")]
```

## Plain-R timing (outside AD tape)

Call both implementations as ordinary R functions with double (non-AD)
inputs. This shows the raw algorithmic cost before any AD overhead is added.

```{r bench-plain-r}
# Pre-compute catch_pred_fya at default par_sel = 0 for each subset
sel_ref_s <- get_selectivity(d_small_b, params_bench$par_sel,
                             pla_mle, d_small_b$len_mid)
cpf_s <- array(0, c(d_small_b$n_fishery, d_small_b$n_year, d_small_b$n_age))
for (f in seq_len(d_small_b$n_fishery))
  for (y in seq_len(d_small_b$n_year))
    cpf_s[f, y, ] <- number_ysa_mle[y, 1, ] * sel_ref_s[f, y, ]

sel_ref_f <- get_selectivity(d_full_b, params_bench$par_sel,
                             pla_mle, d_full_b$len_mid)
cpf_f <- array(0, c(d_full_b$n_fishery, d_full_b$n_year, d_full_b$n_age))
for (f in seq_len(d_full_b$n_fishery))
  for (y in seq_len(d_full_b$n_year))
    cpf_f[f, y, ] <- number_ysa_mle[y, 1, ] * sel_ref_f[f, y, ]

make_plain_args <- function(d, cpf) {
  list(
    lf_obs_flat      = d$lf_obs_flat,
    lf_obs_ints      = d$lf_obs_ints,
    lf_obs_prop      = d$lf_obs_prop,
    catch_pred_fya   = cpf,
    pla              = pla_mle,
    lf_n_f           = d$lf_n_f,
    lf_fishery_f     = d$lf_fishery_f,
    lf_year_fi       = split(d$lf_year, d$lf_fishery),
    lf_n_fi          = split(d$lf_n,    d$lf_fishery),
    lf_minbin        = d$lf_minbin,
    lf_maxbin        = d$lf_maxbin,
    removal_switch_f = d$removal_switch_f,
    lf_switch        = d$lf_switch,
    n_len            = d$n_len,
    n_lf             = d$n_lf,
    log_lf_tau       = rep(0, d$n_fishery)
  )
}

args_s <- make_plain_args(d_small_b, cpf_s)
args_f <- make_plain_args(d_full_b,  cpf_f)

bm_plain <- bench::mark(
  loop_small_r    = do.call(get_length_like_loop, args_s),
  batched_small_r = do.call(get_length_like,      args_s),
  loop_full_r     = do.call(get_length_like_loop, args_f),
  batched_full_r  = do.call(get_length_like,      args_f),
  iterations      = 100,
  check           = FALSE
)
bm_plain[, c("expression", "min", "median", "itr/sec", "total_time")]
```

# Summary

```{r summary-table}
fn_med <- as.numeric(bm_fn$median)
gr_med <- as.numeric(bm_gr$median)
r_med  <- as.numeric(bm_plain$median)

n_lf_vec <- c(d_small_b$n_lf, d_small_b$n_lf,
              d_full_b$n_lf,  d_full_b$n_lf)

res <- data.frame(
  config      = c("loop_small", "batched_small", "loop_full", "batched_full"),
  n_lf        = n_lf_vec,
  plain_ms    = round(r_med  * 1e3, 2),
  fn_ms       = round(fn_med * 1e3, 2),
  gr_ms       = round(gr_med * 1e3, 2),
  gr_fn_ratio = round(gr_med / fn_med, 1)
)

# Speedup of batched relative to loop within each data-size pair
res$fn_speedup <- c(1, round(fn_med[1] / fn_med[2], 2),
                    1, round(fn_med[3] / fn_med[4], 2))
res$gr_speedup <- c(1, round(gr_med[1] / gr_med[2], 2),
                    1, round(gr_med[3] / gr_med[4], 2))
print(res)
```

## Interpretation guide

- **`plain_ms`** — wall-clock cost with no AD, measuring raw R/BLAS algorithmic
  cost. The ratio `fn_ms / plain_ms` reveals per-call AD tape overhead relative
  to plain R.
- **`fn_ms`** — forward pass cost on the RTMB tape. Scales roughly with `n_lf`
  and with the number of free parameters driving the tape (here: `par_sel`
  for active fisheries).
- **`gr_ms`** — reverse-mode gradient cost. This is determined by the total
  number of elementary scalar operations recorded on the CppAD tape during
  the forward pass. RTMB uses CppAD's operator-overloading AD: **matrix
  multiply records each scalar multiply-add individually**, not as a compact
  DGEMM node. The loop and batched implementations therefore produce tapes of
  identical size for the LF likelihood itself.
- **`gr/fn ratio`** — expected to be ~2–5× for a compact tape. A ratio ≫ 5×
  indicates the tape is large relative to the raw computation, i.e., dominated
  by many chained scalar operations.
- **`fn_speedup` / `gr_speedup`** — speedup of batched over loop within the
  same data size. Any plain-R speedup from batching (fewer BLAS calls) does
  not transfer to the AD gradient because the tape size is unchanged.

## AD tape operation count

Both implementations record identical numbers of scalar multiply-adds for the
LF matmul step, because RTMB/CppAD does not use atomic DGEMM tape nodes:

```{r tape-estimate}
n_a     <- base_data$n_age
n_l     <- base_data$n_len
n_obs_s <- d_small_b$n_lf
n_obs_f <- d_full_b$n_lf
n_fish_s <- length(lf_fish_small)
n_fish_f <- length(lf_fish_all)

cat("Scalar multiply-adds on CppAD tape — LF matmul (same for loop and batched):\n")
cat("  small:", n_obs_s, "obs ×", n_a, "ages ×", n_l, "bins =",
    format(n_obs_s * n_a * n_l, big.mark = ","), "ops\n")
cat("  full: ", n_obs_f, "obs ×", n_a, "ages ×", n_l, "bins =",
    format(n_obs_f * n_a * n_l, big.mark = ","), "ops\n")

cat("\nR-level %*% calls (this IS reduced by batching):\n")
cat("  loop small:   ", n_obs_s, "calls (one per obs)\n")
cat("  batched small:", n_fish_s, "calls (one per fishery)\n")
cat("  loop full:    ", n_obs_f, "calls (one per obs)\n")
cat("  batched full: ", n_fish_f, "calls (one per fishery)\n")

cat("\nDominant tape cost (sel_fya loop, shared by both implementations):\n")
cat(" ", base_data$n_fishery, "fisheries ×", base_data$n_year, "years =",
    base_data$n_fishery * base_data$n_year,
    "double-normal evaluations + N×sel products\n")
```

# Results and Discussion

```{r results-discussion}
fn_speedup_small <- round(fn_med[1] / fn_med[2], 2)
fn_speedup_full  <- round(fn_med[3] / fn_med[4], 2)
gr_speedup_small <- round(gr_med[1] / gr_med[2], 2)
gr_speedup_full  <- round(gr_med[3] / gr_med[4], 2)
plain_speedup_small <- round(r_med[1] / r_med[2], 2)
plain_speedup_full  <- round(r_med[3] / r_med[4], 2)
gr_fn_loop_small    <- round(gr_med[1] / fn_med[1], 0)
gr_fn_loop_full     <- round(gr_med[3] / fn_med[3], 0)

cat("Plain-R speedup  (batched/loop) — small:", plain_speedup_small, "x   full:", plain_speedup_full, "x\n")
cat("fn()    speedup  (batched/loop) — small:", fn_speedup_small,    "x   full:", fn_speedup_full, "x\n")
cat("gr()    speedup  (batched/loop) — small:", gr_speedup_small,    "x   full:", gr_speedup_full, "x\n")
cat("gr/fn ratio (loop) — small:", gr_fn_loop_small, "x   full:", gr_fn_loop_full, "x\n")
```

## What the benchmark shows

**Plain-R (no AD):** batching does deliver a modest speedup
(`r plain_speedup_small`× small, `r plain_speedup_full`× full) by reducing
the number of R-level `%*%` calls from one per observation to one per fishery,
allowing BLAS to operate on larger matrices.

**AD gradient — no improvement from batching:** the `gr()` speedup is
effectively 1× (or even slightly negative) for both subset sizes. This is the
key result. RTMB uses CppAD's operator-overloading AD, which records **every
scalar multiply-add individually** — there is no compact DGEMM node on the
tape. Whether the matmul is spelled as `n_obs` individual row-vector products
or one `[n_obs × n_age]` block multiply, CppAD records the same
`n_obs × n_age × n_len` scalar operations either way. The tape size, and
therefore the reverse-mode sweep time, is identical.

**Where is the tape time actually spent?** The very high `gr/fn` ratio
(~`r gr_fn_loop_small`× small, ~`r gr_fn_loop_full`× full) confirms the tape
is expensive. In this benchmark model, `catch_pred_fya = N × sel_fya` is
computed inside a double loop over all `n_fishery × n_year` combinations with
`par_sel` on the tape, making the double-normal selectivity evaluation the
dominant tape cost — not the LF matmul at all.

## Implications for optimisation

| Approach | Plain-R gain | AD gradient gain | Recommended? |
|---|---|---|---|
| Batch LF matmul per fishery | Yes (~`r plain_speedup_full`×) | No | For plain-R use only |
| Reduce `sel_fya` tape ops | N/A | Yes (dominant cost) | High priority |
| Use CppAD atomic DGEMM | N/A | Would help | Not available in RTMB |
| Reduce active fisheries | Proportional | Proportional | Use `lf_keep_fisheries` |

To reduce gradient cost meaningfully, the target is the `sel_fya` loop
(`n_fishery × n_year` double-normal evaluations on the AD tape), not the
LF matmul. The batching refactor is still worth keeping because it reduces
R-overhead in the forward pass and improves BLAS efficiency when running the
model outside the AD tape (e.g., plain-R diagnostics and posterior predictive
checks in `do_dynamics()`).
