---
title: "'Opakapaka case study: benchmarking opal against Stock Synthesis 3"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{'Opakapaka case study: benchmarking opal against Stock Synthesis 3}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
---

```{r, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

# Introduction

This vignette documents the configuration, fitting, and cross-platform
benchmarking of an `opal` stock assessment model for 'opakapaka
(*Pristipomoides filamentosus*), a Hawaiian deepwater snapper. The reference
model was built in Stock Synthesis 3 (SS3 v3.30.19) as part of the
[Opaka_simulation](https://github.com/noaa-pifsc/Opaka_simulation) project,
which provides both an operating model (OM) and an estimation model (EM).

The primary objective is to verify that `opal`'s population dynamics engine
reproduces SS3's spawning biomass trajectory when given identical inputs. A
secondary objective is to demonstrate model fit to multiple data sources:
index and length composition data.

The model covers the historical period 1949–2023 with three fleets: a
commercial fishery (Comm), a non-commercial fishery (Non_comm), and a
research fishery survey (ResFish).

# Load inputs

Load the `opal` package and the `RTMB` dependency. The `tidyverse` package is
used for data manipulation and plotting.

```{r load-pkg, echo=TRUE, message=FALSE}
library(tidyverse)
library(opal)
library(RTMB)

theme_set(theme_bw())
```

The bundled data object `opaka_data` contains all biological parameters, catch
and CPUE observations, length structure, and prior specifications needed for the
opakapaka assessment model.

SS3 indexes ages from 0 to 43 (44 age classes), while opal uses 1-based
internal indexing. To ensure the growth curve, survivorship, and all
age-structured biology vectors align correctly with SS3, we set `A1 = 0`
(anchoring the Schnute growth parameterization at biological age 0) and
redefine `age_a` accordingly. The `min_age` field is passed to
`get_growth()` so that the age sequence starts at 0 rather than 1.

```{r load-data, echo=TRUE, message=FALSE}
data(opaka_data)
data <- opaka_data
if (is.null(data$n_index)) data$n_index <- 2L
names(data)

# remove priors
data$priors <- NULL

# set ages to anchor growth curve (since SS3 indexs from 0 to 43, but opal uses 1-based indexing)
data$A1 <- 0
data$A2 <- data$A1 + (data$n_age-1)
data$min_age <- 0
ages <- data$age_a <- seq(data$min_age, by = 1, length.out = data$n_age)
```

Key dimensions of the data:

```{r data-dims, echo=TRUE}
cat("Number of ages:", data$n_age, "\n")
cat("Number of years:", data$n_year, "\n")
cat("Number of fisheries:", data$n_fishery, "\n")
cat("Number of length bins:", data$n_len, "\n")
```

## Biological inputs

Natural mortality is constant across all ages at M = 0.135 yr$^{-1}$. Maturity
follows a length-based logistic curve with Mat50 = 40.7 cm. The
length-weight relationship uses a = 1.75 $\times 10^{-5}$ and b = 2.99. Fecundity is
proportional to body weight (SS3 option 1 with a = 1, b = 0).

The sex ratio vector (`sex_ratio`) is set to 1.0 at all ages. Although SS3
applies a female fraction of 0.5 when computing SSB, this scaling is already
implicit in opal's $B_0$ parameterization: opal defines $B_0$ as female-only
spawning biomass (matching SS3's reported SSB_Virgin), so the per-recruit
spawning potential does not require an additional sex-ratio multiplier. See
the dynamics diagnostic section for further discussion.

```{r plot-biology, echo=TRUE, message=FALSE, fig.height=4, fig.width=7}
# Natural mortality
ggplot(data.frame(age = ages, M = data$M), aes(x = age, y = M)) +
  geom_line() + geom_point(size = 0.8) +
  labs(x = "Age (opal internal)", y = "M (annual rate)",
       title = "Natural Mortality-at-Age")

# Maturity at length
len_mid <- seq(data$len_bin_start + data$len_bin_width / 2,
               by = data$len_bin_width, length.out = data$n_len)
ggplot(data.frame(length = len_mid, maturity = data$maturity),
       aes(x = length, y = maturity)) +
  geom_line() + geom_point() +
  labs(x = "Length (cm)", y = "Maturity",
       title = "Maturity-at-Length (logistic, Mat50 = 40.7 cm)")

# define sex ratio vector
data$sex_ratio <- rep(1, data$n_age)
```

## Length composition data

Length composition observations are loaded from the bundled `opaka_lf`
long-format data object and transformed into model-ready arrays. Data are
available for two fleets: Comm (fleet 1, 1949–2023) and ResFish (fleet 3,
2017–2023). Fleet 2 (Non_comm) has no length composition data.

```{r prep-lf, echo=TRUE, message=FALSE}
data(opaka_lf)

# Pivot to wide format: one row per fishery x timestep, bins as columns
lf_wide <- opaka_lf %>%
  tidyr::pivot_wider(
    id_cols = c(fishery, year, month, ts),
    names_from = bin,
    values_from = value,
    values_fill = 0
  ) %>%
  arrange(fishery, ts)

# Define variance adjustment scalars for each fishery
var_adjust_scalars <- 1 / rep(1, data$n_fishery)

data <- prep_lf_data(data, lf_wide,
                     lf_keep_fisheries = c(1, 3),
                     lf_var_adjust = var_adjust_scalars,
                     lf_addtocomp = 1e-10)
```

## Set up SS3-style bias ramps

SS3 applies a bias adjustment ramp to the lognormal recruitment correction
term, following Methot and Taylor (2011). The ramp reduces the correction
$-0.5 \cdot \sigma_R^2$ in years where recruitment deviations are poorly informed by
data, preventing the mean-unbiasing from introducing systematic bias when
deviations are estimated imprecisely. The ramp is defined by four control
years (ascending start, plateau start, plateau end, descending end) and a
maximum bias adjustment fraction.

Two ramps are constructed: one for the initial age cohorts (birth years
1906–1949, applied via `init_bias_adj_a`) and one for the time-series
cohorts (1949–2023, applied via `bias_adj_y`). Both use the SS3 EM's
estimated ramp parameters.

```{r add-bias-ramp, echo=TRUE, message=FALSE}
# 1. INITIAL COHORTS (1906 to 1949)
# Data ages are 1-based, so 1949 - (age - 1) gives the birth year.
# Age_a = 1 (bio age 0) -> Born in 1949.
# Age_a = 44 (bio age 43) -> Born in 1906.
init_cohort_years <- 1949 - (data$age_a - 1)

data$init_bias_adj_a <- get_bias_adj_vector(
  years = init_cohort_years,
  do_rec_bias_ramp = 1,
  bias_years = c(1931, 1974, 2021.88, 2022.43),
  max_bias_adj = 0.869
)

# 2. TIME SERIES COHORTS (1950 to 2024)
# Because rdev_y[1] generates recruits for year 2 (1950),
# the time-series bias ramp must evaluate 1950 to 2024.
calendar_years <- (1949-1) + data$years

data$bias_adj_y <- get_bias_adj_vector(
  years = calendar_years,
  do_rec_bias_ramp = 1,
  bias_years = c(1931, 1974, 2021.88, 2022.43),
  max_bias_adj = 0.869
)
```

# Model setup

## Parameters

Define the initial parameter values. Growth parameters (`log_L1`, `log_L2`,
`log_k`, `log_CV1`, `log_CV2`) and selectivity (`par_sel`) are initialised
from the SS3 estimates. Load these from the bundled `opaka_parameters` data
object:

```{r get-pars, echo=TRUE, message=FALSE}
data(opaka_parameters)

# Route 1949 dev to biological age 0, which is index 1 in opal's 1-based age vector.
# In the SS3 vector, rdev_y[1] is the 1949 deviation; here it seeds the initial
# age-0 cohort, while later years are shifted into the modeled rdev_y time series.
init_rdev_a <- rep(0.0, data$n_age)
init_rdev_a = rev(c(0.0206807,0.00321579,0.00372620,0.0043174,0.00500263,0.00579674,0.00671727,0.00778422,0.00902081,0.0104539,0.0121143,0.0140375,0.0162641,0.0188404,0.0218182,0.0252561,0.0292182,0.0337726,0.0389898,0.0449362,0.0516662,0.0592049,0.0675236,0.0764992,0.0858443,0.0950047,0.102706,0.107687,0.107712,0.0998349,0.0808137,0.0482339,0.00234443,-0.0517563,-0.102395,-0.131458,-0.121101,-0.0714826,-0.00742166,-0.0109308,-0.0385029,0.0366483,0.0165105,-0.166906))
# Remaining initial ages start at equilibrium and are estimated via init_rdev_a.
init_rdev_a[1] <- opaka_parameters$rdev_y[1]

# Shift 1950-2023 into rdev_y and add 0.0 for terminal/forecast year 2024
rdev_y_shifted <- c(opaka_parameters$rdev_y[2:length(opaka_parameters$rdev_y)], 0.0)

parameters <- list(
  log_B0 = 9,
  log_h = as.numeric(opaka_parameters$log_h),
  log_sigma_r = as.numeric(opaka_parameters$log_sigma_r),
  log_cpue_q = as.numeric(opaka_parameters$log_cpue_q),
  log_cpue_tau = rep(log(0.1),data$n_index),
  log_cpue_omega = as.numeric(rep(opaka_parameters$log_cpue_omega, data$n_index)),
  log_lf_tau = as.numeric(log(rep(0.1, data$n_fishery))),
  log_L1 = as.numeric(opaka_parameters$log_L1),
  log_L2 = as.numeric(opaka_parameters$log_L2),
  log_k = as.numeric(opaka_parameters$log_k),
  log_CV1 = as.numeric(opaka_parameters$log_CV1),
  log_CV2 = as.numeric(opaka_parameters$log_CV2),
  par_sel = as.matrix(opaka_parameters$par_sel),
  log_init_F_f = c(log(0.0119122), rep(log(1e-8), data$n_fishery - 1)),
  rdev_y = rdev_y_shifted,
  init_rdev_a = init_rdev_a
)

# convert ss3 parameters to opal parameters (selectivity)
parameters$par_sel = convert_ss3_selex_to_rtmb(parameters$par_sel, data$sel_type_f, data$len_mid)
```

## Priors

Priors are specified using `get_priors()`. The data object already contains
prior center values for growth parameters:

```{r get-priors, echo=TRUE, message=FALSE}
data$priors <- get_priors(parameters = parameters, data = data)
evaluate_priors(parameters = parameters, priors = data$priors)
```

## Parameter map

RTMB's `map` mechanism controls which parameters are estimated vs. fixed.
Parameters mapped to `factor(NA)` are held at their initial values.

The estimated parameter set includes `log_B0` (unfished spawning biomass),
`log_cpue_q` (catchability for each CPUE index), `rdev_y` (annual
recruitment deviations), `init_rdev_a` (initial age-structure deviations),
and selected selectivity parameters for the Comm and ResFish fleets.
Steepness, $\sigma_R$, growth parameters, and some selectivity parameters are fixed
at values from the SS3 OM.

```{r get-map, echo=TRUE, message=FALSE}
map_sel <- matrix(NA, nrow(parameters$par_sel), ncol(parameters$par_sel))
# map_sel[1,1:2 ] <- c(1,2) # Comm fleet selectivity parameters
# map_sel[3,c(1,4,6) ] <- c(3,4,5) # Comm fleet selectivity parameters
map_sel = as.vector(map_sel)

map_lf_tau <- rep(NA, length(parameters$log_lf_tau))

map <- list(
  # log_B0 = factor(NA),
  log_h = factor(NA),
  log_sigma_r = factor(NA),
  # log_cpue_q = factor(NA),
  log_cpue_tau = as.factor(rep(NA, data$n_index)),
  log_cpue_omega = as.factor(rep(NA, data$n_index)),
  log_lf_tau = factor(map_lf_tau),
  log_L1  = factor(NA),
  log_L2  = factor(NA),
  log_k   = factor(NA),
  log_CV1 = factor(NA),
  log_CV2 = factor(NA),
  par_sel = factor(map_sel),
  # log_init_F_f = as.factor(c(1, rep(NA, data$n_fishery - 1))),
  log_init_F_f = rep(factor(NA), data$n_fishery),
  # init_rdev_a = rep(factor(NA), data$n_age)
  init_rdev_a = as.factor(seq_len(data$n_age)),
  rdev_y = as.factor(seq_len(length(parameters$rdev_y)))
)
```

## Build the AD object

Using the `data`, the `parameters`, the parameter `map`, and the model
(`opal_model`), the AD object is created using RTMB's `MakeADFun` function:

```{r make-adfun, echo=TRUE, message=FALSE}
data$lf_switch <- 0 # skip length comps initially (removal only)
# data$lf_switch <- 1L # multinomial likelihood
# data$cpue_switch <- 0L
# Can set random = "rdev_y" to treat recruitment deviations as random effects, but start with fixed effects for simplicity
obj <- MakeADFun(func = cmb(opal_model, data), parameters = parameters, map = map)
unique(names(obj$par))
obj$fn()
obj$gr()
```

Inspect initial model outputs:

```{r init-checks, echo=TRUE, message=FALSE, fig.height=4, fig.width=7}
plot(obj$report()$spawning_biomass_y, type = "l",
     xlab = "Time step", ylab = "Spawning biomass",
     main = "Initial spawning biomass trajectory")
```

## Parameter bounds

```{r get-par-bounds, echo=TRUE, message=FALSE}
Lwr <- rep(-Inf, length(obj$par))
Upr <- rep(Inf, length(obj$par))
Lwr[grep("log_B0", names(obj$par))] <- log(1)
Upr[grep("log_B0", names(obj$par))] <- log(exp(22))
Lwr[grep("log_cpue_q", names(obj$par))] <- log(0.001)
Upr[grep("log_cpue_q", names(obj$par))] <- log(10)
Lwr[grep("rdev_y", names(obj$par))] <- rep(-5, length(grep("rdev_y", names(obj$par))))
Upr[grep("rdev_y", names(obj$par))] <- rep(5, length(grep("rdev_y", names(obj$par))))
Lwr[grep("init_rdev_a", names(obj$par))] <- rep(-5, data$n_age)
Upr[grep("init_rdev_a", names(obj$par))] <- rep(5, data$n_age)
bounds <- data.frame(par = names(obj$par), lower = Lwr, upper = Upr)
```

# ASPM

The Age-Structured Production Model (ASPM) is the simplest opal variant: all
recruitment deviations are fixed at zero, so recruitment follows the
Beverton-Holt stock-recruitment curve exactly. The only estimated parameters
are the population scale ($B_0$) and CPUE catchabilities. The model fits the
two CPUE indices only; length compositions are not included.

## Setup

```{r aspm-setup, echo=TRUE, message=FALSE}
# Build data object for ASPM: inherit lf_switch = 0, no sel_fa_external at this point
data_aspm <- data
data_aspm$cpue_switch <- 1L   # explicitly fit CPUE

# Parameters: all recruitment deviations fixed at 0
parameters_aspm <- parameters
parameters_aspm$rdev_y     <- rep(0.0, length(parameters$rdev_y))
parameters_aspm$init_rdev_a <- rep(0.0, data$n_age)

# Map: fix rdev_y and init_rdev_a in addition to ASPM-R fixed params
map_aspm <- map
map_aspm$rdev_y     <- factor(rep(NA, length(parameters_aspm$rdev_y)))
map_aspm$init_rdev_a <- factor(rep(NA, data$n_age))

obj_aspm <- MakeADFun(func = cmb(opal_model, data_aspm),
                      parameters = parameters_aspm, map = map_aspm)
unique(names(obj_aspm$par))
obj_aspm$fn()
```

## Parameter bounds

```{r aspm-bounds, echo=TRUE, message=FALSE}
Lwr_aspm <- rep(-Inf, length(obj_aspm$par))
Upr_aspm <- rep( Inf, length(obj_aspm$par))
Lwr_aspm[grep("log_B0",     names(obj_aspm$par))] <- log(1)
Upr_aspm[grep("log_B0",     names(obj_aspm$par))] <- log(exp(22))
Lwr_aspm[grep("log_cpue_q", names(obj_aspm$par))] <- log(0.001)
Upr_aspm[grep("log_cpue_q", names(obj_aspm$par))] <- log(10)
```

## Optimisation

```{r aspm-optimize, echo=TRUE, results="hide"}
control <- list(eval.max = 10000, iter.max = 10000)
opt_aspm <- nlminb(start = obj_aspm$par, objective = obj_aspm$fn,
                   gradient = obj_aspm$gr, lower = Lwr_aspm, upper = Upr_aspm,
                   control = control)
opt_aspm <- nlminb(start = opt_aspm$par, objective = obj_aspm$fn,
                   gradient = obj_aspm$gr, lower = Lwr_aspm, upper = Upr_aspm,
                   control = control)
opt_aspm <- nlminb(start = opt_aspm$par, objective = obj_aspm$fn,
                   gradient = obj_aspm$gr, lower = Lwr_aspm, upper = Upr_aspm,
                   control = control)
```

```{r aspm-convergence, echo=TRUE, message=FALSE}
cat("Convergence:", opt_aspm$convergence, "\n")
cat("Max gradient:", max(abs(obj_aspm$gr())), "\n")
obj_aspm$report()$lp_prior
sum(obj_aspm$report()$lp_cpue)
```

Check that all parameters are estimable using the `check_estimability` function:

```{r check-estimability, echo=TRUE, message=FALSE, error = TRUE}
check_estimability(obj = obj_aspm)
```

## Parameter table

```{r aspm-par-table, echo=TRUE, message=FALSE}
get_par_table(obj_aspm, parameters_aspm, map_aspm,
              lower = Lwr_aspm, upper = Upr_aspm, grad_tol = 1e-2, digits = 2L)
```

## CPUE fit

```{r aspm-cpue, echo=TRUE, message=FALSE, fig.height=4, fig.width=7}
cpue_df_aspm <- data_aspm$cpue_data %>%
  mutate(pred = obj_aspm$report()$cpue_pred)

ggplot(cpue_df_aspm, aes(x = ts, y = value)) +
  geom_point(aes(fill = factor(index)), shape = 21, size = 3) +
  geom_line(aes(y = pred, colour = factor(index)), linewidth = 1.25) +
  labs(x = "Time step", y = "CPUE", colour = "Index", fill = "Index",
       title = "ASPM: observed vs predicted CPUE") +
  scale_colour_brewer(type = "qual", palette = "Dark2") +
  scale_fill_brewer(type = "qual", palette = "Dark2") +
  theme_minimal()
```

# ASPM-R

The ASPM-R model extends the ASPM by freely estimating annual recruitment
deviations (`rdev_y`) and initial age-structure deviations (`init_rdev_a`),
allowing the model to track interannual recruitment variability. It fits the
CPUE index only; length compositions remain off.

Optimise using the `nlminb` function, run twice to ensure convergence:

```{r run-nlminb, echo=TRUE, results="hide"}
control <- list(eval.max = 10000, iter.max = 10000)
opt <- nlminb(start = obj$par, objective = obj$fn, gradient = obj$gr,
              hessian = obj$he, lower = Lwr, upper = Upr, control = control)
opt <- nlminb(start = opt$par, objective = obj$fn, gradient = obj$gr,
              hessian = obj$he, lower = Lwr, upper = Upr, control = control)
opt <- nlminb(start = opt$par, objective = obj$fn, gradient = obj$gr,
              hessian = obj$he, lower = Lwr, upper = Upr, control = control)
obj$fn()
max(abs(obj$gr()))
obj$report()$lp_prior
obj$report()$lp_penalty
obj$report()$lp_rec
sum(obj$report()$lp_cpue)
sum(obj$report()$lp_lf)
```

Compare initial and estimated parameter values:

```{r opt-results, echo=TRUE, message=FALSE}
get_par_table(obj, parameters, map, lower = Lwr, upper = Upr, grad_tol = 1e-2, digits = 2L)
```

## Diagnostics

Check that all parameters are estimable using the `check_estimability` function:

```{r check-estimability, echo=TRUE, message=FALSE, error = TRUE}
check_estimability(obj = obj)
```

Look for any parameters with high correlations (absolute correlation > 0.95)
using the `get_cor_pairs` function:

```{r check-cor, echo=TRUE, message=FALSE}
get_cor_pairs(obj = obj, threshold = 0.95)
```

Calculate standard deviations of all model parameters:

```{r run-sdreport, echo=TRUE, message=FALSE}
Report <- RTMB::sdreport(obj)
```

## Data fits

The CPUE series is set up using RTMB's `OBS` mechanism inside the model, which
allows simulation via `obj$simulate()`.

```{r simulate, echo=TRUE, message=FALSE, fig.height=4, fig.width=7}
set.seed(42)
sims <- replicate(20, obj$simulate()$cpue_log_obs, simplify = FALSE)

# again include index so points and lines are drawn per series
cpue_df <- data$cpue_data %>%
  mutate(pred = obj$report()$cpue_pred)

# plot simulated replicates with index grouping
base_plot <- ggplot(cpue_df, aes(x = ts, y = value)) +
  geom_point(aes(fill = factor(index)),shape=21,size=3) +
  geom_line(aes(y = pred, color = factor(index)),linewidth=1.25) +
  labs(x = "Time step", y = "CPUE", color = "Index", fill = "Index",
       title = "Simulated vs observed CPUE (20 replicates)") +
  scale_color_brewer(type = "qual", palette = "Dark2") +
  scale_fill_brewer(type = "qual", palette = "Dark2") +
  theme_minimal()

# add simulated series in gray
sim_df <- map_dfr(sims, ~ data.frame(ts = data$cpue_data$ts,
                                     index = as.factor(data$cpue_data$index),
                                     sim = exp(.x)), .id = "iter")

base_plot +
  geom_line(data = sim_df, aes(x = ts, y = sim, group = iter,color=index), alpha = 0.5)

```

The length compositions are also set up using RTMB's `OBS` mechanism inside the model, which
allows simulation via `obj$simulate()`.

```{r lf-diagnostics, echo=TRUE, message=FALSE, fig.height=6, fig.width=7}
if (data$lf_switch == 0) {
  cat("Length composition likelihood is switched off (lf_switch = 0), so no predicted length compositions to show.\n")
} else {
  lf_rep <- obj$report()
  lf_pred <- lf_rep$lf_pred
  nbins <- data$lf_maxbin[1] - data$lf_minbin[1] + 1

  # Name the list elements and columns
  names(lf_pred) <- data$lf_fishery_f
  for (k in seq_along(lf_pred)) colnames(lf_pred[[k]]) <- data$len_mid

  # Helper: convert a flat OBS vector back to list-of-matrices (proportions)
  flat_to_list <- function(v, lf_n_f, nbins) {
    out <- vector("list", length(lf_n_f))
    offset <- 0L
    for (k in seq_along(lf_n_f)) {
      m <- matrix(v[(offset + 1):(offset + lf_n_f[k] * nbins)], nrow = lf_n_f[k], ncol = nbins, byrow = TRUE)
      # Normalise each row to proportions
      rs <- rowSums(m)
      rs[rs == 0] <- 1
      out[[k]] <- m / rs
      offset <- offset + lf_n_f[k] * nbins
    }
    out
  }

  # Split year index by fishery (must match length of lf_year)
  lf_year_list <- split(data$lf_year, data$lf_fishery)

  # ---- Predicted: long data frame ----
  df_pred <- map_dfr(seq_along(lf_pred), function(i) {
    m <- lf_pred[[i]]
    yrs <- lf_year_list[[i]]
    as.data.frame.table(m, responseName = "pred", stringsAsFactors = FALSE) %>%
      rename(id = Var1, length = Var2) %>%
      mutate(fishery = names(lf_pred)[i],
             id = as.integer(factor(id, levels = unique(id))),
             year = yrs[id],
             length = as.numeric(length))
  })

  # ---- Observed: long data frame ----
  df_obs <- data.frame(data$lf_obs_in)
  names(df_obs) <- data$len_mid
  df_obs <- df_obs %>%
    mutate(year = data$lf_year, fishery = as.character(data$lf_fishery)) %>%
    group_by(fishery) %>%
    mutate(id = row_number()) %>%
    ungroup() %>%
    pivot_longer(cols = -c(year, fishery, id), names_to = "length", values_to = "obs") %>%
    mutate(length = as.numeric(length))

  # ---- Simulations: generate n_sim draws and convert to long format ----
  n_sim <- 5
  df_sim_all <- map_dfr(seq_len(n_sim), function(s) {
    # Determine which OBS vector was active based on lf_switch
    sim <- obj$simulate()
    if (data$lf_switch == 1) {
      sim_vec <- sim$lf_obs_flat
    } else if (data$lf_switch == 2) {
      sim_vec <- sim$lf_obs_prop
    } else {
      sim_vec <- sim$lf_obs_ints
    }
    sim_list <- flat_to_list(sim_vec, data$lf_n_f, nbins)
    names(sim_list) <- data$lf_fishery_f
    for (k in seq_along(sim_list)) colnames(sim_list[[k]]) <- data$len_mid
    map_dfr(seq_along(sim_list), function(i) {
      m <- sim_list[[i]]
      yrs <- lf_year_list[[i]]
      as.data.frame.table(m, responseName = "sim", stringsAsFactors = FALSE) %>%
        rename(id = Var1, length = Var2) %>%
        mutate(fishery = names(sim_list)[i],
               id = as.integer(factor(id, levels = unique(id))),
               year = yrs[id],
               length = as.numeric(length),
               sim_id = s)
    })
  })

  # ---- Join observed + predicted ----
  df <- left_join(df_obs, df_pred, by = c("id", "fishery", "length")) %>%
    select(-year.y) %>%
    rename(year = year.x)

  # ---- Plot fishery 1 (Comm) ----
  yrs_plot_1 <- seq(from=11,to=81,by=10)
  p1 <- ggplot(df %>% filter(fishery == "1", year %in% yrs_plot_1), aes(x = length)) +
    geom_col(aes(y = obs), fill = "grey70", width = 2) +
    geom_line(data = df_sim_all %>% filter(fishery == "1", year %in% yrs_plot_1),
              aes(y = sim, group = sim_id), colour = "steelblue", alpha = 0.3, linewidth = 0.4) +
    geom_line(aes(y = pred), colour = "red3", linewidth = 0.7) +
    facet_wrap(~ year, scales = "free_y", ncol = 5) +
    labs(x = "Length (cm)", y = "Proportion",
         title = "Length composition: Comm (Fishery 1)",
         subtitle = "Grey bars = observed, red = predicted, blue = simulated") +
    theme(strip.text = element_text(size = 7),
          axis.text = element_text(size = 6))
  print(p1)
}
```

## Model estimates

Visualise the selectivity curves by fleet (at length):

```{r plot-selectivity-length, echo=TRUE, message=FALSE, fig.height=4, fig.width=10, fig.cap="Selectivity-at-Length by Fleet"}
par_sel_est <- obj$env$parList()$par_sel

fleet_names <- c("Comm", "Non_comm", "ResFish")

sel_len_df <- do.call(rbind, lapply(seq_len(data$n_fishery), function(f) {
  par_f <- par_sel_est[f, ]
  sel_l <- if (data$sel_type_f[f] == 1L) {
    sel_logistic(data$len_mid, par_f)
  } else {
    sel_double_normal(data$len_mid, par_f)
  }
  data.frame(
    fishery    = f,
    fleet_name = fleet_names[f],
    length     = data$len_mid,
    selectivity = as.numeric(sel_l)
  )
}))

ggplot(sel_len_df, aes(x = length, y = selectivity)) +
  geom_line() +
  facet_wrap(~fleet_name, ncol = 3) +
  labs(x = "Length (cm)", y = "Selectivity") +
  ylim(0, 1)
```

Visualise the selectivity curves by fleet (at age):

```{r plot-selectivity, echo=TRUE, message=FALSE, fig.height=4, fig.width=10, fig.cap="Selectivity-at-Age by Fleet"}
rep <- obj$report()
sel_fya <- rep$sel_fya

sel_df <- expand.grid(fishery = 1:data$n_fishery, age = 1:data$n_age)
sel_df$selectivity <- sapply(1:nrow(sel_df), function(i) {
  sel_fya[sel_df$fishery[i], 1, sel_df$age[i]]
})
sel_df$real_age <- data$age_a[sel_df$age]
sel_df$fleet_name <- fleet_names[sel_df$fishery]

ggplot(sel_df, aes(x = real_age, y = selectivity)) +
  geom_line() +
  facet_wrap(~fleet_name, ncol = 3) +
  labs(x = "Age", y = "Selectivity") +
  ylim(0, 1)
```

# Cross-platform comparison

## Estimated spawning biomass

The figure below compares opal's estimated spawning biomass trajectory
(blue, with 95% confidence intervals) to three reference trajectories from
SS3: the operating model truth (black), the SS3 estimation model with 3
growth platoons (red), and an SS3 estimation model without platoons (dark
green). The SS3 EM without platoons provides the most direct comparison to
opal, which uses a single growth track with continuous PLA integration
rather than discrete platoons.

Opal's estimated trajectory tracks the SS3 EM (no platoon) closely in shape,
with a persistent negative bias in scale of approximately 10–20%. This bias
is an estimation-level difference, not a structural dynamics error, as
demonstrated by the dynamics diagnostic below.
```{r opt-spbio, echo=TRUE, message=FALSE, fig.height=4, fig.width=7}

spawning_biomass_est <- as.data.frame(summary(Report)[which(rownames(summary(Report)) == "spawning_biomass_y"), c("Estimate","Std. Error")])
colnames(spawning_biomass_est) <- c("estimate", "std_error")
spawning_biomass_est <- spawning_biomass_est[1:data$n_year,]
spawning_biomass_est$year <- calendar_years
spawning_biomass_est$lower <- spawning_biomass_est$estimate - 1.96 * spawning_biomass_est$std_error
spawning_biomass_est$upper <- spawning_biomass_est$estimate + 1.96 * spawning_biomass_est$std_error

# bring in true spawning biomass from the SS3 model (from the data object)
data(opaka_truth)

# prepare plotting limits
ybmax <- max(spawning_biomass_est$upper,
             obj$report()$spawning_biomass_y,
             opaka_truth$om$sprseries$SSB,
             opaka_truth$em$sprseries$SSB,
             na.rm = TRUE)

ggplot(spawning_biomass_est, aes(x = year)) +
  geom_ribbon(aes(ymin = lower, ymax = upper), fill = "steelblue", alpha = 0.25) +
  geom_line(aes(y = estimate), colour = "steelblue", linewidth = 1) +
  geom_line(data=subset(opaka_truth$om$sprseries, Yr %in% calendar_years), aes(x=Yr, y=SSB), colour = "black", linewidth = 1) +
  geom_line(data=subset(opaka_truth$em$sprseries, Yr %in% calendar_years), aes(x=Yr, y=SSB), colour = "red", linewidth = 1) +
  labs(x = "Year", y = "Spawning biomass",
       title = "Estimated spawning biomass trajectory",
       subtitle = "Black = SS3 OM truth, Red = SS3 EM, Coloured = opal estimates") +
  # scale_y_continuous(limits = c(0, ybmax * 1.2)) +
  theme_minimal()
```

## Dynamics diagnostic

To understand where opal's estimates differ from SS3, we build a sequence of
four model variants that progressively relax constraints:

1. **`obj`** (ASPM-R): The main opal assessment model structured as an age-structured production model (ASPM) and only fitting to the index. $B_0$, CPUE
   catchabilities, all 44 initial age-structure deviations (`init_rdev_a`), and 74 annual recruitment deviations (`rdev_y`)
   are freely estimated. Selectivity is computed through opal's PLA mechanism
   using SS3-derived length selectivity parameters.

2. **`obj_check_nosel`** (PLA selectivity; scale fixed): All parameters are
   fixed at SS3's MLE values, with selectivity computed via opal's
   probability-of-length-at-age (PLA) integration. If the trajectory matches SS3, any remaining gap in `obj`
   is an estimation-level difference.

3. **`obj_check`** (SS3 selectivity; scale fixed): Same as above but with SS3's
   exact age-selectivity matrix (from the `AGE_SELEX` report) injected directly,
   bypassing PLA. Comparing to `obj_check_nosel` reveals how much of any gap is
   attributable to the PLA approximation vs. other model structure.

4. **`obj_est`** (PLA selectivity; scale estimated): Starting from SS3's MLE
   values, $B_0$ and CPUE catchabilities are freely re-estimated while all other
   parameters (growth, selectivity, recruitment deviations) remain fixed at SS3
   values. This isolates the effect of estimating scale from the SS3 MLE starting
   point.


```{r dynamics-check, echo=TRUE, message=FALSE, fig.height=5, fig.width=8}
# ---- Dynamics diagnostic: fix all parameters at SS3 MLE values ----
# This isolates opal's population dynamics from estimation differences.
# If the forward projection matches SS3, the dynamics are correct and any
# gap in the estimated model is due to parameter estimation (different
# likelihood surface, selectivity parameterization, etc.). If the projection
# doesn't match, a structural issue remains.

# SS3 MLE values from ss.par and derived quantities
# SS3's SSB_Virgin = 4785.86, ln(R0) = 5.67638520257, h = 0.76, sigma_r = 0.52
# SS3's init_F[1] = 0.0119122, Q_parm = c(-3.74414655161, -6.23387027572)

# Build parameter list with all values at SS3 MLEs
parameters_ss3 <- parameters  # start from the current parameter list

# B0: SS3's SSB_Virgin = 4785.86 
parameters_ss3$log_B0 <- log(4785.86)

# Steepness, sigma_r 
parameters_ss3$log_h <- log(0.76) 
parameters_ss3$log_sigma_r <- log(0.52) 

# CPUE catchability (from ss.par LnQ_base)
parameters_ss3$log_cpue_q <- c(-3.74414655161, -6.23387027572) 

# Recruitment deviations: use SS3's full recdev vector
ss3_early <- c(
  -0.100388081942, -0.0133231316797, -0.0205218842101, -0.0243787483951, -0.0254464045204, 
  -0.0309043433115, -0.0323683089752, -0.0347861084933, -0.0410012292256, -0.0465287465557, 
  -0.0483985193582, -0.0650709835384, -0.0730960186062, -0.0793601697985, -0.0916013103317, 
  -0.0711755058129, -0.0873595560193, -0.0955446758745, -0.0806933151000, -0.0368260177671, 
  0.00663148238291, -0.0369263951583
) 

ss3_recdev2 <- c(
  0.0898798722060, -0.0919739062107, -0.0488564460371, -0.0797260689764, 1.83829650229, 
  -0.873301204449, -0.490428267008, -0.157171136291, -0.232120826122, 0.953858054349, 
  -0.737573581568, -1.08461065089, -0.637868159763, -0.934868322300, -1.72922585297, 
  -0.699373643035, -2.02175315411, -2.20933860510, 1.36337011943, -1.07372072783, 
  0.00685259273854, 0.0182242930198, 0.0427099426272, 0.0662006349269, 0.0694910737975, 
  0.0758887519085, 0.0793469633128, 0.0726814957641, 0.0758684448321, 0.0756207512152, 
  0.0688521101610, 0.0828900343037, 0.0715624551603, 0.0730163471792, 0.0890257135687, 
  0.0783664017535, 0.0806897553991, -0.755963354768, -0.0525403469453, 0.666734524461, 
  -0.511914821670, 0.600711769603, 0.483259386264, -0.124994792101, -0.628092867573, 
  0.335410237277, 0.572806369958, -0.896710824827, 0.367139586377, -0.139086836557, 
  -0.0176418339395, -0.313022461971, -0.284929128456, 0.297020376982, -0.545617764725, 
  -1.20081598781, -0.482796197420, -1.15050091740, -0.183455169268, -1.27372662907, 
  0.562755219094, -0.0884643712166, -0.678641658554, -0.181153422133, 0.223682687681, 
  0.503069168537, -0.240753189196, 0.310816321640, 0.589877497159, -0.0449985095147, 
  -0.525425283716, -0.102773027444, -0.810755125650, -0.0540447620912, -0.0641558428572, 
  -0.226062803588, 0.251881337494, 0.314568302935, 0.460729603143, 0.546857137016, 
  0.328741809160, 0.0844776134106, -0.00794988734173, -0.621421981067, -0.816127773538, 
  -0.0339354219406, -1.25151257570, -0.245244520498, 0.978242288327, 0.0661896749837, 
  -0.340074963625, -1.20599618914, 0.0272569302965, 0.795699552551, -0.955887084821, 
  0.939729029175, 0.657807067048, 0.820398283059, -0.0250147568033, -0.471654518218, 
  -0.182452156358, -0.0406356927574, 0.153207399783, -0.0964843056519, -0.0170828802073, 
  -0.201316841963, 0.0793882544085, 0.320280198818, 0.304378032386, -0.570128219659, 
  0.577409302896, 0.473936474269, 0.106050739082, 0.185694724022, -0.0214486454327, 
  -0.304247844565, -0.794738878920, 0.360593778732, -0.440429885505, -0.0979735757953, 0.00000000000
) 

parameters_ss3$init_rdev_a <- rev(c(ss3_early, ss3_recdev2[1:22]))
parameters_ss3$init_rdev_a[1] <- ss3_recdev2[22]  # 1949 dev at age 0

# Time-series rdevs: recdev2[23:end] are 1950 onwards, plus 0.0 for terminal
parameters_ss3$rdev_y <- c(ss3_recdev2[23:length(ss3_recdev2)], 0.0)

# Selectivity: use the converted SS3 values (already in parameters$par_sel)
parameters_ss3$par_sel <- parameters$par_sel

# Initial F
parameters_ss3$log_init_F_f <- c(log(0.0119122), rep(log(1e-8), data$n_fishery - 1)) 

# ---- Platoon-weighted selectivity-at-age override ----
# Values extracted from SS3 AGE_SELEX report (Asel2)
platoon_frac <- c(1)

# Fleet 1 (Comm) - Asel2 for single platoon (ages 0-43)
sel1_p1 <- c(
  0, 0.000743477, 0.28038, 0.883339, 0.988619, 0.998576, 0.999738, 0.999931, 
  0.999976, 0.999989, 0.999994, 0.999997, 0.999998, 0.999998, 0.999999, 0.999999, 
  0.999999, 0.999999, 0.999999, 0.999999, 0.999999, 0.999999, 0.999999, 0.999999, 
  0.999999, 0.999999, 0.999999, 0.999999, 0.999999, 0.999999, 0.999999, 0.999999, 
  0.999999, 0.999999, 0.999999, 0.999999, 0.999999, 0.999999, 0.999999, 0.999999, 
  0.999999, 0.999999, 0.999999, 0.999999
)

# Fleet 2 (Non_comm) - Asel2 for single platoon (ages 0-43)
sel2_p1 <- c(
  0, 0.0191573, 0.189067, 0.56362, 0.818265, 0.923212, 0.963411, 0.980038, 
  0.987707, 0.991629, 0.993818, 0.99513, 0.995963, 0.996516, 0.996897, 0.997166, 
  0.997362, 0.997505, 0.997613, 0.997694, 0.997755, 0.997803, 0.997839, 0.997867, 
  0.997889, 0.997906, 0.997919, 0.99793, 0.997938, 0.997944, 0.997949, 0.997953, 
  0.997956, 0.997958, 0.99796, 0.997962, 0.997963, 0.997964, 0.997964, 0.997965, 
  0.997965, 0.997966, 0.997966, 0.997966
)

# Fleet 3 (ResFish) - Asel2 for single platoon (ages 0-43)
sel3_p1 <- c(
  0, 0.989487, 0.999614, 0.906725, 0.599267, 0.369316, 0.263631, 0.219743, 
  0.200867, 0.192082, 0.187625, 0.185176, 0.183734, 0.182834, 0.182246, 0.181847, 
  0.181568, 0.181367, 0.181221, 0.181112, 0.181031, 0.180969, 0.180922, 0.180885, 
  0.180857, 0.180835, 0.180819, 0.180805, 0.180795, 0.180787, 0.180781, 0.180776, 
  0.180772, 0.180769, 0.180767, 0.180765, 0.180763, 0.180762, 0.180761, 0.180761, 
  0.18076, 0.18076, 0.180759, 0.180759
)

# Platoon-weighted selectivity-at-age by fleet
sel_fa <- matrix(0, nrow = 3, ncol = 44)
sel_fa[1, ] <- platoon_frac[1] * sel1_p1 
sel_fa[2, ] <- platoon_frac[1] * sel2_p1 
sel_fa[3, ] <- platoon_frac[1] * sel3_p1 

data$sel_fa_external <- sel_fa

# ---- Fix ALL parameters (no estimation) ----
map_fixed <- list(
  log_B0       = factor(NA),
  log_h        = factor(NA),
  log_sigma_r  = factor(NA),
  log_cpue_q   = factor(rep(NA, length(parameters_ss3$log_cpue_q))),
  log_cpue_tau = factor(rep(NA, data$n_index)),
  log_cpue_omega = factor(rep(NA, data$n_index)),
  log_lf_tau   = factor(rep(NA, length(parameters_ss3$log_lf_tau))),
  log_L1       = factor(NA),
  log_L2       = factor(NA),
  log_k        = factor(NA),
  log_CV1      = factor(NA),
  log_CV2      = factor(NA),
  par_sel      = factor(rep(NA, length(parameters_ss3$par_sel))),
  log_init_F_f = factor(rep(NA, data$n_fishery)),
  rdev_y       = factor(rep(NA, length(parameters_ss3$rdev_y))),
  init_rdev_a  = factor(rep(NA, data$n_age))
)

# ---- Build and evaluate: without selectivity override ----
data_no_sel <- data
data_no_sel$sel_fa_external <- NULL
obj_check_nosel <- MakeADFun(func = cmb(opal_model, data_no_sel),
                             parameters = parameters_ss3, map = map_fixed)
rep_check_nosel <- obj_check_nosel$report()

# ---- Build and evaluate: with selectivity override ----
obj_check <- MakeADFun(func = cmb(opal_model, data),
                       parameters = parameters_ss3, map = map_fixed)
rep_check <- obj_check$report()

cat("Without selectivity override:\n")
cat("  B0:", rep_check_nosel$B0, " R0:", rep_check_nosel$R0,
    " SPR0:", rep_check_nosel$B0 / rep_check_nosel$R0, "\n")
cat("With selectivity override:\n")
cat("  B0:", rep_check$B0, " R0:", rep_check$R0,
    " SPR0:", rep_check$B0 / rep_check$R0, "\n\n")

# Maximum absolute difference in spawning biomass
ssb_diff <- abs(rep_check$spawning_biomass_y[1:data$n_year] -
                rep_check_nosel$spawning_biomass_y[1:data$n_year])
cat("Max |SSB difference| between PLA and SS3 selectivity:",
    round(max(ssb_diff), 1), "\n")

# ---- Comparison plot ----
data(opaka_truth)

ssb_nosel <- data.frame(
  year = calendar_years,
  estimate = rep_check_nosel$spawning_biomass_y[1:data$n_year],
  model = "opal (PLA selectivity; scale fixed)"
)
ssb_sel <- data.frame(
  year = calendar_years,
  estimate = rep_check$spawning_biomass_y[1:data$n_year],
  model = "opal (SS3 selectivity; scale fixed)"
)
ssb_full <- data.frame(
  year = calendar_years,
  estimate = obj$report()$spawning_biomass_y[1:data$n_year],
  model = "opal (ASPM-R)"
)
ssb_df <- rbind(ssb_nosel, ssb_sel, ssb_full)

ybmax <- max(ssb_df$estimate,
             opaka_truth$om$sprseries$SSB,
             opaka_truth$em$sprseries$SSB, na.rm = TRUE)

ggplot() +
  geom_line(data = ssb_df, aes(x = year, y = estimate, colour = model),
            linewidth = 1) +
  geom_line(data = subset(opaka_truth$om$sprseries, Yr %in% calendar_years),
            aes(x = Yr, y = SSB, colour = "SS3 OM"), linewidth = 1) +
  geom_line(data = subset(opaka_truth$em$sprseries, Yr %in% calendar_years),
            aes(x = Yr, y = SSB, colour = "SS3 EM"), linewidth = 1) +
  scale_colour_manual(values = c("opal (SS3 selectivity; scale fixed)" = "steelblue",
                                 "opal (PLA selectivity; scale fixed)" = "darkorange",
                                 "opal (ASPM-R)" = "purple",
                                 "SS3 OM" = "black",
                                 "SS3 EM" = "red")) +
  labs(x = "Year", y = "Spawning biomass",
       title = "Dynamics diagnostic: opal at SS3 MLE parameters",
       subtitle = "No estimation — all parameters fixed at SS3 values",
       colour = NULL) +
  scale_y_continuous(limits = c(0, ybmax * 1.2)) +
  theme_minimal() +
  theme(legend.position = "bottom")
```

```{r dynamics-estimate, echo=TRUE, message=FALSE, fig.height=5, fig.width=8}
# ---- Dynamics estimate: optimize B0, init / time rec devs, and init F ----
map_est <- map_fixed
map_est$log_B0 <- NULL
map_est$log_cpue_q <- NULL
# map_est$rdev_y <- NULL
# map_est$init_rdev_a <- NULL
# map_est$log_init_F_f <- factor(c(1, rep(NA, data$n_fishery - 1)))

# Build object using opal PLA model (with selectivity override)
obj_est <- MakeADFun(func = cmb(opal_model, data_no_sel),
                     parameters = parameters_ss3,
                     map = map_est)

# Parameter bounds (relevant parameters only)
Lwr <- rep(-Inf, length(obj_est$par))
Upr <- rep(Inf, length(obj_est$par))
Lwr[grep("log_B0", names(obj_est$par))] <- log(1)
Upr[grep("log_B0", names(obj_est$par))] <- log(exp(22))
Lwr[grep("rdev_y", names(obj_est$par))] <- rep(-5, length(grep("rdev_y", names(obj_est$par))))
Upr[grep("rdev_y", names(obj_est$par))] <- rep(5, length(grep("rdev_y", names(obj_est$par))))
Lwr[grep("init_rdev_a", names(obj_est$par))] <- rep(-5, data$n_age)
Upr[grep("init_rdev_a", names(obj_est$par))] <- rep(5, data$n_age)
Lwr[grep("log_init_F_f", names(obj_est$par))] <- log(1e-12)
Upr[grep("log_init_F_f", names(obj_est$par))] <- log(3)

# Optimize (2-stage run for stability)
control <- list(eval.max = 10000, iter.max = 10000)
opt_est <- nlminb(start = obj_est$par, objective = obj_est$fn,
                  gradient = obj_est$gr, lower = Lwr, upper = Upr,
                  control = control)
opt_est <- nlminb(start = opt_est$par, objective = obj_est$fn,
                  gradient = obj_est$gr, lower = Lwr, upper = Upr,
                  control = control)

cat("estimation convergence:", opt_est$convergence, "\n")
cat("max gradient:", max(abs(obj_est$gr())), "\n")

rep_est <- obj_est$report()

ssb_est <- data.frame(
  year = calendar_years,
  estimate = rep_est$spawning_biomass_y[1:data$n_year],
  model = "opal (PLA selectivity; scale estimated)"
)

ssb_df_est <- rbind(ssb_df, ssb_est)

ybmax_est <- max(ssb_df_est$estimate,
                 opaka_truth$om$sprseries$SSB,
                 opaka_truth$em$sprseries$SSB,
                 na.rm = TRUE)

# Extended comparison plot
ggplot() +
  geom_line(data = ssb_df_est, aes(x = year, y = estimate, colour = model), linewidth = 1) +
  geom_line(data = subset(opaka_truth$om$sprseries, Yr %in% calendar_years),
            aes(x = Yr, y = SSB, colour = "SS3 OM"), linewidth = 1) +
  geom_line(data = subset(opaka_truth$em$sprseries, Yr %in% calendar_years),
            aes(x = Yr, y = SSB, colour = "SS3 EM"), linewidth = 1) +
  scale_colour_manual(values = c("opal (PLA selectivity; scale fixed)" = "steelblue",
                                 "opal (SS3 selectivity; scale fixed)" = "darkorange",
                                 "opal (PLA selectivity; scale estimated)" = "green4",
                                 "opal (ASPM-R)" = "purple",
                                 "SS3 OM" = "black",
                                 "SS3 EM" = "red")) +
  labs(x = "Year", y = "Spawning biomass",
       title = "Dynamics diagnostic + estimation comparison",
       subtitle = "Fixed SS3 MLE (PLA/SS3 selectivity) vs estimated PLA fit",
       colour = NULL) +
  scale_y_continuous(limits = c(0, ybmax_est * 1.2)) +
  theme_minimal() +
  theme(legend.position = "bottom")
```

# Model with length compositions

The ASPM-R model (`obj`) estimated $B_0$, CPUE catchabilities, and initial
age-structure deviations, but held selectivity fixed at SS3's MLE values.
Turning on the length composition likelihood allows the data to inform
selectivity directly and relaxes that constraint.

The new model (`obj_lf`) extends `obj` with two additional changes:

- **`lf_switch = 1L`** (multinomial): length composition observations from
  Comm (fleet 1, 1949--2023) and ResFish (fleet 3, 2017--2023) enter the
  likelihood.
- **Selectivity estimated** for fleets 1 and 3: the two logistic parameters
  for Comm and all six double-normal parameters for ResFish are freed.
  Fleet 2 (Non_comm) retains its fixed SS3 selectivity as it has no LF data.

All other parameters match `obj`: $\sigma_R$, steepness, and growth remain
fixed at SS3 values; $B_0$, catchabilities, `rdev_y`, and `init_rdev_a` are
freely estimated.

## Setup

```{r lf-model-setup, echo=TRUE, message=FALSE}
# data_no_sel has no sel_fa_external (set in dynamics-check); use it as base
data_lf <- data_no_sel
data_lf$lf_switch <- 1L

map_sel <- matrix(NA, nrow(parameters$par_sel), ncol(parameters$par_sel))
map_sel[1,1:2 ] <- c(1,2) # Comm fleet selectivity parameters
map_sel[3,c(1,4,6) ] <- c(3,4,5) # Comm fleet selectivity parameters
map_sel = as.vector(map_sel)

map_lf <- map
map_lf$par_sel <- factor(map_sel)
```

## Build the AD object

```{r lf-make-adfun, echo=TRUE, message=FALSE}
obj_lf <- MakeADFun(func = cmb(opal_model, data_lf),
                    parameters = parameters,
                    map = map_lf)
unique(names(obj_lf$par))
obj_lf$fn()   # should be finite
```

## Parameter bounds

```{r lf-bounds, echo=TRUE, message=FALSE}
Lwr_lf <- rep(-Inf, length(obj_lf$par))
Upr_lf <- rep( Inf, length(obj_lf$par))

Lwr_lf[grep("log_B0",      names(obj_lf$par))] <- log(1)
Upr_lf[grep("log_B0",      names(obj_lf$par))] <- log(exp(22))
Lwr_lf[grep("log_cpue_q",  names(obj_lf$par))] <- log(0.001)
Upr_lf[grep("log_cpue_q",  names(obj_lf$par))] <- log(10)
Lwr_lf[grep("rdev_y",      names(obj_lf$par))] <- -5
Upr_lf[grep("rdev_y",      names(obj_lf$par))] <-  5
Lwr_lf[grep("init_rdev_a", names(obj_lf$par))] <- -5
Upr_lf[grep("init_rdev_a", names(obj_lf$par))] <-  5
Lwr_lf[grep("par_sel",     names(obj_lf$par))] <- -7
Upr_lf[grep("par_sel",     names(obj_lf$par))] <-  7
```

## Optimisation

```{r lf-optimize, echo=TRUE, results="hide"}
control <- list(eval.max = 10000, iter.max = 10000)
opt_lf <- nlminb(start = obj_lf$par, objective = obj_lf$fn,
                 gradient = obj_lf$gr, lower = Lwr_lf, upper = Upr_lf,
                 control = control)
opt_lf <- nlminb(start = opt_lf$par, objective = obj_lf$fn,
                 gradient = obj_lf$gr, lower = Lwr_lf, upper = Upr_lf,
                 control = control)
opt_lf <- nlminb(start = opt_lf$par, objective = obj_lf$fn,
                 gradient = obj_lf$gr, lower = Lwr_lf, upper = Upr_lf,
                 control = control)
```

```{r lf-convergence, echo=TRUE, message=FALSE}
cat("Convergence:", opt_lf$convergence, "\n")
cat("Max gradient:", max(abs(obj_lf$gr())), "\n")
obj_lf$report()$lp_prior
obj_lf$report()$lp_rec
sum(obj_lf$report()$lp_cpue)
sum(obj_lf$report()$lp_lf)
check_estimability(obj = obj_lf)
```

## Parameter table

```{r lf-par-table, echo=TRUE, message=FALSE}
get_par_table(obj_lf, parameters, map_lf,lower = Lwr_lf, upper = Upr_lf, grad_tol = 1e-2, digits = 2L,include="all_est")
```

## Length composition fits

```{r lf-data-fits, echo=TRUE, message=FALSE, fig.height=6, fig.width=9}
lf_rep_lf  <- obj_lf$report()
lf_pred_lf <- lf_rep_lf$lf_pred

names(lf_pred_lf) <- data_lf$lf_fishery_f
for (k in seq_along(lf_pred_lf)) colnames(lf_pred_lf[[k]]) <- data_lf$len_mid

lf_year_list_lf <- split(data_lf$lf_year, data_lf$lf_fishery)

df_pred_lf <- map_dfr(seq_along(lf_pred_lf), function(i) {
  m   <- lf_pred_lf[[i]]
  yrs <- lf_year_list_lf[[i]]
  as.data.frame.table(m, responseName = "pred", stringsAsFactors = FALSE) %>%
    rename(id = Var1, length = Var2) %>%
    mutate(fishery = as.character(names(lf_pred_lf)[i]),
           id      = as.integer(factor(id, levels = unique(id))),
           year    = yrs[id],
           length  = as.numeric(length))
})

df_obs_lf <- data.frame(data_lf$lf_obs_in)
names(df_obs_lf) <- data_lf$len_mid
df_obs_lf <- df_obs_lf %>%
  mutate(year = data_lf$lf_year, fishery = as.character(data_lf$lf_fishery)) %>%
  group_by(fishery) %>%
  mutate(id = row_number()) %>%
  ungroup() %>%
  pivot_longer(cols = -c(year, fishery, id), names_to = "length", values_to = "obs") %>%
  mutate(length = as.numeric(length))

df_lf <- left_join(df_obs_lf, df_pred_lf, by = c("id", "fishery", "length")) %>%
  select(-year.y) %>%
  rename(year = year.x) %>%
  mutate(cal_year = 1948L + year)   # convert model timestep to calendar year

# Fleet 1 (Comm): every 10 timesteps
ts_comm <- seq(min(data_lf$lf_year[data_lf$lf_fishery == 1]),
               max(data_lf$lf_year[data_lf$lf_fishery == 1]),
               by = 10L)
p_comm <- ggplot(df_lf %>% filter(fishery == "1", year %in% ts_comm), aes(x = length)) +
  geom_col(aes(y = obs), fill = "grey70", width = 2) +
  geom_line(aes(y = pred), colour = "cyan4", linewidth = 0.8) +
  facet_wrap(~cal_year, scales = "free_y", ncol = 5) +
  labs(x = "Length (cm)", y = "Proportion",
       title = "Length composition fit: Comm (Fleet 1)",
       subtitle = "Grey = observed, cyan = predicted") +
  theme(strip.text = element_text(size = 7), axis.text = element_text(size = 6))
print(p_comm)

# Fleet 3 (ResFish): all available timesteps
ts_res <- sort(unique(data_lf$lf_year[data_lf$lf_fishery == 3]))
p_res <- ggplot(df_lf %>% filter(fishery == "3", year %in% ts_res), aes(x = length)) +
  geom_col(aes(y = obs), fill = "grey70", width = 2) +
  geom_line(aes(y = pred), colour = "cyan4", linewidth = 0.8) +
  facet_wrap(~cal_year, scales = "free_y", ncol = 4) +
  labs(x = "Length (cm)", y = "Proportion",
       title = "Length composition fit: ResFish (Fleet 3)",
       subtitle = "Grey = observed, cyan = predicted") +
  theme(strip.text = element_text(size = 7), axis.text = element_text(size = 6))
print(p_res)
```

## Selectivity comparison

```{r lf-selectivity, echo=TRUE, message=FALSE, fig.height=4, fig.width=10}
par_sel_lf  <- obj_lf$env$parList()$par_sel
par_sel_obj <- obj$env$parList()$par_sel

sel_comp_df <- do.call(rbind, lapply(seq_len(data_lf$n_fishery), function(f) {
  sel_lf_l <- if (data_lf$sel_type_f[f] == 1L) {
    sel_logistic(data_lf$len_mid, par_sel_lf[f, ])
  } else {
    sel_double_normal(data_lf$len_mid, par_sel_lf[f, ])
  }
  sel_obj_l <- if (data$sel_type_f[f] == 1L) {
    sel_logistic(data$len_mid, par_sel_obj[f, ])
  } else {
    sel_double_normal(data$len_mid, par_sel_obj[f, ])
  }
  rbind(
    data.frame(fleet = fleet_names[f], length = data_lf$len_mid,
               selectivity = as.numeric(sel_lf_l),  model = "opal (with LF)"),
    data.frame(fleet = fleet_names[f], length = data$len_mid,
               selectivity = as.numeric(sel_obj_l), model = "opal (ASPM-R)")
  )
}))

ggplot(sel_comp_df, aes(x = length, y = selectivity, colour = model)) +
  geom_line(linewidth = 0.9) +
  facet_wrap(~fleet, ncol = 3) +
  scale_colour_manual(values = c("opal (with LF)" = "cyan4",
                                 "opal (ASPM-R)"  = "purple")) +
  labs(x = "Length (cm)", y = "Selectivity",
       title = "Selectivity-at-length: ASPM-R vs model with LF",
       colour = NULL) +
  ylim(0, 1) +
  theme_minimal() +
  theme(legend.position = "bottom")
```

## Spawning biomass comparison

```{r lf-biomass, echo=TRUE, message=FALSE, fig.height=5, fig.width=8}
rep_lf <- obj_lf$report()

ssb_lf_df <- data.frame(
  year     = calendar_years,
  estimate = rep_lf$spawning_biomass_y[1:data$n_year],
  model    = "opal (with LF)"
)

ssb_df_all <- rbind(ssb_df_est, ssb_lf_df)

ybmax_all <- max(ssb_df_all$estimate,
                 opaka_truth$om$sprseries$SSB,
                 opaka_truth$em$sprseries$SSB,
                 na.rm = TRUE)

ggplot() +
  geom_line(data = ssb_df_all,
            aes(x = year, y = estimate, colour = model), linewidth = 1) +
  geom_line(data = subset(opaka_truth$om$sprseries, Yr %in% calendar_years),
            aes(x = Yr, y = SSB, colour = "SS3 OM"), linewidth = 1) +
  geom_line(data = subset(opaka_truth$em$sprseries, Yr %in% calendar_years),
            aes(x = Yr, y = SSB, colour = "SS3 EM"), linewidth = 1) +
  scale_colour_manual(values = c(
    "opal (PLA selectivity; scale fixed)"     = "steelblue",
    "opal (SS3 selectivity; scale fixed)"     = "darkorange",
    "opal (PLA selectivity; scale estimated)" = "green4",
    "opal (ASPM-R)"                           = "purple",
    "opal (with LF)"                          = "cyan4",
    "SS3 OM"                                  = "black",
    "SS3 EM"                                  = "red"
  )) +
  labs(x = "Year", y = "Spawning biomass",
       title = "Full model comparison",
       subtitle = "Fixed diagnostics \u2192 ASPM-R \u2192 model with LF + estimated selectivity",
       colour = NULL) +
  scale_y_continuous(limits = c(0, ybmax_all * 1.2)) +
  theme_minimal() +
  theme(legend.position = "right")
```

# Catch Curve

The catch-curve model uses the same parameter structure as the opal with LF
model (selectivity estimated for Comm and ResFish, recruitment deviations
freely estimated) but removes the CPUE index from the likelihood. Instead,
information about population scale comes entirely from the length composition
data and the catch history. Because catchability ($q$) is not informed by
any index, it is fixed at its SS3 MLE value to avoid an unidentified
parameter.

## Setup

```{r cc-setup, echo=TRUE, message=FALSE}
# Inherit lf_switch = 1L and prepared LF arrays from data_lf
data_cc <- data_lf
data_cc$cpue_switch <- 0L   # exclude CPUE from likelihood

# Fix log_cpue_q since it is unidentified without CPUE data
map_cc <- map_lf
map_cc$log_cpue_q <- factor(rep(NA, length(parameters$log_cpue_q)))

obj_cc <- MakeADFun(func = cmb(opal_model, data_cc),
                    parameters = parameters,
                    map = map_cc)
unique(names(obj_cc$par))
obj_cc$fn()   # should be finite
```

## Parameter bounds

```{r cc-bounds, echo=TRUE, message=FALSE}
Lwr_cc <- rep(-Inf, length(obj_cc$par))
Upr_cc <- rep( Inf, length(obj_cc$par))

Lwr_cc[grep("log_B0",      names(obj_cc$par))] <- log(1)
Upr_cc[grep("log_B0",      names(obj_cc$par))] <- log(exp(22))
Lwr_cc[grep("rdev_y",      names(obj_cc$par))] <- -5
Upr_cc[grep("rdev_y",      names(obj_cc$par))] <-  5
Lwr_cc[grep("init_rdev_a", names(obj_cc$par))] <- -5
Upr_cc[grep("init_rdev_a", names(obj_cc$par))] <-  5
Lwr_cc[grep("par_sel",     names(obj_cc$par))] <- -7
Upr_cc[grep("par_sel",     names(obj_cc$par))] <-  7
```

## Optimisation

```{r cc-optimize, echo=TRUE, results="hide"}
control <- list(eval.max = 10000, iter.max = 10000)
opt_cc <- nlminb(start = obj_cc$par, objective = obj_cc$fn,
                 gradient = obj_cc$gr, lower = Lwr_cc, upper = Upr_cc,
                 control = control)
opt_cc <- nlminb(start = opt_cc$par, objective = obj_cc$fn,
                 gradient = obj_cc$gr, lower = Lwr_cc, upper = Upr_cc,
                 control = control)
opt_cc <- nlminb(start = opt_cc$par, objective = obj_cc$fn,
                 gradient = obj_cc$gr, lower = Lwr_cc, upper = Upr_cc,
                 control = control)
```

```{r cc-convergence, echo=TRUE, message=FALSE}
cat("Convergence:", opt_cc$convergence, "\n")
cat("Max gradient:", max(abs(obj_cc$gr())), "\n")
obj_cc$report()$lp_prior
obj_cc$report()$lp_rec
sum(obj_cc$report()$lp_lf)
```

## Diagnostics

```{r cc-diagnostics, echo=TRUE, message=FALSE, error=TRUE}
get_par_table(obj_cc, parameters, map_cc,
              lower = Lwr_cc, upper = Upr_cc, grad_tol = 1e-2, digits = 2L,
              include = "all_est")
check_estimability(obj = obj_cc)
```

## Length composition fits

```{r cc-lf-fits, echo=TRUE, message=FALSE, fig.height=6, fig.width=9}
rep_cc       <- obj_cc$report()
lf_pred_cc   <- rep_cc$lf_pred

names(lf_pred_cc) <- data_cc$lf_fishery_f
for (k in seq_along(lf_pred_cc)) colnames(lf_pred_cc[[k]]) <- data_cc$len_mid

lf_year_list_cc <- split(data_cc$lf_year, data_cc$lf_fishery)

df_pred_cc <- map_dfr(seq_along(lf_pred_cc), function(i) {
  m   <- lf_pred_cc[[i]]
  yrs <- lf_year_list_cc[[i]]
  as.data.frame.table(m, responseName = "pred", stringsAsFactors = FALSE) %>%
    rename(id = Var1, length = Var2) %>%
    mutate(fishery = as.character(names(lf_pred_cc)[i]),
           id      = as.integer(factor(id, levels = unique(id))),
           year    = yrs[id],
           length  = as.numeric(length))
})

df_obs_cc <- data.frame(data_cc$lf_obs_in)
names(df_obs_cc) <- data_cc$len_mid
df_obs_cc <- df_obs_cc %>%
  mutate(year = data_cc$lf_year, fishery = as.character(data_cc$lf_fishery)) %>%
  group_by(fishery) %>%
  mutate(id = row_number()) %>%
  ungroup() %>%
  pivot_longer(cols = -c(year, fishery, id), names_to = "length", values_to = "obs") %>%
  mutate(length = as.numeric(length))

df_cc <- left_join(df_obs_cc, df_pred_cc, by = c("id", "fishery", "length")) %>%
  select(-year.y) %>%
  rename(year = year.x) %>%
  mutate(cal_year = 1948L + year)

# Fleet 1 (Comm): decadal subset
ts_comm_cc <- seq(min(data_cc$lf_year[data_cc$lf_fishery == 1]),
                  max(data_cc$lf_year[data_cc$lf_fishery == 1]),
                  by = 10L)
p_cc_comm <- ggplot(df_cc %>% filter(fishery == "1", year %in% ts_comm_cc),
                    aes(x = length)) +
  geom_col(aes(y = obs), fill = "grey70", width = 2) +
  geom_line(aes(y = pred), colour = "darkorange", linewidth = 0.8) +
  facet_wrap(~cal_year, scales = "free_y", ncol = 5) +
  labs(x = "Length (cm)", y = "Proportion",
       title = "Catch curve: length composition fit — Comm (Fleet 1)",
       subtitle = "Grey = observed, orange = predicted") +
  theme(strip.text = element_text(size = 7), axis.text = element_text(size = 6))
print(p_cc_comm)

# Fleet 3 (ResFish): all available timesteps
ts_res_cc <- sort(unique(data_cc$lf_year[data_cc$lf_fishery == 3]))
p_cc_res <- ggplot(df_cc %>% filter(fishery == "3", year %in% ts_res_cc),
                   aes(x = length)) +
  geom_col(aes(y = obs), fill = "grey70", width = 2) +
  geom_line(aes(y = pred), colour = "darkorange", linewidth = 0.8) +
  facet_wrap(~cal_year, scales = "free_y", ncol = 4) +
  labs(x = "Length (cm)", y = "Proportion",
       title = "Catch curve: length composition fit — ResFish (Fleet 3)",
       subtitle = "Grey = observed, orange = predicted") +
  theme(strip.text = element_text(size = 7), axis.text = element_text(size = 6))
print(p_cc_res)
```

## Spawning biomass

```{r cc-biomass, echo=TRUE, message=FALSE, fig.height=5, fig.width=8}
ssb_cc_df <- data.frame(
  year     = calendar_years,
  estimate = rep_cc$spawning_biomass_y[1:data$n_year],
  model    = "opal (catch curve)"
)

ssb_df_cc <- rbind(
  data.frame(year = calendar_years,
             estimate = obj$report()$spawning_biomass_y[1:data$n_year],
             model = "opal (ASPM-R)"),
  data.frame(year = calendar_years,
             estimate = rep_lf$spawning_biomass_y[1:data$n_year],
             model = "opal (with LF)"),
  ssb_cc_df
)

ybmax_cc <- max(ssb_df_cc$estimate,
                opaka_truth$om$sprseries$SSB,
                opaka_truth$em$sprseries$SSB, na.rm = TRUE)

ggplot() +
  geom_line(data = ssb_df_cc, aes(x = year, y = estimate, colour = model),
            linewidth = 1) +
  geom_line(data = subset(opaka_truth$om$sprseries, Yr %in% calendar_years),
            aes(x = Yr, y = SSB, colour = "SS3 OM"), linewidth = 1) +
  geom_line(data = subset(opaka_truth$em$sprseries, Yr %in% calendar_years),
            aes(x = Yr, y = SSB, colour = "SS3 EM"), linewidth = 1) +
  scale_colour_manual(values = c(
    "opal (ASPM-R)"     = "purple",
    "opal (with LF)"    = "cyan4",
    "opal (catch curve)" = "darkorange",
    "SS3 OM"            = "black",
    "SS3 EM"            = "red"
  )) +
  labs(x = "Year", y = "Spawning biomass",
       title = "Spawning biomass: ASPM-R vs opal with LF vs catch curve",
       colour = NULL) +
  scale_y_continuous(limits = c(0, ybmax_cc * 1.2)) +
  theme_minimal() +
  theme(legend.position = "bottom")
```

# Final summary

This section collects the four summary plots across all primary model
variants: ASPM (index only, no recruitment deviations), ASPM-R (index only,
recruitment deviations estimated), catch curve (LF only, no CPUE), and
opal with LF (index + LF, selectivity estimated). SS3 OM and SS3 EM
trajectories are included in the biomass panel for reference.

A shared colour palette is used throughout:

```{r summary-palette, echo=TRUE}
summary_colours <- c(
  "ASPM"           = "#E69F00",  # amber
  "ASPM-R"         = "purple",
  "Catch curve"    = "darkorange",
  "opal (with LF)" = "cyan4",
  "SS3 OM"         = "black",
  "SS3 EM"         = "red"
)
```

## CPUE fits

CPUE predicted trajectories for ASPM, ASPM-R, and opal with LF. The catch
curve is excluded as it does not fit the CPUE index.

```{r summary-cpue, echo=TRUE, message=FALSE, fig.height=4, fig.width=8}
make_cpue_df <- function(obj, model_label, cpue_data) {
  cpue_data %>%
    mutate(pred  = obj$report()$cpue_pred,
           model = model_label)
}

cpue_summary <- bind_rows(
  make_cpue_df(obj_aspm, "ASPM",           data_aspm$cpue_data),
  make_cpue_df(obj,      "ASPM-R",          data$cpue_data),
  make_cpue_df(obj_lf,   "opal (with LF)", data_lf$cpue_data)
)

ggplot(cpue_summary, aes(x = ts)) +
  geom_point(data = cpue_summary %>% filter(model == "ASPM-R"),
             aes(y = value, fill = factor(index)),
             shape = 21, size = 2.5) +
  geom_line(aes(y = pred, colour = model, linetype = factor(index)),
            linewidth = 1) +
  scale_colour_manual(values = summary_colours) +
  scale_fill_brewer(type = "qual", palette = "Dark2") +
  facet_wrap(~index, ncol = 2, scales = "free_y",
             labeller = labeller(index = c("1" = "Index 1 (Comm)", "2" = "Index 2 (Survey)"))) +
  labs(x = "Time step", y = "CPUE", colour = "Model", fill = "Index",
       linetype = "Index",
       title = "Summary: CPUE fits",
       subtitle = "Points = observed; lines = model predictions") +
  theme_minimal() +
  theme(legend.position = "bottom")
```

## Length composition fits

Observed length compositions (grey bars) with predicted proportions from
opal with LF (cyan) and the catch curve (orange). Shown for Comm (decadal
subset) and all ResFish years.

```{r summary-lf, echo=TRUE, message=FALSE, fig.height=7, fig.width=10}
# Build long-format predicted LF for both models, labelled
build_lf_pred_df <- function(obj_x, data_x, model_label) {
  pred_list <- obj_x$report()$lf_pred
  names(pred_list) <- data_x$lf_fishery_f
  for (k in seq_along(pred_list)) colnames(pred_list[[k]]) <- data_x$len_mid
  yr_list <- split(data_x$lf_year, data_x$lf_fishery)
  map_dfr(seq_along(pred_list), function(i) {
    m   <- pred_list[[i]]
    yrs <- yr_list[[i]]
    as.data.frame.table(m, responseName = "pred", stringsAsFactors = FALSE) %>%
      rename(id = Var1, length = Var2) %>%
      mutate(fishery   = as.character(names(pred_list)[i]),
             id        = as.integer(factor(id, levels = unique(id))),
             year      = yrs[id],
             cal_year  = 1948L + year,
             length    = as.numeric(length),
             model     = model_label)
  })
}

df_pred_sum_lf <- build_lf_pred_df(obj_lf, data_lf, "opal (with LF)")
df_pred_sum_cc <- build_lf_pred_df(obj_cc, data_cc, "Catch curve")
df_pred_summary <- bind_rows(df_pred_sum_lf, df_pred_sum_cc)

# Observed (use data_lf; same obs for both models)
df_obs_sum <- data.frame(data_lf$lf_obs_in)
names(df_obs_sum) <- data_lf$len_mid
df_obs_sum <- df_obs_sum %>%
  mutate(year = data_lf$lf_year, fishery = as.character(data_lf$lf_fishery)) %>%
  group_by(fishery) %>% mutate(id = row_number()) %>% ungroup() %>%
  pivot_longer(cols = -c(year, fishery, id), names_to = "length", values_to = "obs") %>%
  mutate(length = as.numeric(length), cal_year = 1948L + year)

# Comm: decadal years
ts_comm_sum <- seq(min(data_lf$lf_year[data_lf$lf_fishery == 1]),
                   max(data_lf$lf_year[data_lf$lf_fishery == 1]), by = 10L)
obs_comm <- df_obs_sum %>% filter(fishery == "1", year %in% ts_comm_sum)
pred_comm <- df_pred_summary %>% filter(fishery == "1", year %in% ts_comm_sum)

p_sum_comm <- ggplot(obs_comm, aes(x = length)) +
  geom_col(aes(y = obs), fill = "grey70", width = 2) +
  geom_line(data = pred_comm,
            aes(y = pred, colour = model), linewidth = 0.8) +
  scale_colour_manual(values = c("opal (with LF)" = "cyan4",
                                 "Catch curve"    = "darkorange")) +
  facet_wrap(~cal_year, scales = "free_y", ncol = 5) +
  labs(x = "Length (cm)", y = "Proportion", colour = "Model",
       title = "Summary: LF fits — Comm (Fleet 1, decadal)") +
  theme(strip.text = element_text(size = 7),
        axis.text  = element_text(size = 6),
        legend.position = "bottom")
print(p_sum_comm)

# ResFish: all years
ts_res_sum <- sort(unique(data_lf$lf_year[data_lf$lf_fishery == 3]))
obs_res  <- df_obs_sum %>% filter(fishery == "3", year %in% ts_res_sum)
pred_res <- df_pred_summary %>% filter(fishery == "3", year %in% ts_res_sum)

p_sum_res <- ggplot(obs_res, aes(x = length)) +
  geom_col(aes(y = obs), fill = "grey70", width = 2) +
  geom_line(data = pred_res,
            aes(y = pred, colour = model), linewidth = 0.8) +
  scale_colour_manual(values = c("opal (with LF)" = "cyan4",
                                 "Catch curve"    = "darkorange")) +
  facet_wrap(~cal_year, scales = "free_y", ncol = 4) +
  labs(x = "Length (cm)", y = "Proportion", colour = "Model",
       title = "Summary: LF fits — ResFish (Fleet 3, all years)") +
  theme(strip.text = element_text(size = 7),
        axis.text  = element_text(size = 6),
        legend.position = "bottom")
print(p_sum_res)
```

## Spawning biomass

```{r summary-biomass, echo=TRUE, message=FALSE, fig.height=5, fig.width=8}
ssb_summary <- bind_rows(
  data.frame(year = calendar_years,
             estimate = obj_aspm$report()$spawning_biomass_y[1:data$n_year],
             model = "ASPM"),
  data.frame(year = calendar_years,
             estimate = obj$report()$spawning_biomass_y[1:data$n_year],
             model = "ASPM-R"),
  data.frame(year = calendar_years,
             estimate = rep_cc$spawning_biomass_y[1:data$n_year],
             model = "Catch curve"),
  data.frame(year = calendar_years,
             estimate = rep_lf$spawning_biomass_y[1:data$n_year],
             model = "opal (with LF)")
)

ybmax_sum <- max(ssb_summary$estimate,
                 opaka_truth$om$sprseries$SSB,
                 opaka_truth$em$sprseries$SSB, na.rm = TRUE)

ggplot() +
  geom_line(data = ssb_summary,
            aes(x = year, y = estimate, colour = model), linewidth = 1) +
  geom_line(data = subset(opaka_truth$om$sprseries, Yr %in% calendar_years),
            aes(x = Yr, y = SSB, colour = "SS3 OM"), linewidth = 1) +
  geom_line(data = subset(opaka_truth$em$sprseries, Yr %in% calendar_years),
            aes(x = Yr, y = SSB, colour = "SS3 EM"), linewidth = 1) +
  scale_colour_manual(values = summary_colours) +
  labs(x = "Year", y = "Spawning biomass",
       title = "Summary: spawning biomass trajectories",
       colour = NULL) +
  scale_y_continuous(limits = c(0, ybmax_sum * 1.2)) +
  theme_minimal() +
  theme(legend.position = "bottom")
```

## Dynamic depletion

Dynamic depletion ($B_t / B_{0,\text{dynamic}}$) for each opal model variant.
The SS3 OM and SS3 EM are not shown as their depletion is defined relative to
a different reference point and is not directly comparable to opal's
`dynamic_depletion_y` metric.

```{r summary-depletion, echo=TRUE, message=FALSE, fig.height=4, fig.width=8}
dep_summary <- bind_rows(
  data.frame(year = calendar_years,
             depletion = obj_aspm$report()$dynamic_depletion_y[1:data$n_year],
             model = "ASPM"),
  data.frame(year = calendar_years,
             depletion = obj$report()$dynamic_depletion_y[1:data$n_year],
             model = "ASPM-R"),
  data.frame(year = calendar_years,
             depletion = rep_cc$dynamic_depletion_y[1:data$n_year],
             model = "Catch curve"),
  data.frame(year = calendar_years,
             depletion = rep_lf$dynamic_depletion_y[1:data$n_year],
             model = "opal (with LF)")
)

ggplot(dep_summary, aes(x = year, y = depletion, colour = model)) +
  geom_line(linewidth = 1) +
  geom_hline(yintercept = 0.4, linetype = "dashed", colour = "grey50") +
  scale_colour_manual(values = summary_colours) +
  labs(x = "Year", y = expression(B[t] / B[0*","*dynamic]),
       title = "Summary: dynamic depletion trajectories",
       subtitle = "Dashed line = 40% depletion reference",
       colour = NULL) +
  scale_y_continuous(limits = c(0, 1)) +
  theme_minimal() +
  theme(legend.position = "bottom")
```

