---
title: "The 'Opakapaka model"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{The 'Opakapaka model}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
---

```{r, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

# Introduction

`opal` is an open, modular R package for fisheries stock assessment built on
[RTMB](https://github.com/kaskr/RTMB) for automatic differentiation. This
vignette demonstrates model setup, fitting, and diagnostics using bundled data
for ['opakapaka](https://github.com/noaa-pifsc/Opaka_simulation), extracted from a Stock Synthesis 3 (SS3) model and converted to
opal conventions. The model covers the historical period 1949–2023 with three
fleets: a commercial fishery (Comm), a non-commercial fishery (Non_comm), and a
research fishery survey (ResFish).

# Load inputs

Load the `opal` package and the `RTMB` dependency. The `tidyverse` package is
used for data manipulation and plotting.

```{r load-pkg, echo=TRUE, message=FALSE}
library(tidyverse)
library(opal)
library(RTMB)

theme_set(theme_bw())
```

The bundled data object `opaka_data` contains all biological parameters, catch
and CPUE observations, length structure, and prior specifications needed for the
opakapaka assessment model:

```{r load-data, echo=TRUE, message=FALSE}
data(opaka_data)
data <- opaka_data
if (is.null(data$n_index)) data$n_index <- 2L
names(data)
```

Key dimensions of the data:

```{r data-dims, echo=TRUE}
cat("Number of ages:", data$n_age, "\n")
cat("Number of years:", data$n_year, "\n")
cat("Number of fisheries:", data$n_fishery, "\n")
cat("Number of length bins:", data$n_len, "\n")
```

## Biological inputs

Natural mortality, maturity at length, and length-weight parameters are all
contained in the data object. Unlike the BET case study where M varies with
age, opakapaka uses a constant M = 0.135 across all ages.

```{r plot-biology, echo=TRUE, message=FALSE, fig.height=4, fig.width=7}
ages <- data$age_a

# Natural mortality
ggplot(data.frame(age = ages, M = data$M), aes(x = age, y = M)) +
  geom_line() + geom_point(size = 0.8) +
  labs(x = "Age (opal internal)", y = "M (annual rate)",
       title = "Natural Mortality-at-Age")

# Maturity at length
len_mid <- seq(data$len_bin_start + data$len_bin_width / 2,
               by = data$len_bin_width, length.out = data$n_len)
ggplot(data.frame(length = len_mid, maturity = data$maturity),
       aes(x = length, y = maturity)) +
  geom_line() + geom_point() +
  labs(x = "Length (cm)", y = "Maturity",
       title = "Maturity-at-Length (logistic, Mat50 = 40.7 cm)")
```

## Length composition data

Length composition observations are loaded from the bundled `opaka_lf`
long-format data object and transformed into model-ready arrays. Data are
available for two fleets: Comm (fleet 1, 1949–2023) and ResFish (fleet 3,
2017–2023). Fleet 2 (Non_comm) has no length composition data.

```{r prep-lf, echo=TRUE, message=FALSE}
data(opaka_lf)

# Pivot to wide format: one row per fishery x timestep, bins as columns
lf_wide <- opaka_lf %>%
  tidyr::pivot_wider(
    id_cols = c(fishery, year, month, ts),
    names_from = bin,
    values_from = value,
    values_fill = 0
  ) %>%
  arrange(fishery, ts)

# Define variance adjustment scalars for each fishery
var_adjust_scalars <- 1 / rep(1, data$n_fishery)

data <- prep_lf_data(data, lf_wide,
                     lf_keep_fisheries = c(1, 3),
                     lf_var_adjust = var_adjust_scalars)
```

## Add data free years

Add "data free" years at the model start to allow the population to depart from the
steady-state age structure without influencing the likelihood.  This requires
adjusting every time‑series array so that the first `n_early` years are blank
(zeros) and any time-step columns/rows are shifted accordingly.

```{r add-data-free, echo=TRUE, message=FALSE}
# number of years of data to prepend
n_early <- 10

# extend the year vector and related indices
data$n_year <- data$n_year + n_early
data$years <- seq(min(data$years), by = 1, length.out = data$n_year)
data$first_yr <- data$years[1]
data$last_yr <- data$years[data$n_year]
# catch series starts after the data‑free years
data$first_yr_catch <- data$years[n_early + 1]

# shift CPUE time steps so they line up with the expanded year vector
data$cpue_data <- data$cpue_data %>%
  mutate(ts = ts + n_early)

# prepend zeros along the first (year) dimension of the 3‑D catch array
catch_dim <- dim(data$catch_obs_ysf)
zeros <- array(0, dim = c(n_early, catch_dim[2], catch_dim[3]))
new_catch <- array(0, dim = c(catch_dim[1] + n_early,
                               catch_dim[2], catch_dim[3]))
new_catch[(n_early + 1):nrow(new_catch), , ] <- data$catch_obs_ysf

data$catch_obs_ysf <- new_catch

data$lf_year <- data$lf_year + n_early
```


# Model setup

## Parameters

Define the initial parameter values. Growth parameters (`log_L1`, `log_L2`,
`log_k`, `log_CV1`, `log_CV2`) and selectivity (`par_sel`) are initialised
from the SS3 estimates. Load these from the bundled `opaka_parameters` data
object:

```{r get-pars, echo=TRUE, message=FALSE}
data(opaka_parameters)

parameters <- list(
  log_B0 = 15,
  log_h = as.numeric(opaka_parameters$log_h),
  log_sigma_r = as.numeric(opaka_parameters$log_sigma_r),
  log_cpue_q = as.numeric(opaka_parameters$log_cpue_q),
  cpue_creep = as.numeric(rep(opaka_parameters$cpue_creep, data$n_index)),
  log_cpue_tau = rep(log(0.1),data$n_index),
  log_cpue_omega = as.numeric(rep(opaka_parameters$log_cpue_omega, data$n_index)),
  log_lf_tau = as.numeric(log(rep(0.1, data$n_fishery))),
  log_L1 = as.numeric(opaka_parameters$log_L1),
  log_L2 = as.numeric(opaka_parameters$log_L2),
  log_k = as.numeric(opaka_parameters$log_k),
  log_CV1 = as.numeric(opaka_parameters$log_CV1),
  log_CV2 = as.numeric(opaka_parameters$log_CV2),
  par_sel = as.matrix(opaka_parameters$par_sel),
  rdev_y = as.numeric(rep(0, data$n_year))
)
```

## Priors

Priors are specified using `get_priors()`. The data object already contains
prior center values for growth parameters:

```{r get-priors, echo=TRUE, message=FALSE}
data$priors <- get_priors(parameters = parameters, data = data)
evaluate_priors(parameters = parameters, priors = data$priors)
```

## Parameter map

Use RTMB's `map` option to turn parameters on/off. Parameters mapped to
`factor(NA)` are fixed at their initial values. For the initial run, we fix
most parameters and estimate only `log_B0`, `log_cpue_q`, and `rdev_y`:

```{r get-map, echo=TRUE, message=FALSE}
map_sel <- matrix(NA, nrow(parameters$par_sel), ncol(parameters$par_sel))
# map_sel[1,1:2 ] <- c(1, 2) # Comm fleet selectivity parameters
# map_sel[3,c(1,3,4) ] <- c(3,4,5) # Comm fleet selectivity parameters

map_lf_tau <- rep(NA, length(parameters$log_lf_tau))

map <- list(
  # log_B0 = factor(NA),
  log_h = factor(NA),
  log_sigma_r = factor(NA),
  # log_cpue_q = factor(NA),
  cpue_creep = as.factor(rep(NA, data$n_index)),
  log_cpue_tau = as.factor(rep(NA, data$n_index)),
  log_cpue_omega = as.factor(rep(NA, data$n_index)),
  log_lf_tau = factor(map_lf_tau),
  log_L1  = factor(NA),
  log_L2  = factor(NA),
  log_k   = factor(NA),
  log_CV1 = factor(NA),
  log_CV2 = factor(NA),
  par_sel = factor(map_sel)
  # rdev_y = as.factor(rep(NA, length(parameters$rdev_y)))
)
```

## Build the AD object

Using the `data`, the `parameters`, the parameter `map`, and the model
(`opal_model`), the AD object is created using RTMB's `MakeADFun` function:

```{r make-adfun, echo=TRUE, message=FALSE}
data$lf_switch <- 0 # skip length comps initially (removal only)
# data$lf_switch <- 1L # multinomial likelihood

obj <- MakeADFun(func = cmb(opal_model, data), parameters = parameters, map = map)
unique(names(obj$par))
obj$fn()
obj$gr()
```

Inspect initial model outputs:

```{r init-checks, echo=TRUE, message=FALSE, fig.height=4, fig.width=7}
plot(obj$report()$spawning_biomass_y, type = "l",
     xlab = "Time step", ylab = "Spawning biomass",
     main = "Initial spawning biomass trajectory")
```

## Parameter bounds

```{r get-par-bounds, echo=TRUE, message=FALSE}
Lwr <- rep(-Inf, length(obj$par))
Upr <- rep(Inf, length(obj$par))
Lwr[grep("log_B0", names(obj$par))] <- log(1)
Upr[grep("log_B0", names(obj$par))] <- log(exp(22))
Lwr[grep("log_cpue_q", names(obj$par))] <- log(0.001)
Upr[grep("log_cpue_q", names(obj$par))] <- log(10)
Lwr[grep("rdev_y", names(obj$par))] <- rep(-5, length(grep("rdev_y", names(obj$par))))
Upr[grep("rdev_y", names(obj$par))] <- rep(5, length(grep("rdev_y", names(obj$par))))
bounds <- data.frame(par = names(obj$par), lower = Lwr, upper = Upr)
```

# Optimisation

Optimise using the `nlminb` function, run twice to ensure convergence:

```{r run-nlminb, echo=TRUE, results="hide"}
control <- list(eval.max = 10000, iter.max = 10000)
opt <- nlminb(start = obj$par, objective = obj$fn, gradient = obj$gr,
              hessian = obj$he, lower = Lwr, upper = Upr, control = control)
opt <- nlminb(start = opt$par, objective = obj$fn, gradient = obj$gr,
              hessian = obj$he, lower = Lwr, upper = Upr, control = control)
obj$fn()
max(abs(obj$gr()))
obj$report()$lp_prior
obj$report()$lp_penalty
obj$report()$lp_rec
sum(obj$report()$lp_cpue)
sum(obj$report()$lp_lf)
```

Compare initial and estimated parameter values:

```{r opt-results, echo=TRUE, message=FALSE}
get_par_table(obj, parameters, map, lower = Lwr, upper = Upr, grad_tol = 1e-2, digits = 2L)
```

Inspect fitted CPUE and catch:

```{r opt-cpue, echo=TRUE, message=FALSE, fig.height=4, fig.width=7}
# combine observations and predictions and respect the index column
cpue_df <- data$cpue_data %>%
  mutate(pred = obj$report()$cpue_pred)

# simple ggplot version that splits by index (could also facet by season/fishery)
ggplot(cpue_df, aes(x = ts, y = value)) +
  geom_point(aes(color = factor(index), shape = factor(index))) +
  geom_line(aes(y = pred, color = factor(index))) +
  labs(x = "Time step", y = "CPUE", color = "Index", shape = "Index",
       title = "CPUE: observed vs predicted") +
  scale_color_brewer(type = "qual", palette = "Dark2") +
  theme_minimal()
```

Inspect estimated spawning biomass trajectory:

```{r opt-spbio, echo=TRUE, message=FALSE, fig.height=4, fig.width=7}
plot(obj$report()$spawning_biomass_y, type = "l",
     xlab = "Time step", ylab = "Spawning biomass",
     main = "Estimated spawning biomass trajectory",
     ylim = c(0, max(obj$report()$spawning_biomass_y) * 1.2))
```

## Selectivity

Visualise the selectivity curves by fleet:

```{r plot-selectivity, echo=TRUE, message=FALSE, fig.height=4, fig.width=10, fig.cap="Selectivity-at-Age by Fleet"}
rep <- obj$report()
sel_fya <- rep$sel_fya

fleet_names <- c("Comm", "Non_comm", "ResFish")

sel_df <- expand.grid(fishery = 1:data$n_fishery, age = 1:data$n_age)
sel_df$selectivity <- sapply(1:nrow(sel_df), function(i) {
  sel_fya[sel_df$fishery[i], 1, sel_df$age[i]]
})
sel_df$fleet_name <- fleet_names[sel_df$fishery]

ggplot(sel_df, aes(x = age, y = selectivity)) +
  geom_line() +
  facet_wrap(~fleet_name, ncol = 3) +
  labs(x = "Age (opal internal)", y = "Selectivity") +
  ylim(0, 1)
```

## Diagnostics

Check that all parameters are estimable using the `check_estimability` function:

```{r check-estimability, echo=TRUE, message=FALSE, error = TRUE}
check_estimability(obj = obj)
```

Look for any parameters with high correlations (absolute correlation > 0.95)
using the `get_cor_pairs` function:

```{r check-cor, echo=TRUE, message=FALSE}
get_cor_pairs(obj = obj, threshold = 0.95)
```

Calculate standard deviations of all model parameters:

```{r run-sdreport, echo=TRUE, message=FALSE}
Report <- sdreport(obj)
```

## Data fits

Inspect predicted vs observed length compositions:

```{r lf-diagnostics, echo=TRUE, message=FALSE, fig.height=6, fig.width=7}
if (data$lf_switch == 0) {
  cat("Length composition likelihood is switched off (lf_switch = 0),",
      "so no predicted length compositions to show.\n")
} else {
  lf_rep <- obj$report()
  lf_pred <- lf_rep$lf_pred
  nbins <- data$lf_maxbin[1] - data$lf_minbin[1] + 1

  names(lf_pred) <- data$lf_fishery_f
  for (k in seq_along(lf_pred)) colnames(lf_pred[[k]]) <- data$len_mid

  lf_year_list <- split(data$lf_year, data$lf_fishery)

  df_pred <- map_dfr(seq_along(lf_pred), function(i) {
    m <- lf_pred[[i]]
    yrs <- lf_year_list[[i]]
    as.data.frame.table(m, responseName = "pred", stringsAsFactors = FALSE) %>%
      rename(id = Var1, length = Var2) %>%
      mutate(fishery = names(lf_pred)[i],
             id = as.integer(factor(id, levels = unique(id))),
             year = yrs[id],
             length = as.numeric(length),
             pred = as.numeric(pred))
  })

  print(
    ggplot(df_pred, aes(x = length, y = pred)) +
      geom_line() +
      facet_wrap(~fishery + year, scales = "free_y", ncol = 5) +
      labs(x = "Length (cm)", y = "Predicted proportion",
           title = "Predicted length compositions by fleet and year")
  )
}
```

# Simulation

The CPUE series is set up using RTMB's `OBS` mechanism inside the model, which
allows simulation via `obj$simulate()`.

```{r simulate, echo=TRUE, message=FALSE, fig.height=4, fig.width=7}
set.seed(42)
sims <- replicate(20, obj$simulate()$cpue_log_obs, simplify = FALSE)

# again include index so points and lines are drawn per series
cpue_df <- data$cpue_data %>%
  mutate(pred = obj$report()$cpue_pred)

# plot simulated replicates with index grouping
base_plot <- ggplot(cpue_df, aes(x = ts, y = value)) +
  geom_point(aes(color = factor(index), shape = factor(index))) +
  geom_line(aes(y = pred, color = factor(index))) +
  labs(x = "Time step", y = "CPUE", color = "Index", shape = "Index",
       title = "Simulated vs observed CPUE (20 replicates)") +
  scale_color_brewer(type = "qual", palette = "Dark2") +
  theme_minimal()

# add simulated series in gray
sim_df <- map_dfr(sims, ~ data.frame(ts = data$cpue_data$ts,
                                     index = as.factor(data$cpue_data$index),
                                     sim = exp(.x)), .id = "iter")

base_plot +
  geom_line(data = sim_df, aes(x = ts, y = sim, group = iter,color=index), alpha = 0.5)

```
