---
title: "Review an Opakapaka Fit"
format: html
execute:
  eval: true
  warning: false
  message: false
vignette: >
  %\VignetteIndexEntry{Review an Opakapaka Fit}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

```{r}
#| include: false
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

# Review an `opal_fit`

This vignette examines the portable fixed-effects opakapaka fit created for
the Quickstart. It rebuilds the transient RTMB objective only to obtain model
predictions, so no optimisation is performed here.

The bundled fit may have been saved under a different R version. Portable
integrity permits that difference while still checking model compatibility and
requiring the rebuilt objective to match the saved value.

```{r}
#| label: load-fit
library(opal)

fit_path <- system.file("extdata", "opaka_quickstart_fit.rds", package = "opal")
if (!nzchar(fit_path)) {
  stop("The bundled opakapaka quickstart fit is unavailable.", call. = FALSE)
}
fit <- read_opal_fit(fit_path, strict = TRUE, integrity = "portable")
fit
summary(fit)
```

## CPUE fit

```{r}
#| label: cpue-fit
#| fig-width: 8
#| fig-height: 5
object <- opal_fit_object(fit, integrity = "portable")
plot_data <- fit$data
plot_data$cpue_data$season <- 1L
plot_cpue(plot_data, object)
```

## Diagnostics

The compact estimability result and maximum absolute gradient were saved with
the fit. They can be inspected without creating a new objective.

```{r}
#| label: diagnostics
list(
  estimability = fit$fit$estimability,
  maximum_gradient = fit$fit$diagnostics$max_gradient
)
```