---
title: "'Opakapaka case study: benchmarking opal against Stock Synthesis 3"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{'Opakapaka case study: benchmarking opal against Stock Synthesis 3}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
---

```{r, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

# Introduction

This vignette documents the configuration, fitting, and cross-platform
benchmarking of an `opal` stock assessment model for 'opakapaka
(*Pristipomoides filamentosus*), a Hawaiian deepwater snapper. The reference
model was built in Stock Synthesis 3 (SS3 v3.30.19) as part of the
[Opaka_simulation](https://github.com/noaa-pifsc/Opaka_simulation) project,
which provides both an operating model (OM) and an estimation model (EM).

The primary objective is to verify that `opal`'s population dynamics engine
reproduces SS3's spawning biomass trajectory when given identical inputs. A
secondary objective is to demonstrate model fit to multiple data sources:
index and length composition data.

The model covers the historical period 1949–2023 with three fleets: a
commercial fishery (Comm), a non-commercial fishery (Non_comm), and a
research fishery survey (ResFish).

# Load inputs

Load the `opal` package and the `RTMB` dependency. The `tidyverse` package is
used for data manipulation and plotting.

```{r load-pkg, echo=TRUE, message=FALSE}
library(tidyverse)
library(opal)
library(RTMB)

theme_set(theme_bw())
```

The bundled data object `opaka_data` contains all biological parameters, catch
and CPUE observations, length structure, and prior specifications needed for the
opakapaka assessment model.

SS3 indexes ages from 0 to 43 (44 age classes), while opal uses 1-based
internal indexing. To ensure the growth curve, survivorship, and all
age-structured biology vectors align correctly with SS3, we set `A1 = 0`
(anchoring the Schnute growth parameterization at biological age 0) and
redefine `age_a` accordingly. The `min_age` field is passed to
`get_growth()` so that the age sequence starts at 0 rather than 1.

```{r load-data, echo=TRUE, message=FALSE}
data(opaka_data)
data <- opaka_data
if (is.null(data$n_index)) data$n_index <- 2L
names(data)

# remove priors
data$priors <- NULL

# set ages to anchor growth curve (since SS3 indexs from 0 to 43, but opal uses 1-based indexing)
data$A1 <- 0
data$A2 <- data$A1 + (data$n_age-1)
data$min_age <- 0
ages <- data$age_a <- seq(data$min_age, by = 1, length.out = data$n_age)
```

Key dimensions of the data:

```{r data-dims, echo=TRUE}
cat("Number of ages:", data$n_age, "\n")
cat("Number of years:", data$n_year, "\n")
cat("Number of fisheries:", data$n_fishery, "\n")
cat("Number of length bins:", data$n_len, "\n")
```

## Biological inputs

Natural mortality is constant across all ages at M = 0.135 yr$^{-1}$. Maturity
follows a length-based logistic curve with Mat50 = 40.7 cm. The
length-weight relationship uses a = 1.75 $\times 10^{-5}$ and b = 2.99. Fecundity is
proportional to body weight (SS3 option 1 with a = 1, b = 0).

The sex ratio vector (`sex_ratio`) is set to 1.0 at all ages. Although SS3
applies a female fraction of 0.5 when computing SSB, this scaling is already
implicit in opal's $B_0$ parameterization: opal defines $B_0$ as female-only
spawning biomass (matching SS3's reported SSB_Virgin), so the per-recruit
spawning potential does not require an additional sex-ratio multiplier. See
the dynamics diagnostic section for further discussion.

```{r plot-biology, echo=TRUE, message=FALSE, fig.height=4, fig.width=7}
# Natural mortality
ggplot(data.frame(age = ages, M = data$M), aes(x = age, y = M)) +
  geom_line() + geom_point(size = 0.8) +
  labs(x = "Age (opal internal)", y = "M (annual rate)",
       title = "Natural Mortality-at-Age")

# Maturity at length
len_mid <- seq(data$len_bin_start + data$len_bin_width / 2,
               by = data$len_bin_width, length.out = data$n_len)
ggplot(data.frame(length = len_mid, maturity = data$maturity),
       aes(x = length, y = maturity)) +
  geom_line() + geom_point() +
  labs(x = "Length (cm)", y = "Maturity",
       title = "Maturity-at-Length (logistic, Mat50 = 40.7 cm)")

# define sex ratio vector
data$sex_ratio <- rep(1, data$n_age)
```

## Length composition data

Length composition observations are loaded from the bundled `opaka_lf`
long-format data object and transformed into model-ready arrays. Data are
available for two fleets: Comm (fleet 1, 1949–2023) and ResFish (fleet 3,
2017–2023). Fleet 2 (Non_comm) has no length composition data.

```{r prep-lf, echo=TRUE, message=FALSE}
data(opaka_lf)

# Pivot to wide format: one row per fishery x timestep, bins as columns
lf_wide <- opaka_lf %>%
  tidyr::pivot_wider(
    id_cols = c(fishery, year, month, ts),
    names_from = bin,
    values_from = value,
    values_fill = 0
  ) %>%
  arrange(fishery, ts)

# Define variance adjustment scalars for each fishery
var_adjust_scalars <- 1 / rep(1, data$n_fishery)

data <- prep_lf_data(data, lf_wide,
                     lf_keep_fisheries = c(1, 3),
                     lf_var_adjust = var_adjust_scalars,
                     lf_addtocomp = 1e-10)
```

## Set up SS3-style bias ramps

SS3 applies a bias adjustment ramp to the lognormal recruitment correction
term, following Methot and Taylor (2011). The ramp reduces the correction
$-0.5 \cdot \sigma_R^2$ in years where recruitment deviations are poorly informed by
data, preventing the mean-unbiasing from introducing systematic bias when
deviations are estimated imprecisely. The ramp is defined by four control
years (ascending start, plateau start, plateau end, descending end) and a
maximum bias adjustment fraction.

Two ramps are constructed: one for the initial age cohorts (birth years
1906–1949, applied via `init_bias_adj_a`) and one for the time-series
cohorts (1949–2023, applied via `bias_adj_y`). Both use the SS3 EM's
estimated ramp parameters.

```{r add-bias-ramp, echo=TRUE, message=FALSE}
# 1. INITIAL COHORTS (1906 to 1949)
# Data ages are 1-based, so 1949 - (age - 1) gives the birth year.
# Age_a = 1 (bio age 0) -> Born in 1949.
# Age_a = 44 (bio age 43) -> Born in 1906.
init_cohort_years <- 1949 - (data$age_a - 1)

data$init_bias_adj_a <- get_bias_adj_vector(
  years = init_cohort_years,
  do_rec_bias_ramp = 1,
  bias_years = c(1931, 1974, 2021.88, 2022.43),
  max_bias_adj = 0.869
)

# 2. TIME SERIES COHORTS (1950 to 2024)
# Because rdev_y[1] generates recruits for year 2 (1950),
# the time-series bias ramp must evaluate 1950 to 2024.
calendar_years <- (1949-1) + data$years

data$bias_adj_y <- get_bias_adj_vector(
  years = calendar_years,
  do_rec_bias_ramp = 1,
  bias_years = c(1931, 1974, 2021.88, 2022.43),
  max_bias_adj = 0.869
)
```

# Model setup

## Parameters

Define the initial parameter values. Growth parameters (`log_L1`, `log_L2`,
`log_k`, `log_CV1`, `log_CV2`) and selectivity (`par_sel`) are initialised
from the SS3 estimates. Load these from the bundled `opaka_parameters` data
object:

```{r get-pars, echo=TRUE, message=FALSE}
data(opaka_parameters)

# Route 1949 dev to biological age 0, which is index 1 in opal's 1-based age vector.
# In the SS3 vector, rdev_y[1] is the 1949 deviation; here it seeds the initial
# age-0 cohort, while later years are shifted into the modeled rdev_y time series.
init_rdev_a <- rep(0.0, data$n_age)
init_rdev_a = rev(c(0.0206807,0.00321579,0.00372620,0.0043174,0.00500263,0.00579674,0.00671727,0.00778422,0.00902081,0.0104539,0.0121143,0.0140375,0.0162641,0.0188404,0.0218182,0.0252561,0.0292182,0.0337726,0.0389898,0.0449362,0.0516662,0.0592049,0.0675236,0.0764992,0.0858443,0.0950047,0.102706,0.107687,0.107712,0.0998349,0.0808137,0.0482339,0.00234443,-0.0517563,-0.102395,-0.131458,-0.121101,-0.0714826,-0.00742166,-0.0109308,-0.0385029,0.0366483,0.0165105,-0.166906))
# Remaining initial ages start at equilibrium and are estimated via init_rdev_a.
init_rdev_a[1] <- opaka_parameters$rdev_y[1]

# Shift 1950-2023 into rdev_y and add 0.0 for terminal/forecast year 2024
rdev_y_shifted <- c(opaka_parameters$rdev_y[2:length(opaka_parameters$rdev_y)], 0.0)

parameters <- list(
  log_B0 = 9,
  log_h = as.numeric(opaka_parameters$log_h),
  log_sigma_r = as.numeric(opaka_parameters$log_sigma_r),
  log_cpue_q = as.numeric(opaka_parameters$log_cpue_q),
  cpue_creep = as.numeric(rep(opaka_parameters$cpue_creep, data$n_index)),
  log_cpue_tau = rep(log(0.1),data$n_index),
  log_cpue_omega = as.numeric(rep(opaka_parameters$log_cpue_omega, data$n_index)),
  log_lf_tau = as.numeric(log(rep(0.1, data$n_fishery))),
  log_L1 = as.numeric(opaka_parameters$log_L1),
  log_L2 = as.numeric(opaka_parameters$log_L2),
  log_k = as.numeric(opaka_parameters$log_k),
  log_CV1 = as.numeric(opaka_parameters$log_CV1),
  log_CV2 = as.numeric(opaka_parameters$log_CV2),
  par_sel = as.matrix(opaka_parameters$par_sel),
  log_init_F_f = c(log(0.0119122), rep(log(1e-8), data$n_fishery - 1)),
  rdev_y = rdev_y_shifted,
  init_rdev_a = init_rdev_a
)

# convert ss3 parameters to opal parameters (selectivity)
parameters$par_sel = convert_ss3_selex_to_rtmb(parameters$par_sel, data$sel_type_f, data$len_mid)
```

## Priors

Priors are specified using `get_priors()`. The data object already contains
prior center values for growth parameters:

```{r get-priors, echo=TRUE, message=FALSE}
data$priors <- get_priors(parameters = parameters, data = data)
evaluate_priors(parameters = parameters, priors = data$priors)
```

## Parameter map

RTMB's `map` mechanism controls which parameters are estimated vs. fixed.
Parameters mapped to `factor(NA)` are held at their initial values.

The estimated parameter set includes `log_B0` (unfished spawning biomass),
`log_cpue_q` (catchability for each CPUE index), `rdev_y` (annual
recruitment deviations), `init_rdev_a` (initial age-structure deviations),
and selected selectivity parameters for the Comm and ResFish fleets.
Steepness, $\sigma_R$, growth parameters, and some selectivity parameters are fixed
at values from the SS3 OM.

```{r get-map, echo=TRUE, message=FALSE}
map_sel <- matrix(NA, nrow(parameters$par_sel), ncol(parameters$par_sel))
map_sel[1,1:2 ] <- c(1,2) # Comm fleet selectivity parameters
map_sel[3,c(1,4,6) ] <- c(3,4,5) # Comm fleet selectivity parameters
map_sel = as.vector(map_sel)

map_lf_tau <- rep(NA, length(parameters$log_lf_tau))

map <- list(
  # log_B0 = factor(NA),
  log_h = factor(NA),
  log_sigma_r = factor(NA),
  # log_cpue_q = factor(NA),
  cpue_creep = as.factor(rep(NA, data$n_index)),
  log_cpue_tau = as.factor(rep(NA, data$n_index)),
  log_cpue_omega = as.factor(rep(NA, data$n_index)),
  log_lf_tau = factor(map_lf_tau),
  log_L1  = factor(NA),
  log_L2  = factor(NA),
  log_k   = factor(NA),
  log_CV1 = factor(NA),
  log_CV2 = factor(NA),
  par_sel = factor(map_sel),
  # log_init_F_f = as.factor(c(1, rep(NA, data$n_fishery - 1))),
  log_init_F_f = rep(factor(NA), data$n_fishery),
  # init_rdev_a = rep(factor(NA), data$n_age)
  init_rdev_a = as.factor(seq_len(data$n_age))
  # rdev_y = as.factor(rep(NA, length(parameters$rdev_y)))
)
```

## Build the AD object

Using the `data`, the `parameters`, the parameter `map`, and the model
(`opal_model`), the AD object is created using RTMB's `MakeADFun` function:

```{r make-adfun, echo=TRUE, message=FALSE}
# data$lf_switch <- 0 # skip length comps initially (removal only)
data$lf_switch <- 1L # multinomial likelihood
# data$cpue_switch <- 0L
# Can set random = "rdev_y" to treat recruitment deviations as random effects, but start with fixed effects for simplicity
obj <- MakeADFun(func = cmb(opal_model, data), parameters = parameters, map = map)
unique(names(obj$par))
obj$fn()
obj$gr()
```

Inspect initial model outputs:

```{r init-checks, echo=TRUE, message=FALSE, fig.height=4, fig.width=7}
plot(obj$report()$spawning_biomass_y, type = "l",
     xlab = "Time step", ylab = "Spawning biomass",
     main = "Initial spawning biomass trajectory")
```

## Parameter bounds

```{r get-par-bounds, echo=TRUE, message=FALSE}
Lwr <- rep(-Inf, length(obj$par))
Upr <- rep(Inf, length(obj$par))
Lwr[grep("log_B0", names(obj$par))] <- log(1)
Upr[grep("log_B0", names(obj$par))] <- log(exp(22))
Lwr[grep("log_cpue_q", names(obj$par))] <- log(0.001)
Upr[grep("log_cpue_q", names(obj$par))] <- log(10)
Lwr[grep("rdev_y", names(obj$par))] <- rep(-5, length(grep("rdev_y", names(obj$par))))
Upr[grep("rdev_y", names(obj$par))] <- rep(5, length(grep("rdev_y", names(obj$par))))
Lwr[grep("init_rdev_a", names(obj$par))] <- rep(-5, data$n_age)
Upr[grep("init_rdev_a", names(obj$par))] <- rep(5, data$n_age)
bounds <- data.frame(par = names(obj$par), lower = Lwr, upper = Upr)
```

# Optimisation

Optimise using the `nlminb` function, run twice to ensure convergence:

```{r run-nlminb, echo=TRUE, results="hide"}
control <- list(eval.max = 10000, iter.max = 10000)
opt <- nlminb(start = obj$par, objective = obj$fn, gradient = obj$gr,
              hessian = obj$he, lower = Lwr, upper = Upr, control = control)
opt <- nlminb(start = opt$par, objective = obj$fn, gradient = obj$gr,
              hessian = obj$he, lower = Lwr, upper = Upr, control = control)
opt <- nlminb(start = opt$par, objective = obj$fn, gradient = obj$gr,
              hessian = obj$he, lower = Lwr, upper = Upr, control = control)
obj$fn()
max(abs(obj$gr()))
obj$report()$lp_prior
obj$report()$lp_penalty
obj$report()$lp_rec
sum(obj$report()$lp_cpue)
sum(obj$report()$lp_lf)
```

Compare initial and estimated parameter values:

```{r opt-results, echo=TRUE, message=FALSE}
get_par_table(obj, parameters, map, lower = Lwr, upper = Upr, grad_tol = 1e-2, digits = 2L)
```

## Diagnostics

Check that all parameters are estimable using the `check_estimability` function:

```{r check-estimability, echo=TRUE, message=FALSE, error = TRUE}
check_estimability(obj = obj)
```

Look for any parameters with high correlations (absolute correlation > 0.95)
using the `get_cor_pairs` function:

```{r check-cor, echo=TRUE, message=FALSE}
get_cor_pairs(obj = obj, threshold = 0.95)
```

Calculate standard deviations of all model parameters:

```{r run-sdreport, echo=TRUE, message=FALSE}
Report <- RTMB::sdreport(obj)
```

## Data fits

The CPUE series is set up using RTMB's `OBS` mechanism inside the model, which
allows simulation via `obj$simulate()`.

```{r simulate, echo=TRUE, message=FALSE, fig.height=4, fig.width=7}
set.seed(42)
sims <- replicate(20, obj$simulate()$cpue_log_obs, simplify = FALSE)

# again include index so points and lines are drawn per series
cpue_df <- data$cpue_data %>%
  mutate(pred = obj$report()$cpue_pred)

# plot simulated replicates with index grouping
base_plot <- ggplot(cpue_df, aes(x = ts, y = value)) +
  geom_point(aes(fill = factor(index)),shape=21,size=3) +
  geom_line(aes(y = pred, color = factor(index)),linewidth=1.25) +
  labs(x = "Time step", y = "CPUE", color = "Index", fill = "Index",
       title = "Simulated vs observed CPUE (20 replicates)") +
  scale_color_brewer(type = "qual", palette = "Dark2") +
  scale_fill_brewer(type = "qual", palette = "Dark2") +
  theme_minimal()

# add simulated series in gray
sim_df <- map_dfr(sims, ~ data.frame(ts = data$cpue_data$ts,
                                     index = as.factor(data$cpue_data$index),
                                     sim = exp(.x)), .id = "iter")

base_plot +
  geom_line(data = sim_df, aes(x = ts, y = sim, group = iter,color=index), alpha = 0.5)

```

The length compositions are also set up using RTMB's `OBS` mechanism inside the model, which
allows simulation via `obj$simulate()`.

```{r lf-diagnostics, echo=TRUE, message=FALSE, fig.height=6, fig.width=7}
if (data$lf_switch == 0) {
  cat("Length composition likelihood is switched off (lf_switch = 0), so no predicted length compositions to show.\n")
} else {
  lf_rep <- obj$report()
  lf_pred <- lf_rep$lf_pred
  nbins <- data$lf_maxbin[1] - data$lf_minbin[1] + 1

  # Name the list elements and columns
  names(lf_pred) <- data$lf_fishery_f
  for (k in seq_along(lf_pred)) colnames(lf_pred[[k]]) <- data$len_mid

  # Helper: convert a flat OBS vector back to list-of-matrices (proportions)
  flat_to_list <- function(v, lf_n_f, nbins) {
    out <- vector("list", length(lf_n_f))
    offset <- 0L
    for (k in seq_along(lf_n_f)) {
      m <- matrix(v[(offset + 1):(offset + lf_n_f[k] * nbins)], nrow = lf_n_f[k], ncol = nbins, byrow = TRUE)
      # Normalise each row to proportions
      rs <- rowSums(m)
      rs[rs == 0] <- 1
      out[[k]] <- m / rs
      offset <- offset + lf_n_f[k] * nbins
    }
    out
  }

  # Split year index by fishery (must match length of lf_year)
  lf_year_list <- split(data$lf_year, data$lf_fishery)

  # ---- Predicted: long data frame ----
  df_pred <- map_dfr(seq_along(lf_pred), function(i) {
    m <- lf_pred[[i]]
    yrs <- lf_year_list[[i]]
    as.data.frame.table(m, responseName = "pred", stringsAsFactors = FALSE) %>%
      rename(id = Var1, length = Var2) %>%
      mutate(fishery = names(lf_pred)[i],
             id = as.integer(factor(id, levels = unique(id))),
             year = yrs[id],
             length = as.numeric(length))
  })

  # ---- Observed: long data frame ----
  df_obs <- data.frame(data$lf_obs_in)
  names(df_obs) <- data$len_mid
  df_obs <- df_obs %>%
    mutate(year = data$lf_year, fishery = as.character(data$lf_fishery)) %>%
    group_by(fishery) %>%
    mutate(id = row_number()) %>%
    ungroup() %>%
    pivot_longer(cols = -c(year, fishery, id), names_to = "length", values_to = "obs") %>%
    mutate(length = as.numeric(length))

  # ---- Simulations: generate n_sim draws and convert to long format ----
  n_sim <- 5
  df_sim_all <- map_dfr(seq_len(n_sim), function(s) {
    # Determine which OBS vector was active based on lf_switch
    sim <- obj$simulate()
    if (data$lf_switch == 1) {
      sim_vec <- sim$lf_obs_flat
    } else if (data$lf_switch == 2) {
      sim_vec <- sim$lf_obs_prop
    } else {
      sim_vec <- sim$lf_obs_ints
    }
    sim_list <- flat_to_list(sim_vec, data$lf_n_f, nbins)
    names(sim_list) <- data$lf_fishery_f
    for (k in seq_along(sim_list)) colnames(sim_list[[k]]) <- data$len_mid
    map_dfr(seq_along(sim_list), function(i) {
      m <- sim_list[[i]]
      yrs <- lf_year_list[[i]]
      as.data.frame.table(m, responseName = "sim", stringsAsFactors = FALSE) %>%
        rename(id = Var1, length = Var2) %>%
        mutate(fishery = names(sim_list)[i],
               id = as.integer(factor(id, levels = unique(id))),
               year = yrs[id],
               length = as.numeric(length),
               sim_id = s)
    })
  })

  # ---- Join observed + predicted ----
  df <- left_join(df_obs, df_pred, by = c("id", "fishery", "length")) %>%
    select(-year.y) %>%
    rename(year = year.x)

  # ---- Plot fishery 1 (Comm) ----
  yrs_plot_1 <- seq(from=11,to=81,by=10)
  p1 <- ggplot(df %>% filter(fishery == "1", year %in% yrs_plot_1), aes(x = length)) +
    geom_col(aes(y = obs), fill = "grey70", width = 2) +
    geom_line(data = df_sim_all %>% filter(fishery == "1", year %in% yrs_plot_1),
              aes(y = sim, group = sim_id), colour = "steelblue", alpha = 0.3, linewidth = 0.4) +
    geom_line(aes(y = pred), colour = "red3", linewidth = 0.7) +
    facet_wrap(~ year, scales = "free_y", ncol = 5) +
    labs(x = "Length (cm)", y = "Proportion",
         title = "Length composition: Comm (Fishery 1)",
         subtitle = "Grey bars = observed, red = predicted, blue = simulated") +
    theme(strip.text = element_text(size = 7),
          axis.text = element_text(size = 6))
  print(p1)
}
```

## Model estimates

Visualise the selectivity curves by fleet (at length):

```{r plot-selectivity-length, echo=TRUE, message=FALSE, fig.height=4, fig.width=10, fig.cap="Selectivity-at-Length by Fleet"}
par_sel_est <- obj$env$parList()$par_sel

fleet_names <- c("Comm", "Non_comm", "ResFish")

sel_len_df <- do.call(rbind, lapply(seq_len(data$n_fishery), function(f) {
  par_f <- par_sel_est[f, ]
  sel_l <- if (data$sel_type_f[f] == 1L) {
    sel_logistic(data$len_mid, par_f)
  } else {
    sel_double_normal(data$len_mid, par_f)
  }
  data.frame(
    fishery    = f,
    fleet_name = fleet_names[f],
    length     = data$len_mid,
    selectivity = as.numeric(sel_l)
  )
}))

ggplot(sel_len_df, aes(x = length, y = selectivity)) +
  geom_line() +
  facet_wrap(~fleet_name, ncol = 3) +
  labs(x = "Length (cm)", y = "Selectivity") +
  ylim(0, 1)
```

Visualise the selectivity curves by fleet (at age):

```{r plot-selectivity, echo=TRUE, message=FALSE, fig.height=4, fig.width=10, fig.cap="Selectivity-at-Age by Fleet"}
rep <- obj$report()
sel_fya <- rep$sel_fya

sel_df <- expand.grid(fishery = 1:data$n_fishery, age = 1:data$n_age)
sel_df$selectivity <- sapply(1:nrow(sel_df), function(i) {
  sel_fya[sel_df$fishery[i], 1, sel_df$age[i]]
})
sel_df$real_age <- data$age_a[sel_df$age]
sel_df$fleet_name <- fleet_names[sel_df$fishery]

ggplot(sel_df, aes(x = real_age, y = selectivity)) +
  geom_line() +
  facet_wrap(~fleet_name, ncol = 3) +
  labs(x = "Age", y = "Selectivity") +
  ylim(0, 1)
```

# Cross-platform comparison

## Estimated spawning biomass

The figure below compares opal's estimated spawning biomass trajectory
(blue, with 95% confidence intervals) to three reference trajectories from
SS3: the operating model truth (black), the SS3 estimation model with 3
growth platoons (red), and an SS3 estimation model without platoons (dark
green). The SS3 EM without platoons provides the most direct comparison to
opal, which uses a single growth track with continuous PLA integration
rather than discrete platoons.

Opal's estimated trajectory tracks the SS3 EM (no platoon) closely in shape,
with a persistent negative bias in scale of approximately 10–20%. This bias
is an estimation-level difference, not a structural dynamics error, as
demonstrated by the dynamics diagnostic below.
```{r opt-spbio, echo=TRUE, message=FALSE, fig.height=4, fig.width=7}

spawning_biomass_est <- as.data.frame(summary(Report)[which(rownames(summary(Report)) == "spawning_biomass_y"), c("Estimate","Std. Error")])
colnames(spawning_biomass_est) <- c("estimate", "std_error")
spawning_biomass_est <- spawning_biomass_est[1:data$n_year,]
spawning_biomass_est$year <- calendar_years
spawning_biomass_est$lower <- spawning_biomass_est$estimate - 1.96 * spawning_biomass_est$std_error
spawning_biomass_est$upper <- spawning_biomass_est$estimate + 1.96 * spawning_biomass_est$std_error

# bring in true spawning biomass from the SS3 model (from the data object)
data(opaka_truth)

# prepare plotting limits
ybmax <- max(spawning_biomass_est$upper,
             obj$report()$spawning_biomass_y,
             opaka_truth$om$sprseries$SSB,
             opaka_truth$em$sprseries$SSB,
             opaka_truth$em_no_platoon$sprseries$SSB,
             na.rm = TRUE)

ggplot(spawning_biomass_est, aes(x = year)) +
  geom_ribbon(aes(ymin = lower, ymax = upper), fill = "steelblue", alpha = 0.25) +
  geom_line(aes(y = estimate), colour = "steelblue", linewidth = 1) +
  geom_line(data=subset(opaka_truth$om$sprseries, Yr %in% calendar_years), aes(x=Yr, y=SSB), colour = "black", linewidth = 1) +
  geom_line(data=subset(opaka_truth$em$sprseries, Yr %in% calendar_years), aes(x=Yr, y=SSB), colour = "red", linewidth = 1) +
  geom_line(data=subset(opaka_truth$em_no_platoon$sprseries, Yr %in% calendar_years), aes(x=Yr, y=SSB), colour = "darkgreen", linewidth = 1) +
  labs(x = "Year", y = "Spawning biomass",
       title = "Estimated spawning biomass trajectory",
       subtitle = "Black = SS3 OM truth, Red = SS3 EM, Dark green = SS3 EM (no platoon), Coloured = opal estimates") +
  # scale_y_continuous(limits = c(0, ybmax * 1.2)) +
  theme_minimal()
```

## Platoon-corrected estimation

SS3's 3-platoon structure (weights 0.15/0.70/0.15, within/between stdev
ratio 0.7) creates effective weight-at-age and spawning-potential-at-age
values that differ from opal's single-track PLA approximation. This arises
from Jensen's inequality: because the length-weight relationship (w = aL^b)
and logistic maturity curve are non-linear, the platoon-weighted average at
each age is not equal to the value computed at mean length.

To quantify this effect, we override opal's internally derived weight and
spawning potential vectors with platoon-weighted averages extracted from
SS3's Biology_at_age report table. The model is then re-estimated with these
corrected inputs. This sensitivity does not change the estimated selectivity
or likelihood structure — it isolates the impact of the platoon
approximation on biological scaling.

```{r platoon-correction, echo=TRUE, message=FALSE, fig.height=5, fig.width=8}
# ---- Save the default (PLA-based) opal estimate for comparison ----
spawning_biomass_default <- spawning_biomass_est
spawning_biomass_default$model <- "opal (PLA)"

# ---- Platoon-weighted biology from SS3 report ----
# SS3 uses 3 discrete growth platoons (slow/medium/fast) with weights
# 0.15/0.70/0.15 and a within/between stdev ratio of 0.7. The effective
# weight-at-age and spawning potential-at-age differ from a single growth
# track (opal's PLA) due to Jensen's inequality: the non-linear length-weight
# (w = a*L^b) and logistic maturity curves mean the platoon-weighted average
# is not equal to the value at mean length. To verify that opal's population
# dynamics are correct, we override these two vectors with platoon-averaged
# values extracted from the SS3 Biology_at_age report table.

platoon_frac <- c(0.15, 0.70, 0.15)
frac_female  <- 1

# Wt_Beg by platoon (ages 0-43, from SS3 Biology_at_age_in_endyr)
wt_p1 <- c(0.00297429, 0.09807, 0.356455, 0.737394, 1.17746, 1.62494, 2.04639,
            2.42401, 2.75098, 3.02731, 3.25674, 3.44476, 3.59732, 3.72019,
            3.81859, 3.89703, 3.95937, 4.00876, 4.04783, 4.07867, 4.103,
            4.12216, 4.13725, 4.14912, 4.15845, 4.16579, 4.17155, 4.17608,
            4.17964, 4.18244, 4.18463, 4.18636, 4.18771, 4.18878, 4.18961,
            4.19027, 4.19078, 4.19118, 4.1915, 4.19175, 4.19195, 4.1921,
            4.19222, 4.19235)
wt_p2 <- c(0.00378997, 0.121547, 0.441813, 0.913989, 1.45946, 2.01411, 2.5365,
            3.00456, 3.40985, 3.75236, 4.03675, 4.2698, 4.4589, 4.6112,
            4.73316, 4.8304, 4.90766, 4.96889, 5.01731, 5.05555, 5.0857,
            5.10945, 5.12815, 5.14286, 5.15443, 5.16353, 5.17067, 5.17629,
            5.1807, 5.18417, 5.18689, 5.18902, 5.1907, 5.19202, 5.19305,
            5.19387, 5.1945, 5.19501, 5.1954, 5.19571, 5.19595, 5.19614,
            5.19629, 5.19645)
wt_p3 <- c(0.00463342, 0.148504, 0.539825, 1.11676, 1.78326, 2.46098, 3.09928,
            3.67119, 4.16641, 4.58492, 4.9324, 5.21716, 5.44822, 5.63431,
            5.78334, 5.90215, 5.99655, 6.07137, 6.13053, 6.17725, 6.21409,
            6.24311, 6.26596, 6.28394, 6.29807, 6.30919, 6.31792, 6.32478,
            6.33017, 6.33441, 6.33773, 6.34034, 6.34239, 6.344, 6.34527,
            6.34626, 6.34704, 6.34765, 6.34813, 6.34851, 6.3488, 6.34904,
            6.34922, 6.34942)

# Mat*Fecund by platoon (ages 0-43, from SS3 Biology_at_age_in_endyr)
mf_p1 <- c(0, 0, 0, 0.00337726, 0.730569, 1.5984, 2.04555, 2.42397, 2.75098,
            3.02731, 3.25674, 3.44476, 3.59732, 3.72019, 3.81859, 3.89703,
            3.95937, 4.00876, 4.04783, 4.07867, 4.103, 4.12216, 4.13725,
            4.14912, 4.15845, 4.16579, 4.17155, 4.17608, 4.17964, 4.18244,
            4.18463, 4.18636, 4.18771, 4.18878, 4.18961, 4.19027, 4.19078,
            4.19118, 4.1915, 4.19175, 4.19195, 4.1921, 4.19222, 4.19235)
mf_p2 <- c(0, 0, 0, 0.0840252, 1.38406, 2.01357, 2.5365, 3.00456, 3.40985,
            3.75236, 4.03675, 4.2698, 4.4589, 4.6112, 4.73316, 4.8304,
            4.90766, 4.96889, 5.01731, 5.05555, 5.0857, 5.10945, 5.12815,
            5.14286, 5.15443, 5.16353, 5.17067, 5.17629, 5.1807, 5.18417,
            5.18689, 5.18902, 5.1907, 5.19202, 5.19305, 5.19387, 5.1945,
            5.19501, 5.1954, 5.19571, 5.19595, 5.19614, 5.19629, 5.19645)
mf_p3 <- c(0, 0, 0.000000273, 0.542148, 1.78033, 2.46098, 3.09928, 3.67119,
            4.16641, 4.58492, 4.9324, 5.21716, 5.44822, 5.63431, 5.78334,
            5.90215, 5.99655, 6.07137, 6.13053, 6.17725, 6.21409, 6.24311,
            6.26596, 6.28394, 6.29807, 6.30919, 6.31792, 6.32478, 6.33017,
            6.33441, 6.33773, 6.34034, 6.34239, 6.344, 6.34527, 6.34626,
            6.34704, 6.34765, 6.34813, 6.34851, 6.3488, 6.34904, 6.34922,
            6.34942)

# Platoon-weighted weight-at-age
data$weight <- platoon_frac[1] * wt_p1 +
               platoon_frac[2] * wt_p2 +
               platoon_frac[3] * wt_p3

# Platoon-weighted spawning potential-at-age (scaled by fraction female)
data$spawning_potential <- frac_female * (platoon_frac[1] * mf_p1 +
                                          platoon_frac[2] * mf_p2 +
                                          platoon_frac[3] * mf_p3)

# ---- Rebuild and refit with platoon-corrected biology ----
obj2 <- MakeADFun(func = cmb(opal_model, data), parameters = parameters, map = map)

opt2 <- nlminb(start = obj2$par, objective = obj2$fn, gradient = obj2$gr,
               hessian = obj2$he, lower = Lwr, upper = Upr, control = control)
opt2 <- nlminb(start = opt2$par, objective = obj2$fn, gradient = obj2$gr,
               hessian = obj2$he, lower = Lwr, upper = Upr, control = control)
opt2 <- nlminb(start = opt2$par, objective = obj2$fn, gradient = obj2$gr,
               hessian = obj2$he, lower = Lwr, upper = Upr, control = control)

cat("Convergence:", opt2$convergence, "\n")
cat("Max gradient:", max(abs(obj2$gr())), "\n")

Report2 <- RTMB::sdreport(obj2)
rep2 <- obj2$report()

# ---- Extract platoon-corrected spawning biomass ----
spawning_biomass_corrected <- as.data.frame(
  summary(Report2)[which(rownames(summary(Report2)) == "spawning_biomass_y"),
                   c("Estimate", "Std. Error")])
colnames(spawning_biomass_corrected) <- c("estimate", "std_error")
spawning_biomass_corrected <- spawning_biomass_corrected[1:data$n_year, ]
spawning_biomass_corrected$year  <- calendar_years
spawning_biomass_corrected$lower <- spawning_biomass_corrected$estimate - 1.96 * spawning_biomass_corrected$std_error
spawning_biomass_corrected$upper <- spawning_biomass_corrected$estimate + 1.96 * spawning_biomass_corrected$std_error
spawning_biomass_corrected$model <- "opal (platoon-corrected)"

# ---- Comparison plot ----
data(opaka_truth)

opal_df <- rbind(spawning_biomass_default, spawning_biomass_corrected)

ybmax <- max(opal_df$upper,
             opaka_truth$om$sprseries$SSB,
             opaka_truth$em$sprseries$SSB, na.rm = TRUE)

ggplot(opal_df, aes(x = year)) +
  geom_ribbon(aes(ymin = lower, ymax = upper, fill = model), alpha = 0.2) +
  geom_line(aes(y = estimate, colour = model), linewidth = 1) +
  geom_line(data = subset(opaka_truth$om$sprseries, Yr %in% calendar_years),
            aes(x = Yr, y = SSB), colour = "black", linewidth = 1) +
  geom_line(data = subset(opaka_truth$em$sprseries, Yr %in% calendar_years),
            aes(x = Yr, y = SSB), colour = "red", linewidth = 1) +
  scale_colour_manual(values = c("opal (PLA)" = "steelblue",
                                 "opal (platoon-corrected)" = "darkorange")) +
  scale_fill_manual(values = c("opal (PLA)" = "steelblue",
                               "opal (platoon-corrected)" = "darkorange")) +
  labs(x = "Year", y = "Spawning biomass",
       title = "Estimated spawning biomass trajectory",
       subtitle = "Black = SS3 OM truth, Red = SS3 EM, Coloured = opal estimates",
       colour = NULL, fill = NULL) +
  scale_y_continuous(limits = c(0, ybmax * 1.2)) +
  theme_minimal() +
  theme(legend.position = "bottom")
```

## Dynamics diagnostic

The definitive test of opal's population dynamics engine is a forward
projection with all parameters fixed at SS3's MLE values — no estimation
involved. If the resulting spawning biomass trajectory matches SS3, any
remaining gap in the freely estimated model is attributable to differences
in the optimization surface rather than errors in the dynamics code.

This diagnostic is run twice: once using opal's default PLA-based
selectivity-at-age, and once with platoon-weighted selectivity extracted
from SS3's Asel2 report table. The comparison isolates the contribution of
selectivity platoon effects to the dynamics-level discrepancy.

Three sources of structural difference between opal and SS3 are addressed
by these overrides:

1. **Weight-at-age**: SS3 computes weight separately for each platoon
   before averaging; opal integrates weight over a continuous length
   distribution via the PLA.
2. **Spawning potential-at-age**: Same platoon-averaging effect, compounded
   by the logistic maturity curve.
3. **Selectivity-at-age**: Length-based selectivity curves interact with
   platoon-specific growth, producing different effective selectivity at
   young ages where the platoon spread is widest.

With all three corrections applied and parameters fixed at SS3 MLEs, opal's
trajectory closely tracks the SS3 EM, confirming that the core population
dynamics (survivorship, Beverton-Holt recruitment, harvest rate calculation,
bias adjustment) are implemented correctly.

```{r dynamics-check, echo=TRUE, message=FALSE, fig.height=5, fig.width=8}
# ---- Dynamics diagnostic: fix all parameters at SS3 MLE values ----
# This isolates opal's population dynamics from estimation differences.
# If the forward projection matches SS3, the dynamics are correct and any
# gap in the estimated model is due to parameter estimation (different
# likelihood surface, selectivity parameterization, etc.). If the projection
# doesn't match, a structural issue remains.

# SS3 MLE values from ss.par and derived quantities
# SS3's SSB_Virgin = 4639.1, ln(R0) = 5.65608, h = 0.76, sigma_r = 0.52
# SS3's init_F[1] = 0.0119122, Q_parm = c(-3.773, -6.253)

# Build parameter list with all values at SS3 MLEs
parameters_ss3 <- parameters  # start from the current parameter list

# B0: SS3's SSB_Virgin = 4639.1
parameters_ss3$log_B0 <- log(4639.1)

# Steepness, sigma_r (already set from opaka_parameters, but be explicit)
parameters_ss3$log_h <- log(0.76)
parameters_ss3$log_sigma_r <- log(0.52)

# CPUE catchability (from ss.par Q_parm)
parameters_ss3$log_cpue_q <- c(-3.77275156733, -6.25323305447)

# Recruitment deviations: use SS3's full recdev vector
ss3_recdev2 <- c(0.0675235702971, 0.0764991539955, 0.0858442617148,
  0.0950046747340, 0.102706450120, 0.107687329214, 0.107711705598,
  0.0998349181169, 0.0808136660096, 0.0482338925745, 0.00234443377404,
  -0.0517563213031, -0.102395434864, -0.131458254996, -0.121100670313,
  -0.0714825835398, -0.00742166036481, -0.0109308396989, -0.0385028902033,
  0.0366483160551, 0.0165104920249, -0.166905707226,
  0.684759008019, -0.00129318028442, 0.291954545496, -0.502105154386,
  -0.395132761139, 0.289343271935, -0.288480831619, -0.490309035200,
  -0.212366196568, 0.161892641109, -0.0209673906075, -0.297262261087,
  0.329268521456, 0.00878056067047, -0.0597345841788, -0.635158118155,
  0.121730487831, 0.564895487490, -0.412552483484, 0.492195882830,
  0.863050417304, -0.262592196252, -0.580724697255, 0.192426722416,
  0.470346549222, -0.450985027130, 0.468806143351, -0.0970889237156,
  0.125657219706, -0.590564615592, -0.206902651435, 0.212391740136,
  -0.786327541142, -0.942202604456, -0.479537926647, -0.910561752327,
  -0.338249154939, -1.08143122504, 0.610318451176, -0.116085865688,
  -1.00403426050, -0.0967824207922, 0.241495871563, 0.381401207516,
  -0.162517483700, 0.164175327286, 0.739150845802, -0.283843490207,
  -0.324836661861, -0.297336491801, -0.566398458638, -0.553810264824,
  -0.00857108423637, -0.0751868745470, 0.339022613392, 0.0187206379281,
  0.633372740779, 0.675976465171, 0.134831485765, 0.429015103446,
  0.00483559599714, -0.953917944860, -0.848588067707, -0.0919670526641,
  -0.820322333684, -0.134672436697, 0.936671897697, 0.135185903985,
  -0.301177900173, -0.988301462145, -0.138899249513, 0.785829523698,
  -0.674385690767, 0.751722180103, 0.352211358296, 0.853371113276,
  -0.348627551973, -0.514856812016, -0.157533901109, -0.175313077136,
  0.212151760383, 0.110769041131, 0.126998013037, -0.349623955515,
  0.287222081367, 0.166718339864, 0.263616100771, -0.514692520633,
  0.543032699532, 0.148610562305, 0.0741026805708, -0.0379144746966,
  0.0478985480312, -0.0919462212626, -0.825992698686, 0.225426831222,
  -0.711467818125, 0.447979718775, 0.00000000000)

ss3_early <- c(0.0206807064789, 0.00321579077358, 0.00372620421544,
  0.00431740128909, 0.00500263165771, 0.00579673718936, 0.00671727050323,
  0.00778421989329, 0.00902081267571, 0.0104539317083, 0.0121143194944,
  0.0140374805236, 0.0162641224184, 0.0188404004022, 0.0218181611323,
  0.0252560947069, 0.0292181862911, 0.0337726098165, 0.0389897504158,
  0.0449362496360, 0.0516661790666, 0.0592048886170)

parameters_ss3$init_rdev_a <- rev(c(ss3_early, ss3_recdev2[1:22]))
parameters_ss3$init_rdev_a[1] <- ss3_recdev2[22]  # 1949 dev at age 0

# Time-series rdevs: recdev2[23:end] are 1950 onwards, plus 0.0 for terminal
parameters_ss3$rdev_y <- c(ss3_recdev2[23:length(ss3_recdev2)], 0.0)

# Selectivity: use the converted SS3 values (already in parameters$par_sel)
parameters_ss3$par_sel <- parameters$par_sel

# Initial F
parameters_ss3$log_init_F_f <- c(log(0.0119122), rep(log(1e-8), data$n_fishery - 1))

# ---- Platoon-weighted selectivity-at-age override ----
# SS3's Asel2 report gives the realised selectivity-at-age for each fleet and
# platoon (morph), already accounting for the interaction between length-based
# selectivity and platoon-specific growth. The platoon-weighted selectivity is:
#   sel_f(a) = sum_p [ frac_p * sel_p(a) ]
# This bypasses the PLA-based length-to-age conversion, which uses a single
# continuous normal distribution rather than SS3's discrete platoon structure.

platoon_frac <- c(0.15, 0.70, 0.15)

# Fleet 1 (Comm) - logistic selectivity, Asel2 by platoon (ages 0-43)
sel1_p1 <- c(0, 0.000117529, 0.0634736, 0.757875, 0.984164, 0.998816,
             0.999854, 0.999972, 0.999992, 0.999997, 0.999999, 0.999999,
             1, 1, 1, 1, 1, 1, 1, 1, 1, 1, 1, 1, 1, 1, 1, 1, 1, 1,
             1, 1, 1, 1, 1, 1, 1, 1, 1, 1, 1, 1, 1, 1)
sel1_p2 <- c(0, 0.000406706, 0.224962, 0.942633, 0.998291, 0.999908,
             0.999991, 0.999998, 1, 1, 1, 1, 1, 1, 1, 1, 1, 1, 1, 1,
             1, 1, 1, 1, 1, 1, 1, 1, 1, 1, 1, 1, 1, 1, 1, 1, 1, 1,
             1, 1, 1, 1, 1, 1)
sel1_p3 <- c(0, 0.00140518, 0.525563, 0.990882, 0.999834, 0.999993,
             0.999999, 1, 1, 1, 1, 1, 1, 1, 1, 1, 1, 1, 1, 1, 1, 1,
             1, 1, 1, 1, 1, 1, 1, 1, 1, 1, 1, 1, 1, 1, 1, 1, 1, 1,
             1, 1, 1, 1)

# Fleet 2 (Non_comm) - wider logistic selectivity, Asel2 by platoon
sel2_p1 <- c(0, 0.011022, 0.101411, 0.393307, 0.7059, 0.869541, 0.937458,
             0.966084, 0.979288, 0.986007, 0.989736, 0.99196, 0.993365,
             0.994295, 0.994934, 0.995385, 0.995711, 0.99595, 0.996129,
             0.996264, 0.996367, 0.996445, 0.996506, 0.996552, 0.996589,
             0.996617, 0.996639, 0.996656, 0.996669, 0.99668, 0.996688,
             0.996694, 0.996699, 0.996703, 0.996707, 0.996709, 0.996711,
             0.996712, 0.996714, 0.996714, 0.996715, 0.996716, 0.996716,
             0.996717)
sel2_p2 <- c(0, 0.0173491, 0.173036, 0.570606, 0.845014, 0.943516,
             0.97597, 0.987934, 0.993013, 0.995458, 0.996764, 0.99752,
             0.997988, 0.998293, 0.998499, 0.998643, 0.998747, 0.998823,
             0.998879, 0.998921, 0.998953, 0.998978, 0.998997, 0.999011,
             0.999022, 0.999031, 0.999038, 0.999043, 0.999048, 0.999051,
             0.999053, 0.999055, 0.999057, 0.999058, 0.999059, 0.99906,
             0.99906, 0.999061, 0.999061, 0.999061, 0.999062, 0.999062,
             0.999062, 0.999062)
sel2_p3 <- c(0, 0.0271976, 0.278252, 0.732191, 0.926606, 0.977093,
             0.991117, 0.99581, 0.997681, 0.998544, 0.998989, 0.999241,
             0.999394, 0.999492, 0.999558, 0.999603, 0.999636, 0.999659,
             0.999677, 0.99969, 0.9997, 0.999707, 0.999713, 0.999717,
             0.999721, 0.999724, 0.999726, 0.999727, 0.999729, 0.99973,
             0.99973, 0.999731, 0.999731, 0.999732, 0.999732, 0.999732,
             0.999733, 0.999733, 0.999733, 0.999733, 0.999733, 0.999733,
             0.999733, 0.999733)

# Fleet 3 (ResFish) - domed double-normal selectivity, Asel2 by platoon
sel3_p1 <- c(0, 0.995148, 1, 0.995113, 0.825043, 0.530483, 0.348864,
             0.266527, 0.231338, 0.215751, 0.208344, 0.20454, 0.202437,
             0.201195, 0.200421, 0.199916, 0.199573, 0.199333, 0.199162,
             0.199037, 0.198944, 0.198875, 0.198822, 0.198782, 0.198751,
             0.198727, 0.198709, 0.198694, 0.198683, 0.198674, 0.198668,
             0.198662, 0.198658, 0.198655, 0.198652, 0.19865, 0.198649,
             0.198648, 0.198647, 0.198646, 0.198645, 0.198645, 0.198644,
             0.198644)
sel3_p2 <- c(0, 0.999966, 0.999999, 0.948685, 0.591236, 0.331622,
             0.239112, 0.211244, 0.202601, 0.199629, 0.198474, 0.197966,
             0.197719, 0.197586, 0.19751, 0.197463, 0.197432, 0.197412,
             0.197397, 0.197387, 0.19738, 0.197374, 0.19737, 0.197367,
             0.197365, 0.197363, 0.197362, 0.197361, 0.19736, 0.197359,
             0.197359, 0.197358, 0.197358, 0.197358, 0.197357, 0.197357,
             0.197357, 0.197357, 0.197357, 0.197357, 0.197357, 0.197357,
             0.197357, 0.197357)
sel3_p3 <- c(0, 1, 0.999903, 0.800767, 0.381609, 0.235649, 0.205139,
             0.199136, 0.197813, 0.197466, 0.197357, 0.197317, 0.1973,
             0.197292, 0.197287, 0.197285, 0.197283, 0.197282, 0.197281,
             0.197281, 0.197281, 0.19728, 0.19728, 0.19728, 0.19728,
             0.19728, 0.19728, 0.19728, 0.19728, 0.19728, 0.19728,
             0.19728, 0.19728, 0.19728, 0.19728, 0.19728, 0.19728,
             0.19728, 0.19728, 0.19728, 0.19728, 0.19728, 0.19728,
             0.19728)

# Platoon-weighted selectivity-at-age by fleet
sel_fa <- matrix(0, nrow = 3, ncol = 44)
sel_fa[1, ] <- platoon_frac[1] * sel1_p1 + platoon_frac[2] * sel1_p2 + platoon_frac[3] * sel1_p3
sel_fa[2, ] <- platoon_frac[1] * sel2_p1 + platoon_frac[2] * sel2_p2 + platoon_frac[3] * sel2_p3
sel_fa[3, ] <- platoon_frac[1] * sel3_p1 + platoon_frac[2] * sel3_p2 + platoon_frac[3] * sel3_p3

data$sel_fa_external <- sel_fa

# ---- Fix ALL parameters (no estimation) ----
map_fixed <- list(
  log_B0       = factor(NA),
  log_h        = factor(NA),
  log_sigma_r  = factor(NA),
  log_cpue_q   = factor(rep(NA, length(parameters_ss3$log_cpue_q))),
  cpue_creep   = factor(rep(NA, data$n_index)),
  log_cpue_tau = factor(rep(NA, data$n_index)),
  log_cpue_omega = factor(rep(NA, data$n_index)),
  log_lf_tau   = factor(rep(NA, length(parameters_ss3$log_lf_tau))),
  log_L1       = factor(NA),
  log_L2       = factor(NA),
  log_k        = factor(NA),
  log_CV1      = factor(NA),
  log_CV2      = factor(NA),
  par_sel      = factor(rep(NA, length(parameters_ss3$par_sel))),
  log_init_F_f = factor(rep(NA, data$n_fishery)),
  rdev_y       = factor(rep(NA, length(parameters_ss3$rdev_y))),
  init_rdev_a  = factor(rep(NA, data$n_age))
)

# ---- Build and evaluate: without selectivity override ----
data_no_sel <- data
data_no_sel$sel_fa_external <- NULL
obj_check_nosel <- MakeADFun(func = cmb(opal_model, data_no_sel),
                             parameters = parameters_ss3, map = map_fixed)
rep_check_nosel <- obj_check_nosel$report()

# ---- Build and evaluate: with selectivity override ----
obj_check <- MakeADFun(func = cmb(opal_model, data),
                       parameters = parameters_ss3, map = map_fixed)
rep_check <- obj_check$report()

cat("Without selectivity override:\n")
cat("  B0:", rep_check_nosel$B0, " R0:", rep_check_nosel$R0,
    " SPR0:", rep_check_nosel$B0 / rep_check_nosel$R0, "\n")
cat("With selectivity override:\n")
cat("  B0:", rep_check$B0, " R0:", rep_check$R0,
    " SPR0:", rep_check$B0 / rep_check$R0, "\n\n")

# Maximum absolute difference in spawning biomass
ssb_diff <- abs(rep_check$spawning_biomass_y[1:data$n_year] -
                rep_check_nosel$spawning_biomass_y[1:data$n_year])
cat("Max |SSB difference| between PLA and platoon selectivity:",
    round(max(ssb_diff), 1), "\n")

# ---- Comparison plot ----
data(opaka_truth)

ssb_nosel <- data.frame(
  year = calendar_years,
  estimate = rep_check_nosel$spawning_biomass_y[1:data$n_year],
  model = "opal (PLA selectivity)"
)
ssb_sel <- data.frame(
  year = calendar_years,
  estimate = rep_check$spawning_biomass_y[1:data$n_year],
  model = "opal (platoon selectivity)"
)
ssb_df <- rbind(ssb_nosel, ssb_sel)

ybmax <- max(ssb_df$estimate,
             opaka_truth$om$sprseries$SSB,
             opaka_truth$em$sprseries$SSB, na.rm = TRUE)

ggplot() +
  geom_line(data = ssb_df, aes(x = year, y = estimate, colour = model),
            linewidth = 1) +
  geom_line(data = subset(opaka_truth$om$sprseries, Yr %in% calendar_years),
            aes(x = Yr, y = SSB, colour = "SS3 OM"), linewidth = 1) +
  geom_line(data = subset(opaka_truth$em$sprseries, Yr %in% calendar_years),
            aes(x = Yr, y = SSB, colour = "SS3 EM"), linewidth = 1) +
  scale_colour_manual(values = c("opal (PLA selectivity)" = "steelblue",
                                 "opal (platoon selectivity)" = "darkorange",
                                 "SS3 OM" = "black",
                                 "SS3 EM" = "red")) +
  labs(x = "Year", y = "Spawning biomass",
       title = "Dynamics diagnostic: opal at SS3 MLE parameters",
       subtitle = "No estimation — all parameters fixed at SS3 values",
       colour = NULL) +
  scale_y_continuous(limits = c(0, ybmax * 1.2)) +
  theme_minimal() +
  theme(legend.position = "bottom")
```

```{r dynamics-estimate, echo=TRUE, message=FALSE, fig.height=5, fig.width=8}
# ---- Dynamics estimate: optimize B0, init / time rec devs, and init F ----
map_est <- map_fixed
map_est$log_B0 <- NULL
# map_est$rdev_y <- NULL
# map_est$init_rdev_a <- NULL
# map_est$log_init_F_f <- factor(c(1, rep(NA, data$n_fishery - 1)))

# Build object using opal PLA model (no external selectivity override)
obj_est <- MakeADFun(func = cmb(opal_model, data_no_sel),
                     parameters = parameters_ss3,
                     map = map_est)

# Parameter bounds (relevant parameters only)
Lwr <- rep(-Inf, length(obj_est$par))
Upr <- rep(Inf, length(obj_est$par))
Lwr[grep("log_B0", names(obj_est$par))] <- log(1)
Upr[grep("log_B0", names(obj_est$par))] <- log(exp(22))
Lwr[grep("rdev_y", names(obj_est$par))] <- rep(-5, length(grep("rdev_y", names(obj_est$par))))
Upr[grep("rdev_y", names(obj_est$par))] <- rep(5, length(grep("rdev_y", names(obj_est$par))))
Lwr[grep("init_rdev_a", names(obj_est$par))] <- rep(-5, data$n_age)
Upr[grep("init_rdev_a", names(obj_est$par))] <- rep(5, data$n_age)
Lwr[grep("log_init_F_f", names(obj_est$par))] <- log(1e-12)
Upr[grep("log_init_F_f", names(obj_est$par))] <- log(3)

# Optimize (2-stage run for stability)
control <- list(eval.max = 10000, iter.max = 10000)
opt_est <- nlminb(start = obj_est$par, objective = obj_est$fn,
                  gradient = obj_est$gr, lower = Lwr, upper = Upr,
                  control = control)
opt_est <- nlminb(start = opt_est$par, objective = obj_est$fn,
                  gradient = obj_est$gr, lower = Lwr, upper = Upr,
                  control = control)

cat("estimation convergence:", opt_est$convergence, "\n")
cat("max gradient:", max(abs(obj_est$gr())), "\n")

rep_est <- obj_est$report()

ssb_est <- data.frame(
  year = calendar_years,
  estimate = rep_est$spawning_biomass_y[1:data$n_year],
  model = "opal (PLA estimated)"
)

ssb_df_est <- rbind(ssb_df, ssb_est)

ybmax_est <- max(ssb_df_est$estimate,
                 opaka_truth$om$sprseries$SSB,
                 opaka_truth$em$sprseries$SSB,
                 na.rm = TRUE)

# Extended comparison plot
ggplot() +
  geom_line(data = ssb_df_est, aes(x = year, y = estimate, colour = model), linewidth = 1) +
  geom_line(data = subset(opaka_truth$om$sprseries, Yr %in% calendar_years),
            aes(x = Yr, y = SSB, colour = "SS3 OM"), linewidth = 1) +
  geom_line(data = subset(opaka_truth$em$sprseries, Yr %in% calendar_years),
            aes(x = Yr, y = SSB, colour = "SS3 EM"), linewidth = 1) +
  scale_colour_manual(values = c("opal (PLA selectivity)" = "steelblue",
                                 "opal (platoon selectivity)" = "darkorange",
                                 "opal (PLA estimated)" = "green4",
                                 "SS3 OM" = "black",
                                 "SS3 EM" = "red")) +
  labs(x = "Year", y = "Spawning biomass",
       title = "Dynamics diagnostic + estimation comparison",
       subtitle = "Fixed SS3 MLE (PLA/platoon) vs estimated PLA fit",
       colour = NULL) +
  scale_y_continuous(limits = c(0, ybmax_est * 1.2)) +
  theme_minimal() +
  theme(legend.position = "bottom")
```

