---
title: "Review an Opakapaka Fit"
format: html
execute:
  eval: true
  warning: false
  message: false
vignette: >
  %\VignetteIndexEntry{Review an Opakapaka Fit}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

```{r}
#| include: false
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

# Review a saved assessment

This vignette upgrades the bundled legacy `opal_fit` to `opal_obj`. It
examines the portable fixed-effects opakapaka fit created for the
[Quickstart](quickstart.html). It rebuilds the transient RTMB objective to
verify the stored objective and obtain predictions, without optimisation.

The bundled fit may have been saved under a different R version. Portable
integrity permits that difference while still checking model compatibility and
requiring the rebuilt objective to match the saved value.

```{r}
#| label: load-fit
library(opal)

fit_path <- system.file("extdata", "opaka_quickstart_fit.rds", package = "opal")
if (!nzchar(fit_path)) {
  stop("The bundled opakapaka quickstart fit is unavailable.", call. = FALSE)
}
fit <- opal_read(fit_path)
fit
summary(fit)
```

## CPUE fit

```{r}
#| label: cpue-fit
#| fig-width: 8
#| fig-height: 5
plot_cpue(fit)
```

## Diagnostics

The compact estimability result and maximum absolute gradient were saved with
the fit. They can be inspected without creating a new objective.

```{r}
#| label: diagnostics
list(
  estimability = fit$fit$estimability,
  maximum_gradient = fit$fit$diagnostics$max_gradient
)
```

Refresh the reusable fit checks with `opal_check()`. The stage remains
`fitted` whether checks pass or fail; inspect the validation record before
interpreting the assessment.

```{r}
#| label: refresh-checks
fit <- opal_check(fit, scope = "fit")
fit$validation$fit$metrics
summary(fit)
```

The bundled legacy fit's maximum gradient can exceed the default `0.001`
tolerance even with optimiser convergence code zero. That is a failed
diagnostic check, not a read or migration failure. This review preserves the
saved point; the [Quickstart](quickstart.html) and
[projection guide](projections.html) demonstrate bounded continuation from
that point before using a fit that passes the checks.

## Continue and save the assessment

Metadata changes preserve the fitted result. Configuration changes through
`opal_update()` clear dependent results so that another fit or sampler run
cannot accidentally use stale outputs.

```{r}
#| label: save-reviewed
fit <- opal_update(fit, metadata = list(label = "Reviewed baseline"))
path <- tempfile(fileext = ".rds")
opal_save(fit, path)
restored <- opal_read(path, rebuild = TRUE)
stopifnot(isTRUE(all.equal(opal_report(fit), opal_report(restored))))
unlink(path)
```

See the [projection guide](projections.html) to carry this object into a
future-catch scenario, or `opal_mcmc()` to add posterior sampling.

## OSA residuals and further analyses

Use `fit <- opal_osa(fit)` followed by `plot_osa_sdnr(fit)` to compare observation
residual dispersion across datasets. The [assessment-tools guide](assessment-tools.html)
shows this figure, composition fits, posterior analysis, and sensitivity workflows.
