---
title: "Review an Opakapaka Fit"
format: html
execute:
  eval: true
  warning: false
  message: false
vignette: >
  %\VignetteIndexEntry{Review an Opakapaka Fit}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

```{r}
#| include: false
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

# Review an `opal_fit`

This vignette examines the portable fixed-effects opakapaka fit created for
the Quickstart. It rebuilds the transient RTMB objective only to obtain model
predictions, so no optimization is performed here.

```{r}
#| label: load-fit
library(opal)

fit_path <- system.file("extdata", "opaka_quickstart_fit.rds", package = "opal")
if (!nzchar(fit_path)) {
  stop("The bundled opakapaka quickstart fit is unavailable.", call. = FALSE)
}
fit <- read_opal_fit(fit_path, strict = TRUE)
fit
summary(fit)
```

## CPUE fit

```{r}
#| label: cpue-fit
#| fig-width: 8
#| fig-height: 5
object <- opal_fit_object(fit)
plot_data <- fit$data
plot_data$cpue_data$season <- 1L
plot_cpue(plot_data, object)
```

## Diagnostics

The compact estimability result and maximum absolute gradient were saved with
the fit. They can be inspected without creating a new objective.

```{r}
#| label: diagnostics
list(
  estimability = fit$fit$estimability,
  maximum_gradient = fit$fit$diagnostics$max_gradient
)
```