---
title: "The 'Opakapaka model"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{The 'Opakapaka model}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
---

```{r, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

# Introduction

`opal` is an open, modular R package for fisheries stock assessment built on
[RTMB](https://github.com/kaskr/RTMB) for automatic differentiation. This
vignette demonstrates model setup, fitting, and diagnostics using bundled data
for ['opakapaka](https://github.com/noaa-pifsc/Opaka_simulation), extracted from a Stock Synthesis 3 (SS3) model and converted to
opal conventions. The model covers the historical period 1949–2023 with three
fleets: a commercial fishery (Comm), a non-commercial fishery (Non_comm), and a
research fishery survey (ResFish).

# Load inputs

Load the `opal` package and the `RTMB` dependency. The `tidyverse` package is
used for data manipulation and plotting.

```{r load-pkg, echo=TRUE, message=FALSE}
library(tidyverse)
library(opal)
library(RTMB)

theme_set(theme_bw())
```

The bundled data object `opaka_data` contains all biological parameters, catch
and CPUE observations, length structure, and prior specifications needed for the
opakapaka assessment model:

```{r load-data, echo=TRUE, message=FALSE}
data(opaka_data)
data <- opaka_data
if (is.null(data$n_index)) data$n_index <- 2L
names(data)

# set ages to anchor growth curve (since SS3 indexs from 0 to 43, but opal uses 1-based indexing)
data$A1 <- 0
data$A2 <- data$A1 + (data$n_age-1)
data$min_age <- 0
ages <- data$age_a <- seq(data$min_age, by = 1, length.out = data$n_age)
```

Key dimensions of the data:

```{r data-dims, echo=TRUE}
cat("Number of ages:", data$n_age, "\n")
cat("Number of years:", data$n_year, "\n")
cat("Number of fisheries:", data$n_fishery, "\n")
cat("Number of length bins:", data$n_len, "\n")
```

## Biological inputs

Natural mortality, maturity at length, and length-weight parameters are all
contained in the data object. Unlike the BET case study where M varies with
age, opakapaka uses a constant M = 0.135 across all ages.

```{r plot-biology, echo=TRUE, message=FALSE, fig.height=4, fig.width=7}
# Natural mortality
ggplot(data.frame(age = ages, M = data$M), aes(x = age, y = M)) +
  geom_line() + geom_point(size = 0.8) +
  labs(x = "Age (opal internal)", y = "M (annual rate)",
       title = "Natural Mortality-at-Age")

# Maturity at length
len_mid <- seq(data$len_bin_start + data$len_bin_width / 2,
               by = data$len_bin_width, length.out = data$n_len)
ggplot(data.frame(length = len_mid, maturity = data$maturity),
       aes(x = length, y = maturity)) +
  geom_line() + geom_point() +
  labs(x = "Length (cm)", y = "Maturity",
       title = "Maturity-at-Length (logistic, Mat50 = 40.7 cm)")

# define sex ratio vector
data$sex_ratio <- rep(1, data$n_age)
```

## Length composition data

Length composition observations are loaded from the bundled `opaka_lf`
long-format data object and transformed into model-ready arrays. Data are
available for two fleets: Comm (fleet 1, 1949–2023) and ResFish (fleet 3,
2017–2023). Fleet 2 (Non_comm) has no length composition data.

```{r prep-lf, echo=TRUE, message=FALSE}
data(opaka_lf)

# Pivot to wide format: one row per fishery x timestep, bins as columns
lf_wide <- opaka_lf %>%
  tidyr::pivot_wider(
    id_cols = c(fishery, year, month, ts),
    names_from = bin,
    values_from = value,
    values_fill = 0
  ) %>%
  arrange(fishery, ts)

# Define variance adjustment scalars for each fishery
var_adjust_scalars <- 1 / rep(1, data$n_fishery)

data <- prep_lf_data(data, lf_wide,
                     lf_keep_fisheries = c(1, 3),
                     lf_var_adjust = var_adjust_scalars)
```

## Set up SS3-style bias ramps

```{r add-bias-ramp, echo=TRUE, message=FALSE}
# 1. INITIAL COHORTS (1906 to 1949)
# Data ages are 1-based, so 1949 - (age - 1) gives the birth year.
# Age_a = 1 (bio age 0) -> Born in 1949.
# Age_a = 44 (bio age 43) -> Born in 1906.
init_cohort_years <- 1949 - (data$age_a - 1)

data$init_bias_adj_a <- get_bias_adj_vector(
  years = init_cohort_years,
  do_rec_bias_ramp = 1,
  bias_years = c(1931, 1974, 2021.88, 2022.43),
  max_bias_adj = 0.869
)

# 2. TIME SERIES COHORTS (1950 to 2024)
# Because rdev_y[1] generates recruits for year 2 (1950),
# the time-series bias ramp must evaluate 1950 to 2024.
calendar_years <- (1949-1) + data$years

data$bias_adj_y <- get_bias_adj_vector(
  years = calendar_years,
  do_rec_bias_ramp = 1,
  bias_years = c(1931, 1974, 2021.88, 2022.43),
  max_bias_adj = 0.869
)
```

# Model setup

## Parameters

Define the initial parameter values. Growth parameters (`log_L1`, `log_L2`,
`log_k`, `log_CV1`, `log_CV2`) and selectivity (`par_sel`) are initialised
from the SS3 estimates. Load these from the bundled `opaka_parameters` data
object:

```{r get-pars, echo=TRUE, message=FALSE}
data(opaka_parameters)

# Route 1949 dev to biological age 0, which is index 1 in opal's 1-based age vector.
# In the SS3 vector, rdev_y[1] is the 1949 deviation; here it seeds the initial
# age-0 cohort, while later years are shifted into the modeled rdev_y time series.
init_rdev_a <- rep(0.0, data$n_age)
init_rdev_a = rev(c(0.0206807,0.00321579,0.00372620,0.0043174,0.00500263,0.00579674,0.00671727,0.00778422,0.00902081,0.0104539,0.0121143,0.0140375,0.0162641,0.0188404,0.0218182,0.0252561,0.0292182,0.0337726,0.0389898,0.0449362,0.0516662,0.0592049,0.0675236,0.0764992,0.0858443,0.0950047,0.102706,0.107687,0.107712,0.0998349,0.0808137,0.0482339,0.00234443,-0.0517563,-0.102395,-0.131458,-0.121101,-0.0714826,-0.00742166,-0.0109308,-0.0385029,0.0366483,0.0165105,-0.166906))
# Remaining initial ages start at equilibrium and are estimated via init_rdev_a.
init_rdev_a[1] <- opaka_parameters$rdev_y[1]

# Shift 1950-2023 into rdev_y and add 0.0 for terminal/forecast year 2024
rdev_y_shifted <- c(opaka_parameters$rdev_y[2:length(opaka_parameters$rdev_y)], 0.0)

parameters <- list(
  log_B0 = 9,
  log_h = as.numeric(opaka_parameters$log_h),
  log_sigma_r = as.numeric(opaka_parameters$log_sigma_r),
  log_cpue_q = as.numeric(opaka_parameters$log_cpue_q),
  cpue_creep = as.numeric(rep(opaka_parameters$cpue_creep, data$n_index)),
  log_cpue_tau = rep(log(0.1),data$n_index),
  log_cpue_omega = as.numeric(rep(opaka_parameters$log_cpue_omega, data$n_index)),
  log_lf_tau = as.numeric(log(rep(0.1, data$n_fishery))),
  log_L1 = as.numeric(opaka_parameters$log_L1),
  log_L2 = as.numeric(opaka_parameters$log_L2),
  log_k = as.numeric(opaka_parameters$log_k),
  log_CV1 = as.numeric(opaka_parameters$log_CV1),
  log_CV2 = as.numeric(opaka_parameters$log_CV2),
  par_sel = as.matrix(opaka_parameters$par_sel),
  log_init_F_f = c(log(0.0119122), rep(log(1e-8), data$n_fishery - 1)),
  rdev_y = rdev_y_shifted,
  init_rdev_a = init_rdev_a
)

# convert ss3 parameters to opal parameters (selectivity)
parameters$par_sel = convert_ss3_selex_to_rtmb(parameters$par_sel, data$sel_type_f, data$len_mid)
```

## Priors

Priors are specified using `get_priors()`. The data object already contains
prior center values for growth parameters:

```{r get-priors, echo=TRUE, message=FALSE}
data$priors <- get_priors(parameters = parameters, data = data)
evaluate_priors(parameters = parameters, priors = data$priors)
```

## Parameter map

Use RTMB's `map` option to turn parameters on/off. Parameters mapped to
`factor(NA)` are fixed at their initial values. For the initial run, we fix
most parameters and estimate only `log_B0`, `log_cpue_q`, `rdev_y`, and `init_rdev_a`:

```{r get-map, echo=TRUE, message=FALSE}
map_sel <- matrix(NA, nrow(parameters$par_sel), ncol(parameters$par_sel))
map_sel[1,1:2 ] <- c(1,2) # Comm fleet selectivity parameters
map_sel[3,c(1,4,6) ] <- c(3,4,5) # Comm fleet selectivity parameters
map_sel = as.vector(map_sel)

map_lf_tau <- rep(NA, length(parameters$log_lf_tau))

map <- list(
  # log_B0 = factor(NA),
  log_h = factor(NA),
  log_sigma_r = factor(NA),
  # log_cpue_q = factor(NA),
  cpue_creep = as.factor(rep(NA, data$n_index)),
  log_cpue_tau = as.factor(rep(NA, data$n_index)),
  log_cpue_omega = as.factor(rep(NA, data$n_index)),
  log_lf_tau = factor(map_lf_tau),
  log_L1  = factor(NA),
  log_L2  = factor(NA),
  log_k   = factor(NA),
  log_CV1 = factor(NA),
  log_CV2 = factor(NA),
  par_sel = factor(map_sel),
  # log_init_F_f = as.factor(c(1, rep(NA, data$n_fishery - 1))),
  log_init_F_f = rep(factor(NA), data$n_fishery),
  # init_rdev_a = rep(factor(NA), data$n_age)
  init_rdev_a = as.factor(seq_len(data$n_age))
  # rdev_y = as.factor(rep(NA, length(parameters$rdev_y)))
)
```

## Build the AD object

Using the `data`, the `parameters`, the parameter `map`, and the model
(`opal_model`), the AD object is created using RTMB's `MakeADFun` function:

```{r make-adfun, echo=TRUE, message=FALSE}
# data$lf_switch <- 0 # skip length comps initially (removal only)
data$lf_switch <- 1L # multinomial likelihood
# data$cpue_switch <- 0L
# Can set random = "rdev_y" to treat recruitment deviations as random effects, but start with fixed effects for simplicity
obj <- MakeADFun(func = cmb(opal_model, data), parameters = parameters, map = map)
unique(names(obj$par))
obj$fn()
obj$gr()
```

Inspect initial model outputs:

```{r init-checks, echo=TRUE, message=FALSE, fig.height=4, fig.width=7}
plot(obj$report()$spawning_biomass_y, type = "l",
     xlab = "Time step", ylab = "Spawning biomass",
     main = "Initial spawning biomass trajectory")
```

## Parameter bounds

```{r get-par-bounds, echo=TRUE, message=FALSE}
Lwr <- rep(-Inf, length(obj$par))
Upr <- rep(Inf, length(obj$par))
Lwr[grep("log_B0", names(obj$par))] <- log(1)
Upr[grep("log_B0", names(obj$par))] <- log(exp(22))
Lwr[grep("log_cpue_q", names(obj$par))] <- log(0.001)
Upr[grep("log_cpue_q", names(obj$par))] <- log(10)
Lwr[grep("rdev_y", names(obj$par))] <- rep(-5, length(grep("rdev_y", names(obj$par))))
Upr[grep("rdev_y", names(obj$par))] <- rep(5, length(grep("rdev_y", names(obj$par))))
Lwr[grep("init_rdev_a", names(obj$par))] <- rep(-5, data$n_age)
Upr[grep("init_rdev_a", names(obj$par))] <- rep(5, data$n_age)
bounds <- data.frame(par = names(obj$par), lower = Lwr, upper = Upr)
```

# Optimisation

Optimise using the `nlminb` function, run twice to ensure convergence:

```{r run-nlminb, echo=TRUE, results="hide"}
control <- list(eval.max = 10000, iter.max = 10000)
opt <- nlminb(start = obj$par, objective = obj$fn, gradient = obj$gr,
              hessian = obj$he, lower = Lwr, upper = Upr, control = control)
opt <- nlminb(start = opt$par, objective = obj$fn, gradient = obj$gr,
              hessian = obj$he, lower = Lwr, upper = Upr, control = control)
opt <- nlminb(start = opt$par, objective = obj$fn, gradient = obj$gr,
              hessian = obj$he, lower = Lwr, upper = Upr, control = control)
obj$fn()
max(abs(obj$gr()))
obj$report()$lp_prior
obj$report()$lp_penalty
obj$report()$lp_rec
sum(obj$report()$lp_cpue)
sum(obj$report()$lp_lf)
```

Compare initial and estimated parameter values:

```{r opt-results, echo=TRUE, message=FALSE}
get_par_table(obj, parameters, map, lower = Lwr, upper = Upr, grad_tol = 1e-2, digits = 2L)
```

## Diagnostics

Check that all parameters are estimable using the `check_estimability` function:

```{r check-estimability, echo=TRUE, message=FALSE, error = TRUE}
check_estimability(obj = obj)
```

Look for any parameters with high correlations (absolute correlation > 0.95)
using the `get_cor_pairs` function:

```{r check-cor, echo=TRUE, message=FALSE}
get_cor_pairs(obj = obj, threshold = 0.95)
```

Calculate standard deviations of all model parameters:

```{r run-sdreport, echo=TRUE, message=FALSE}
Report <- RTMB::sdreport(obj)
```

## Data fits

The CPUE series is set up using RTMB's `OBS` mechanism inside the model, which
allows simulation via `obj$simulate()`.

```{r simulate, echo=TRUE, message=FALSE, fig.height=4, fig.width=7}
set.seed(42)
sims <- replicate(20, obj$simulate()$cpue_log_obs, simplify = FALSE)

# again include index so points and lines are drawn per series
cpue_df <- data$cpue_data %>%
  mutate(pred = obj$report()$cpue_pred)

# plot simulated replicates with index grouping
base_plot <- ggplot(cpue_df, aes(x = ts, y = value)) +
  geom_point(aes(fill = factor(index)),shape=21,size=3) +
  geom_line(aes(y = pred, color = factor(index)),linewidth=1.25) +
  labs(x = "Time step", y = "CPUE", color = "Index", fill = "Index",
       title = "Simulated vs observed CPUE (20 replicates)") +
  scale_color_brewer(type = "qual", palette = "Dark2") +
  scale_fill_brewer(type = "qual", palette = "Dark2") +
  theme_minimal()

# add simulated series in gray
sim_df <- map_dfr(sims, ~ data.frame(ts = data$cpue_data$ts,
                                     index = as.factor(data$cpue_data$index),
                                     sim = exp(.x)), .id = "iter")

base_plot +
  geom_line(data = sim_df, aes(x = ts, y = sim, group = iter,color=index), alpha = 0.5)

```

The length compositions are also set up using RTMB's `OBS` mechanism inside the model, which
allows simulation via `obj$simulate()`.

```{r lf-diagnostics, echo=TRUE, message=FALSE, fig.height=6, fig.width=7}
if (data$lf_switch == 0) {
  cat("Length composition likelihood is switched off (lf_switch = 0), so no predicted length compositions to show.\n")
} else {
  lf_rep <- obj$report()
  lf_pred <- lf_rep$lf_pred
  nbins <- data$lf_maxbin[1] - data$lf_minbin[1] + 1

  # Name the list elements and columns
  names(lf_pred) <- data$lf_fishery_f
  for (k in seq_along(lf_pred)) colnames(lf_pred[[k]]) <- data$len_mid

  # Helper: convert a flat OBS vector back to list-of-matrices (proportions)
  flat_to_list <- function(v, lf_n_f, nbins) {
    out <- vector("list", length(lf_n_f))
    offset <- 0L
    for (k in seq_along(lf_n_f)) {
      m <- matrix(v[(offset + 1):(offset + lf_n_f[k] * nbins)], nrow = lf_n_f[k], ncol = nbins, byrow = TRUE)
      # Normalise each row to proportions
      rs <- rowSums(m)
      rs[rs == 0] <- 1
      out[[k]] <- m / rs
      offset <- offset + lf_n_f[k] * nbins
    }
    out
  }

  # Split year index by fishery (must match length of lf_year)
  lf_year_list <- split(data$lf_year, data$lf_fishery)

  # ---- Predicted: long data frame ----
  df_pred <- map_dfr(seq_along(lf_pred), function(i) {
    m <- lf_pred[[i]]
    yrs <- lf_year_list[[i]]
    as.data.frame.table(m, responseName = "pred", stringsAsFactors = FALSE) %>%
      rename(id = Var1, length = Var2) %>%
      mutate(fishery = names(lf_pred)[i],
             id = as.integer(factor(id, levels = unique(id))),
             year = yrs[id],
             length = as.numeric(length))
  })

  # ---- Observed: long data frame ----
  df_obs <- data.frame(data$lf_obs_in)
  names(df_obs) <- data$len_mid
  df_obs <- df_obs %>%
    mutate(year = data$lf_year, fishery = as.character(data$lf_fishery)) %>%
    group_by(fishery) %>%
    mutate(id = row_number()) %>%
    ungroup() %>%
    pivot_longer(cols = -c(year, fishery, id), names_to = "length", values_to = "obs") %>%
    mutate(length = as.numeric(length))

  # ---- Simulations: generate n_sim draws and convert to long format ----
  n_sim <- 5
  df_sim_all <- map_dfr(seq_len(n_sim), function(s) {
    # Determine which OBS vector was active based on lf_switch
    sim <- obj$simulate()
    if (data$lf_switch == 1) {
      sim_vec <- sim$lf_obs_flat
    } else if (data$lf_switch == 2) {
      sim_vec <- sim$lf_obs_prop
    } else {
      sim_vec <- sim$lf_obs_ints
    }
    sim_list <- flat_to_list(sim_vec, data$lf_n_f, nbins)
    names(sim_list) <- data$lf_fishery_f
    for (k in seq_along(sim_list)) colnames(sim_list[[k]]) <- data$len_mid
    map_dfr(seq_along(sim_list), function(i) {
      m <- sim_list[[i]]
      yrs <- lf_year_list[[i]]
      as.data.frame.table(m, responseName = "sim", stringsAsFactors = FALSE) %>%
        rename(id = Var1, length = Var2) %>%
        mutate(fishery = names(sim_list)[i],
               id = as.integer(factor(id, levels = unique(id))),
               year = yrs[id],
               length = as.numeric(length),
               sim_id = s)
    })
  })

  # ---- Join observed + predicted ----
  df <- left_join(df_obs, df_pred, by = c("id", "fishery", "length")) %>%
    select(-year.y) %>%
    rename(year = year.x)

  # ---- Plot fishery 1 (Comm) ----
  yrs_plot_1 <- seq(from=11,to=81,by=10)
  p1 <- ggplot(df %>% filter(fishery == "1", year %in% yrs_plot_1), aes(x = length)) +
    geom_col(aes(y = obs), fill = "grey70", width = 2) +
    geom_line(data = df_sim_all %>% filter(fishery == "1", year %in% yrs_plot_1),
              aes(y = sim, group = sim_id), colour = "steelblue", alpha = 0.3, linewidth = 0.4) +
    geom_line(aes(y = pred), colour = "red3", linewidth = 0.7) +
    facet_wrap(~ year, scales = "free_y", ncol = 5) +
    labs(x = "Length (cm)", y = "Proportion",
         title = "Length composition: Comm (Fishery 1)",
         subtitle = "Grey bars = observed, red = predicted, blue = simulated") +
    theme(strip.text = element_text(size = 7),
          axis.text = element_text(size = 6))
  print(p1)
}
```

## Model estimates

Visualise the selectivity curves by fleet (at length):

```{r plot-selectivity-length, echo=TRUE, message=FALSE, fig.height=4, fig.width=10, fig.cap="Selectivity-at-Length by Fleet"}
par_sel_est <- obj$env$parList()$par_sel

fleet_names <- c("Comm", "Non_comm", "ResFish")

sel_len_df <- do.call(rbind, lapply(seq_len(data$n_fishery), function(f) {
  par_f <- par_sel_est[f, ]
  sel_l <- if (data$sel_type_f[f] == 1L) {
    sel_logistic(data$len_mid, par_f)
  } else {
    sel_double_normal(data$len_mid, par_f)
  }
  data.frame(
    fishery    = f,
    fleet_name = fleet_names[f],
    length     = data$len_mid,
    selectivity = as.numeric(sel_l)
  )
}))

ggplot(sel_len_df, aes(x = length, y = selectivity)) +
  geom_line() +
  facet_wrap(~fleet_name, ncol = 3) +
  labs(x = "Length (cm)", y = "Selectivity") +
  ylim(0, 1)
```

Visualise the selectivity curves by fleet (at age):

```{r plot-selectivity, echo=TRUE, message=FALSE, fig.height=4, fig.width=10, fig.cap="Selectivity-at-Age by Fleet"}
rep <- obj$report()
sel_fya <- rep$sel_fya

sel_df <- expand.grid(fishery = 1:data$n_fishery, age = 1:data$n_age)
sel_df$selectivity <- sapply(1:nrow(sel_df), function(i) {
  sel_fya[sel_df$fishery[i], 1, sel_df$age[i]]
})
sel_df$real_age <- data$age_a[sel_df$age]
sel_df$fleet_name <- fleet_names[sel_df$fishery]

ggplot(sel_df, aes(x = real_age, y = selectivity)) +
  geom_line() +
  facet_wrap(~fleet_name, ncol = 3) +
  labs(x = "Age", y = "Selectivity") +
  ylim(0, 1)
```

# Compare to known values

Inspect estimated spawning biomass trajectory from *opal* (blue) relative to the simulated 
"truth" from the *SS3* operating model (black) and the estimated trajectory from the *SS3* 
estimation model (red).
The base case fixes initial equilibrium fishing mortality using SS3-implied values:
fleet 1 uses `InitF = 0.0119122` and fleets without initial catch use a negligible
placeholder (`1e-8`). Estimating `init_F_f` can be treated as a future sensitivity
when early composition data are informative about the initial age structure.

```{r opt-spbio, echo=TRUE, message=FALSE, fig.height=4, fig.width=7}

spawning_biomass_est <- as.data.frame(summary(Report)[which(rownames(summary(Report)) == "spawning_biomass_y"), c("Estimate","Std. Error")])
colnames(spawning_biomass_est) <- c("estimate", "std_error")
spawning_biomass_est <- spawning_biomass_est[1:data$n_year,]
spawning_biomass_est$year <- calendar_years
spawning_biomass_est$lower <- spawning_biomass_est$estimate - 1.96 * spawning_biomass_est$std_error
spawning_biomass_est$upper <- spawning_biomass_est$estimate + 1.96 * spawning_biomass_est$std_error

# bring in true spawning biomass from the SS3 model (from the data object)
data(opaka_truth)

# prepare plotting limits
ybmax <- max(spawning_biomass_est$upper, obj$report()$spawning_biomass_y, na.rm = TRUE)

ggplot(spawning_biomass_est, aes(x = year)) +
  geom_ribbon(aes(ymin = lower, ymax = upper), fill = "steelblue", alpha = 0.25) +
  geom_line(aes(y = estimate), colour = "steelblue", linewidth = 1) +
  geom_line(data=subset(opaka_truth$om$sprseries, Yr %in% calendar_years), aes(x=Yr, y=SSB), colour = "black", linewidth = 1) +
  geom_line(data=subset(opaka_truth$em$sprseries, Yr %in% calendar_years), aes(x=Yr, y=SSB), colour = "red", linewidth = 1) +
  labs(x = "Year", y = "Spawning biomass",
    title = "Estimated spawning biomass trajectory") +
  # scale_y_continuous(limits = c(0, ybmax * 1.2)) +
  theme_minimal()
```

```{r platoon-correction, echo=TRUE, message=FALSE, fig.height=5, fig.width=8}
# ---- Save the default (PLA-based) opal estimate for comparison ----
spawning_biomass_default <- spawning_biomass_est
spawning_biomass_default$model <- "opal (PLA)"

# ---- Platoon-weighted biology from SS3 report ----
# SS3 uses 3 discrete growth platoons (slow/medium/fast) with weights
# 0.15/0.70/0.15 and a within/between stdev ratio of 0.7. The effective
# weight-at-age and spawning potential-at-age differ from a single growth
# track (opal's PLA) due to Jensen's inequality: the non-linear length-weight
# (w = a*L^b) and logistic maturity curves mean the platoon-weighted average
# is not equal to the value at mean length. To verify that opal's population
# dynamics are correct, we override these two vectors with platoon-averaged
# values extracted from the SS3 Biology_at_age report table.

platoon_frac <- c(0.15, 0.70, 0.15)
frac_female  <- 1

# Wt_Beg by platoon (ages 0-43, from SS3 Biology_at_age_in_endyr)
wt_p1 <- c(0.00297429, 0.09807, 0.356455, 0.737394, 1.17746, 1.62494, 2.04639,
            2.42401, 2.75098, 3.02731, 3.25674, 3.44476, 3.59732, 3.72019,
            3.81859, 3.89703, 3.95937, 4.00876, 4.04783, 4.07867, 4.103,
            4.12216, 4.13725, 4.14912, 4.15845, 4.16579, 4.17155, 4.17608,
            4.17964, 4.18244, 4.18463, 4.18636, 4.18771, 4.18878, 4.18961,
            4.19027, 4.19078, 4.19118, 4.1915, 4.19175, 4.19195, 4.1921,
            4.19222, 4.19235)
wt_p2 <- c(0.00378997, 0.121547, 0.441813, 0.913989, 1.45946, 2.01411, 2.5365,
            3.00456, 3.40985, 3.75236, 4.03675, 4.2698, 4.4589, 4.6112,
            4.73316, 4.8304, 4.90766, 4.96889, 5.01731, 5.05555, 5.0857,
            5.10945, 5.12815, 5.14286, 5.15443, 5.16353, 5.17067, 5.17629,
            5.1807, 5.18417, 5.18689, 5.18902, 5.1907, 5.19202, 5.19305,
            5.19387, 5.1945, 5.19501, 5.1954, 5.19571, 5.19595, 5.19614,
            5.19629, 5.19645)
wt_p3 <- c(0.00463342, 0.148504, 0.539825, 1.11676, 1.78326, 2.46098, 3.09928,
            3.67119, 4.16641, 4.58492, 4.9324, 5.21716, 5.44822, 5.63431,
            5.78334, 5.90215, 5.99655, 6.07137, 6.13053, 6.17725, 6.21409,
            6.24311, 6.26596, 6.28394, 6.29807, 6.30919, 6.31792, 6.32478,
            6.33017, 6.33441, 6.33773, 6.34034, 6.34239, 6.344, 6.34527,
            6.34626, 6.34704, 6.34765, 6.34813, 6.34851, 6.3488, 6.34904,
            6.34922, 6.34942)

# Mat*Fecund by platoon (ages 0-43, from SS3 Biology_at_age_in_endyr)
mf_p1 <- c(0, 0, 0, 0.00337726, 0.730569, 1.5984, 2.04555, 2.42397, 2.75098,
            3.02731, 3.25674, 3.44476, 3.59732, 3.72019, 3.81859, 3.89703,
            3.95937, 4.00876, 4.04783, 4.07867, 4.103, 4.12216, 4.13725,
            4.14912, 4.15845, 4.16579, 4.17155, 4.17608, 4.17964, 4.18244,
            4.18463, 4.18636, 4.18771, 4.18878, 4.18961, 4.19027, 4.19078,
            4.19118, 4.1915, 4.19175, 4.19195, 4.1921, 4.19222, 4.19235)
mf_p2 <- c(0, 0, 0, 0.0840252, 1.38406, 2.01357, 2.5365, 3.00456, 3.40985,
            3.75236, 4.03675, 4.2698, 4.4589, 4.6112, 4.73316, 4.8304,
            4.90766, 4.96889, 5.01731, 5.05555, 5.0857, 5.10945, 5.12815,
            5.14286, 5.15443, 5.16353, 5.17067, 5.17629, 5.1807, 5.18417,
            5.18689, 5.18902, 5.1907, 5.19202, 5.19305, 5.19387, 5.1945,
            5.19501, 5.1954, 5.19571, 5.19595, 5.19614, 5.19629, 5.19645)
mf_p3 <- c(0, 0, 0.000000273, 0.542148, 1.78033, 2.46098, 3.09928, 3.67119,
            4.16641, 4.58492, 4.9324, 5.21716, 5.44822, 5.63431, 5.78334,
            5.90215, 5.99655, 6.07137, 6.13053, 6.17725, 6.21409, 6.24311,
            6.26596, 6.28394, 6.29807, 6.30919, 6.31792, 6.32478, 6.33017,
            6.33441, 6.33773, 6.34034, 6.34239, 6.344, 6.34527, 6.34626,
            6.34704, 6.34765, 6.34813, 6.34851, 6.3488, 6.34904, 6.34922,
            6.34942)

# Platoon-weighted weight-at-age
data$weight <- platoon_frac[1] * wt_p1 +
               platoon_frac[2] * wt_p2 +
               platoon_frac[3] * wt_p3

# Platoon-weighted spawning potential-at-age (scaled by fraction female)
data$spawning_potential <- frac_female * (platoon_frac[1] * mf_p1 +
                                          platoon_frac[2] * mf_p2 +
                                          platoon_frac[3] * mf_p3)

# ---- Rebuild and refit with platoon-corrected biology ----
obj2 <- MakeADFun(func = cmb(opal_model, data), parameters = parameters, map = map)

opt2 <- nlminb(start = obj2$par, objective = obj2$fn, gradient = obj2$gr,
               hessian = obj2$he, lower = Lwr, upper = Upr, control = control)
opt2 <- nlminb(start = opt2$par, objective = obj2$fn, gradient = obj2$gr,
               hessian = obj2$he, lower = Lwr, upper = Upr, control = control)
opt2 <- nlminb(start = opt2$par, objective = obj2$fn, gradient = obj2$gr,
               hessian = obj2$he, lower = Lwr, upper = Upr, control = control)

cat("Convergence:", opt2$convergence, "\n")
cat("Max gradient:", max(abs(obj2$gr())), "\n")

Report2 <- RTMB::sdreport(obj2)
rep2 <- obj2$report()

# ---- Extract platoon-corrected spawning biomass ----
spawning_biomass_corrected <- as.data.frame(
  summary(Report2)[which(rownames(summary(Report2)) == "spawning_biomass_y"),
                   c("Estimate", "Std. Error")])
colnames(spawning_biomass_corrected) <- c("estimate", "std_error")
spawning_biomass_corrected <- spawning_biomass_corrected[1:data$n_year, ]
spawning_biomass_corrected$year  <- calendar_years
spawning_biomass_corrected$lower <- spawning_biomass_corrected$estimate - 1.96 * spawning_biomass_corrected$std_error
spawning_biomass_corrected$upper <- spawning_biomass_corrected$estimate + 1.96 * spawning_biomass_corrected$std_error
spawning_biomass_corrected$model <- "opal (platoon-corrected)"

# ---- Comparison plot ----
data(opaka_truth)

opal_df <- rbind(spawning_biomass_default, spawning_biomass_corrected)

ybmax <- max(opal_df$upper,
             opaka_truth$om$sprseries$SSB,
             opaka_truth$em$sprseries$SSB, na.rm = TRUE)

ggplot(opal_df, aes(x = year)) +
  geom_ribbon(aes(ymin = lower, ymax = upper, fill = model), alpha = 0.2) +
  geom_line(aes(y = estimate, colour = model), linewidth = 1) +
  geom_line(data = subset(opaka_truth$om$sprseries, Yr %in% calendar_years),
            aes(x = Yr, y = SSB), colour = "black", linewidth = 1) +
  geom_line(data = subset(opaka_truth$em$sprseries, Yr %in% calendar_years),
            aes(x = Yr, y = SSB), colour = "red", linewidth = 1) +
  scale_colour_manual(values = c("opal (PLA)" = "steelblue",
                                 "opal (platoon-corrected)" = "darkorange")) +
  scale_fill_manual(values = c("opal (PLA)" = "steelblue",
                               "opal (platoon-corrected)" = "darkorange")) +
  labs(x = "Year", y = "Spawning biomass",
       title = "Estimated spawning biomass trajectory",
       subtitle = "Black = SS3 OM truth, Red = SS3 EM, Coloured = opal estimates",
       colour = NULL, fill = NULL) +
  scale_y_continuous(limits = c(0, ybmax * 1.2)) +
  theme_minimal() +
  theme(legend.position = "bottom")
```

```{r dynamics-check, echo=TRUE, message=FALSE, fig.height=5, fig.width=8}
# ---- Dynamics diagnostic: fix all parameters at SS3 MLE values ----
# This isolates opal's population dynamics from estimation differences.
# If the forward projection matches SS3, the dynamics are correct and any
# gap in the estimated model is due to parameter estimation (different
# likelihood surface, selectivity parameterization, etc.). If the projection
# doesn't match, a structural issue remains.

# SS3 MLE values from ss.par and derived quantities
# SS3's SSB_Virgin = 4639.1, ln(R0) = 5.65608, h = 0.76, sigma_r = 0.52
# SS3's init_F[1] = 0.0119122, Q_parm = c(-3.773, -6.253)

# Build parameter list with all values at SS3 MLEs
parameters_ss3 <- parameters  # start from the current parameter list

# B0: SS3's SSB_Virgin = 4639.1
parameters_ss3$log_B0 <- log(4639.1)

# Steepness, sigma_r (already set from opaka_parameters, but be explicit)
parameters_ss3$log_h <- log(0.76)
parameters_ss3$log_sigma_r <- log(0.52)

# CPUE catchability (from ss.par Q_parm)
parameters_ss3$log_cpue_q <- c(-3.77275156733, -6.25323305447)

# Recruitment deviations: use SS3's full recdev2 vector
# recdev2 covers the 22 early-period devs (ages) + 75 main-period devs
# init_rdev_a: reversed early devs (already set in the parameter block above)
# rdev_y: shifted main devs (already set in the parameter block above)
# These should already match SS3 from the original parameter setup, but
# ensure they are the unmodified SS3 values:
ss3_recdev2 <- c(0.0675235702971, 0.0764991539955, 0.0858442617148,
  0.0950046747340, 0.102706450120, 0.107687329214, 0.107711705598,
  0.0998349181169, 0.0808136660096, 0.0482338925745, 0.00234443377404,
  -0.0517563213031, -0.102395434864, -0.131458254996, -0.121100670313,
  -0.0714825835398, -0.00742166036481, -0.0109308396989, -0.0385028902033,
  0.0366483160551, 0.0165104920249, -0.166905707226,
  0.684759008019, -0.00129318028442, 0.291954545496, -0.502105154386,
  -0.395132761139, 0.289343271935, -0.288480831619, -0.490309035200,
  -0.212366196568, 0.161892641109, -0.0209673906075, -0.297262261087,
  0.329268521456, 0.00878056067047, -0.0597345841788, -0.635158118155,
  0.121730487831, 0.564895487490, -0.412552483484, 0.492195882830,
  0.863050417304, -0.262592196252, -0.580724697255, 0.192426722416,
  0.470346549222, -0.450985027130, 0.468806143351, -0.0970889237156,
  0.125657219706, -0.590564615592, -0.206902651435, 0.212391740136,
  -0.786327541142, -0.942202604456, -0.479537926647, -0.910561752327,
  -0.338249154939, -1.08143122504, 0.610318451176, -0.116085865688,
  -1.00403426050, -0.0967824207922, 0.241495871563, 0.381401207516,
  -0.162517483700, 0.164175327286, 0.739150845802, -0.283843490207,
  -0.324836661861, -0.297336491801, -0.566398458638, -0.553810264824,
  -0.00857108423637, -0.0751868745470, 0.339022613392, 0.0187206379281,
  0.633372740779, 0.675976465171, 0.134831485765, 0.429015103446,
  0.00483559599714, -0.953917944860, -0.848588067707, -0.0919670526641,
  -0.820322333684, -0.134672436697, 0.936671897697, 0.135185903985,
  -0.301177900173, -0.988301462145, -0.138899249513, 0.785829523698,
  -0.674385690767, 0.751722180103, 0.352211358296, 0.853371113276,
  -0.348627551973, -0.514856812016, -0.157533901109, -0.175313077136,
  0.212151760383, 0.110769041131, 0.126998013037, -0.349623955515,
  0.287222081367, 0.166718339864, 0.263616100771, -0.514692520633,
  0.543032699532, 0.148610562305, 0.0741026805708, -0.0379144746966,
  0.0478985480312, -0.0919462212626, -0.825992698686, 0.225426831222,
  -0.711467818125, 0.447979718775, 0.00000000000)

# First 22 values of recdev2 are the initial age devs (reverse into init_rdev_a)
# The remaining values are the main time-series devs
# init_rdev_a[1] gets the 1949 dev (first of recdev2 after the early devs)
ss3_early <- c(0.0206807064789, 0.00321579077358, 0.00372620421544,
  0.00431740128909, 0.00500263165771, 0.00579673718936, 0.00671727050323,
  0.00778421989329, 0.00902081267571, 0.0104539317083, 0.0121143194944,
  0.0140374805236, 0.0162641224184, 0.0188404004022, 0.0218181611323,
  0.0252560947069, 0.0292181862911, 0.0337726098165, 0.0389897504158,
  0.0449362496360, 0.0516661790666, 0.0592048886170)

parameters_ss3$init_rdev_a <- rev(c(ss3_early, ss3_recdev2[1:22]))
parameters_ss3$init_rdev_a[1] <- ss3_recdev2[22]  # 1949 dev at age 0

# Time-series rdevs: recdev2[23:end] are 1950 onwards, plus 0.0 for terminal
parameters_ss3$rdev_y <- c(ss3_recdev2[23:length(ss3_recdev2)], 0.0)

# Selectivity: use the converted SS3 values (already in parameters$par_sel)
parameters_ss3$par_sel <- parameters$par_sel

# Initial F
parameters_ss3$log_init_F_f <- c(log(0.0119122), rep(log(1e-8), data$n_fishery - 1))

# ---- Fix ALL parameters (no estimation) ----
map_fixed <- list(
  log_B0       = factor(NA),
  log_h        = factor(NA),
  log_sigma_r  = factor(NA),
  log_cpue_q   = factor(rep(NA, length(parameters_ss3$log_cpue_q))),
  cpue_creep   = factor(rep(NA, data$n_index)),
  log_cpue_tau = factor(rep(NA, data$n_index)),
  log_cpue_omega = factor(rep(NA, data$n_index)),
  log_lf_tau   = factor(rep(NA, length(parameters_ss3$log_lf_tau))),
  log_L1       = factor(NA),
  log_L2       = factor(NA),
  log_k        = factor(NA),
  log_CV1      = factor(NA),
  log_CV2      = factor(NA),
  par_sel      = factor(rep(NA, length(parameters_ss3$par_sel))),
  log_init_F_f = factor(rep(NA, data$n_fishery)),
  rdev_y       = factor(rep(NA, length(parameters_ss3$rdev_y))),
  init_rdev_a  = factor(rep(NA, data$n_age))
)

# ---- Build and evaluate (no optimization) ----
obj_check <- MakeADFun(func = cmb(opal_model, data),
                       parameters = parameters_ss3, map = map_fixed)
rep_check <- obj_check$report()

cat("B0:", rep_check$B0, "\n")
cat("R0:", rep_check$R0, "\n")
cat("SPR0 (B0/R0):", rep_check$B0 / rep_check$R0, "\n")
cat("Initial SSB:", rep_check$spawning_biomass_y[1], "\n")

# ---- Comparison: opal dynamics at SS3 MLEs vs SS3 trajectories ----
data(opaka_truth)

ssb_check <- data.frame(
  year = calendar_years,
  estimate = rep_check$spawning_biomass_y[1:data$n_year]
)

ybmax <- max(ssb_check$estimate,
             opaka_truth$om$sprseries$SSB,
             opaka_truth$em$sprseries$SSB, na.rm = TRUE)

ggplot() +
  geom_line(data = ssb_check, aes(x = year, y = estimate, colour = "opal (SS3 MLEs)"),
            linewidth = 1) +
  geom_line(data = subset(opaka_truth$om$sprseries, Yr %in% calendar_years),
            aes(x = Yr, y = SSB, colour = "SS3 OM"), linewidth = 1) +
  geom_line(data = subset(opaka_truth$em$sprseries, Yr %in% calendar_years),
            aes(x = Yr, y = SSB, colour = "SS3 EM"), linewidth = 1) +
  scale_colour_manual(values = c("opal (SS3 MLEs)" = "darkorange",
                                 "SS3 OM" = "black",
                                 "SS3 EM" = "red")) +
  labs(x = "Year", y = "Spawning biomass",
       title = "Dynamics diagnostic: opal at SS3 MLE parameters",
       subtitle = "No estimation — all parameters fixed at SS3 values",
       colour = NULL) +
  scale_y_continuous(limits = c(0, ybmax * 1.2)) +
  theme_minimal() +
  theme(legend.position = "bottom")
```
