---
title: "Opakapaka selectivity: double-normal and double Richards"
format: html
execute:
  eval: false
vignette: >
  %\VignetteIndexEntry{Opakapaka selectivity: double-normal and double Richards}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{quarto::html}
---

```{r}
#| include: false
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

# Introduction

This example compares the existing SS3 pattern-24 double-normal selectivity
with the six-parameter double Richards curve for the ResFish fleet in the
bundled opakapaka assessment. Both models retain the commercial logistic
selectivity, use the same data and likelihoods, and estimate the same
selectivity parameters. This is an illustrative curve-family comparison, not
an assessment recommendation or an evaluation of parameter identifiability.

The double Richards curve has an ascending and descending generalized-logistic
limb. Its six real-line parameters respectively control the ascending 50% point,
ascending width, ascending shape, separation to the descending 50% point,
descending width, and descending shape. Shape parameters of zero give logistic
limbs. Unlike the double-normal curve, the double Richards curve is smooth in
both length and its parameters. Its raw peak is not rescaled, so overlapping
limbs can produce a peak below one.

# Packages and data

```{r}
#| label: packages
#| message: false
library(dplyr)
library(ggplot2)
library(tidyr)
library(opal)
library(RTMB)

data(opaka_data)
data(opaka_lf)
data(opaka_parameters)

data_base <- opaka_data
data_base$n_index <- 2L
data_base$priors <- NULL
data_base$A1 <- 0
data_base$A2 <- data_base$A1 + data_base$n_age - 1
data_base$min_age <- 0
data_base$age_a <- seq(data_base$min_age, by = 1, length.out = data_base$n_age)
data_base$sex_ratio <- rep(1, data_base$n_age)

lf_wide <- opaka_lf |>
  pivot_wider(id_cols = c(fishery, year, month, ts),
              names_from = bin, values_from = value, values_fill = 0) |>
  arrange(fishery, ts)
data_base <- prep_lf_data(
  data_base, lf_wide, lf_keep_fisheries = c(1, 3),
  lf_var_adjust = rep(1, data_base$n_fishery), lf_addtocomp = 1e-10
)
data_base$lf_switch <- 1L
data_base$init_bias_adj_a <- get_bias_adj_vector(
  1949 - (data_base$age_a - 1), 1, c(1931, 1974, 2021.88, 2022.43), 0.869
)
data_base$bias_adj_y <- get_bias_adj_vector(
  (1949 - 1) + data_base$years, 1, c(1931, 1974, 2021.88, 2022.43), 0.869
)
```

# Paired model setup

SS3 parameters are converted before changing the ResFish type because the
double Richards curve has no SS3 equivalent. The Richards starting values are
obtained by fitting its curve to the converted double-normal curve on the
length grid. This makes the comparison begin from similar selectivity shapes.

```{r}
#| label: paired-setup
#| message: false
init_rdev_a <- rep(0, data_base$n_age)
init_rdev_a[1] <- opaka_parameters$rdev_y[1]
rdev_y <- c(opaka_parameters$rdev_y[-1], 0)

parameters_base <- list(
  log_B0 = 9,
  log_h = as.numeric(opaka_parameters$log_h),
  log_sigma_r = as.numeric(opaka_parameters$log_sigma_r),
  log_cpue_q = as.numeric(opaka_parameters$log_cpue_q),
  cpue_creep = as.numeric(rep(opaka_parameters$cpue_creep, data_base$n_index)),
  log_cpue_tau = rep(log(0.1), data_base$n_index),
  log_cpue_omega = as.numeric(rep(opaka_parameters$log_cpue_omega, data_base$n_index)),
  log_lf_tau = rep(log(0.1), data_base$n_fishery),
  log_L1 = as.numeric(opaka_parameters$log_L1),
  log_L2 = as.numeric(opaka_parameters$log_L2),
  log_k = as.numeric(opaka_parameters$log_k),
  log_CV1 = as.numeric(opaka_parameters$log_CV1),
  log_CV2 = as.numeric(opaka_parameters$log_CV2),
  par_sel = convert_ss3_selex_to_rtmb(
    as.matrix(opaka_parameters$par_sel), data_base$sel_type_f, data_base$len_mid
  ),
  log_init_F_f = c(log(0.0119122), rep(log(1e-8), data_base$n_fishery - 1)),
  rdev_y = rdev_y,
  init_rdev_a = init_rdev_a
)

fit_richards_start <- MakeADFun(
  function(par) sum((
    sel_double_richards(data_base$len_mid, par) -
      sel_double_normal(data_base$len_mid, parameters_base$par_sel[3, ])
  )^2),
  c(parameters_base$par_sel[3, 1], 0, 0, 0, 0, 0)
)
richards_start <- nlminb(
  fit_richards_start$par, fit_richards_start$fn, fit_richards_start$gr,
  lower = rep(-5, 6), upper = rep(5, 6)
)$par

data_dn <- data_base
parameters_dn <- parameters_base

data_dr <- data_base
data_dr$sel_type_f[3] <- 3L
parameters_dr <- parameters_base
parameters_dr$par_sel[3, ] <- richards_start
```

For a fair structural comparison, both fits estimate the two commercial
logistic parameters and all six ResFish selectivity parameters. Other
parameters follow the opakapaka length-composition configuration.

```{r}
#| label: fit-models
#| message: false
make_map <- function(parameters, data) {
  selectivity_map <- matrix(NA_integer_, nrow(parameters$par_sel), ncol(parameters$par_sel))
  selectivity_map[1, 1:2] <- 1:2
  selectivity_map[3, ] <- 3:8
  list(
    log_h = factor(NA), log_sigma_r = factor(NA),
    cpue_creep = factor(rep(NA, data$n_index)),
    log_cpue_tau = factor(rep(NA, data$n_index)),
    log_cpue_omega = factor(rep(NA, data$n_index)),
    log_lf_tau = factor(rep(NA, length(parameters$log_lf_tau))),
    log_L1 = factor(NA), log_L2 = factor(NA), log_k = factor(NA),
    log_CV1 = factor(NA), log_CV2 = factor(NA),
    par_sel = factor(as.vector(selectivity_map)),
    log_init_F_f = factor(rep(NA, data$n_fishery)),
    init_rdev_a = factor(seq_len(data$n_age)),
    rdev_y = factor(seq_len(length(parameters$rdev_y)))
  )
}

fit_model <- function(data, parameters, richards = FALSE) {
  data$priors <- get_priors(parameters, data)
  map <- make_map(parameters, data)
  obj <- MakeADFun(cmb(opal_model, data), parameters, map = map, silent = TRUE)
  bounds <- get_bounds(obj, parameters)
  lower <- bounds$lower
  upper <- bounds$upper
  if (richards) {
    selectivity_index <- grep("par_sel", names(obj$par))
    lower[selectivity_index] <- -5
    upper[selectivity_index] <- 5
  }
  control <- list(eval.max = 10000, iter.max = 10000)
  opt <- nlminb(obj$par, obj$fn, obj$gr, lower = lower, upper = upper, control = control)
  opt <- nlminb(opt$par, obj$fn, obj$gr, lower = lower, upper = upper, control = control)
  list(data = data, parameters = parameters, map = map, obj = obj, opt = opt,
       lower = lower, upper = upper, report = obj$report())
}

fit_dn <- fit_model(data_dn, parameters_dn)
fit_dr <- fit_model(data_dr, parameters_dr, richards = TRUE)

data.frame(
  model = c("Double-normal", "Double Richards"),
  convergence = c(fit_dn$opt$convergence, fit_dr$opt$convergence),
  maximum_gradient = c(max(abs(fit_dn$obj$gr())), max(abs(fit_dr$obj$gr()))),
  nll = c(fit_dn$opt$objective, fit_dr$opt$objective)
)
```

```{r}
#| label: fit-diagnostics
#| output: false
diagnostics <- lapply(
  list(double_normal = fit_dn, double_richards = fit_dr),
  function(fit) list(
    estimability = check_estimability(fit$obj),
    high_correlations = get_cor_pairs(fit$obj, threshold = 0.95)
  )
)
```

# Selectivity comparison

The curves below are raw selectivity. In particular, the double Richards
maximum is informative: a value below one reflects overlap between its limbs
and is not a normalization error.

```{r}
#| label: compare-selectivity-length
#| fig-cap: "ResFish selectivity-at-length under the two curve families."
par_dn <- fit_dn$obj$env$parList()$par_sel[3, ]
par_dr <- fit_dr$obj$env$parList()$par_sel[3, ]
selectivity_length <- bind_rows(
  data.frame(length = data_base$len_mid,
             selectivity = sel_double_normal(data_base$len_mid, par_dn),
             model = "Double-normal"),
  data.frame(length = data_base$len_mid,
             selectivity = sel_double_richards(data_base$len_mid, par_dr),
             model = "Double Richards")
)
ggplot(selectivity_length, aes(length, selectivity, colour = model)) +
  geom_line(linewidth = 0.8) +
  labs(x = "Length (cm)", y = "Raw selectivity", colour = NULL) +
  coord_cartesian(ylim = c(0, 1)) +
  theme_bw()
```

```{r}
#| label: richards-natural-parameters
natural <- double_richards_natural(data_base$len_mid, par_dr)
data.frame(parameter = names(natural), value = unname(natural))
```

`l50_asc` and `l50_desc` are directly interpretable 50% points. The widths
equal a logistic 50-to-95% width only when their associated shape is one
(`par[3]` or `par[6]` equals zero); otherwise they scale inverse steepness.

```{r}
#| label: compare-selectivity-age
#| fig-cap: "PLA-integrated ResFish selectivity-at-age."
selectivity_age <- bind_rows(
  data.frame(age = data_base$age_a, selectivity = fit_dn$report$sel_fya[3, 1, ],
             model = "Double-normal"),
  data.frame(age = data_base$age_a, selectivity = fit_dr$report$sel_fya[3, 1, ],
             model = "Double Richards")
)
ggplot(selectivity_age, aes(age, selectivity, colour = model)) +
  geom_line(linewidth = 0.8) +
  labs(x = "Age", y = "Selectivity", colour = NULL) +
  coord_cartesian(ylim = c(0, 1)) +
  theme_bw()
```

# Fits and population consequences

The fitted length compositions are compared for ResFish observations. The
models share the same observations; only their selectivity family differs.

```{r}
#| label: compare-length-compositions
#| fig-cap: "Observed and fitted ResFish length compositions."
composition_data <- function(fit, model) {
  index <- which(fit$data$lf_fishery_f == 3L)
  predicted <- fit$report$lf_pred[[index]]
  observed <- fit$data$lf_obs_in[fit$data$lf_fishery == 3L, , drop = FALSE]
  observed <- observed / rowSums(observed)
  bind_rows(
    data.frame(year = rep(fit$data$lf_year[fit$data$lf_fishery == 3L], each = ncol(predicted)),
               length = rep(fit$data$len_mid, times = nrow(predicted)),
               value = as.vector(t(observed)), series = "Observed", model = model),
    data.frame(year = rep(fit$data$lf_year[fit$data$lf_fishery == 3L], each = ncol(predicted)),
               length = rep(fit$data$len_mid, times = nrow(predicted)),
               value = as.vector(t(predicted)), series = "Fitted", model = model)
  )
}
composition <- bind_rows(
  composition_data(fit_dn, "Double-normal"),
  composition_data(fit_dr, "Double Richards")
)
ggplot(composition, aes(length, value, colour = series)) +
  geom_line() +
  facet_grid(model ~ year) +
  labs(x = "Length (cm)", y = "Proportion", colour = NULL) +
  theme_bw()
```

```{r}
#| label: compare-objective-and-biomass
component_sum <- function(fit, component) sum(fit$report[[component]])
fit_summary <- data.frame(
  model = c("Double-normal", "Double Richards"),
  nll = c(fit_dn$opt$objective, fit_dr$opt$objective),
  cpue_nll = c(component_sum(fit_dn, "lp_cpue"), component_sum(fit_dr, "lp_cpue")),
  length_nll = c(component_sum(fit_dn, "lp_lf"), component_sum(fit_dr, "lp_lf")),
  terminal_spawning_biomass = c(tail(fit_dn$report$spawning_biomass_y, 1),
                                tail(fit_dr$report$spawning_biomass_y, 1))
)
fit_summary

biomass <- bind_rows(
  data.frame(year = (1949 - 1) + data_base$years,
             spawning_biomass = fit_dn$report$spawning_biomass_y,
             model = "Double-normal"),
  data.frame(year = (1949 - 1) + data_base$years,
             spawning_biomass = fit_dr$report$spawning_biomass_y,
             model = "Double Richards")
)
ggplot(biomass, aes(year, spawning_biomass, colour = model)) +
  geom_line(linewidth = 0.8) +
  labs(x = "Year", y = "Spawning biomass", colour = NULL) +
  theme_bw()
```

Use convergence status, gradients, `check_estimability()`, and
`get_cor_pairs()` before drawing inferential conclusions. The flexible shape
parameters can be weakly identified by real composition data, so this
comparison should be accompanied by sensitivity analyses and uncertainty
estimation before adopting double Richards selectivity in an assessment.
