---
title: "Profiling & Benchmarking"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Profiling & Benchmarking}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
---

```{r, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

# Introduction

This vignette establishes performance baselines for the `opal` BET model prior
to efficiency refactoring. It profiles the full
model evaluation, isolates key sub-functions, and records AD tape metrics. After
each refactoring change, re-run this vignette and compare against the saved
baseline to verify numerical equivalence and measure speedup.

The profiling workflow uses:

- `profvis` — interactive flame graph profiler for identifying hotspots
- `bench` — precise microbenchmarking for comparing before/after
- `system.time` — simple wall-clock timing for quick checks

# Load inputs

Load the `opal` package, `RTMB`, and profiling dependencies:

```{r load-pkg, echo=TRUE, message=FALSE}
library(dplyr)
library(RTMB)
library(profvis)
library(bench)
devtools::load_all()
```

The bundled data object `wcpo_bet_data` contains all biological parameters,
catch and CPUE observations, length structure, and prior specifications needed
for the BET assessment model:

```{r load-data, echo=TRUE, message=FALSE}
data(wcpo_bet_data)
data <- wcpo_bet_data
```

## Length composition data

Length composition observations are loaded from the bundled `wcpo_bet_lf`
long-format data object and transformed into model-ready arrays:

```{r prep-lf, echo=TRUE, message=FALSE}
data(wcpo_bet_lf)

lf_wide <- wcpo_bet_lf %>%
  tidyr::pivot_wider(
    id_cols     = c(fishery, year, month, ts),
    names_from  = bin,
    values_from = value,
    values_fill = 0
  ) %>%
  dplyr::arrange(fishery, ts)

var_adjust_scalars <- 1 / rep(1, data$n_fishery)

data <- prep_lf_data(data, lf_wide,
                     lf_keep_fisheries = c(8, 9),
                     lf_var_adjust     = var_adjust_scalars)
data$lf_switch <- 1L
```

## Weight composition data

Weight composition observations are loaded from the bundled `wcpo_bet_wf`
long-format data object and transformed into model-ready arrays. The
`prep_wf_data()` function also precomputes the rebinning matrix that maps
predicted length compositions to weight bins:

```{r prep-wf, echo=TRUE, message=FALSE}
data(wcpo_bet_wf)

# Weight bin scalars (1 kg bins, 1–200 kg)
data$wt_bin_start <- 1
data$wt_bin_width <- 1
data$n_wt         <- 200L

wf_wide <- wcpo_bet_wf |>
  tidyr::pivot_wider(
    id_cols     = c(fishery, year, month, ts),
    names_from  = bin,
    values_from = value,
    values_fill = 0
  ) |>
  dplyr::arrange(fishery, ts)

data <- prep_wf_data(data, wf_wide,
                     wf_keep_fisheries = c(2),
                     wf_switch = 1L)
```

# Model setup

## Parameters

Load initial parameter values from the bundled `wcpo_bet_parameters` data
object:

```{r get-pars, echo=TRUE, message=FALSE}
data(wcpo_bet_parameters)

parameters <- list(
  log_B0         = 20,
  log_h          = as.numeric(wcpo_bet_parameters$log_h),
  log_sigma_r    = as.numeric(wcpo_bet_parameters$log_sigma_r),
  log_cpue_q     = as.numeric(wcpo_bet_parameters$log_cpue_q),
  cpue_creep     = as.numeric(wcpo_bet_parameters$cpue_creep),
  log_cpue_tau   = as.numeric(wcpo_bet_parameters$log_cpue_tau),
  log_cpue_omega = as.numeric(wcpo_bet_parameters$log_cpue_omega),
  log_lf_tau     = log(rep(0.1, data$n_fishery)),
  log_wf_tau     = rep(0, data$n_fishery),
  log_L1         = as.numeric(wcpo_bet_parameters$log_L1),
  log_L2         = as.numeric(wcpo_bet_parameters$log_L2),
  log_k          = as.numeric(wcpo_bet_parameters$log_k),
  log_CV1        = as.numeric(wcpo_bet_parameters$log_CV1),
  log_CV2        = as.numeric(wcpo_bet_parameters$log_CV2),
  par_sel        = as.matrix(wcpo_bet_parameters$par_sel),
  rdev_y         = as.numeric(wcpo_bet_parameters$rdev_y)
)
```

## Priors

```{r get-priors, echo=TRUE, message=FALSE}
data$priors <- get_priors(parameters = parameters, data = data)
```

## Parameter map

Fix growth, selectivity, and some CPUE parameters at their initial values:

```{r get-map, echo=TRUE, message=FALSE}
map_sel <- matrix(NA, nrow(parameters$par_sel), ncol(parameters$par_sel))

map <- list(
  log_h          = factor(NA),
  log_sigma_r    = factor(NA),
  cpue_creep     = factor(NA),
  log_cpue_tau   = factor(NA),
  log_cpue_omega = factor(NA),
  log_lf_tau     = factor(rep(NA, length(parameters$log_lf_tau))),
  log_wf_tau     = factor(rep(NA, length(parameters$log_wf_tau))),
  log_L1         = factor(NA),
  log_L2         = factor(NA),
  log_k          = factor(NA),
  log_CV1        = factor(NA),
  log_CV2        = factor(NA),
  par_sel        = factor(map_sel)
)
```

## Build the AD object

```{r make-adfun, echo=TRUE, message=FALSE}
t_build <- system.time({
  obj <- MakeADFun(func = cmb(opal_model, data),
                   parameters = parameters, map = map)
})
cat("MakeADFun build time:", round(t_build["elapsed"], 2), "sec\n")
cat("Estimated parameters:", length(obj$par), "\n")
cat("Parameter names:", paste(unique(names(obj$par)), collapse = ", "), "\n")
```

# Model dimensions

These dimensions drive the computational cost. Every unnecessary scalar loop or
function call inside the dynamics is amplified by these values:

```{r dims, echo=TRUE}
cat("n_year:    ", data$n_year, "\n")
cat("n_age:     ", data$n_age, "\n")
cat("n_fishery: ", data$n_fishery, "\n")
cat("n_season:  ", data$n_season, "\n")
cat("n_len:     ", data$n_len, "\n")
cat("n_lf obs:  ", data$n_lf, "\n")
cat("lf_switch: ", data$lf_switch, "\n")
cat("n_wt:      ", data$n_wt, "\n")
cat("n_wf obs:  ", data$n_wf, "\n")
cat("wf_switch: ", data$wf_switch, "\n")
```

# Baseline timing

Warm up the AD tape (the first call constructs it), then time `fn()` and
`gr()` separately:

```{r baseline-warmup, echo=TRUE, message=FALSE}
invisible(obj$fn(obj$par))
invisible(obj$gr(obj$par))
```

```{r baseline-timing, echo=TRUE}
t_fn <- system.time(nll <- obj$fn(obj$par))
t_gr <- system.time(gr  <- obj$gr(obj$par))
cat("obj$fn():", round(t_fn["elapsed"], 4), "sec  (NLL =", round(nll, 4), ")\n")
cat("obj$gr():", round(t_gr["elapsed"], 4), "sec  (max|gr| =", round(max(abs(gr)), 6), ")\n")
cat("gr/fn ratio:", round(t_gr["elapsed"] / max(t_fn["elapsed"], 1e-6), 1), "x\n")
```

For a well-optimised AD tape, the gradient should cost roughly 2--4× the
function evaluation. A much higher ratio suggests redundant operations on the
tape that compound in reverse mode.

Repeated timing with `bench::mark` gives more stable estimates:

```{r baseline-bench, echo=TRUE}
bm_core <- bench::mark(
  fn = obj$fn(obj$par),
  gr = obj$gr(obj$par),
  iterations = 20,
  check = FALSE
)
bm_core[, c("expression", "min", "median", "itr/sec","total_time")]
```

# AD tape diagnostics

The AD tape is the sequence of elementary operations that RTMB records during
the first `fn()` call. Tape size directly determines gradient cost: fewer
operations = faster `gr()`.

```{r tape-diag, echo=TRUE}
tryCatch({
  tape_env <- obj$env
  if (!is.null(tape_env$ADFun)) {
    cat("ADFun object found in obj$env\n\n")
    capture.output(print(tape_env$ADFun)) |> head(20) |> cat(sep = "\n")
  }
}, error = function(e) {
  cat("Could not access tape info:", e$message, "\n")
})

cat("\nobj total size:  ", round(object.size(obj) / 1e6, 1), "MB\n")
cat("obj$env size:    ", round(object.size(obj$env) / 1e6, 1), "MB\n")
```

# profvis — full model evaluation

`profvis` samples the R call stack at regular intervals and produces an
interactive flame graph. Run 10 iterations of `fn()` + `gr()` to accumulate
enough samples:

```{r profvis-run, echo=TRUE, eval=FALSE}
prof <- profvis({
  for (i in 1:10) {
    obj$fn(obj$par)
    obj$gr(obj$par)
  }
}, simplify = FALSE)

htmlwidgets::saveWidget(prof, "profile_full.html", selfcontained = TRUE)
```

Open `profile_full.html` in a browser to inspect the flame graph. Key things to
look for:

- Time in `get_harvest_rate` vs `do_dynamics` vs `opal_model` — if
  `get_harvest_rate` is a significant fraction, Change 1 (inlining) will help.
- Time in `get_length_like` / `get_weight_like` / `get_cpue_like` — if the
  composition likelihoods dominate, Changes 7--9 are important. The weight
  likelihood adds an extra rebinning matrix multiply per observation.
- Time in `getAll()` calls — indicates overhead from unpacking data/parameter
  lists.
- Time in `.Call` (C++ / RTMB internals) vs R code — if most time is in
  `.Call`, the bottleneck is tape complexity rather than R dispatch overhead.

# Component-level profiling

Benchmark individual functions in plain R (outside the AD tape) to understand
their base cost. These timings won't include AD overhead but show the
algorithmic cost of each component.

## `opal_model`

The highest-level entry point bundles all individual components together: growth,
selectivity, dynamics, and likelihoods. Timing `opal_model()` directly outside the
AD tape gives a sense of the un-taped work that the tape must eventually record.

```{r bench-opal-model, echo=TRUE}
bm_opal_model <- bench::mark(
  opal_model = opal_model(parameters = parameters, data = data),
  iterations = 20,
  check = FALSE
)
bm_opal_model[, c("expression", "min", "median", "itr/sec","total_time")]

prof_opal <- profvis({
  for (i in 1:10) {
    opal_model(parameters = parameters, data = data)
  }
}, simplify = FALSE)

htmlwidgets::saveWidget(prof_opal, "profile_opal.html", selfcontained = TRUE)
```

## `get_pla`

The probability-of-length-at-age matrix is a `n_len × n_age` matrix computed
via `pnorm`. Currently uses a scalar inner loop over length bins (see Change 4):

```{r bench-pla, echo=TRUE}
par_list <- obj$env$parList(obj$env$last.par.best)
L1  <- exp(par_list$log_L1)
L2  <- exp(par_list$log_L2)
k   <- exp(par_list$log_k)
CV1 <- exp(par_list$log_CV1)
CV2 <- exp(par_list$log_CV2)

len_lower <- seq(from = data$len_bin_start, by = data$len_bin_width,
                 length.out = data$n_len)
len_upper <- len_lower + data$len_bin_width
len_mid   <- len_lower + data$len_bin_width / 2

mu_a <- get_growth(data$n_age, data$A1, data$A2, L1, L2, k)
sd_a <- get_sd_at_age(mu_a, L1, L2, CV1, CV2)

bm_pla <- bench::mark(
  get_pla = get_pla(len_lower, len_upper, mu_a, sd_a),
  iterations = 50,
  check = FALSE
)
bm_pla[, c("expression", "min", "median", "itr/sec","total_time")]

prof_pla <- profvis({
  for (i in 1:10) {
    get_pla(len_lower, len_upper, mu_a, sd_a)
  }
}, simplify = FALSE)
```

## `get_selectivity`

Selectivity is computed per fishery via logistic or double-normal curves applied
through the PLA matrix:

```{r bench-sel, echo=TRUE}
pla <- get_pla(len_lower, len_upper, mu_a, sd_a)

bm_sel <- bench::mark(
  get_selectivity = get_selectivity(data, par_list$par_sel, pla, len_mid),
  iterations = 50,
  check = FALSE
)
bm_sel[, c("expression", "min", "median", "itr/sec","total_time")]
```

## PLA matrix multiply

The core operation inside `get_length_like` — a `n_len × n_age` matrix-vector
multiply that converts catch-at-age to catch-at-length. This executes once per
LF observation:

```{r bench-pla-mult, echo=TRUE}
catch_a <- rep(1, data$n_age)

bm_pla_mult <- bench::mark(
  `pla %*% catch_a` = c(pla %*% catch_a),
  iterations = 200,
  check = FALSE
)
bm_pla_mult[, c("expression", "min", "median", "itr/sec","total_time")]
```

## `get_length_like`

The full length composition likelihood, called with plain-R arguments (not
AD-taped). This includes the per-observation matrix multiply, bin aggregation,
normalisation, and `dmultinom` evaluation:

```{r bench-lf, echo=TRUE}
invisible(obj$fn(obj$par))
rep <- obj$report()

lf_year_fi <- split(data$lf_year, data$lf_fishery)
lf_n_fi    <- split(data$lf_n, data$lf_fishery)

bm_lf <- bench::mark(
  get_length_like = get_length_like(
    lf_obs_flat      = data$lf_obs_flat,
    lf_obs_ints      = data$lf_obs_ints,
    lf_obs_prop      = data$lf_obs_prop,
    catch_pred_fya   = rep$catch_pred_fya,
    pla              = pla,
    lf_n_f           = data$lf_n_f,
    lf_fishery_f     = data$lf_fishery_f,
    lf_year_fi       = lf_year_fi,
    lf_n_fi          = lf_n_fi,
    lf_minbin        = data$lf_minbin,
    lf_maxbin        = data$lf_maxbin,
    removal_switch_f = data$removal_switch_f,
    lf_switch        = data$lf_switch,
    n_len            = data$n_len,
    n_lf             = data$n_lf,
    log_lf_tau       = par_list$log_lf_tau
  ),
  iterations = 50,
  check = FALSE
)
bm_lf[, c("expression", "min", "median", "itr/sec","total_time")]
```

## `split()` overhead

These `split()` calls currently execute inside the AD-taped function on every
model evaluation (see Change 8):

```{r bench-split, echo=TRUE}
bm_split <- bench::mark(
  split_year = split(data$lf_year, data$lf_fishery),
  split_n    = split(data$lf_n, data$lf_fishery),
  iterations = 500,
  check = FALSE
)
bm_split[, c("expression", "min", "median", "itr/sec","total_time")]
```

## Rebinning matrix multiply

The weight likelihood adds an extra matrix multiply per WF observation:
`wf_rebin_matrix %*% pred_at_length` converts the `n_len`-vector of predicted
length composition to an `n_wt`-vector of predicted weight composition. This
is a `n_wt × n_len` matrix-vector product:

```{r bench-rebin-mult, echo=TRUE}
pred_at_length <- rep(1, data$n_len)

bm_rebin_mult <- bench::mark(
  `rebin %*% pred_l` = c(data$wf_rebin_matrix %*% pred_at_length),
  iterations = 200,
  check = FALSE
)
bm_rebin_mult[, c("expression", "min", "median", "itr/sec","total_time")]
```

## `get_weight_like`

The full weight composition likelihood, called with plain-R arguments (not
AD-taped). This includes the per-observation PLA multiply, rebinning to
weight bins, normalisation, and likelihood evaluation:

```{r bench-wf, echo=TRUE}
invisible(obj$fn(obj$par))
rep <- obj$report()

wf_year_fi <- split(data$wf_year, data$wf_fishery)
wf_n_fi    <- split(data$wf_n, data$wf_fishery)

bm_wf <- bench::mark(
  get_weight_like = get_weight_like(
    wf_obs_flat      = data$wf_obs_flat,
    wf_obs_ints      = data$wf_obs_ints,
    wf_obs_prop      = data$wf_obs_prop,
    catch_pred_fya   = rep$catch_pred_fya,
    pla              = pla,
    wf_rebin_matrix  = data$wf_rebin_matrix,
    wf_n_f           = data$wf_n_f,
    wf_fishery_f     = data$wf_fishery_f,
    wf_year_fi       = wf_year_fi,
    wf_n_fi          = wf_n_fi,
    wf_minbin        = data$wf_minbin,
    wf_maxbin        = data$wf_maxbin,
    removal_switch_f = data$removal_switch_f,
    wf_switch        = data$wf_switch,
    n_wt             = data$n_wt,
    n_wf             = data$n_wf,
    log_wf_tau       = par_list$log_wf_tau
  ),
  iterations = 50,
  check = FALSE
)
bm_wf[, c("expression", "min", "median", "itr/sec","total_time")]
```

## WF `split()` overhead

These `split()` calls for WF data also execute inside the AD-taped function
on every model evaluation, mirroring the LF split overhead:

```{r bench-wf-split, echo=TRUE}
bm_wf_split <- bench::mark(
  split_wf_year = split(data$wf_year, data$wf_fishery),
  split_wf_n    = split(data$wf_n, data$wf_fishery),
  iterations = 500,
  check = FALSE
)
bm_wf_split[, c("expression", "min", "median", "itr/sec","total_time")]
```

# Likelihood component breakdown

Inspect the contribution of each likelihood component to the total NLL:

```{r nll-breakdown, echo=TRUE}
cat("lp_prior:  ", round(rep$lp_prior, 4), "\n")
cat("lp_penalty:", round(rep$lp_penalty, 4), "\n")
cat("lp_rec:    ", round(rep$lp_rec, 4), "\n")
cat("lp_cpue:   ", round(sum(rep$lp_cpue), 4),
    " (sum of", length(rep$lp_cpue), "obs)\n")
cat("lp_lf:     ", round(sum(rep$lp_lf), 4),
    " (sum of", length(rep$lp_lf), "obs)\n")
cat("lp_wf:     ", round(sum(rep$lp_wf), 4),
    " (sum of", length(rep$lp_wf), "obs)\n")
cat("Total NLL: ", round(nll, 4), "\n")
```

# Estimated tape operation counts

Before profiling, we can estimate the theoretical number of scalar AD tape
operations from the model dimensions. These estimates help prioritise which
refactoring changes will have the largest impact:

```{r tape-estimates, echo=TRUE}
n_yr  <- data$n_year
n_a   <- data$n_age
n_f   <- data$n_fishery
n_s   <- data$n_season
n_l   <- data$n_len
n_obs <- data$n_lf
n_w   <- data$n_wt
n_wobs <- data$n_wf

cat("Dynamics loop:\n")
cat("  Harvest rate function calls: ", n_yr, "×", n_s, "=", n_yr * n_s, "\n")
cat("  Scalar catch pred ops:       ", n_f, "×", n_a, "×", n_yr * n_s,
    "=", n_f * n_a * n_yr * n_s, "\n")

cat("\nTime-invariant array overhead:\n")
cat("  sel_fya year replication:  ", n_f, "×", n_yr, "=", n_f * n_yr, "\n")
cat("  weight_fya replication:    ", n_f, "×", n_yr, "=", n_f * n_yr, "\n")

cat("\nPLA construction:\n")
cat("  Current (scalar pnorm):    ", n_a, "×", n_l, "=", n_a * n_l, "\n")
cat("  After vectorisation:       ", n_a, "× 2 =", n_a * 2, "\n")

cat("\nLength likelihood:\n")
cat("  Matrix multiplies:         ", n_obs, "×", n_l, "×", n_a, "=",
    n_obs * n_l * n_a, "multiply-adds\n")
cat("  if-branch evaluations:     ", n_obs, "× 3 =", n_obs * 3, "\n")

cat("\nWeight likelihood:\n")
cat("  PLA multiplies:            ", n_wobs, "×", n_l, "×", n_a, "=",
    n_wobs * n_l * n_a, "multiply-adds\n")
cat("  Rebin multiplies:          ", n_wobs, "×", n_w, "×", n_l, "=",
    n_wobs * n_w * n_l, "multiply-adds\n")
cat("  if-branch evaluations:     ", n_wobs, "× 3 =", n_wobs * 3, "\n")

scalar_total <- n_f * n_a * n_yr * n_s + n_a * n_l + n_f * n_yr * 2 +
                n_wobs * n_w * n_l
cat("\nTotal estimated scalar tape ops: ~", scalar_total, "\n")
```

# Optimisation timing

Uncomment the block below to time a full `nlminb` optimisation run. This is
useful for measuring end-to-end improvement but takes several minutes:

```{r opt-timing, echo=TRUE, eval=FALSE}
Lwr <- rep(-Inf, length(obj$par))
Upr <- rep(Inf, length(obj$par))
Lwr[grep("log_B0", names(obj$par))] <- log(1)
Upr[grep("log_B0", names(obj$par))] <- log(exp(22))
Lwr[grep("log_cpue_q", names(obj$par))] <- log(0.1)
Upr[grep("log_cpue_q", names(obj$par))] <- log(10)
Lwr[grep("rdev_y", names(obj$par))] <- -5
Upr[grep("rdev_y", names(obj$par))] <- 5

control <- list(eval.max = 10000, iter.max = 10000)

t_opt <- system.time({
  opt <- nlminb(start = obj$par, objective = obj$fn, gradient = obj$gr,
                hessian = obj$he, lower = Lwr, upper = Upr, control = control)
})
cat("Optimization time: ", round(t_opt["elapsed"], 1), "sec\n")
cat("Convergence:       ", ifelse(opt$convergence == 0, "TRUE", "FALSE"), "\n")
cat("Final NLL:         ", round(opt$objective, 4), "\n")
cat("Function evals:    ", opt$evaluations["function"], "\n")
cat("Gradient evals:    ", opt$evaluations["gradient"], "\n")
cat("Sec per fn eval:   ", round(t_opt["elapsed"] / opt$evaluations["function"], 4), "\n")
cat("Sec per gr eval:   ", round(t_opt["elapsed"] / opt$evaluations["gradient"], 4), "\n")
```

# Save baseline

Save the current NLL, gradient, report outputs, and timing to an RDS file.
After each refactoring change, load this baseline and compare numerically to
verify equivalence:

```{r save-baseline, echo=TRUE}
baseline <- list(
  timestamp   = Sys.time(),
  description = "Pre-refactoring baseline",
  dimensions  = list(
    n_year    = data$n_year,
    n_age     = data$n_age,
    n_fishery = data$n_fishery,
    n_season  = data$n_season,
    n_len     = data$n_len,
    n_lf      = data$n_lf,
    n_wt      = data$n_wt,
    n_wf      = data$n_wf
  ),
  nll         = nll,
  max_gr      = max(abs(gr)),
  timing      = list(
    fn_median = as.numeric(bm_core$median[1]),
    gr_median = as.numeric(bm_core$median[2])
  ),
  report      = list(
    spawning_biomass_y = rep$spawning_biomass_y,
    catch_pred_fya     = rep$catch_pred_fya,
    lp_cpue            = rep$lp_cpue,
    lp_lf              = rep$lp_lf,
    lp_wf              = rep$lp_wf,
    number_ysa         = rep$number_ysa
  ),
  par_values  = obj$par
)

saveRDS(baseline, "opal_baseline.rds")
```

To compare after refactoring:

```{r compare-baseline, echo=TRUE, eval=FALSE}
old <- readRDS("opal_baseline.rds")

# Numerical equivalence (should be ~1e-14)
max(abs(rep$number_ysa - old$report$number_ysa))
max(abs(rep$catch_pred_fya - old$report$catch_pred_fya))
max(abs(rep$spawning_biomass_y - old$report$spawning_biomass_y))
abs(nll - old$nll)

# Speedup
cat("fn speedup:", round(old$timing$fn_median / as.numeric(bm_core$median[1]), 1), "x\n")
cat("gr speedup:", round(old$timing$gr_median / as.numeric(bm_core$median[2]), 1), "x\n")
```
