---
title: "Projections and uncertainty: MVN vs MCMC"
format: html
execute:
  eval: true
  echo: true
  message: false
  warning: false
vignette: >
  %\VignetteIndexEntry{Projections and uncertainty: MVN vs MCMC}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

# Overview

This vignette demonstrates the full simulation-estimation-projection workflow in
`opal`:

1. **Fit** a model to real data (WCPO bigeye tuna, CPUE likelihood only).
2. **Construct** an alternative catch scenario in which fishing effort is reduced
   by 50 % over the last 20 years, producing a biomass recovery signal.
3. **Simulate** a synthetic CPUE series from the recovery scenario, using the
   original MLE parameters and the modified catch — establishing a known "truth".
4. **Refit** the model to the simulated data so both MVN and MCMC approaches are
   evaluated against a known trajectory.
5. **Project** population dynamics forward under two approaches to parameter
   uncertainty:
   - **MVN simulation** — multivariate-normal draws from the Hessian-derived
     variance-covariance matrix; no MCMC required.
   - **MCMC posteriors** — draws from the posterior distribution obtained via
     SparseNUTS (code shown; set `eval: true` on those chunks to run).
6. **Compare** the resulting uncertainty envelopes against the true simulated
   trajectory.

See `vignette("bet")` for a detailed walkthrough of the BET model setup.

---

# Setup

```{r packages}
library(opal)
library(RTMB)
library(SparseNUTS)
library(ggplot2)
library(dplyr)
library(tidyr)
library(knitr)
library(posterior)

theme_set(theme_bw())
```

## Load data and parameters

```{r data}
data(wcpo_bet_data)
data(wcpo_bet_parameters)

data <- wcpo_bet_data

# Use CPUE likelihood only for speed
data$lf_switch   <- 0L
data$wf_switch   <- 0L
data$cpue_switch <- 1L

# Length bin bounds and midpoints (required by the PLA inside the model)
data$len_lower <- seq(from = data$len_bin_start, by = data$len_bin_width,
                      length.out = data$n_len)
data$len_upper <- data$len_lower + data$len_bin_width
data$len_mid   <- data$len_lower + data$len_bin_width / 2

# Removal switch: 0 = fishery active (required by project_selectivity)
data$removal_switch_f <- rep(0L, data$n_fishery)
```

```{r parameters}
parameters <- list(
  log_B0         = 20,
  log_h          = as.numeric(wcpo_bet_parameters$log_h),
  log_sigma_r    = as.numeric(wcpo_bet_parameters$log_sigma_r),
  log_cpue_q     = as.numeric(wcpo_bet_parameters$log_cpue_q),
  cpue_creep     = as.numeric(wcpo_bet_parameters$cpue_creep),
  log_cpue_tau   = as.numeric(wcpo_bet_parameters$log_cpue_tau),
  log_cpue_omega = as.numeric(wcpo_bet_parameters$log_cpue_omega),
  log_lf_tau     = log(rep(0.1, data$n_fishery)),
  log_wf_tau     = rep(0, data$n_fishery),
  log_L1         = as.numeric(wcpo_bet_parameters$log_L1),
  log_L2         = as.numeric(wcpo_bet_parameters$log_L2),
  log_k          = as.numeric(wcpo_bet_parameters$log_k),
  log_CV1        = as.numeric(wcpo_bet_parameters$log_CV1),
  log_CV2        = as.numeric(wcpo_bet_parameters$log_CV2),
  par_sel        = as.matrix(wcpo_bet_parameters$par_sel),
  rdev_y         = as.numeric(wcpo_bet_parameters$rdev_y)
)
```

Fix all parameters except `log_B0`, `log_cpue_q`, and the recruitment deviates
`rdev_y`:

```{r map}
map <- list(
  log_h          = factor(NA),
  log_sigma_r    = factor(NA),
  cpue_creep     = factor(NA),
  log_cpue_tau   = factor(NA),
  log_cpue_omega = factor(NA),
  log_lf_tau     = factor(rep(NA, data$n_fishery)),
  log_wf_tau     = factor(rep(NA, data$n_fishery)),
  log_L1         = factor(NA),
  log_L2         = factor(NA),
  log_k          = factor(NA),
  log_CV1        = factor(NA),
  log_CV2        = factor(NA),
  par_sel        = factor(matrix(NA, nrow(parameters$par_sel),
                                     ncol(parameters$par_sel)))
)
```

## Build and optimise

```{r make-obj}
#| output: false
#| message: false
obj <- MakeADFun(func = cmb(opal_model, data), parameters = parameters, map = map)

Lwr <- rep(-Inf, length(obj$par))
Upr <- rep( Inf, length(obj$par))
Lwr[grep("log_B0",     names(obj$par))] <- log(1)
Upr[grep("log_B0",     names(obj$par))] <- log(exp(22))
Lwr[grep("log_cpue_q", names(obj$par))] <- log(0.1)
Upr[grep("log_cpue_q", names(obj$par))] <- log(10)
Lwr[grep("rdev_y",     names(obj$par))] <- -5
Upr[grep("rdev_y",     names(obj$par))] <-  5

control <- list(eval.max = 5000, iter.max = 5000)
opt <- nlminb(start = obj$par, objective = obj$fn, gradient = obj$gr,
              lower = Lwr, upper = Upr, control = control)
opt <- nlminb(start = opt$par, objective = obj$fn, gradient = obj$gr,
              lower = Lwr, upper = Upr, control = control)

cat("Max gradient:", max(abs(obj$gr())), "\n")
cat("NLL:", opt$objective, "\n")
```

---

# Alternative catch scenario and simulation

To give the model a distinctive and informative signal about the current stock
state, we modify the catch time-series for the **last 80 quarters (20 years)** so that
effort declines linearly to 50 % of its original level by the terminal quarter.  The
resulting biomass recovery produces a rising CPUE trend that constrains the
current state more tightly, reducing posterior uncertainty and leading to
closer agreement between the MVN and MCMC projections.

```{r alt-catch}
data_alt <- data
n_yr     <- data$n_year

for (y in seq(n_yr - 79L, n_yr)) {
  # Linear ramp from 100 % (quarter n_yr − 79) to 50 % (terminal quarter)
  frac <- 1 - 0.5 * (y - (n_yr - 80L)) / 80
  data_alt$catch_obs_ysf[y, , ] <- data$catch_obs_ysf[y, , ] * frac
}
```

```{r alt-catch-plot}
#| fig-height: 3.5
#| fig-width: 7
#| fig-cap: "Total annual catch (all fisheries and seasons): original series (grey) and the recovery scenario with a 50 % reduction over the last 80 quarters (20 years) (blue)."
catch_orig <- rowSums(data$catch_obs_ysf,     dims = 1)
catch_alt  <- rowSums(data_alt$catch_obs_ysf, dims = 1)
years_all  <- seq_len(n_yr)
plot(years_all, catch_orig, type = "l", col = "grey50", lwd = 1,
     xlab = "Quarter", ylab = "Total catch (mt)",
     main = "Original vs reduced-catch scenario")
lines(years_all, catch_alt, col = "steelblue", lwd = 1.5)
abline(v = n_yr - 79, lty = 2, col = "grey70")
legend("topright", c("Original", "Reduced (recovery scenario)"),
       col = c("grey50", "steelblue"), lwd = c(1, 1.5))
```

Simulate one CPUE realisation from the recovery scenario.  We build an AD
object with the modified catch, evaluate it at the original MLE parameter
vector (`opt$par`) so the simulated trajectory reflects recovery under those
parameters, then draw a single CPUE realisation:

```{r simulate}
#| message: false
# Build model object with modified catch
obj_alt <- MakeADFun(func = cmb(opal_model, data_alt), parameters = parameters,
                     map = map)

# Evaluate at original MLE so internal state = recovery trajectory
invisible(obj_alt$fn(opt$par))

# True SSB trajectory under the recovery scenario (the "ground truth")
rep_true_alt <- obj_alt$report()
sbio_true    <- rep_true_alt$spawning_biomass_y   # length n_year + 1

# Draw one CPUE realisation consistent with the recovery trajectory
set.seed(42)
sim_alt <- obj_alt$simulate()

# Use modified catch + simulated CPUE as the scenario data
data_sim <- data_alt
data_sim$cpue_data$value <- exp(sim_alt$cpue_log_obs)
```

Compare the original and simulated CPUE series — the rising trend in the final
20 years is the recovery signal we want the model to detect:

```{r plot-sim}
#| fig-height: 4
#| fig-width: 7
#| fig-cap: "Original observed CPUE (grey) and simulated CPUE from the recovery scenario (blue). The rising trend in the final 20 years reflects the reduced catch."
plot(data$cpue_data$value, type = "l", col = "grey50", lwd = 1,
     xlab = "Time step", ylab = "CPUE",
     main = "Original (grey) vs recovery-scenario simulated (blue) CPUE")
lines(data_sim$cpue_data$value, col = "steelblue", lwd = 1.5)
abline(v = n_yr - 79, lty = 2, col = "grey70")
legend("topright", c("Original observed", "Recovery scenario"),
       col = c("grey50", "steelblue"), lwd = c(1, 1.5))
```

---

# Refit to simulated data

Build a new AD object from the simulated data and re-optimise from the original
starting values.  This recovers parameters consistent with the simulated
observations and gives us a known "truth" for the projection comparison:

```{r refit}
#| output: false
#| message: false
obj_sim <- MakeADFun(func = cmb(opal_model, data_sim),
                     parameters = parameters, map = map)

opt_sim <- nlminb(start = obj_sim$par, objective = obj_sim$fn,
                  gradient = obj_sim$gr,
                  lower = Lwr, upper = Upr, control = control)
opt_sim <- nlminb(start = opt_sim$par, objective = obj_sim$fn,
                  gradient = obj_sim$gr,
                  lower = Lwr, upper = Upr, control = control)

cat("Max gradient (refitted):", max(abs(obj_sim$gr())), "\n")
```

Add data-driven prior centres using the refitted MLE estimates, then rebuild
`obj_sim` so the AD tape carries the priors for MCMC sampling:

```{r wire-priors}
#| output: false
# Prior centres anchored to the refitted MLE (avoids the default par1=0 for
# log_B0 which would be wildly wrong for a stock of this size)
data_sim$prior_log_B0_mean     <- opt_sim$par["log_B0"]
data_sim$prior_log_cpue_q_mean <- opt_sim$par["log_cpue_q"]

# Build the priors list using the full structured parameters list so that
# get_priors() produces indices that match the RTMB AD tape layout.
# The data-driven centres above are used by get_priors(); the parameter
# values themselves only serve as fallback and do not matter here.
data_sim$priors <- get_priors(parameters, data_sim)

# Rebuild obj_sim with updated data (AD tape must capture data$priors)
obj_sim <- MakeADFun(func = cmb(opal_model, data_sim),
                     parameters = parameters, map = map)
invisible(obj_sim$fn(opt_sim$par))
```

Compute Hessian-based standard errors for derived quantities via the delta
method.  These are used later to draw the historical uncertainty ribbon:

```{r sdreport}
#| output: false
# ADREPORT(spawning_biomass_y) is registered inside do_dynamics(), so sdreport
# propagates Hessian uncertainty through the full historical trajectory via
# automatic differentiation.
sdr_sim   <- RTMB::sdreport(obj_sim)
sdr_sum   <- summary(sdr_sim, "report")
sbio_rows <- which(rownames(sdr_sum) == "spawning_biomass_y")
sbio_mle_sd <- sdr_sum[sbio_rows, "Estimate"]
sbio_se_sd  <- sdr_sum[sbio_rows, "Std. Error"]
```

Compare the true recovery-scenario trajectory against the trajectory estimated
by refitting to the simulated data.  Close agreement shows the model can
recover the true trajectory from the simulated observations:

```{r sbio-check}
#| fig-height: 4
#| fig-width: 7
rep_sim  <- obj_sim$report()

ylim <- range(c(sbio_true, rep_sim$spawning_biomass_y)) * c(0, 1.15)
plot(sbio_true, type = "l", col = "black", lwd = 2, lty = 2,
     ylim = ylim,
     xlab = "Time step", ylab = "Spawning biomass (mt)",
     main = "True recovery trajectory (black dashed) vs refitted MLE (blue)")
lines(rep_sim$spawning_biomass_y, col = "steelblue", lwd = 2)
legend("topright", c("True (recovery scenario)", "Refitted MLE"),
       col = c("black", "steelblue"), lwd = 2, lty = c(2, 1))
```

---

# Parameter estimates and uncertainty

The refitted model has three groups of free parameters: the unfished spawning
biomass (`log_B0`), the CPUE catchability coefficient (`log_cpue_q`), and the
annual log-recruitment deviates (`rdev_y`).  Standard errors are computed from
the square root of the diagonal of the inverse Hessian at the MLE.

```{r param-table}
H_sim   <- obj_sim$he()
Sigma   <- solve(H_sim)
SE_vec  <- sqrt(diag(Sigma))
mu_vec  <- obj_sim$env$last.par.best

# Extract scalar free parameters and convert to natural scale
scalar_params <- data.frame(
  Parameter  = c("B0 (mt)", "CPUE q"),
  MLE        = c(exp(mu_vec["log_B0"]),
                 exp(mu_vec["log_cpue_q"])),
  SE_log     = c(SE_vec[names(mu_vec) == "log_B0"][1],
                 SE_vec[names(mu_vec) == "log_cpue_q"][1]),
  CI_lo95    = c(exp(mu_vec["log_B0"]  - 1.96 * SE_vec[names(mu_vec) == "log_B0"][1]),
                 exp(mu_vec["log_cpue_q"] - 1.96 * SE_vec[names(mu_vec) == "log_cpue_q"][1])),
  CI_hi95    = c(exp(mu_vec["log_B0"]  + 1.96 * SE_vec[names(mu_vec) == "log_B0"][1]),
                 exp(mu_vec["log_cpue_q"] + 1.96 * SE_vec[names(mu_vec) == "log_cpue_q"][1]))
)
scalar_params[, 2:5] <- round(scalar_params[, 2:5], 4)
scalar_params$MLE[1]     <- round(scalar_params$MLE[1])
scalar_params$CI_lo95[1] <- round(scalar_params$CI_lo95[1])
scalar_params$CI_hi95[1] <- round(scalar_params$CI_hi95[1])
knitr::kable(scalar_params,
             col.names = c("Parameter", "MLE", "SE (log scale)", "95% CI lower", "95% CI upper"),
             caption   = "MLE estimates and Hessian-based 95% confidence intervals.")
```

The recruitment deviates `rdev_y` have one value per model year
(`r data_sim$n_year` parameters total).  Each deviate is centred on zero with a
prior `sd = sigma_r = `r round(rep_sim$sigma_r, 2)``.

```{r rdev-plot}
#| fig-height: 3.5
#| fig-width: 8
#| fig-cap: "Estimated log-recruitment deviates with approximate 95% point-wise confidence intervals (±1.96 × Hessian SE)."
rdev_idx <- which(names(mu_vec) == "rdev_y")
rdev_mle <- mu_vec[rdev_idx]
rdev_se  <- SE_vec[rdev_idx]
rdev_df  <- data.frame(
  year  = seq_len(data_sim$n_year),
  rdev  = rdev_mle,
  lo95  = rdev_mle - 1.96 * rdev_se,
  hi95  = rdev_mle + 1.96 * rdev_se
)
ggplot(rdev_df, aes(x = year)) +
  geom_ribbon(aes(ymin = lo95, ymax = hi95), fill = "steelblue", alpha = 0.25) +
  geom_line(aes(y = rdev), colour = "steelblue", linewidth = 0.6) +
  geom_hline(yintercept = 0, linetype = "dashed", colour = "grey50") +
  labs(x = "Model year", y = "log-recruitment deviate",
       title = "Estimated recruitment deviates (rdev_y)")
```

---

# Forward projections

## Common setup

```{r proj-setup}
#| output: false
n_proj <- 10   # projection quarters
n_iter <- 200  # uncertainty iterations

# Projected recruitment deviates — ARIMA-based simulation from the fitted series
rdev_proj <- project_rec_devs(
  data     = data_sim,
  obj      = obj_sim,
  first_yr = data_sim$first_yr,
  n_proj   = n_proj,
  n_iter   = n_iter
)

# Projected selectivity — replicate mean selectivity over the last 10 fitted years
sel_proj <- project_selectivity(
  data     = data_sim,
  obj      = obj_sim,
  first_yr = data_sim$last_yr - 9,
  n_proj   = n_proj,
  n_iter   = n_iter
)

# Catch scenario: hold catch constant at the last observed year's level
catch_ysf_proj <- array(0, dim = c(n_proj, data_sim$n_season, data_sim$n_fishery))
for (y in seq_len(n_proj)) {
  catch_ysf_proj[y, , ] <- data_sim$catch_obs_ysf[data_sim$n_year, , ]
}
```

## Approach 1: MVN simulation

When `mcmc = NULL`, `project_dynamics()` samples parameter sets from a
multivariate-normal distribution centred on the MLE, using the inverse Hessian
as the covariance matrix.  The sampling is performed on working (unconstrained)
scales — log for positive parameters, logit for autocorrelation parameters —
with the covariance adjusted via the delta method, then back-transformed to the
native parameter space before each call to `obj$report()`.

```{r proj-mvn}
#| output: false
proj_mvn <- project_dynamics(
  data      = data_sim,
  object    = obj_sim,
  mcmc      = NULL,          # triggers MVN path
  n_proj    = n_proj,
  n_iter    = n_iter,
  rdev_y    = rdev_proj$rdev_y,
  sel_fya   = sel_proj,
  catch_ysf = catch_ysf_proj
)
```

## Approach 2: MCMC posteriors

Run a No-U-Turn sampler (NUTS) chain via `SparseNUTS`.  The `bet_globals()`
call exports all helper functions needed for likelihood evaluation inside the
MCMC workers.

```{r mcmc-sampling}
#| output: false
n_chains_mcmc <- 8
mcmc_fit <- SparseNUTS::sample_snuts(
  obj         = obj_sim,
  globals     = bet_globals(),
  num_samples = 500,
  num_warmup  = 1000,
  chains      = n_chains_mcmc,
  cores       = min(n_chains_mcmc, parallel::detectCores()),
  metric      = "dense"     # full Hessian-based mass matrix — captures B0/q/rdev_y correlations
)
```

Use the posterior draws for projection.  The same pre-generated recruitment
deviates and selectivity arrays are reused so that the only difference between
the two approaches is how historical parameter uncertainty is represented:

```{r proj-mcmc}
#| output: false
proj_mcmc <- project_dynamics(
  data        = data_sim,
  object      = obj_sim,
  mcmc        = mcmc_fit,      # triggers MCMC path
  n_proj      = n_proj,
  n_iter      = n_iter,
  rdev_y      = rdev_proj$rdev_y,
  sel_fya     = sel_proj,
  catch_ysf   = catch_ysf_proj,
  return_hist = TRUE           # collect historical SB per MCMC draw
)
```

## MCMC diagnostics

Before comparing projection envelopes, we check whether the NUTS chains have
converged.  With 1000 warmup steps and 8 independent chains, the dense-metric
SparseNUTS sampler adapts a full Hessian-based mass matrix that captures
correlations between B0, q, and the rdev_y parameters in the 270-dimensional
parameter space.  Split-chain Rhat and per-chain ESS
provide reliable convergence diagnostics across chains.

```{r extract-post}
# Extract posterior draws (post-warmup only, all chains merged row-by-row)
post_samples  <- SparseNUTS::extract_samples(mcmc_fit)
B0_mcmc       <- exp(post_samples[, "log_B0"])
# Subset to the n_iter draws that project_dynamics used (for per-draw depletion)
B0_mcmc_iter  <- B0_mcmc[seq_len(n_iter)]

# Reconstruct chain / iteration labels (chains concatenated in order)
n_post_total  <- nrow(post_samples)
n_per_chain   <- n_post_total / n_chains_mcmc
chain_id      <- rep(seq_len(n_chains_mcmc), each = n_per_chain)
iter_id       <- rep(seq_len(n_per_chain),   times = n_chains_mcmc)

# NUTS-level diagnostics: divergences and max-treedepth violations
SparseNUTS::check_snuts_diagnostics(mcmc_fit)
```

```{r mcmc-traces}
#| fig-height: 5
#| fig-width: 8
#| fig-cap: "MCMC traces for scalar parameters across all chains (colours). Red dashed = MLE estimate. Chains overlapping and stationary around a common mean indicates good convergence."
trace_df <- data.frame(
  iter       = iter_id,
  chain      = factor(chain_id),
  log_B0     = post_samples[, "log_B0"],
  log_cpue_q = post_samples[, "log_cpue_q"]
) |>
  tidyr::pivot_longer(-c(iter, chain), names_to = "param", values_to = "value")

mle_ref <- data.frame(
  param = c("log_B0", "log_cpue_q"),
  mle   = mu_vec[c("log_B0", "log_cpue_q")]
)

ggplot(trace_df, aes(x = iter, y = value, colour = chain)) +
  geom_line(linewidth = 0.3, alpha = 0.7) +
  geom_hline(data = mle_ref, aes(yintercept = mle),
             linetype = "dashed", colour = "red", linewidth = 0.6) +
  facet_wrap(~ param, scales = "free_y", ncol = 1,
             labeller = as_labeller(
               c(log_B0 = "log~B[0]", log_cpue_q = "log~q[CPUE]"),
               default = label_parsed)) +
  scale_colour_discrete(name = "Chain") +
  labs(x = "Post-warmup iteration", y = "Parameter value",
       title = "MCMC traces — scalar parameters",
       subtitle = sprintf("Red dashed = MLE; %d post-warmup samples × %d chains",
                          n_per_chain, n_chains_mcmc))
```

Without informative priors on `log_B0` and `log_cpue_q`, the joint likelihood
has a ridge: high $B_0$ paired with low $q$ fits the CPUE as well as low $B_0$
with high $q$.  MCMC correctly samples this flat posterior, but the posterior
mean can lie far from the MLE (which sits at the mode).  Enabling data-driven
priors (centred on the refitted MLE) regularises the posterior toward the MLE.
The prior vs posterior comparison below illustrates this.

```{r mcmc-diag-table}
# Scalar parameters + 5 evenly-spaced rdev_y as a spread-check
rdev_idx_all  <- grep("rdev_y", colnames(post_samples))
rdev_idx_show <- rdev_idx_all[round(seq(1, length(rdev_idx_all), length.out = 5))]
diag_cols     <- c("log_B0", "log_cpue_q", colnames(post_samples)[rdev_idx_show])

# Build draws_df with chain / iteration metadata for correct split-chain Rhat
draws_df <- as.data.frame(post_samples[, diag_cols, drop = FALSE])
draws_df[[".chain"]]     <- chain_id
draws_df[[".iteration"]] <- iter_id
draws_obj <- posterior::as_draws_df(draws_df)
diag_tbl  <- posterior::summarise_draws(
  draws_obj,
  posterior::default_summary_measures(),
  posterior::default_convergence_measures()
)
# Add MLE column for easy comparison
diag_tbl$mle <- mu_vec[match(diag_tbl$variable, names(mu_vec))]

knitr::kable(
  diag_tbl[, c("variable", "mle", "mean", "sd", "q5", "q95",
               "ess_bulk", "ess_tail", "rhat")],
  digits  = 3,
  caption = paste0(
    "MCMC diagnostics for key parameters (split-chain Rhat and ESS). ",
    "'mle' = MLE estimate for comparison — large mle-mean discrepancy ",
    "indicates prior-likelihood tension or poor mixing. ",
    "ESS < 100 or Rhat > 1.01 indicate poor convergence."
  )
)
```

```{r prior-post-plot}
#| fig-height: 4
#| fig-width: 8
#| fig-cap: "Prior (blue curve) vs MCMC posterior (grey histogram) for the two confounded scalar parameters. Red dashed line = MLE estimate. Priors are centred on the MLE with sd = 1.0 (log B0) and 1.5 (log q). Overlap of the posterior with the prior and agreement between the posterior mean and MLE indicate that the priors are doing their job of regularising the B0/q ridge."
prior_params <- list(
  log_B0     = list(mean = mu_vec["log_B0"],     sd = 1.0),
  log_cpue_q = list(mean = mu_vec["log_cpue_q"], sd = 1.5)
)
par_labels <- c(log_B0 = "log~B[0]", log_cpue_q = "log~q[CPUE]")

prior_lines <- do.call(rbind, lapply(names(prior_params), function(nm) {
  pp <- prior_params[[nm]]
  x  <- post_samples[, nm]
  xr <- range(c(x, pp$mean + c(-4, 4) * pp$sd))
  xs <- seq(xr[1], xr[2], length.out = 300)
  data.frame(x = xs, density = dnorm(xs, pp$mean, pp$sd), param = nm)
}))
post_points <- do.call(rbind, lapply(names(prior_params), function(nm) {
  data.frame(x = post_samples[, nm], param = nm)
}))
mle_ref2 <- data.frame(
  param = names(prior_params),
  mle   = mu_vec[names(prior_params)]
)

ggplot() +
  geom_histogram(data = post_points,
                 aes(x = x, y = after_stat(density)),
                 bins = 30, fill = "grey70", colour = "white") +
  geom_line(data = prior_lines,
            aes(x = x, y = density),
            colour = "steelblue", linewidth = 0.8) +
  geom_vline(data = mle_ref2, aes(xintercept = mle),
             colour = "red", linetype = "dashed", linewidth = 0.7) +
  facet_wrap(~ param, scales = "free",
             labeller = as_labeller(par_labels, default = label_parsed)) +
  labs(x = "Parameter value", y = "Density",
       title = "Prior (blue) vs MCMC posterior (grey) for scalar parameters",
       subtitle = "Red dashed = MLE estimate")
```

---

# Comparison: MVN vs MCMC uncertainty

Extract projected spawning biomass from each iteration and compute quantile
envelopes.

```{r extract-sbio}
spa <- obj_sim$report()$spawning_potential_a  # [n_age] maturity × fecundity

# Helper: turn a list of do_dynamics() results into a long data frame.
# number_ysa[1,,] is the initial state (= model end-state); include it as
# proj_year = 0 so it bridges continuously to the historical trajectory.
sbio_df <- function(proj_list, label) {
  do.call(rbind, lapply(seq_along(proj_list), function(i) {
    dyn <- proj_list[[i]]
    sb  <- sapply(1:(n_proj + 1), function(y) sum(dyn$number_ysa[y, 1, ] * spa))
    data.frame(iter = i, proj_year = 0:n_proj, sbio = sb, method = label)
  }))
}

df_mvn  <- sbio_df(proj_mvn,       "MVN")   # plain list (no return_hist)
df_mcmc <- sbio_df(proj_mcmc$dyn,  "MCMC")  # list(dyn=..., hist_sbio=...)
df_all  <- rbind(df_mvn, df_mcmc)
```

```{r summarise-proj}
# MLE historical trajectory (from sdreport estimates, identical to rep_sim$spawning_biomass_y)
hist_df <- data.frame(
  timestep = seq_len(data_sim$n_year + 1),
  sbio     = sbio_mle_sd[seq_len(data_sim$n_year + 1)]
)

n_hist_show <- 100  # 100 quarters ≈ 25 years (covers the full 80-quarter recovery window)
idx         <- (data_sim$n_year + 1 - n_hist_show + 1):(data_sim$n_year + 1)

# MVN historical ribbon: delta-method Gaussian CI from the Hessian (no Jensen bias).
# Drawing all 268 rdev_y from their MVN marginals and propagating through the
# exponential recruitment function inflates E[SBY] ≈ 76 % above the MLE — the
# delta method avoids this by linearising at the MLE.
df_hist_mvn <- data.frame(
  timestep = idx,
  med      = sbio_mle_sd[idx],
  lo50     = sbio_mle_sd[idx] - qnorm(0.75) * sbio_se_sd[idx],
  hi50     = sbio_mle_sd[idx] + qnorm(0.75) * sbio_se_sd[idx],
  lo90     = sbio_mle_sd[idx] - qnorm(0.95) * sbio_se_sd[idx],
  hi90     = sbio_mle_sd[idx] + qnorm(0.95) * sbio_se_sd[idx],
  method   = "MVN"
)

# MCMC historical ribbon: actual posterior trajectory quantiles
hist_ribbon <- function(hist_sbio, label, n_show) {
  n_ts <- ncol(hist_sbio)
  idx2 <- (n_ts - n_show + 1):n_ts
  do.call(rbind, lapply(idx2, function(t) {
    v <- hist_sbio[, t]
    data.frame(timestep = t,
               med  = median(v),
               lo50 = quantile(v, 0.25), hi50 = quantile(v, 0.75),
               lo90 = quantile(v, 0.05), hi90 = quantile(v, 0.95),
               method = label)
  }))
}
df_hist_mcmc <- hist_ribbon(proj_mcmc$hist_sbio, "MCMC", n_hist_show)

df_hist_ribbon <- rbind(df_hist_mvn, df_hist_mcmc)

# Projection quantile ribbons (proj_year = 0 is the bridge / join point)
df_ribbon <- df_all |>
  group_by(method, proj_year) |>
  summarise(
    med  = median(sbio),
    lo50 = quantile(sbio, 0.25),
    hi50 = quantile(sbio, 0.75),
    lo90 = quantile(sbio, 0.05),
    hi90 = quantile(sbio, 0.95),
    .groups = "drop"
  ) |>
  mutate(timestep = data_sim$n_year + 1 + proj_year)

hist_zoom <- hist_df[hist_df$timestep > (data_sim$n_year + 1 - n_hist_show), ]

# "True" SSB from the recovery scenario: obj_alt evaluated at opt$par
# (the trajectory that actually generated the simulated CPUE observations)
hist_zoom_true <- data.frame(
  timestep = idx,
  sbio     = sbio_true[idx]
)

# Depletion = SBY / B0
B0_est  <- rep_sim$B0    # estimated MLE B0 from obj_sim
B0_true <- rep_true_alt$B0  # true B0 used to generate the simulated data

# MVN historical depletion: divide by B0_est (conditional-expectation approach
# keeps B0 as the primary scalar uncertainty; approximately correct)
df_hist_mvn_dep  <- df_hist_mvn |>
  mutate(across(c(med, lo50, hi50, lo90, hi90), ~ . / B0_est))

# MCMC historical depletion: per-draw B0 (B0_mcmc_iter defined in extract-post)
hist_sbio_dep <- sweep(proj_mcmc$hist_sbio, 1, B0_mcmc_iter, "/")
df_hist_mcmc_dep <- hist_ribbon(hist_sbio_dep, "MCMC", n_hist_show)
df_hist_ribbon_dep <- rbind(df_hist_mvn_dep, df_hist_mcmc_dep)

# MVN projection depletion: divide by B0_est
df_ribbon_mvn_dep <- df_ribbon |>
  filter(method == "MVN") |>
  mutate(across(c(med, lo50, hi50, lo90, hi90), ~ . / B0_est))

# MCMC projection depletion: per-draw B0
df_ribbon_mcmc_dep <- df_mcmc |>
  mutate(sbio = sbio / B0_mcmc[iter]) |>
  group_by(method, proj_year) |>
  summarise(
    med  = median(sbio),
    lo50 = quantile(sbio, 0.25), hi50 = quantile(sbio, 0.75),
    lo90 = quantile(sbio, 0.05), hi90 = quantile(sbio, 0.95),
    .groups = "drop"
  ) |>
  mutate(timestep = data_sim$n_year + 1 + proj_year)

df_ribbon_dep <- rbind(df_ribbon_mvn_dep, df_ribbon_mcmc_dep)

# MLE and true depletion reference lines
hist_zoom_dep      <- hist_zoom      |> mutate(sbio = sbio / B0_est)
hist_zoom_true_dep <- hist_zoom_true |> mutate(sbio = sbio / B0_true)
```

```{r mcmc-traj-diag}
# Sanity checks for MCMC trajectory vs MLE
n_t <- ncol(proj_mcmc$hist_sbio)  # = n_year + 1

# 1. First timestep: SBY[1] / B0 should equal 1.0 (unfished equilibrium)
sby1_B0_ratio <- proj_mcmc$hist_sbio[, 1] / B0_mcmc_iter
cat("Check 1: SBY[t=1] / B0 (unfished eq, should be ~1.0 for all draws)\n")
cat(sprintf("  MLE: %.4f\n", rep_sim$spawning_biomass_y[1] / B0_est))
cat(sprintf("  MCMC: mean=%.4f, median=%.4f, sd=%.4f\n",
            mean(sby1_B0_ratio), median(sby1_B0_ratio), sd(sby1_B0_ratio)))

# 2. Terminal timestep depletion
dep_mle  <- sbio_mle_sd[n_t] / B0_est
dep_mcmc <- proj_mcmc$hist_sbio[, n_t] / B0_mcmc_iter
cat("\nCheck 2: Terminal depletion SBY[last] / B0\n")
cat(sprintf("  MLE:  %.4f\n", dep_mle))
cat(sprintf("  MCMC: mean=%.4f, median=%.4f, sd=%.4f\n",
            mean(dep_mcmc), median(dep_mcmc), sd(dep_mcmc)))

# 3. Absolute SBY at terminal timestep (before B0 division)
cat("\nCheck 3: Absolute SBY at terminal timestep\n")
cat(sprintf("  MLE SBY[last]:  %.1f\n", sbio_mle_sd[n_t]))
cat(sprintf("  MCMC SBY[last]: mean=%.1f, median=%.1f\n",
            mean(proj_mcmc$hist_sbio[, n_t]), median(proj_mcmc$hist_sbio[, n_t])))
cat(sprintf("  MLE B0:  %.1f\n", B0_est))
cat(sprintf("  MCMC B0: mean=%.1f, median=%.1f\n",
            mean(B0_mcmc), median(B0_mcmc)))

# 4. rdev_y: are MCMC posterior means close to MLE?
rdev_y_mle  <- mu_vec[names(mu_vec) == "rdev_y"]
post_mat    <- SparseNUTS::extract_samples(mcmc_fit)
rdev_cols   <- grep("rdev_y", colnames(post_mat))
rdev_y_post_mean <- colMeans(post_mat[, rdev_cols])
cat("\nCheck 4: rdev_y MLE vs MCMC mean (first 5 years)\n")
cat(sprintf("  MLE:       %s\n", paste(round(rdev_y_mle[1:5], 3), collapse = "  ")))
cat(sprintf("  MCMC mean: %s\n", paste(round(rdev_y_post_mean[1:5], 3), collapse = "  ")))
cat(sprintf("  Max |MLE - MCMC mean| across all rdev_y: %.4f\n",
            max(abs(rdev_y_mle - rdev_y_post_mean))))
```

```{r rdev-diag}
#| fig-height: 4
#| fig-width: 10
#| fig-cap: "Recruitment deviates (log scale) for MLE (black) vs MCMC posterior (blue ribbon = 90% CI, blue line = mean). Horizontal dashed lines = ±sigma_r."

# A. rdev_y scale confirmation
# rdev_y is on the LOG scale: get_recruitment() uses exp(rdev - 0.5*sigma_r^2).
# rdev_y = 0 → median recruitment; rdev_y = sigma_r → e^0.6 ≈ 1.82 × median.
sigma_r_val <- exp(parameters$log_sigma_r)
cat("A. rdev_y scale (log scale):\n")
cat(sprintf("  sigma_r = %.3f\n", sigma_r_val))
cat(sprintf("  MLE rdev_y[1]  = %.3f  (%.2f sigma_r from 0)\n",
            rdev_y_mle[1], rdev_y_mle[1] / sigma_r_val))
cat(sprintf("  MCMC mean rdev_y[1] = %.3f  (%.2f sigma_r from 0)\n",
            rdev_y_post_mean[1], rdev_y_post_mean[1] / sigma_r_val))

# B. NLL at MLE, MCMC posterior mean, and MCMC posterior mode
mcmc_mean_par  <- colMeans(post_mat)
nll_mle        <- obj_sim$fn(opt_sim$par)
nll_mcmc_mean  <- obj_sim$fn(mcmc_mean_par)

# Posterior mode = MCMC sample with the lowest NLL (MAP estimate from chain)
nll_post      <- apply(post_mat, 1, function(par) obj_sim$fn(as.numeric(par)))
mode_idx      <- which.min(nll_post)
mcmc_mode_par <- as.numeric(post_mat[mode_idx, ])
nll_mcmc_mode <- nll_post[mode_idx]

rdev_prior_mle  <- sum(dnorm(rdev_y_mle,                                0, sigma_r_val, log = TRUE))
rdev_prior_mean <- sum(dnorm(mcmc_mean_par[names(mu_vec) == "rdev_y"],  0, sigma_r_val, log = TRUE))
rdev_prior_mode <- sum(dnorm(mcmc_mode_par[names(mu_vec) == "rdev_y"],  0, sigma_r_val, log = TRUE))

cat("\nB. NLL penalty comparison:\n")
cat(sprintf("  NLL at MLE:               %.2f\n", nll_mle))
cat(sprintf("  NLL at MCMC post. mode:   %.2f  (diff vs MLE: %+.2f)\n",
            nll_mcmc_mode, nll_mcmc_mode - nll_mle))
cat(sprintf("  NLL at MCMC post. mean:   %.2f  (diff vs MLE: %+.2f)\n",
            nll_mcmc_mean, nll_mcmc_mean - nll_mle))
cat(sprintf("  rdev log-prior at MLE:           %.2f\n", rdev_prior_mle))
cat(sprintf("  rdev log-prior at MCMC mode:     %.2f  (diff: %+.2f)\n",
            rdev_prior_mode, rdev_prior_mode - rdev_prior_mle))
cat(sprintf("  rdev log-prior at MCMC mean:     %.2f  (diff: %+.2f)\n",
            rdev_prior_mean, rdev_prior_mean - rdev_prior_mle))

# C. rdev_y pattern plot
rdev_q05 <- apply(post_mat[, rdev_cols], 2, quantile, 0.05)
rdev_q95 <- apply(post_mat[, rdev_cols], 2, quantile, 0.95)
yr_seq   <- seq_len(data_sim$n_year)

df_rdev <- data.frame(
  year      = yr_seq,
  mle       = rdev_y_mle,
  mcmc_mean = rdev_y_post_mean,
  mcmc_q05  = rdev_q05,
  mcmc_q95  = rdev_q95
)

ggplot(df_rdev, aes(x = year)) +
  geom_ribbon(aes(ymin = mcmc_q05, ymax = mcmc_q95), fill = "steelblue", alpha = 0.2) +
  geom_line(aes(y = mcmc_mean), colour = "steelblue", linewidth = 0.7) +
  geom_line(aes(y = mle), colour = "black", linewidth = 0.8) +
  geom_hline(yintercept = c(-sigma_r_val, sigma_r_val),
             linetype = "dashed", colour = "grey50", linewidth = 0.4) +
  labs(x = "Year (index)", y = "rdev_y (log scale)",
       title = "Recruitment deviates: MLE (black) vs MCMC posterior mean + 90% CI (blue)")
```

```{r comparison-plot}
#| fig-height: 5
#| fig-width: 8
#| fig-cap: "Spawning biomass over the last 100 time-steps and 10-time-step projection. Black dashed = true SSB used to generate simulated data. Grey solid = estimated MLE (obj_sim). Coloured ribbons: 50 % (darker) and 90 % (lighter) uncertainty envelopes by method. Orange median line shows MCMC posterior central tendency in the historical period. Vertical dashed line = end of historical period."
pal <- c(MVN = "steelblue", MCMC = "darkorange")

ggplot() +
  # Historical 90 % ribbons — coloured by method
  geom_ribbon(data = df_hist_ribbon,
              aes(x = timestep, ymin = lo90, ymax = hi90, fill = method),
              alpha = 0.12) +
  # Historical 50 % ribbons
  geom_ribbon(data = df_hist_ribbon,
              aes(x = timestep, ymin = lo50, ymax = hi50, fill = method),
              alpha = 0.25) +
  # MCMC historical median line (coloured; MVN median = MLE so not separately shown)
  geom_line(data = subset(df_hist_ribbon, method == "MCMC"),
            aes(x = timestep, y = med, colour = method),
            linewidth = 0.75) +
  # Estimated MLE trajectory (obj_sim)
  geom_line(data = hist_zoom,
            aes(x = timestep, y = sbio),
            colour = "grey20", linewidth = 0.9) +
  # True simulated SSB (ground truth from recovery scenario)
  geom_line(data = hist_zoom_true,
            aes(x = timestep, y = sbio),
            colour = "black", linewidth = 0.8, linetype = "dashed") +
  # End of historical period
  geom_vline(xintercept = data_sim$n_year + 1,
             linetype = "dashed", colour = "grey50") +
  # Projection 90 % ribbons
  geom_ribbon(data = df_ribbon,
              aes(x = timestep, ymin = lo90, ymax = hi90, fill = method),
              alpha = 0.15) +
  # Projection 50 % ribbons
  geom_ribbon(data = df_ribbon,
              aes(x = timestep, ymin = lo50, ymax = hi50, fill = method),
              alpha = 0.30) +
  # Projection median lines
  geom_line(data = df_ribbon,
            aes(x = timestep, y = med, colour = method),
            linewidth = 1) +
  scale_fill_manual(values = pal) +
  scale_colour_manual(values = pal) +
  labs(
    x        = "Time step",
    y        = "Spawning biomass (mt)",
    colour   = "Method",
    fill     = "Method",
    title    = "Spawning biomass: historical uncertainty + projections",
    subtitle = paste0(
      "Last ", n_hist_show, " time-steps shown; ",
      n_proj, "-time-step constant-catch projection; ",
      n_iter, " draws per method"
    )
  )
```

```{r depletion-plot}
#| fig-height: 5
#| fig-width: 8
#| fig-cap: "Relative depletion (SB / B₀) over the last 100 time-steps and projection. All model-output quantities are divided by the estimated B₀ from obj_sim; the true-trajectory line uses the true B₀ from the recovery scenario. Dotted red line = 20 % B₀ limit reference point."
ggplot() +
  geom_ribbon(data = df_hist_ribbon_dep,
              aes(x = timestep, ymin = lo90, ymax = hi90, fill = method),
              alpha = 0.12) +
  geom_ribbon(data = df_hist_ribbon_dep,
              aes(x = timestep, ymin = lo50, ymax = hi50, fill = method),
              alpha = 0.25) +
  geom_line(data = subset(df_hist_ribbon_dep, method == "MCMC"),
            aes(x = timestep, y = med, colour = method),
            linewidth = 0.75) +
  geom_line(data = hist_zoom_dep,
            aes(x = timestep, y = sbio),
            colour = "grey20", linewidth = 0.9) +
  geom_line(data = hist_zoom_true_dep,
            aes(x = timestep, y = sbio),
            colour = "black", linewidth = 0.8, linetype = "dashed") +
  geom_vline(xintercept = data_sim$n_year + 1,
             linetype = "dashed", colour = "grey50") +
  geom_ribbon(data = df_ribbon_dep,
              aes(x = timestep, ymin = lo90, ymax = hi90, fill = method),
              alpha = 0.15) +
  geom_ribbon(data = df_ribbon_dep,
              aes(x = timestep, ymin = lo50, ymax = hi50, fill = method),
              alpha = 0.30) +
  geom_line(data = df_ribbon_dep,
            aes(x = timestep, y = med, colour = method),
            linewidth = 1) +
  geom_hline(yintercept = 0.2, linetype = "dotted",
             colour = "red", linewidth = 0.5) +
  coord_cartesian(ylim = c(0, NA)) +
  scale_fill_manual(values = pal) +
  scale_colour_manual(values = pal) +
  labs(
    x        = "Time step",
    y        = expression("Relative depletion (SB / B"[0]*")"),
    colour   = "Method",
    fill     = "Method",
    title    = "Relative depletion: historical uncertainty + projections",
    subtitle = paste0(
      "Last ", n_hist_show, " time-steps shown; ",
      n_proj, "-time-step constant-catch projection; ",
      n_iter, " draws per method"
    )
  )
```

**Reading the bridge point.**

*MVN projections (blue)*: `log_B0` and `log_cpue_q` are drawn from the MVN;
each `rdev_y` is then set to its conditional expectation given those scalar draws,
`E[rdev_y | scalar_draw] = μ + Σ_{rdev,sc} Σ_{sc,sc}⁻¹ (draw − μ)`.
This accounts for the negative `log_B0` ↔ `rdev_y` correlation in the Hessian
(higher B₀ requires lower historical deviates to fit the same declining
trajectory) and avoids a Jensen's-inequality bias — drawing all 268 deviates
simultaneously from their marginal distributions would inflate E[SBY] ≈ 76 %
above the MLE.  The bridge-point ribbon therefore aligns with the delta-method
(sdreport) historical ribbon in the left panel, both centred on the MLE and
spanning similar uncertainty.

*MCMC projections (orange)*: each draw comes from the full joint posterior,
including all `rdev_y`.  The bridge-point spread therefore reflects the actual
posterior uncertainty in the current biomass state.

In both cases, future process uncertainty (projected recruitment deviates and
selectivity) widens the bands further over the projection horizon.

**Why is the future uncertainty moderate?**  Because bigeye tuna mature slowly
(spawning potential is near zero for ages 1–4), new recruits take several years
before they contribute meaningfully to spawning biomass.  Under constant catch the
existing adult stock declines almost deterministically; the divergence from future
recruitment and selectivity uncertainty accumulates gradually.  Longer projection
horizons or more variable catch scenarios will produce wider bands.

```{r summary-table}
# Numerical summary at the end of the projection
df_all |>
  filter(proj_year == n_proj) |>
  group_by(method) |>
  summarise(
    mean   = mean(sbio),
    sd     = sd(sbio),
    p05    = quantile(sbio, 0.05),
    p25    = quantile(sbio, 0.25),
    median = median(sbio),
    p75    = quantile(sbio, 0.75),
    p95    = quantile(sbio, 0.95),
    .groups = "drop"
  )
```
