---
title: "Project a saved assessment"
format: html
execute:
  eval: true
  warning: false
  message: false
vignette: >
  %\VignetteIndexEntry{Project a saved assessment}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

This example carries one `opal_obj` from a saved fixed-effects assessment
through a five-year projection and a save/read round trip. Ten trajectories
keep the example quick; they demonstrate the interface and do not provide
assessment-quality uncertainty estimates. See the [Quickstart](quickstart.qmd)
for fitting and the [saved-fit review](opakapaka-fit-review.qmd) for diagnostics.

## Load and check the assessment

```{r}
#| label: load-assessment
library(opal)
assessment <- opal_read(system.file(
  "extdata", "opaka_quickstart_fit.rds", package = "opal"
))
assessment <- opal_check(assessment, scope = "fit")
summary(assessment)
```

## Choose future inputs

This illustrative scenario repeats the final observed annual catch, uses
the last ten model years to generate future selectivity, and simulates
recruitment deviations from an autoregressive model of the fitted history.
These are explicit scenario assumptions that require review for each stock.

```{r}
#| label: future-inputs
set.seed(42)
n_proj <- 5L
n_iter <- 10L
data <- assessment$data

recruitment <- project_rec_devs(
  assessment, uncertainty = "fit", arima = FALSE,
  first_yr = data$last_yr - 9L, n_proj = n_proj, n_iter = n_iter
)
selectivity <- project_selectivity(
  assessment, first_yr = data$last_yr - 9L,
  n_proj = n_proj, n_iter = n_iter
)
future_catch <- array(0, c(n_proj, data$n_season, data$n_fishery))
for (year in seq_len(n_proj)) {
  future_catch[year, , ] <- data$catch_obs_ysf[data$n_year, , ]
}
```

Recruitment has dimensions `[iteration, future year]`; selectivity has
`[iteration, fishery, future year, age]`; catch has
`[future year, season, fishery]`. Historical windows use the model-year
coordinates in `data$first_yr` and `data$last_yr`; the bundled Quickstart uses
indices 1--75, rather than calendar years. `project_selectivity()` uses the stored
point report and simulates variation independently by age and fishery; it
does not propagate posterior uncertainty in selectivity parameters.

## Project and retain the result

`uncertainty = "mvn"` draws parameters using the Hessian at the fitted
point. This path supports fixed-effects models. `opal_project()` retains
the output, future inputs, seed, and source identity inside the assessment.
Its seed controls the dynamics draws; the seed above controls recruitment
and selectivity generation.

```{r}
#| label: project
#| results: hide
assessment <- opal_project(
  assessment, name = "constant_catch", uncertainty = "mvn", seed = 43,
  n_proj = n_proj, n_iter = n_iter,
  rdev_y = recruitment$rdev_y, sel_fya = selectivity,
  catch_ysf = future_catch, return_hist = TRUE
)
```

```{r}
#| label: inspect-projection
projection <- assessment$derived$constant_catch$result
names(projection)
assessment$derived$constant_catch$uncertainty
length(projection$dyn)
```

The first projected state is the terminal historical state for the same
parameter draw. Include this bridge point when plotting a continuous
trajectory from the historical period into the future.

```{r}
#| label: projection-plot
#| fig-width: 8
#| fig-height: 4
spawning_biomass <- vapply(
  projection$dyn, function(draw) as.numeric(draw$spawning_biomass_y),
  numeric(n_proj + 1L)
)
matplot(0:n_proj, spawning_biomass,
        type = "l", lty = 1, col = grDevices::adjustcolor("#00797F", 0.4),
        xlab = "Projection year (0 = terminal historical state)",
        ylab = "Spawning biomass")
```

## Save the assessment and its projection

```{r}
#| label: save-projection
path <- tempfile(fileext = ".rds")
opal_save(assessment, path)
restored <- opal_read(path)
stopifnot(identical(restored$derived$constant_catch$result, projection))
unlink(path)
```

Changing model configuration through `opal_update()`, or replacing the
source fit or posterior, clears stored projections. Change a label through
`metadata` to preserve compatible results.

## Use posterior draws

For an assessment with MCMC, select the uncertainty source explicitly.
Recruitment and dynamics consume the same first `n_iter` retained draws in
the same order. Random-effect models require complete joint posterior
draws; marginal-only draws and random-effect MVN projections are unsupported.

The following recipe is not executed during documentation builds. Run it
after choosing sampling effort and reviewing the MCMC diagnostics.

```{r}
#| eval: false
assessment <- opal_mcmc(assessment, seed = 44, chains = 4, cores = 4,
                        num_warmup = 1000, num_samples = 500)
summary(assessment)
assessment$validation$mcmc

set.seed(45)
recruitment <- project_rec_devs(
  assessment, uncertainty = "mcmc", arima = FALSE,
  first_yr = data$last_yr - 9L, n_proj = n_proj, n_iter = n_iter
)
assessment <- opal_project(
  assessment, name = "posterior_constant_catch", uncertainty = "mcmc",
  seed = 46, n_proj = n_proj, n_iter = n_iter,
  rdev_y = recruitment$rdev_y, sel_fya = selectivity,
  catch_ysf = future_catch, return_hist = TRUE
)
opal_save(assessment, "assessment-projected.rds")
```
