---
title: "'Opakapaka projections: MVN vs MCMC uncertainty"
format: html
execute:
  eval: false
vignette: >
  %\VignetteIndexEntry{'Opakapaka projections: MVN vs MCMC uncertainty}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{quarto::html}
editor_options:
  chunk_output_type: console
---

```{r}
#| include: false
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

# Overview

This vignette carries the 'opakapaka (*Pristipomoides filamentosus*) case study
forward from estimation into projection. It reuses the *opal with length
compositions* model (`obj_lf`) developed in `vignette("opakapaka")` — the
assessment that fits both CPUE indices and length compositions and estimates
selectivity — and projects the stock forward under a constant-catch scenario,
propagating parameter uncertainty two ways:

1. **MVN simulation** — multivariate-normal draws from the Hessian-derived
   variance-covariance matrix at the MLE; no MCMC required.
2. **MCMC posteriors** — draws from the joint posterior obtained via
   `SparseNUTS` (code shown; a cached fit is loaded by default).

The historical spawning-biomass uncertainty is shown with the delta method
(`sdreport`) for the MVN case and with posterior trajectory quantiles for the
MCMC case. Both are bridged continuously into the projection period so the
terminal-state uncertainty and the future process uncertainty are visible on a
single axis. The Stock Synthesis 3 operating-model (OM) and estimation-model
(EM) spawning-biomass trajectories are overlaid over the historical period for
reference.

See `vignette("opakapaka")` for the full model configuration, biological
inputs, and cross-platform benchmarking; only the elements needed to build
`obj_lf` are reproduced here.

---

# Setup

```{r}
#| label: packages
#| message: false
library(opal)
library(RTMB)
library(SparseNUTS)
library(tidyverse)
library(posterior)

theme_set(theme_bw())

# The MCMC fit is pre-computed and loaded from the package (see the
# `mcmc-sampling` chunk, which is disabled); n_chains_mcmc must match the
# number of chains used to produce the cached fit.
n_chains_mcmc <- 8L
```

## Data, parameters, and map

The data preparation, parameter list, priors, and parameter map are identical
to the *opal with LF* model in `vignette("opakapaka")`. They are reproduced
compactly below.

```{r}
#| label: data
#| message: false
data(opaka_data)
data <- opaka_data
if (is.null(data$n_index)) data$n_index <- 2L
data$priors  <- NULL

# Anchor the growth curve at biological age 0 (SS3 indexes ages 0-43)
data$A1      <- 0
data$A2      <- data$A1 + (data$n_age - 1)
data$min_age <- 0
ages <- data$age_a <- seq(data$min_age, by = 1, length.out = data$n_age)
data$sex_ratio <- rep(1, data$n_age)

# Length composition data (Comm = fleet 1, ResFish = fleet 3)
data(opaka_lf)
lf_wide <- opaka_lf %>%
  tidyr::pivot_wider(id_cols = c(fishery, year, month, ts),
                     names_from = bin, values_from = value, values_fill = 0) %>%
  arrange(fishery, ts)
data <- prep_lf_data(data, lf_wide,
                     lf_keep_fisheries = c(1, 3),
                     lf_var_adjust     = 1 / rep(1, data$n_fishery),
                     lf_addtocomp      = 1e-10)

# Calendar years for the modelled period (1949-2023)
calendar_years <- (1949 - 1) + data$years

# SS3-style recruitment bias-adjustment ramps
init_cohort_years <- 1949 - (data$age_a - 1)
data$init_bias_adj_a <- get_bias_adj_vector(
  years = init_cohort_years, do_rec_bias_ramp = 1,
  bias_years = c(1931, 1974, 2021.88, 2022.43), max_bias_adj = 0.869)
data$bias_adj_y <- get_bias_adj_vector(
  years = calendar_years, do_rec_bias_ramp = 1,
  bias_years = c(1931, 1974, 2021.88, 2022.43), max_bias_adj = 0.869)
```

```{r}
#| label: parameters
#| message: false
data(opaka_parameters)

# Route the 1949 deviate to the age-0 cohort; shift 1950-2023 into rdev_y and
# append a terminal 0 (see the opakapaka vignette for the full rationale).
init_rdev_a <- rev(c(0.0206807,0.00321579,0.00372620,0.0043174,0.00500263,
  0.00579674,0.00671727,0.00778422,0.00902081,0.0104539,0.0121143,0.0140375,
  0.0162641,0.0188404,0.0218182,0.0252561,0.0292182,0.0337726,0.0389898,
  0.0449362,0.0516662,0.0592049,0.0675236,0.0764992,0.0858443,0.0950047,
  0.102706,0.107687,0.107712,0.0998349,0.0808137,0.0482339,0.00234443,
  -0.0517563,-0.102395,-0.131458,-0.121101,-0.0714826,-0.00742166,-0.0109308,
  -0.0385029,0.0366483,0.0165105,-0.166906))
init_rdev_a[1]   <- opaka_parameters$rdev_y[1]
rdev_y_shifted   <- c(opaka_parameters$rdev_y[2:length(opaka_parameters$rdev_y)], 0.0)

parameters <- list(
  log_B0         = 9,
  log_h          = as.numeric(opaka_parameters$log_h),
  log_sigma_r    = as.numeric(opaka_parameters$log_sigma_r),
  log_cpue_q     = as.numeric(opaka_parameters$log_cpue_q),
  log_cpue_tau   = rep(log(0.1), data$n_index),
  log_cpue_omega = as.numeric(rep(opaka_parameters$log_cpue_omega, data$n_index)),
  log_lf_tau     = as.numeric(log(rep(0.1, data$n_fishery))),
  log_L1         = as.numeric(opaka_parameters$log_L1),
  log_L2         = as.numeric(opaka_parameters$log_L2),
  log_k          = as.numeric(opaka_parameters$log_k),
  log_CV1        = as.numeric(opaka_parameters$log_CV1),
  log_CV2        = as.numeric(opaka_parameters$log_CV2),
  par_sel        = as.matrix(opaka_parameters$par_sel),
  log_init_F_f   = c(log(0.0119122), rep(log(1e-8), data$n_fishery - 1)),
  rdev_y         = rdev_y_shifted,
  init_rdev_a    = init_rdev_a
)
parameters$par_sel <- convert_ss3_selex_to_rtmb(parameters$par_sel,
                                                data$sel_type_f, data$len_mid)

data$priors <- get_priors(parameters = parameters, data = data)
```

The parameter map frees $B_0$, the two CPUE catchabilities, all recruitment
and initial-age deviations, and the two Comm logistic + three ResFish
double-normal selectivity parameters. Steepness, $\sigma_R$, growth, and the
non-commercial selectivity are fixed at SS3 values.

```{r}
#| label: map
#| message: false
map_sel <- matrix(NA, nrow(parameters$par_sel), ncol(parameters$par_sel))
map_sel[1, 1:2]        <- c(1, 2)   # Comm logistic
map_sel[3, c(1, 4, 6)] <- c(3, 4, 5) # ResFish double-normal
map_sel <- as.vector(map_sel)

map <- list(
  log_h          = factor(NA),
  log_sigma_r    = factor(NA),
  log_cpue_tau   = factor(rep(NA, data$n_index)),
  log_cpue_omega = factor(rep(NA, data$n_index)),
  log_lf_tau     = factor(rep(NA, length(parameters$log_lf_tau))),
  log_L1  = factor(NA), log_L2 = factor(NA), log_k = factor(NA),
  log_CV1 = factor(NA), log_CV2 = factor(NA),
  par_sel      = factor(map_sel),
  log_init_F_f = rep(factor(NA), data$n_fishery),
  init_rdev_a  = as.factor(seq_len(data$n_age)),
  rdev_y       = as.factor(seq_len(length(parameters$rdev_y)))
)
```

## Build and optimise

```{r}
#| label: fit
#| output: false
#| message: false
data$lf_switch <- 1L   # multinomial length-composition likelihood

obj_lf <- MakeADFun(func = cmb(opal_model, data), parameters = parameters,
                    map = map)

Lwr <- rep(-Inf, length(obj_lf$par)); Upr <- rep(Inf, length(obj_lf$par))
Lwr[grep("log_B0",      names(obj_lf$par))] <- log(1)
Upr[grep("log_B0",      names(obj_lf$par))] <- log(exp(22))
Lwr[grep("log_cpue_q",  names(obj_lf$par))] <- log(0.001)
Upr[grep("log_cpue_q",  names(obj_lf$par))] <- log(10)
Lwr[grep("rdev_y",      names(obj_lf$par))] <- -5
Upr[grep("rdev_y",      names(obj_lf$par))] <-  5
Lwr[grep("init_rdev_a", names(obj_lf$par))] <- -5
Upr[grep("init_rdev_a", names(obj_lf$par))] <-  5
Lwr[grep("par_sel",     names(obj_lf$par))] <- -7
Upr[grep("par_sel",     names(obj_lf$par))] <-  7

control <- list(eval.max = 10000, iter.max = 10000)
opt <- nlminb(obj_lf$par, obj_lf$fn, obj_lf$gr, lower = Lwr, upper = Upr,
              control = control)
opt <- nlminb(opt$par,    obj_lf$fn, obj_lf$gr, lower = Lwr, upper = Upr,
              control = control)
opt <- nlminb(opt$par,    obj_lf$fn, obj_lf$gr, lower = Lwr, upper = Upr,
              control = control)

cat("Convergence:", opt$convergence, "\n")
cat("Max gradient:", max(abs(obj_lf$gr())), "\n")
cat("NLL:", opt$objective, "\n")
```

## Historical uncertainty (delta method)

`ADREPORT(spawning_biomass_y)` is registered inside `do_dynamics()`, so
`sdreport()` propagates Hessian uncertainty through the full historical
spawning-biomass trajectory via automatic differentiation. These delta-method
standard errors are used later for the MVN historical ribbon.

```{r}
#| label: sdreport
#| output: false
#| message: false
sdr       <- RTMB::sdreport(obj_lf)
sdr_sum   <- summary(sdr, "report")
sbio_rows <- which(rownames(sdr_sum) == "spawning_biomass_y")
sbio_mle_sd <- sdr_sum[sbio_rows, "Estimate"]
sbio_se_sd  <- sdr_sum[sbio_rows, "Std. Error"]

# Scalar parameter SEs from the inverse Hessian (for reference)
H_lf   <- obj_lf$he(); Sigma <- solve(H_lf); SE_vec <- sqrt(diag(Sigma))
mu_vec <- obj_lf$env$last.par.best
```

The MLE $B_0$ and the estimated terminal spawning biomass:

```{r}
#| label: mle-summary
B0_est   <- obj_lf$report()$B0
term_idx <- data$n_year + 1
cat("B0 (MLE):", round(B0_est), "mt\n")
cat("Terminal SB:", round(sbio_mle_sd[term_idx]),
    "+/-", round(sbio_se_sd[term_idx]), "mt (delta-method SE)\n")
cat("Terminal depletion SB/B0:", round(sbio_mle_sd[term_idx] / B0_est, 3), "\n")
```

---

# Forward projections

## Common setup

The projection holds catch constant at the terminal year's level across a
10-year horizon. Projected recruitment deviates are simulated from an ARIMA
model fitted to the estimated `rdev_y` series; projected selectivity replicates
the mean selectivity over the last ten estimated years.

```{r}
#| label: proj-setup
#| output: false
#| message: false
# Fields required by the projection helpers
data$first_yr         <- 1949L
data$last_yr          <- 2023L
data$removal_switch_f <- rep(0L, data$n_fishery)

n_proj <- 10    # projection years
n_iter <- 200   # uncertainty iterations

# Projected recruitment deviates (ARIMA on the fitted MLE series)
rdev_proj <- project_rec_devs(
  data = data, obj = obj_lf, first_yr = data$first_yr,
  n_proj = n_proj, n_iter = n_iter)

# Projected selectivity (mean of the last 10 estimated years, per iteration)
sel_proj <- project_selectivity(
  data = data, obj = obj_lf, first_yr = data$last_yr - 9,
  n_proj = n_proj, n_iter = n_iter)

# Constant-catch scenario: hold catch at the last observed year's level
catch_ysf_proj <- array(0, dim = c(n_proj, data$n_season, data$n_fishery))
for (y in seq_len(n_proj)) {
  catch_ysf_proj[y, , ] <- data$catch_obs_ysf[data$n_year, , ]
}
```

## Approach 1: MVN simulation

With `mcmc = NULL`, `project_dynamics()` draws parameter sets from a
multivariate-normal centred on the MLE, using the inverse Hessian as the
covariance. Positive parameters already on the log scale (`log_B0`,
`log_cpue_q`) are drawn in log-space; the recruitment and initial-age
deviations (`rdev_y`, `init_rdev_a`) and the selectivity working-scale
parameters (`par_sel`) are drawn on their natural unbounded scale. To avoid a
Jensen's-inequality inflation of mean biomass, each deviation is set to its
conditional expectation given the drawn scalar parameters, which preserves the
negative $B_0 \leftrightarrow$ deviation correlation and aligns the bridge
point with the delta-method historical ribbon.

```{r}
#| label: proj-mvn
#| output: false
#| message: false
proj_mvn <- project_dynamics(
  data      = data,
  object    = obj_lf,
  mcmc      = NULL,             # triggers the MVN path
  n_proj    = n_proj,
  n_iter    = n_iter,
  rdev_y    = rdev_proj$rdev_y,
  sel_fya   = sel_proj,
  catch_ysf = catch_ysf_proj
)
```

## Approach 2: MCMC posteriors

A No-U-Turn sampler chain is run via `SparseNUTS` with a dense mass matrix.
The `bet_globals()` call exports the helper functions needed for likelihood
evaluation inside the MCMC workers (it is model-agnostic despite the name).
Because both CPUE indices and the length compositions inform the fit, the
$B_0$/$q$ scale is well identified and no additional priors are needed beyond
those already in the model.

```{r}
#| label: mcmc-sampling
#| eval: false
#| output: false
n_chains_mcmc <- 8
mcmc_fit <- SparseNUTS::sample_snuts(
  obj         = obj_lf,
  globals     = bet_globals(),
  num_samples = 750,
  num_warmup  = 1500,
  chains      = n_chains_mcmc,
  cores       = min(n_chains_mcmc, parallel::detectCores()),
  metric      = "dense",
  seed        = 42,
  # A high adapt_delta and deeper trees are needed for this posterior: with the
  # default settings the survey length-composition geometry produces divergences
  # and elevated Rhat.
  control     = list(adapt_delta = 0.99, max_treedepth = 12)
)
saveRDS(mcmc_fit, "inst/extdata/opaka_mcmc_fit.rds")
```

```{r}
#| label: mcmc-load
#| message: false
mcmc_fit <- readRDS(system.file("extdata", "opaka_mcmc_fit.rds", package = "opal"))
```

For the MCMC path, projected recruitment deviates are generated by fitting an
ARIMA model to each posterior draw's own historical `rdev_y` series, so the
projected recruitment pattern stays consistent with the historical pattern for
that draw.

```{r}
#| label: rdev-proj-mcmc
#| output: false
#| message: false
rdev_proj_mcmc <- project_rec_devs(
  data = data, obj = obj_lf, mcmc = mcmc_fit,
  first_yr = data$first_yr, n_proj = n_proj, n_iter = n_iter)
```

```{r}
#| label: proj-mcmc
#| output: false
#| message: false
proj_mcmc <- project_dynamics(
  data        = data,
  object      = obj_lf,
  mcmc        = mcmc_fit,       # triggers the MCMC path
  n_proj      = n_proj,
  n_iter      = n_iter,
  rdev_y      = rdev_proj_mcmc$rdev_y,
  sel_fya     = sel_proj,
  catch_ysf   = catch_ysf_proj,
  return_hist = TRUE            # collect historical SB per posterior draw
)
```

## MCMC diagnostics

Split-chain $\hat{R}$ and effective sample sizes for the scalar parameters and
a spread of recruitment deviates confirm convergence before the projection
envelopes are compared.

```{r}
#| label: mcmc-diagnostics
#| message: false
post_samples <- SparseNUTS::extract_samples(mcmc_fit)
n_post_total <- nrow(post_samples)
n_per_chain  <- n_post_total / n_chains_mcmc
chain_id     <- rep(seq_len(n_chains_mcmc), each = n_per_chain)
iter_id      <- rep(seq_len(n_per_chain),   times = n_chains_mcmc)

rdev_idx_all  <- grep("rdev_y", colnames(post_samples))
rdev_idx_show <- rdev_idx_all[round(seq(1, length(rdev_idx_all), length.out = 4))]
diag_cols     <- c("log_B0", "log_cpue_q[1]", "log_cpue_q[2]",
                   colnames(post_samples)[rdev_idx_show])

draws_df <- as.data.frame(post_samples[, diag_cols, drop = FALSE])
draws_df[[".chain"]]     <- chain_id
draws_df[[".iteration"]] <- iter_id
draws_obj <- posterior::as_draws_df(draws_df)
diag_tbl  <- posterior::summarise_draws(
  draws_obj,
  posterior::default_summary_measures(),
  posterior::default_convergence_measures())
diag_tbl$mle <- mu_vec[match(diag_tbl$variable, names(mu_vec))]

knitr::kable(
  diag_tbl[, c("variable", "mle", "mean", "sd", "q5", "q95",
               "ess_bulk", "ess_tail", "rhat")],
  digits = 3,
  caption = paste0("MCMC diagnostics for key parameters (split-chain Rhat and ",
                   "ESS). ESS < 100 or Rhat > 1.01 indicate poor convergence."))

SparseNUTS::check_snuts_diagnostics(mcmc_fit)
```

```{r}
#| label: mcmc-traces
#| fig-height: 4
#| fig-width: 8
#| fig-cap: "MCMC traces for the scalar parameters (log B0 and the two log CPUE catchabilities) across all chains. Red dashed = MLE. Overlapping, stationary chains indicate good mixing."
trace_df <- data.frame(
  iter  = iter_id, chain = factor(chain_id),
  log_B0        = post_samples[, "log_B0"],
  `log_cpue_q1` = post_samples[, "log_cpue_q[1]"],
  `log_cpue_q2` = post_samples[, "log_cpue_q[2]"],
  check.names = FALSE) |>
  tidyr::pivot_longer(-c(iter, chain), names_to = "param", values_to = "value")

mle_ref <- data.frame(
  param = c("log_B0", "log_cpue_q1", "log_cpue_q2"),
  mle   = c(mu_vec["log_B0"], mu_vec[names(mu_vec) == "log_cpue_q"]))

ggplot(trace_df, aes(iter, value, colour = chain)) +
  geom_line(linewidth = 0.3, alpha = 0.7) +
  geom_hline(data = mle_ref, aes(yintercept = mle),
             linetype = "dashed", colour = "red", linewidth = 0.5) +
  facet_wrap(~ param, scales = "free_y", ncol = 1) +
  labs(x = "Post-warmup iteration", y = "Value", colour = "Chain")
```

---

# Comparison: MVN vs MCMC uncertainty

Projected spawning biomass is extracted from each iteration and summarised as
quantile envelopes. `number_ysa[1, , ]` is the terminal historical state
(`proj_year = 0`), so it bridges continuously to the historical trajectory.

```{r}
#| label: extract-sbio
#| message: false
spa <- obj_lf$report()$spawning_potential_a

sbio_df <- function(proj_list, label) {
  do.call(rbind, lapply(seq_along(proj_list), function(i) {
    dyn <- proj_list[[i]]
    sb  <- sapply(1:(n_proj + 1), function(y) sum(dyn$number_ysa[y, 1, ] * spa))
    data.frame(iter = i, proj_year = 0:n_proj, sbio = sb, method = label)
  }))
}

df_mvn  <- sbio_df(proj_mvn,      "MVN")
df_mcmc <- sbio_df(proj_mcmc$dyn, "MCMC")
df_all  <- rbind(df_mvn, df_mcmc)
```

```{r}
#| label: summarise-proj
#| message: false
# Map model timesteps to calendar years: historical t = 1..(n_year+1) -> 1948+t
# (t = n_year+1 is the terminal state = 2024); projection proj_year p -> 2024+p
n_hist_show <- 60L
hist_idx    <- (term_idx - n_hist_show + 1):term_idx
hist_years  <- 1948L + hist_idx

# MVN historical ribbon: delta-method Gaussian CI (no Jensen bias)
df_hist_mvn <- data.frame(
  year = hist_years,
  med  = sbio_mle_sd[hist_idx],
  lo50 = sbio_mle_sd[hist_idx] - qnorm(0.75) * sbio_se_sd[hist_idx],
  hi50 = sbio_mle_sd[hist_idx] + qnorm(0.75) * sbio_se_sd[hist_idx],
  lo90 = sbio_mle_sd[hist_idx] - qnorm(0.95) * sbio_se_sd[hist_idx],
  hi90 = sbio_mle_sd[hist_idx] + qnorm(0.95) * sbio_se_sd[hist_idx],
  method = "MVN")

# MCMC historical ribbon: posterior trajectory quantiles
hist_ribbon <- function(hist_sbio, label, idx, years) {
  do.call(rbind, lapply(seq_along(idx), function(k) {
    v <- hist_sbio[, idx[k]]
    data.frame(year = years[k],
               med  = median(v, na.rm = TRUE),
               lo50 = quantile(v, 0.25, na.rm = TRUE),
               hi50 = quantile(v, 0.75, na.rm = TRUE),
               lo90 = quantile(v, 0.05, na.rm = TRUE),
               hi90 = quantile(v, 0.95, na.rm = TRUE),
               method = label)
  }))
}
df_hist_mcmc   <- hist_ribbon(proj_mcmc$hist_sbio, "MCMC", hist_idx, hist_years)
df_hist_ribbon <- rbind(df_hist_mvn, df_hist_mcmc)

# Projection quantile ribbons (proj_year = 0 -> 2024 bridge point)
df_ribbon <- df_all |>
  filter(!is.na(sbio)) |>
  group_by(method, proj_year) |>
  summarise(med  = median(sbio, na.rm = TRUE),
            lo50 = quantile(sbio, 0.25, na.rm = TRUE),
            hi50 = quantile(sbio, 0.75, na.rm = TRUE),
            lo90 = quantile(sbio, 0.05, na.rm = TRUE),
            hi90 = quantile(sbio, 0.95, na.rm = TRUE),
            .groups = "drop") |>
  filter(!is.na(med)) |>
  mutate(year = 2024L + proj_year)

# MLE historical line + SS3 references
hist_mle_df <- data.frame(year = hist_years, sbio = sbio_mle_sd[hist_idx])
data(opaka_truth)
ss3_om <- subset(opaka_truth$om$sprseries, Yr %in% hist_years)
ss3_em <- subset(opaka_truth$em$sprseries, Yr %in% hist_years)
```

```{r}
#| label: comparison-plot
#| fig-height: 5
#| fig-width: 8
#| fig-cap: "'Opakapaka spawning biomass: historical uncertainty and projections. Last 60 years shown with a 10-year constant-catch projection (200 draws per method). Grey solid = estimated MLE; black dashed = SS3 OM; red = SS3 EM. Coloured ribbons: 50% (darker) and 90% (lighter) envelopes by method. Orange line = MCMC historical median. Vertical dashed line = end of the historical period (2024)."
pal <- c(MVN = "steelblue", MCMC = "darkorange")

ggplot() +
  geom_ribbon(data = df_hist_ribbon,
              aes(year, ymin = lo90, ymax = hi90, fill = method), alpha = 0.12) +
  geom_ribbon(data = df_hist_ribbon,
              aes(year, ymin = lo50, ymax = hi50, fill = method), alpha = 0.25) +
  geom_line(data = subset(df_hist_ribbon, method == "MCMC"),
            aes(year, med, colour = method), linewidth = 0.75) +
  geom_line(data = hist_mle_df, aes(year, sbio), colour = "grey20", linewidth = 0.9) +
  geom_line(data = ss3_om, aes(Yr, SSB), colour = "black",
            linewidth = 0.8, linetype = "dashed") +
  geom_line(data = ss3_em, aes(Yr, SSB), colour = "red", linewidth = 0.8) +
  geom_vline(xintercept = 2024, linetype = "dashed", colour = "grey50") +
  geom_ribbon(data = df_ribbon,
              aes(year, ymin = lo90, ymax = hi90, fill = method), alpha = 0.15) +
  geom_ribbon(data = df_ribbon,
              aes(year, ymin = lo50, ymax = hi50, fill = method), alpha = 0.30) +
  geom_line(data = df_ribbon, aes(year, med, colour = method), linewidth = 1) +
  scale_fill_manual(values = pal) +
  scale_colour_manual(values = pal) +
  labs(x = "Year", y = "Spawning biomass (mt)", colour = "Method", fill = "Method")
```

```{r}
#| label: depletion-plot
#| fig-height: 5
#| fig-width: 8
#| fig-cap: "'Opakapaka relative depletion (SB / B0): historical uncertainty and projections. Last 60 years shown with a 10-year constant-catch projection (200 draws per method). Model-output quantities use the estimated B0 (MVN) or per-draw B0 (MCMC). Dotted red line = 20% B0; dashed grey = 40% B0."
B0_mcmc      <- exp(post_samples[, "log_B0"])
B0_mcmc_iter <- B0_mcmc[seq_len(n_iter)]

# Historical depletion
df_hist_mvn_dep <- df_hist_mvn |>
  mutate(across(c(med, lo50, hi50, lo90, hi90), ~ . / B0_est))
hist_sbio_dep    <- sweep(proj_mcmc$hist_sbio, 1, B0_mcmc_iter, "/")
df_hist_mcmc_dep <- hist_ribbon(hist_sbio_dep, "MCMC", hist_idx, hist_years)
df_hist_ribbon_dep <- rbind(df_hist_mvn_dep, df_hist_mcmc_dep)

# Projection depletion
df_ribbon_mvn_dep <- df_ribbon |>
  filter(method == "MVN") |>
  mutate(across(c(med, lo50, hi50, lo90, hi90), ~ . / B0_est))
df_ribbon_mcmc_dep <- df_mcmc |>
  mutate(sbio = sbio / B0_mcmc[iter]) |>
  filter(!is.na(sbio)) |>
  group_by(method, proj_year) |>
  summarise(med  = median(sbio, na.rm = TRUE),
            lo50 = quantile(sbio, 0.25, na.rm = TRUE),
            hi50 = quantile(sbio, 0.75, na.rm = TRUE),
            lo90 = quantile(sbio, 0.05, na.rm = TRUE),
            hi90 = quantile(sbio, 0.95, na.rm = TRUE),
            .groups = "drop") |>
  filter(!is.na(med)) |>
  mutate(year = 2024L + proj_year)
df_ribbon_dep <- rbind(df_ribbon_mvn_dep, df_ribbon_mcmc_dep)

hist_mle_dep <- data.frame(year = hist_years, dep = sbio_mle_sd[hist_idx] / B0_est)

ggplot() +
  geom_ribbon(data = df_hist_ribbon_dep,
              aes(year, ymin = lo90, ymax = hi90, fill = method), alpha = 0.12) +
  geom_ribbon(data = df_hist_ribbon_dep,
              aes(year, ymin = lo50, ymax = hi50, fill = method), alpha = 0.25) +
  geom_line(data = subset(df_hist_ribbon_dep, method == "MCMC"),
            aes(year, med, colour = method), linewidth = 0.75) +
  geom_line(data = hist_mle_dep, aes(year, dep), colour = "grey20", linewidth = 0.9) +
  geom_vline(xintercept = 2024, linetype = "dashed", colour = "grey50") +
  geom_ribbon(data = df_ribbon_dep,
              aes(year, ymin = lo90, ymax = hi90, fill = method), alpha = 0.15) +
  geom_ribbon(data = df_ribbon_dep,
              aes(year, ymin = lo50, ymax = hi50, fill = method), alpha = 0.30) +
  geom_line(data = df_ribbon_dep, aes(year, med, colour = method), linewidth = 1) +
  geom_hline(yintercept = 0.2, linetype = "dotted", colour = "red", linewidth = 0.5) +
  geom_hline(yintercept = 0.4, linetype = "dashed", colour = "grey50", linewidth = 0.4) +
  coord_cartesian(ylim = c(0, NA)) +
  scale_fill_manual(values = pal) +
  scale_colour_manual(values = pal) +
  labs(x = "Year", y = expression("Relative depletion (SB / B"[0]*")"),
       colour = "Method", fill = "Method")
```

**Reading the bridge point.** In the MVN case the scalar parameters are drawn
from the Hessian and each deviation is pinned to its conditional expectation,
so the `proj_year = 0` ribbon coincides with the delta-method (`sdreport`)
historical ribbon. In the MCMC case each draw carries its own full history, so
the bridge-point spread reflects the actual posterior uncertainty in the
current biomass state. Because 'opakapaka is long-lived and matures slowly,
the existing adult stock declines nearly deterministically under constant
catch, and the two methods' projection envelopes widen only gradually as future
recruitment and selectivity uncertainty accumulate.

```{r}
#| label: summary-table
#| message: false
df_all |>
  filter(proj_year == n_proj, !is.na(sbio)) |>
  group_by(method) |>
  summarise(mean   = mean(sbio),
            sd     = sd(sbio),
            p05    = quantile(sbio, 0.05),
            median = median(sbio),
            p95    = quantile(sbio, 0.95),
            .groups = "drop") |>
  knitr::kable(digits = 1,
    caption = paste0("Projected spawning biomass (mt) at year ", 2024 + n_proj,
                     " under constant catch, by uncertainty method."))
```
