---
title: "Quickstart Guide"
format: html
execute:
  eval: true
  warning: false
  message: false
vignette: >
  %\VignetteIndexEntry{Quickstart Guide}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

```{r}
#| include: false
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```


This guide fits a fixed-effects assessment for 'opakapaka
(*Pristipomoides filamentosus*) using the bundled data. It uses the same age
convention and selectivity parameterisation as its Stock Synthesis 3 (SS3)
reference model.

One portable S3 class, `opal_obj`, holds the assessment from configuration
through fitting, sampling, and reporting. Its lifecycle stage and diagnostic
status are separate: a fitted object can still fail its convergence checks.
RTMB objectives live outside the saved object and are rebuilt when needed.

## Load the model inputs

The package includes a model-input builder so this vignette and the stored
example fit use the same data preparation and fixed-effects configuration.

```{r}
#| label: setup
library(opal)

inputs <- opaka_quickstart_inputs()
```

The input builder loads the bundled opakapaka catch, CPUE, and length-composition
data; uses biological ages 0--43 to align with SS3; applies the SS3 recruitment
bias ramps; and prepares multinomial length compositions for the commercial
and research fleets. It fixes life-history parameters at their SS3 values and
estimates biomass scale, CPUE catchability, recruitment deviations, initial
age structure, and selectivity for fleets with length data.

```{r}
#| label: model-dimensions
data.frame(
  quantity = c("Years", "Ages", "Fisheries", "Length-composition samples"),
  value = c(
    inputs$data$n_year,
    inputs$data$n_age,
    inputs$data$n_fishery,
    length(inputs$data$lf_year)
  )
)
```

## Fit the model

Create the assessment, then optimise twice using the existing Opal defaults.
The wrapper resolves bounds, stores both optimiser passes, and checks the
final fit. It retains results and warns if the checks fail.

```{r}
#| label: fit
fit <- opal_obj(
  data = inputs$data,
  parameters = inputs$parameters,
  map = inputs$map,
  metadata = list(stock = "Opakapaka", model = "Fixed-effects quickstart")
)
fit <- opal_fit(fit)
fit
fit$validation$fit
```

The optimiser can return convergence code zero while the maximum gradient
still exceeds the default tolerance of `0.001`. Inspect the recorded metrics.
If needed, continue from the stored optimum with a tighter relative objective
tolerance for the optimiser, keeping the same model and diagnostic threshold.
This example checks after each pass, allows up to three further passes, and
stops if the checks still fail; the `fitted` stage alone is not acceptance.

```{r}
#| label: refine-fit
for (attempt in seq_len(3L)) {
  if (isTRUE(fit$validation$fit$passes)) break
  fit <- opal_fit(
    fit, n_passes = 1L,
    control = list(eval.max = 10000L, iter.max = 10000L, rel.tol = 1e-10)
  )
}
stopifnot(isTRUE(fit$validation$fit$passes))
summary(fit)
```

## Save a portable fitted model

`opal_save()` stores ordinary R data with a payload checksum. `opal_read()`
verifies the payload and scientific contract. Neither reading nor rebuilding
runs optimisation or MCMC. The same functions also save unfitted objects.

```{r}
#| label: save-and-read-fit
fit_path <- tempfile("opaka-quickstart-", fileext = ".rds")
opal_save(fit, fit_path)
fit <- opal_read(fit_path, rebuild = TRUE)
summary(fit)
unlink(fit_path)
```

## Inspect spawning biomass

Existing plots accept the assessment object directly. Advanced callers can
obtain a transient RTMB objective with `opal_rtmb(fit, fresh = TRUE)`.

```{r}
#| label: spawning-biomass
#| fig-width: 8
#| fig-height: 4
report <- opal_report(fit)
plot_biomass_spawning(fit, relative = FALSE, labels = "Fixed effects")
```

## Sample and update

Sampling can start before or after fitting. The default sampler uses the
stored point, skips internal optimisation, and honours parameter bounds.
The following example is intentionally not run while building the vignette.
Sampling effort and convergence thresholds need assessment-specific review.

```{r}
#| eval: false
fit <- opal_mcmc(fit, seed = 42, chains = 4, cores = 4,
                 num_warmup = 1000, num_samples = 500)
fit <- opal_check(fit, scope = "mcmc")
summary(fit)
```

Failed checks retain the samples. A failed or unchecked new run does not
replace a previously checked, passing posterior; inspect `fit$mcmc_history`
for earlier and unselected attempts. Refitting the same target preserves its
posterior and records which optimisation originally supplied it.

Use `opal_update()` for changes. Updating data, priors, parameters, maps,
random effects, or bounds clears dependent fits, samples, and projections.
Metadata and optimiser controls preserve them. Direct changes to scientific
fields are rejected by validation. An omitted map resolves defaults, whereas
`map = list()` explicitly leaves all parameters free. Custom stocks should
supply their own parameters; a bundled default requires an explicit `model`.

```{r}
labelled <- opal_update(fit, metadata = list(label = "Baseline"))
summary(labelled)
```

## Projections and older workflows

`project_selectivity()`, `project_rec_devs()`, and `project_dynamics()` accept
objects directly. Choose `uncertainty = "mcmc"` or `"mvn"` for dynamics when
posterior draws exist. `opal_project()` stores a projection with its source
identity and settings; future catches, selectivity, and recruitment remain
explicit inputs. Random-effect dynamics projections require complete joint
posterior draws; marginal-only draws and random-effect MVN projections are
rejected.

`opal_fit(data, obj, opt, ...)` still supports the legacy constructor.
`opal_from_fit()` converts an existing fit, and `opal_read()` upgrades legacy
saved fits after verifying the rebuilt objective, without refitting. The
companion [saved-fit review](opakapaka-fit-review.html) demonstrates this migration.
The [projection guide](projections.html) gives a complete runnable example
with future catch, recruitment, selectivity, and saved projection provenance.
